### 2021-01-21 manuscript/manuscript.tex
adding manuscript

\documentclass[11pt, a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{titlesec}

% Maths tools
\usepackage[tbtags]{amsmath}
\usepackage{amssymb}

% Margin
\usepackage[margin=2.9cm]{geometry}

% Line numbers
\usepackage{lineno}

% Spacing
\usepackage{setspace}
\doublespacing

% Enumeration
\usepackage{enumerate}% http://ctan.org/pkg/enumerate
\usepackage{enumitem}

% indentation
\setlength\parindent{10pt}
\setlength{\parskip}{5pt}

% Figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{epstopdf}
\usepackage{float}
\renewcommand{\thefigure}{\textbf{\arabic{figure}}}
\renewcommand{\figurename}{\textbf{Figure}}

%table
\usepackage{multirow}% http://ctan.org/pkg/multirow
\usepackage{hhline}
\usepackage[table]{xcolor}

\usepackage{authblk}

% References
\usepackage[round]{natbib}
\bibliographystyle{ecology_letters2.bst}

\usepackage{color, xcolor,soul}
%\definecolor{blau}{RGB}{168,221,181}
\definecolor{blau}{RGB}{236,226,240}
\soulregister\cite7
\soulregister\citenum7
\soulregister\citep7
\soulregister\citealt7
\soulregister\citealp7
\soulregister\citet7
\soulregister\ref7

% References and links
\PassOptionsToPackage{hyphens}{url}\usepackage[colorlinks=true,linkcolor=magenta, citecolor=magenta]{hyperref}

%Margin notes
\usepackage{marginnote}
% Foot note distance with text
\usepackage[symbol]{footmisc}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\setlength{\skip\footins}{0.75cm}

%Define colour
\DeclareRobustCommand{\hlc}[1]{{\sethlcolor{blau}\hl{#1}}}

%subsubsection format
\titleformat*{\subsubsection}{\large\it}

\makeatletter
\renewcommand\AB@affilsepx{; \protect\Affilfont}
\makeatother

%Title paper
\title{\vspace{-1cm}
A fresh perspective on distribution modelling: a bayesian framework to understand the distribution of plant species along an environmental gradient}
\author[1,*]{\normalsize Bernat Bramon Mora}
\author[1]{\normalsize Jake M.\ Alexander}
\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
\affil[*]{\footnotesize  bernat.bramon@gmail.com}

\renewcommand\Authands{ and }
\date{}

\begin{document}
\maketitle
\linenumbers
%\section*{Abstract}
% Results are presented for both simulated and empirical data.
% But this approach understates the posterior uncertainty in inference. 
%\section*{Introduction}

\section*{Methods}
\subsection*{Empirical data}
We studied the distribution of alpine plant communities along an elevation gradient. To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.

\subsubsection*{Distribution data}
We studied the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}). Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$. In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species. Additionally, following 30 years (1961–1990) of meteorological data from national weather stations, \citet{scherrerEcologicalIndicatorValues2019} calculated multiple climatic variables for each site at high spatial resolution ($25\,\text{m}$). Here, we focussed on 9 climatic variables, including: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter. %Missing TabsY

\subsubsection*{Floristic data}
To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland. This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits. Species' environmental preferences in this database can be used to inform distribution models---e.g. as an informative prior in a Bayesian framework. These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found and a broad description of their range of variation. These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}). On the other hand, the information regarding species' physiological traits represent general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).

%we independently consider those traits that describe growth and life strategies of the plants---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents. In total, we identify $\sim120$ binary traits that characterize the physiology of species.

%comprising a total of 10 environmental indicator values with 10 types of associations for each of them (i.e. the $r$ associations defined above). Finally, we additionally independently consider species' range of variations for the such indicator values, also defining 10 `traits' with 3 types of associations (I need to see if I can added these to the analyses above).

%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19. Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland. There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
%Based on 30 years (1961–1990) of meteorological data from national weather stations, fourteen climatic variables (Table S5) were calculated according to the method described in Zimmermann & Kienast44 and Zimmerman et al.55. A digital elevation model (DEM) was used to spatially interpolate the climatic data and to create topographic variables (Table S5). Additionally, we used a soil pH map based on Buri et al.43. All environmental data had a spatial resolution of 25 m, a resolution similar to the areas used to calculate the ecological indicator values (EIV; see below). This allowed a straightforward comparison of different sets of predictors composed from environmental variables and EIVs.

%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19. Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland. There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.

{\color{gray}
\subsubsection*{[Trait data]}
This could be Tom's data if we end up using it.}

\subsection*{Distribution model}
There is a long list of model structures well suited to characterize species' distributions (see XX for a review); however, we were interested in a model that explicitly incorporates all information regarding plant's environmental preferences found in the floristic database. More specifically, we wanted to account for the climatic indicator values and range of variation registered for all plants in our dataset. These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions. Therefore, we first formulated a baseline model that directly accounts for such prior information.

\subsubsection*{Baseline model}
Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
\begin{equation} 
\begin{split}
y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
\hat{\alpha}, 
\hat{\lambda^{k}}, 
\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
\end{split}
\label{eq:baseline}
\end{equation}
Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively. While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$. More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e. Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively. Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.

In all cases, all variance-covariance matrices are defined as follows:
\begin{equation} 
\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
\label{eq:covariance}
\end{equation}

where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$. Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions. Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details). For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value. The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species. For any given hyperparameter, we choose adaptive priors across covariance structures. That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$. Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$. We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.

\subsubsection*{Distance matrices}
The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$. In this model, such distance matrices characterize differences between plant species. In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits. More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species. In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries. Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.

More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits. One way to frame this problem is by using a network representation. Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$. Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$. That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.

Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these. In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively. For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups. Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups. Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$. Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
\begin{equation}
Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
\end{equation}
Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$. Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}. In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships. The reason for this is related to the stochastic nature of the expectation-maximization algorithm. This algorithm initially assigns random group memberships to both species and traits. While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$. Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm. In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration. A full section for this is required in the Supplementary Information.}.

The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have. In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e. the extend to which a species is of one type or another. This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$. The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$. Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.

\subsubsection*{Modifying the variance-covariance structures}
The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information. Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}). For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e. two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM. One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g. parameter $\alpha_i$---such that
\begin{equation} 
\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
\label{eq:covariance-complex}
\end{equation}
where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$. Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$. Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.

\section*{Results}

\section*{Discussion}

\bibliography{references}

\end{document}

### 2021-01-21 manuscript/question/question.tex
adding manuscript

--- 2021-01-21 manuscript/manuscript.tex
@@
 % Maths tools
-\usepackage[tbtags]{amsmath}
+\usepackage{amsmath}
 \usepackage{amssymb}
@@
 
-% References and links
 \PassOptionsToPackage{hyphens}{url}\usepackage[colorlinks=true,linkcolor=magenta, citecolor=magenta]{hyperref}
 
-%Margin notes
-\usepackage{marginnote}
-% Foot note distance with text
-\usepackage[symbol]{footmisc}
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\setlength{\skip\footins}{0.75cm}
-
-%Define colour
 \DeclareRobustCommand{\hlc}[1]{{\sethlcolor{blau}\hl{#1}}}
@@
 \title{\vspace{-1cm}
-A fresh perspective on distribution modelling: a bayesian framework to understand the distribution of plant species along an environmental gradient}
-\author[1,*]{\normalsize Bernat Bramon Mora}
-\author[1]{\normalsize Jake M.\ Alexander}
-\affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
-\affil[*]{\footnotesize  bernat.bramon@gmail.com}
+Combining Gaussian Processes}
 
@@
 \linenumbers
-%\section*{Abstract}
-% Results are presented for both simulated and empirical data.
-% But this approach understates the posterior uncertainty in inference. 
-%\section*{Introduction}
+Imagine that you have the ``the varying slopes model'' case described Richard McElreath book (Statistical Rethinking, second edition).
+In that example, you had a coffee robot that was programmed to move among cafés, order coffee, and record the waiting time in the morning and the afternoon.
+In short, it was an example of pooling in multilevel models and covariance among parameters.
+Given the waiting times $W_i$ for all cafés and a binary variable $A_i$ indicating whether or not the waiting time was recorded in the afternoon, the model for this example was the following:
+\begin{equation*} 
+\begin{split}
+W_i & \sim \text{Normal}\left(\mu_{i}, \sigma\right)\\
+\mu_{i} & = \alpha_{\text{\tiny CAFE}_i} + \beta_{\text{\tiny CAFE}_i} A_i\\
+r_n & = x_n-\lambda\\
+\lambda  & \sim \text{MVNormal}\left(\hat{\lambda}, E\right)\\
+\epsilon  & \sim \text{MVNormal}\left(\hat{\epsilon}, V\right)\\
+\alpha  & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
+\hat{\lambda}  & \sim \text{Normal}\left(0,1\right)\\
+\hat{\epsilon}  & \sim \text{Normal}\left(0,1\right)\\
+\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
+\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
+\end{split}
+\end{equation*}
 
-\section*{Methods}
-\subsection*{Empirical data}
-We studied the distribution of alpine plant communities along an elevation gradient.
-To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.
-
-\subsubsection*{Distribution data}
-We studied the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}).
-Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$.
-In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species.
-Additionally, following 30 years (1961–1990) of meteorological data from national weather stations, \citet{scherrerEcologicalIndicatorValues2019} calculated multiple climatic variables for each site at high spatial resolution ($25\,\text{m}$).
-Here, we focussed on 9 climatic variables, including: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter.
-%Missing TabsY
-
-\subsubsection*{Floristic data}
-To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland.
-This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits.
-Species' environmental preferences in this database can be used to inform distribution models---e.g.
-as an informative prior in a Bayesian framework.
-These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found and a broad description of their range of variation.
-These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
-On the other hand, the information regarding species' physiological traits represent general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
-
-%we independently consider those traits that describe growth and life strategies of the plants---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify $\sim120$ binary traits that characterize the physiology of species.
-
-%comprising a total of 10 environmental indicator values with 10 types of associations for each of them (i.e.
-the $r$ associations defined above).
-Finally, we additionally independently consider species' range of variations for the such indicator values, also defining 10 `traits' with 3 types of associations (I need to see if I can added these to the analyses above).
-
-%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19.
-Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland.
-There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
-%Based on 30 years (1961–1990) of meteorological data from national weather stations, fourteen climatic variables (Table S5) were calculated according to the method described in Zimmermann & Kienast44 and Zimmerman et al.55.
-A digital elevation model (DEM) was used to spatially interpolate the climatic data and to create topographic variables (Table S5).
-Additionally, we used a soil pH map based on Buri et al.43.
-All environmental data had a spatial resolution of 25 m, a resolution similar to the areas used to calculate the ecological indicator values (EIV; see below).
-This allowed a straightforward comparison of different sets of predictors composed from environmental variables and EIVs.
-
-%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19.
-Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland.
-There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
-
-{\color{gray}
-\subsubsection*{[Trait data]}
-This could be Tom's data if we end up using it.}
-
-\subsection*{Distribution model}
-There is a long list of model structures well suited to characterize species' distributions (see XX for a review); however, we were interested in a model that explicitly incorporates all information regarding plant's environmental preferences found in the floristic database.
-More specifically, we wanted to account for the climatic indicator values and range of variation registered for all plants in our dataset.
-These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions.
-Therefore, we first formulated a baseline model that directly accounts for such prior information.
-
-\subsubsection*{Baseline model}
-Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
-\begin{equation} 
-\begin{split}
-y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
-\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
-\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
-\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
-\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
-\hat{\alpha}, 
-\hat{\lambda^{k}}, 
-\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
-\end{split}
-\label{eq:baseline}
-\end{equation}
-Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively.
-While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.
-Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
-Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.
-
-In all cases, all variance-covariance matrices are defined as follows:
-\begin{equation} 
-\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
-\label{eq:covariance}
-\end{equation}
-
-where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$.
-Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
-Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
-For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
-The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
-For any given hyperparameter, we choose adaptive priors across covariance structures.
-That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
-Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
-We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
-
-\subsubsection*{Distance matrices}
-The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-In this model, such distance matrices characterize differences between plant species.
-In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits.
-More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
-In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
-Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
-
-More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
-One way to frame this problem is by using a network representation.
-Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
-Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$.
-That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
-
-Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
-In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
-For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
-Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
-Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
-Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
-\begin{equation}
-Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
-\end{equation}
-Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
-Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
-In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
-While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
-The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
-This algorithm initially assigns random group memberships to both species and traits.
-While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
-Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
-In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.
-A full section for this is required in the Supplementary Information.}.
-
-The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
-In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.
-the extend to which a species is of one type or another.
-This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
-The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
-Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
-
-\subsubsection*{Modifying the variance-covariance structures}
-The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information.
-Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}).
-For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.
-two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
-One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.
-parameter $\alpha_i$---such that
-\begin{equation} 
-\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
-\label{eq:covariance-complex}
-\end{equation}
-where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$.
-Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$.
-Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.
-
-\section*{Results}
-
-\section*{Discussion}
-
-\bibliography{references}
+Imagine now that we have additional information on the different cafés.
+For example, suppose that we know how far these cafés are from each other.
+One could hypothesise that different neighbourhoods might have different waiting times (i.e.
+fancy cafés from posh neighbourhoods might have similar waiting times, or cafés from non-residential areas might have longer morning times).
+To incorporate this prior information, we could use Gaussian processes to characterize the different parameters.
+That is, $alpha_i$ could be characterized by a multivariate normal distribution $N(\mu_i, \Sigma)$, where $\Sigma$ is a covariance matrix such that $\Sigma_{ij} = \rho^2 \exp(-\frac{(D_{ij}^{2})}{2\gamma^2}) + \delta_{ij} \sigma^{2}$, and $D_{ij}$ is the distance between cafés $i$ and $j$.
 

### 2021-01-21 manuscript/supplementary_information.tex
adding manuscript

\documentclass[11pt, a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage{authblk}
\usepackage{titlesec}

% Maths tools
\usepackage{amsmath}
\usepackage{amssymb}

% Margin
\usepackage[margin=2.9cm]{geometry}

% Line numbers
\usepackage{lineno}

%Scientific notation
\usepackage{siunitx}

% Spacing
\usepackage{setspace}
\doublespacing

% Enumeration
\usepackage{enumerate}

% indentation
\setlength\parindent{10pt}
\setlength{\parskip}{5pt}

% Figures
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{epstopdf}
\renewcommand{\thefigure}{\textbf{\arabic{figure}}}
\renewcommand{\figurename}{\textbf{Supplementary Figure}}

%table
\usepackage{multirow}
\usepackage{hhline}
\usepackage[table]{xcolor}
\renewcommand{\thetable}{\textbf{\arabic{table}}}
\renewcommand{\tablename}{\textbf{Supplementary Table}}

%footnotes
\usepackage{footmisc}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}

% References
\usepackage[sort&compress, numbers,super]{natbib}
\bibliographystyle{unsrtnat}
\PassOptionsToPackage{hyphens}{url}\usepackage[colorlinks=true,linkcolor=magenta, citecolor=magenta]{hyperref}
\renewcommand\refname{Supplementary References}

%subsubsection format
\titleformat*{\subsubsection}{\large\it}

%highlighting text
\usepackage{color, xcolor,soul}
\definecolor{blau}{RGB}{236,226,240}
\soulregister\cite7
\soulregister\citet7
\soulregister\citealt7
\soulregister\citenum7
\soulregister\citep7
\soulregister\ref7
\DeclareRobustCommand{\hlc}[1]{{\sethlcolor{blau}\hl{#1}}}

\usepackage{titlesec}% http://ctan.org/pkg/titlesec
\titleformat{\section}%
  [hang]% <shape>
  {\normalfont\bfseries\Large}% <format>
  {}% <label>
  {0pt}% <sep>
  {}% <before code>

\renewcommand{\thesection}{}% Remove section references...
\renewcommand{\thesubsection}{}%... from subsections
  \renewcommand{\thesubsubsection}{Supplementary Method \arabic{subsubsection}:}%... from subsections

%Title paper
\title{\vspace{-1cm} \normalsize Supplementary Information\\\vspace{0.2cm}
\LARGE Untangling the seasonal dynamics of plant-pollinator communities}

\author{\textit{Bramon Mora et al.}}

\date{}

\begin{document}
\maketitle
\thispagestyle{empty}

\clearpage

\section*{Supplementary Methods}
%\subsubsection*{Species' structural roles}
%There are multiple ways to characterize the structural roles of species in a community. In this work, we used a definition based on the idea of network motifs\cite{milo2002network}, which characterize the distinct patterns of interactions between $n$ species found within a network. Following the ideas presented by \citet{stouffer2012evolutionary} and \citet{baker2015species}, we defined the role of any species $a$ based on the number of times it appears in any of the unique positions of the $n$-species network motifs. In particular, we used the tools developed by \citet{mora2018pymfinder} to describe the role of $a$ as measure of the number of times $a$ appears in any of the distinct positions found within motifs made of $3$, $4$ and $5$ species. As described in  \citet{mora2018pymfinder}, this measure is designed in such a way that it can incorporate information regarding the interaction strengths between species.

\subsubsection*{Alignment matrix}
We compiled all alignments between the networks in our dataset using the alignment matrix $M$. In particular, $M$ contains information regarding every pairwise alignment between the networks $\{A,B,C,D...\}$ in our dataset---in our case, this corresponds to the pairwise alignment between 44 empirical networks. Given a set of $n$ alignments $\{\lambda\}$ between $A$ and $B$, we calculate a given element $m^{AB}_{ij}$ as $\sum^{n}_{\lambda} k_{\lambda}$, where $k_{\lambda}$ is 0 if species $i\in A$ and $j\in B$ are not paired in the alignment $\lambda$, and $c_{ij}$ otherwise. Notice that $c_{ij}$ characterizes the quality of the pairing $i$-$j$ between $i\in A$ and $j\in B$ presented in the main text. The quality of the pairing $i$-$j$ is calculated as the Pearson's correlation coefficient between the two species' motif-role profiles\citep{mora2018identifying}. Specifically, this coefficient is equal to $1$ if $i$ and $j$ have equivalent roles, $0$ when there is no correlation between them, and $-1$ if they have opposite roles.

\subsubsection*{Relationship between group and position properties}
To compare the different groups of positions, we studied the three position properties described in the main text using a multivariate autoregressive model. Specifically, we studied species' relative degree, neighbours' average relative degree and neighbours' maximum relative degree using species' position group and year of sampling. To do so, we used the R package \textit{brms}\cite{burkner2017brms} to fit a Bayesian multivariate linear regression. As outcome variables, we included, species' relative degree, neighbours' average relative degree, and neighbours' maximum relative degree. As categorical predictors, we included species' position group and year of sampling. We treated species' identity as a random effect, and added a first-order autoregression term using the week of sampling as a covariate and species' identity for every year as a grouping factor. To fit the Bayesian models, we chose weakly informative normally distributed priors---with mean $\mu=0$ and standard deviation $\sigma=1$---for the population-level effects; the default \textit{brms} half Student-\textit{t} priors with 3 degrees of freedom for the standard deviations of the group-level effects; and, the default \textit{brms} flat priors bounded between -1 and 1 for the autocorrelation parameters. To calculate the effect sizes for the comparisons of the different predictions across position groups, we first drew 1000 samples from the different posterior distributions for a given position group and year using the `pp\_expect' function from the R package \textit{brms}\cite{burkner2017brms}. Then, we used the function `cohen.d' of the R package \textit{effsize}\cite{torchiano2017effsize} to calculate the effect sizes for the comparison of every pair of posterior distributions.

\subsubsection*{Species structural dynamics across groups}
As described in the main text, we examined the movement of pollinator and plant species across the different network positions using different probabilistic models. Given $n$ groups of positions, our models describe a scenario in which every species can be found in $n+2$ possible states $\vec{y}$ at time $t$. These include: $n$ states $y_1\dots y_n$ characterizing a species in each of the different groups of positions; a state $y_{n+1}=y_{\text{pre}}$ describing a species that has not yet entered the network; and a state $y_{n+2}=y_{\text{post}}$ describing a species that has already exited the network.

We estimated the rate of movement between the different states using Bayesian multinomial logistic regressions. In particular, our models considers $n+2$ types of events, describing the transition probability $\Pr\left(y_j|s_1\left(y_i,t\right), s_2\left(y_i,t\right),\ldots,s_{n+2}\left(y_i,t\right)\right)$ of species from any state $y_i$ to any possible state $y_j$ at time $t$ following Eq.~(1) from the main text. To infer the scores $s_k\left(y_i,t\right)$ that determine the resulting probabilities, we fixed one of them to serve as an arbitrary baseline---i.e.\,\,we assigned $s_1\left(y_i,t\right)=0$---and estimated the remaining scores as $n+1$ linear models. The form of these linear models characterizes the type of transition probabilities considered. Specifically, we considered the following:
\begin{itemize}
\item[\textbf{i)}] \textbf{Constant transition probabilities.} For this model, we defined the scores as
\begin{equation}
s_k\left(y_i\right)=\alpha_{k1} + \sum_{l=2}^{n+2} \alpha_{kl}\delta_{il},
\end{equation}
where $\alpha_{kl}$ are the parameters inferred by the model, and $\delta_{il}$ is a Kronecker delta that is set to 1 if $y_l=y_i$, and 0 otherwise. For any parameter $\alpha_{kl}$, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.

\item[\textbf{ii)}] \textbf{Constant transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as
\begin{equation}
s_k\left(y_i, m\right)=\alpha_{k1} + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \alpha_{kl}\delta_{il},
\end{equation}
where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise. Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.

\item[\textbf{iii)}] \textbf{Constant probabilities with sampling seasons as group-specific treatments.}  For this model, we defined the scores as
\begin{equation}
s_k\left(y_i, m\right)=\alpha_{k1} + \sum_{n=2}^{3} \gamma_{k1}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl} + \sum_{n=2}^{3} \gamma_{kl}^n\delta_{mn}\right)\delta_{il},
\end{equation}
where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$ and $\gamma_{kl}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise. Notice that the parameters $\gamma_{kl}^n$ are used to treat the three sampling seasons as temporal replicates. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.

\item[\textbf{iv)}] \textbf{Time-dependent transition probabilities}  This model is the one defined in the main text; therefore, scores are defined following Eq.~(2) of the main text.

\item[\textbf{v)}] \textbf{Time-dependent transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as
\begin{equation}
s_k\left(y_i, m, t\right)=\alpha_{k1} +\beta_{k1} t + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} t\right)\delta_{il},
\end{equation}
where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise. Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.

\item[\textbf{vi)}] \textbf{Time-dependent probabilities with sampling seasons as group-specific treatments} For this model, we defined the scores as
\begin{equation}
s_k\left(y_i, m, t\right)=\alpha_{k1} +\beta_{k1} t + \sum_{n=2}^{3} \gamma_{k1}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} t  + \sum_{n=2}^{3} \gamma_{kl}^n\delta_{mn}\right)\delta_{il},
\end{equation}
where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{kl}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise. Notice that the parameters $\gamma_{kl}^n$ are used to treat the three sampling seasons as temporal replicates. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.

%\item[\textbf{vii)}] \hlc{\textbf{Density-dependent transition probabilities.} For this model, we defined the scores as}
%\begin{equation}
%s_k\left(y_i, N_{t-\Delta t}\right)=\alpha_{k1} +\beta_{k1} N_{t-\Delta t} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} N_{t-\Delta t}\right)\delta_{il},
%\end{equation}
%\hlc{where $N_{t-\Delta t}$ describe the total number of species in the network at time $t-\Delta t$; $\alpha_{kl}$ and $\beta_{kl}$ are the parameters inferred by the model; and $\delta_{il}$ is a Kronecker delta that is set to 1 if $y_l=y_i$, and 0 otherwise. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.}

%\item[\textbf{viii)}] \hlc{\textbf{Density-dependent transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as}
%\begin{equation}
%s_k\left(y_i, m, N_{t-\Delta t}\right)=\alpha_{k1} +\beta_{k1} N_{t-\Delta t} + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} N_{t-\Delta t}\right)\delta_{il},
%\end{equation}
%\hlc{where $m={1,2,3}$ is a factor characterizing the sampling season; $N_{t-\Delta t}$ describe the total number of species in the network at time $t-\Delta t$; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise. Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates. For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.}
\end{itemize}

\subsubsection*{Choosing the priors for the Bayesian models}
As described in the main text, we defined the priors for the different parameters of the models as weakly informative normally distributed priors with mean $\mu=0$ and standard deviation $\sigma=10$. We did not have a preconceived notion of what these priors should be, and more restrictive priors--i.e.\,\,normally distributed with $\mu=0$ and $\sigma=1$---produced qualitatively the same results described in Fig.~4 of the main text. That said, we tested the design of model iv using simulated data, and these tests showed that less informative priors were needed in order to produce good estimates for some transition probability matrices.

\subsubsection*{Effect sizes for the comparison of the different transition probabilities}
To understand the differences between the transition probabilities estimated by the Bayesian models, we compared the corresponding posteriors distributions using Cohen's \textit{d}. Given any two transition probabilities described in the Results section of the main text at time $t$, we drew 1000 samples from their posterior distributions using the `extract.samples' function of the R package \textit{rethinking}\cite{mcelreath2018statisticalWAIC}. Then, we used the function `cohen.d' of the R package \textit{effsize}\cite{torchiano2017effsize} to calculate the effect sizes for the comparison of these two posterior distributions.

\section*{Supplementary Notes}
\setcounter{subsubsection}{0}

\subsubsection*{Supplementary Note 1: Relationship between position uniqueness and species richness}
\citet{mora2018identifying} showed that large food webs ($\gtrapprox50$ species) might be more redundant than smaller ones, as these have a greater probability of including trophically identical species as well as species that participate in just a single interaction. That is, species in larger networks might present lower position uniqueness than species in smaller networks. However, it is unclear to what extent uniqueness depends on species richness or is instead an intrinsic property of large ecological networks. For example, if a network of 10 species has a central node individually connected to the remaining 9 species, each of these 9 species will present very low position uniqueness. In contrast, if a network of 10 species is connected as two non-symmetric 5 species motifs, all 10 species will have high position uniqueness. This argument can be made for any number of species $n$. For the sake of this current study, nevertheless, all species show similar levels of position uniqueness, and if larger networks were significantly more redundant, we would expect to see lower levels of uniqueness in Fig.~2a from the main text. That is, controlling for species richness would not change the observed distribution in Fig.~2a in a way that would change our conclusions.

\subsubsection*{Supplementary Note 2: Relationship between species' relative degree and abundances}
To study the link between species' relative degree and abundance, we used the abundance estimates from \citet{caradonna2017interaction}. For plant species, these were measured by independent counting of flowers and flower heads within a series of transects at the study site. On the other side, weekly pollinator abundances were estimated using the number of observation periods in which these were observed during the collection of plant-pollinator interaction data (see \citealt{caradonna2017interaction} for details). Notice that since the abundance data for pollinator species was estimated from the interaction surveys from which their relative degree was calculated, one should expect the abundance measures to be at least somewhat dependent on the pollinators' observed number of interactions.

We studied the relationship between species' relative degree and abundance using two different models. First, and for the sake of simplicity, we used a linear regression to understand the overall trend of the relationship between these variables. We found a positive relationship between relative degree of plants and pollinators and their estimated abundance (Supplementary Fig.~\ref{sfig:abundance}). The regressions presented $R^2$ values of $0.14$ and $0.34$ for plants and pollinators, respectively. Second, we used the Hill function to fit a more realistic scenario, where the relationship is nonlinear and absent species are expected to present 0 relative degree. In this case, we found the model to fit the data with a residual standard error of $0.2627$ on $388$ degrees of freedom for plants, and a residual standard error of $0.1928$ on $850$ degrees of freedom for pollinators (Supplementary Fig.~\ref{sfig:abundance}). Overall, although we observed a positive relationship between relative degree and abundance of species, there is considerable variation ($85.9\%$ and $66.5\%$ according to the linear regression for plants and pollinators, respectively) in species' relative degree that is not explained by abundance (Supplementary Fig.~\ref{sfig:abundance}). Linear regressions including the year of sampling as an additional fixed effect support this observation, with $R^2$ values of $0.17$ and $0.36$ for plants and pollinators, respectively.

\clearpage

\section*{Supplementary Figures and Tables}

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure2-unweighted.png}
    	  \vspace{0.3cm}
	   \caption[Analysis of species' positions across binary networks]{Analysis of species' positions across binary networks. (a) Uniqueness of the position of plant and pollinator species within networks. (b) Variability of species' position across networks, including only plant and pollinator species that appear in multiple networks during a season.}
      \label{sfig:variability-positions}
\end{figure}

\clearpage

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/brms-pollinators.pdf}
        \vspace{0.3cm}
	   \caption[Relationships between group and position properties]{Relationships between group and position properties. Every panel describes the comparison between the different properties studied (species' relative degree, neighbours' average relative degree and neighbours' maximum relative degree) for every pair of pollinator position groups ($A$, $B$, $C$). Each color and shape characterize the predictions of a different position group. The different columns of every panel describe the comparison of the predictions for every year of sampling. The points characterize the mean values, and the errorbars represent the credible intervals (2.5 and 97.5 percent quantiles).The scores along the x axis of every panel describe the mean effect size for each comparison (calculated as Cohen's \textit{d} using 1000 samples from the different posterior distributions; see Supplementary Methods for further details). Just as an example, the panel in the top-left corner of the figure compares the degree of species from groups $A$ and $B$, showing the predictions for every year of sampling.}
      \label{sfig:multivariate}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-3-2013.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of pollinator positions found for the 2013 sampling season]{Distinguishing properties of the different groups of pollinator positions found for the 2013 sampling season. (a) Number of pollinator species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2013}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-3-2014.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of pollinator positions found for the 2014 sampling season]{Distinguishing properties of the different groups of pollinator positions found for the 2014 sampling season. (a) Number of pollinator species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2014}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-3-2013-plants.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of plant positions found for the 2013 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2013 sampling season. (a) Number of plant species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2013-plants}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-3-2014-plants.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of plant positions found for the 2014 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2014 sampling season. (a) Number of plant species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2014-plants}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-3-2015-plants.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of plant positions found for the 2014 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2014 sampling season. (a) Number of plant species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2015-plants}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure3-modularity-7-2015.pdf}
	  \vspace{0.3cm}
	   \caption[Distinguishing properties of the different groups of species' position found for the 2015 sampling season using the multi-level community detection method]{Distinguishing properties of the different groups of species' position found for the 2015 sampling season using the multi-level community detection method (Table 1; \citealt{blondel2008fast}). (a) Number of pollinator species in each group of positions over time. Each color represents a different group of positions. (b) Different properties summarizing the species forming each group. The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group. The second panel shows the neighbours' average relative degree for the species in each group. The third panel shows the relative degree distribution of species in each group. The colored segments depicted under the distributions characterize the mean of each distribution.}
      \label{sfig:groups2015-modularity7}
\end{figure}

\clearpage

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.9\textwidth]{figures/abundance_degree.png}
	   \caption{Relationship between relative degree and abundance of plants and pollinators. The points represent the information of a plant or pollinator for a given week and year. The lines characterize the fit of a linear regression and the Hill function to the data. The shaded areas describe the confidence intervals around the fitted models. The estimated slopes for the linear regression $\num{7.90e-05}\pm \num{9.90e-06}$ and $\num{2.70e-02}\pm \num{0.13}$ show a positive relationship between the relative degree and abundance of plants and pollinators, respectively. The estimated half-saturation density for the Hill function shows that the relationship vanishes at $453.01\pm 90.07$ flower counts and $5.47\pm 0.25$ pollinator observations.}

      \label{sfig:abundance}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/WAIC-models.pdf}
	  \vspace{0.3cm}
	   \caption[WAIC models]{Comparison of the different models using WAIC values. Each row describes one of the models from the `Species dynamics across groups' section described above. Specifically, each row represents a model considering the following: i) constant transition probabilities; ii) constant transition probabilities accounting for sampling seasons as treatments to the intercept; iii) constant probabilities with sampling seasons as group-specific treatments; iv) time-dependent transition probabilities; v) time-dependent transition probabilities accounting for sampling seasons as treatments to the intercept; and vi) time-dependent probabilities with sampling seasons as group-specific treatments. (a) WAIC comparison of the different models. The filled points are the in-sample deviance of each model, and the open points are the estimated WAIC values. The dark line segments represent the standard error of each WAIC. The vertical line references the model with the lowest WAIC value. (b) Akaike weight for each model used to compare predictive accuracy of the different Bayesian models (see \citealt{mcelreath2018statisticalWAIC}). The open points are the mean Akaike weights and the line segments represent the corresponding standard error. The vertical line references the model with the highest Akaike weight. Note that the Akaike weights use the standard error of the difference between each WAIC and the best performing model to compare the relative predictive accuracy of the models. While the uncertainty in the WAIC values overlap between models, the measure shows that model iv) is the best model when accounting for the correlation between these uncertainties. The graphs were generated using some of the functions from the R package \textit{rethinking}\cite{mcelreath2018statisticalWAIC}.}
      \label{sfig:WAIC}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/effectsizes-pollinators.pdf}
	  \vspace{0.3cm}
	   \caption[Effect sizes for the comparison of the transition probabilities reported in the Fig.~4 of the main text]{Effect sizes for the conclusions reported in the ``Results'' section of main text. Every panel represents the effect sizes for the comparison between the time-dependent transition probabilities of Fig.~4. These are calculated based on differences between means, using Cohen's \textit{d} (see Supplementary Methods for further details). Using the notation from Fig.~4, effect sizes are reported for the following comparisons: (a) `enter in A' relative to `enter in B' and `enter in C'; (b) `stay in A' relative to `A$\to$B', `A$\to$C' and `exit from A'; (c) `exit from A' relative to `A$\to$B' and `A$\to$C'; (d) `A$\to$B' relative to `A$\to$C'; (e) `stay in B' relative to `B$\to$A', `B$\to$C' and `exit from B'; (f) `B$\to$A' relative to `B$\to$C' and `exit from B'; (g) `C$\to$A' relative to `C$\to$B', `Stay in C' and `exit from C'; and (h) `exit from C' relative to `C$\to$B' and `Stay in C'. The different lines characterize the estimated effect size, and the shade of each line characterizes the first and third quantiles. The dotted gray line at 1.20 indicates the threshold for which the effect sizes are considered to be very large\cite{sawilowsky2009new}.}
      \label{sfig:effectsizes}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/new-figure4-modularity-3-plants.pdf}
	  \vspace{0.5cm}
	   \caption[Movement of plant species across position groups]{Movement of plant species across position groups. Every panel in the matrix describes the transition probability between the different groups of positions (see model iv from Supplementary Methods). The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time. The different groups are those presented in Supplementary Figs.~\ref{sfig:groups2013-plants}--\ref{sfig:groups2015-plants}. The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups (the low values reflect the fact that plant species rarely leave the networks). The different lines in the graph represent the results found for every sampling year, where the shade of each line characterizes the first and third quantiles. The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles. Notice that the color is chosen based on the recipient group of the transition probability.}
      \label{sfig:dynamics-plants}
\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/new-figure4-modularity-3-control_indepe.pdf}
	  \vspace{0.5cm}
	   \caption[Movement of pollinator species across position groups estimated using a model with constant transition probabilities]{Movement of pollinator species across position groups estimated using a model with constant transition probabilities (see model i from Supplementary Methods). Every panel in the matrix describes the transition probability between the different groups of positions. The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time. The different groups are those presented in Fig.~3 of the main text. The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups. The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles. Notice that the color is chosen based on the recipient group of the transition probability.}
      \label{sfig:dynamics-constant}
\end{figure}

\clearpage

%\begin{figure}[p]
% \centering
%    \vspace{0.5cm}
%    \includegraphics{figures/new-figure4-modularity-3-control_abundance.pdf}
%	  \vspace{0.5cm}
%	   \caption[Movement of pollinator species across position groups estimated using a model with density-dependent transition probabilities]{Movement of pollinator species across position groups estimated using a model with density-dependent transition probabilities (Supplementary Methods). Every panel in the matrix describes the transition probability between the different groups of positions. The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time. The different groups are those presented in Fig.\ 3 of the main text. The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups. The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles. Notice that the color is chosen based on the recipient group of the transition probability.}
%      \label{sfig:dynamics-abundance}
%\end{figure}

\clearpage

\begin{figure}[p]
  \centering
    \vspace{0.5cm}
    \includegraphics{figures/figure4-modularity-7.pdf}
	  \vspace{0.5cm}
	   \caption[Movement of pollinator species across position groups]{Movement of pollinator species across position groups. Every panel in the matrix describes the transition probability between the different groups of positions (see model iv from Supplementary Methods). The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time. The different groups are those presented in Supplementary Fig.~\ref{sfig:groups2015-modularity7}. The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$, $C$ and $D$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups. The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles. Notice that the color is chosen based on the recipient group of the transition probability.}
      \label{sfig:dynamics-modularity7}
\end{figure}

\clearpage

\begin{figure}[ht]
  \centering
    \vspace{0.5cm}
    \includegraphics[width=0.65\textwidth]{figures/sensitivity.png}
    	  \vspace{0.3cm}
	   \caption{Analysis of the sensitivity of species' relative degree and number of links to random removal of observations. (a) Sensitivity of species' relative degree to 5\%, 10\%, 25\% and 50\% random removal of observations for each different year. (b) Percent loss of links following the random removal of 5\%, 10\%, 25\% and 50\% of observations for each different year. The boxes in both panels characterize the median, and the the first and third quantiles; the line segments indicate the minima and maxima; and the whiskers represent the corresponding outliers.}
      \label{sfig:sensitivity}
\end{figure}

\clearpage

\begin{table}[p]
\begin{center}
\def\arraystretch{1.3}
\input{table1_SI.tex}
\end{center}
\caption[Community detection methods summary]{Community detection methods summary. The first column lists the community detection methods used here. The methods used are the following: greedy optimization (greedy\cite{clauset2004finding}); edge betweenness (betweenness\cite{newman2004finding}); short random walks (random walks\cite{pons2006computing});  map equation optimization (infomap\cite{rosvall2009map}); and, multi-level optimization (louvain\cite{blondel2008fast}). The second and third columns describe the number of pollinator (pol.) and plant (pl.) groups found with each method. The rest of the columns provide a pairwise comparison of the results found with the different methods by means of the normalized mutual information criteria\cite{danon2005comparing}. The methods highlighted are the ones used in the paper.}
\label{stable:difference-community}
\end{table}

\clearpage
\bibliography{references_ab}

\end{document}

### 2021-01-22 manuscript/manuscript.tex
new models

@@
 \title{\vspace{-1cm}
-A fresh perspective on distribution modelling: a bayesian framework to understand the distribution of plant species along an environmental gradient}
+Model simplicity breeds contempt: using simple models to answer basic questions on species' distributions}
 \author[1,*]{\normalsize Bernat Bramon Mora}
@@
 \linenumbers
-%\section*{Abstract}
+\section*{Introduction}
+{\color{gray}
+These are just a bunch of thoughts that I had on how to frame the introduction for the manuscript.
+I want to say something along the lines: \textit{we know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
+However, much less is known about how species' distributions compare to each other.
+Here, we use a conceptually more conservative approach to instead understand and compare basic aspects regarding the shape of species' distribution along environmental gradients.}}
+%\citet{austin2002spatial} writes "\textit{there are three major components in any framework for statistical modelling in plant ecology.
+There needs to be an ecological model, a data model, and a statistical model.
+The ecological model consists of the ecological knowledge and theory to be used or tested in the study.
+The data model consists of the decisions made regarding how the data are collected and how the data will be measured or estimated.
+The statistical model involves the choice of statistical method, error function and significance tests.
+Each model interacts in both obvious and subtle ways with the other models to determine the success of any statistical modelling exercise.}" While a lot of work has been developed for all three components outlined by \citet{austin2002spatial}, increasing emphasis has been put into advancing the data and statistical models.
+Indeed,
+
+Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges.
+However, much less attention is paid to how species' distributions compare to each other.
+
+There is no general agreement the shape of species distributions.
+While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994) assume this to be unimodal and symmetric, some have warned that empirical distributions can take many different forms \citep{austin2002spatial}.
+There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems.
+The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies.
+Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
+Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.
+
+To decide among modelling approaches, we first needs to agree on what we know about the system.
+We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.
+Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions.
+One could point out that we also know that many other factors might influence species' presence/absence---e.g.
+the influence of biotic interactions among species.
+However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions.
+As a result, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e.
+the one with the largest entropy---is that such distribution is a Gaussian.
+
+%There is a growing literature on species' distribution models.
+For the most part, however, these models have often focussed on model performance and understanding the predictive power of certain covariates.
+This models often take a response variable, such as presence/absence or abundance data, and fit it with a set of environmental variables and species' ecological or phylogenetic relationships.
+There...
+The shape of distributions on systems of many species is rarely studied.
+That is, simple ...
+%model comparison... 
+%creating maps of relative suitabilities...
+%extrapolate to new data...
+
+%It will be argued in this paper that neglect of ecological knowledge is a limiting factor in the application of statistical modelling in ecology and conservation planning
+
+%The field has shifted towards understanding the effects of biotic factors on such distributions. And while the theory is advancing quickly, the...
+
+%This is a generalised additive model.
 % Results are presented for both simulated and empirical data.
@@
 \end{equation}
-Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively.
+Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively\footnote[2]{I'll rewrite the likelihood function to an ordered categorical as soon as I get things to work properly with count data.}.
 While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
@@
 Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
-In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.
-A full section for this is required in the Supplementary Information.}.
+In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
 
@@
 \section*{Results}
+%We first use the baseline model to estimate the distribution of all species with at least 10 occurrences using all the data that we have with regard to...
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.9\textwidth]{figures/figure1}
+    	  \vspace{0.3cm}
+	   \caption{Relationship between mean and variance of species' distributions.
+These are the results for the main axis of variation for the climatic data (results for the second axis of variation presented in the Supplementary Fig.~2).}
+      \label{fig:correlation}
+\end{figure}
 
-\section*{Discussion}
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.9\textwidth]{figures/invasive}
+    	  \vspace{0.3cm}
+	   \caption{Are there clear geographical patterns for neophytes and for species with decreasing or increasing abundance?}
+      \label{fig:neophytes}
+\end{figure}
 
+%\section*{Discussion}
+\clearpage
 \bibliography{references}

### 2021-01-22 manuscript/supplementary_information.tex
new models

@@
 \title{\vspace{-1cm} \normalsize Supplementary Information\\\vspace{0.2cm}
-\LARGE Untangling the seasonal dynamics of plant-pollinator communities}
+\LARGE Basic aspects of species' distributions}
 
-\author{\textit{Bramon Mora et al.}}
+\author{\textit{Bramon Mora and Alexander}}
 
@@
 
-\section*{Supplementary Methods}
-%\subsubsection*{Species' structural roles}
-%There are multiple ways to characterize the structural roles of species in a community.
-In this work, we used a definition based on the idea of network motifs\cite{milo2002network}, which characterize the distinct patterns of interactions between $n$ species found within a network.
-Following the ideas presented by \citet{stouffer2012evolutionary} and \citet{baker2015species}, we defined the role of any species $a$ based on the number of times it appears in any of the unique positions of the $n$-species network motifs.
-In particular, we used the tools developed by \citet{mora2018pymfinder} to describe the role of $a$ as measure of the number of times $a$ appears in any of the distinct positions found within motifs made of $3$, $4$ and $5$ species.
-As described in  \citet{mora2018pymfinder}, this measure is designed in such a way that it can incorporate information regarding the interaction strengths between species.
-
-\subsubsection*{Alignment matrix}
-We compiled all alignments between the networks in our dataset using the alignment matrix $M$.
-In particular, $M$ contains information regarding every pairwise alignment between the networks $\{A,B,C,D...\}$ in our dataset---in our case, this corresponds to the pairwise alignment between 44 empirical networks.
-Given a set of $n$ alignments $\{\lambda\}$ between $A$ and $B$, we calculate a given element $m^{AB}_{ij}$ as $\sum^{n}_{\lambda} k_{\lambda}$, where $k_{\lambda}$ is 0 if species $i\in A$ and $j\in B$ are not paired in the alignment $\lambda$, and $c_{ij}$ otherwise.
-Notice that $c_{ij}$ characterizes the quality of the pairing $i$-$j$ between $i\in A$ and $j\in B$ presented in the main text.
-The quality of the pairing $i$-$j$ is calculated as the Pearson's correlation coefficient between the two species' motif-role profiles\citep{mora2018identifying}.
-Specifically, this coefficient is equal to $1$ if $i$ and $j$ have equivalent roles, $0$ when there is no correlation between them, and $-1$ if they have opposite roles.
-
-\subsubsection*{Relationship between group and position properties}
-To compare the different groups of positions, we studied the three position properties described in the main text using a multivariate autoregressive model.
-Specifically, we studied species' relative degree, neighbours' average relative degree and neighbours' maximum relative degree using species' position group and year of sampling.
-To do so, we used the R package \textit{brms}\cite{burkner2017brms} to fit a Bayesian multivariate linear regression.
-As outcome variables, we included, species' relative degree, neighbours' average relative degree, and neighbours' maximum relative degree.
-As categorical predictors, we included species' position group and year of sampling.
-We treated species' identity as a random effect, and added a first-order autoregression term using the week of sampling as a covariate and species' identity for every year as a grouping factor.
-To fit the Bayesian models, we chose weakly informative normally distributed priors---with mean $\mu=0$ and standard deviation $\sigma=1$---for the population-level effects; the default \textit{brms} half Student-\textit{t} priors with 3 degrees of freedom for the standard deviations of the group-level effects; and, the default \textit{brms} flat priors bounded between -1 and 1 for the autocorrelation parameters.
-To calculate the effect sizes for the comparisons of the different predictions across position groups, we first drew 1000 samples from the different posterior distributions for a given position group and year using the `pp\_expect' function from the R package \textit{brms}\cite{burkner2017brms}.
-Then, we used the function `cohen.d' of the R package \textit{effsize}\cite{torchiano2017effsize} to calculate the effect sizes for the comparison of every pair of posterior distributions.
-
-\subsubsection*{Species structural dynamics across groups}
-As described in the main text, we examined the movement of pollinator and plant species across the different network positions using different probabilistic models.
-Given $n$ groups of positions, our models describe a scenario in which every species can be found in $n+2$ possible states $\vec{y}$ at time $t$.
-These include: $n$ states $y_1\dots y_n$ characterizing a species in each of the different groups of positions; a state $y_{n+1}=y_{\text{pre}}$ describing a species that has not yet entered the network; and a state $y_{n+2}=y_{\text{post}}$ describing a species that has already exited the network.
-
-We estimated the rate of movement between the different states using Bayesian multinomial logistic regressions.
-In particular, our models considers $n+2$ types of events, describing the transition probability $\Pr\left(y_j|s_1\left(y_i,t\right), s_2\left(y_i,t\right),\ldots,s_{n+2}\left(y_i,t\right)\right)$ of species from any state $y_i$ to any possible state $y_j$ at time $t$ following Eq.~(1) from the main text.
-To infer the scores $s_k\left(y_i,t\right)$ that determine the resulting probabilities, we fixed one of them to serve as an arbitrary baseline---i.e.\,\,we assigned $s_1\left(y_i,t\right)=0$---and estimated the remaining scores as $n+1$ linear models.
-The form of these linear models characterizes the type of transition probabilities considered.
-Specifically, we considered the following:
-\begin{itemize}
-\item[\textbf{i)}] \textbf{Constant transition probabilities.} For this model, we defined the scores as
-\begin{equation}
-s_k\left(y_i\right)=\alpha_{k1} + \sum_{l=2}^{n+2} \alpha_{kl}\delta_{il},
-\end{equation}
-where $\alpha_{kl}$ are the parameters inferred by the model, and $\delta_{il}$ is a Kronecker delta that is set to 1 if $y_l=y_i$, and 0 otherwise.
-For any parameter $\alpha_{kl}$, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.
-
-\item[\textbf{ii)}] \textbf{Constant transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as
-\begin{equation}
-s_k\left(y_i, m\right)=\alpha_{k1} + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \alpha_{kl}\delta_{il},
-\end{equation}
-where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise.
-Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.
-
-\item[\textbf{iii)}] \textbf{Constant probabilities with sampling seasons as group-specific treatments.}  For this model, we defined the scores as
-\begin{equation}
-s_k\left(y_i, m\right)=\alpha_{k1} + \sum_{n=2}^{3} \gamma_{k1}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl} + \sum_{n=2}^{3} \gamma_{kl}^n\delta_{mn}\right)\delta_{il},
-\end{equation}
-where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$ and $\gamma_{kl}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise.
-Notice that the parameters $\gamma_{kl}^n$ are used to treat the three sampling seasons as temporal replicates.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.
-
-\item[\textbf{iv)}] \textbf{Time-dependent transition probabilities}  This model is the one defined in the main text; therefore, scores are defined following Eq.~(2) of the main text.
-
-\item[\textbf{v)}] \textbf{Time-dependent transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as
-\begin{equation}
-s_k\left(y_i, m, t\right)=\alpha_{k1} +\beta_{k1} t + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} t\right)\delta_{il},
-\end{equation}
-where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise.
-Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.
-
-\item[\textbf{vi)}] \textbf{Time-dependent probabilities with sampling seasons as group-specific treatments} For this model, we defined the scores as
-\begin{equation}
-s_k\left(y_i, m, t\right)=\alpha_{k1} +\beta_{k1} t + \sum_{n=2}^{3} \gamma_{k1}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} t  + \sum_{n=2}^{3} \gamma_{kl}^n\delta_{mn}\right)\delta_{il},
-\end{equation}
-where $m={1,2,3}$ is a factor characterizing the sampling season; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{kl}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise.
-Notice that the parameters $\gamma_{kl}^n$ are used to treat the three sampling seasons as temporal replicates.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.
-
-%\item[\textbf{vii)}] \hlc{\textbf{Density-dependent transition probabilities.} For this model, we defined the scores as}
-%\begin{equation}
-%s_k\left(y_i, N_{t-\Delta t}\right)=\alpha_{k1} +\beta_{k1} N_{t-\Delta t} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} N_{t-\Delta t}\right)\delta_{il},
-%\end{equation}
-%\hlc{where $N_{t-\Delta t}$ describe the total number of species in the network at time $t-\Delta t$; $\alpha_{kl}$ and $\beta_{kl}$ are the parameters inferred by the model; and $\delta_{il}$ is a Kronecker delta that is set to 1 if $y_l=y_i$, and 0 otherwise.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.}
-
-%\item[\textbf{viii)}] \hlc{\textbf{Density-dependent transition probabilities accounting for sampling seasons as treatments to the intercept.} For this model, we defined the scores as}
-%\begin{equation}
-%s_k\left(y_i, m, N_{t-\Delta t}\right)=\alpha_{k1} +\beta_{k1} N_{t-\Delta t} + \sum_{n=2}^{3} \gamma_{k}^n\delta_{mn} + \sum_{l=2}^{n+2} \left(\alpha_{kl}+\beta_{kl} N_{t-\Delta t}\right)\delta_{il},
-%\end{equation}
-%\hlc{where $m={1,2,3}$ is a factor characterizing the sampling season; $N_{t-\Delta t}$ describe the total number of species in the network at time $t-\Delta t$; $\alpha_{kl}$, $\beta_{kl}$ and $\gamma_{k}^n$ are the parameters inferred by the model; and, $\delta_{mn}$ and $\delta_{il}$ are Kronecker deltas that are set to 1 if $m=n$ and $y_l=y_i$, respectively, and 0 otherwise.
-Notice that the parameters $\gamma_{k}^n$ are used to treat the three sampling seasons as temporal replicates.
-For any parameter, we chose weakly informative normally distributed priors, with mean $\mu=0$ and standard deviation $\sigma=10$.}
-\end{itemize}
-
-\subsubsection*{Choosing the priors for the Bayesian models}
-As described in the main text, we defined the priors for the different parameters of the models as weakly informative normally distributed priors with mean $\mu=0$ and standard deviation $\sigma=10$.
-We did not have a preconceived notion of what these priors should be, and more restrictive priors--i.e.\,\,normally distributed with $\mu=0$ and $\sigma=1$---produced qualitatively the same results described in Fig.~4 of the main text.
-That said, we tested the design of model iv using simulated data, and these tests showed that less informative priors were needed in order to produce good estimates for some transition probability matrices.
-
-\subsubsection*{Effect sizes for the comparison of the different transition probabilities}
-To understand the differences between the transition probabilities estimated by the Bayesian models, we compared the corresponding posteriors distributions using Cohen's \textit{d}.
-Given any two transition probabilities described in the Results section of the main text at time $t$, we drew 1000 samples from their posterior distributions using the `extract.samples' function of the R package \textit{rethinking}\cite{mcelreath2018statisticalWAIC}.
-Then, we used the function `cohen.d' of the R package \textit{effsize}\cite{torchiano2017effsize} to calculate the effect sizes for the comparison of these two posterior distributions.
-
-\section*{Supplementary Notes}
-\setcounter{subsubsection}{0}
-
-\subsubsection*{Supplementary Note 1: Relationship between position uniqueness and species richness}
-\citet{mora2018identifying} showed that large food webs ($\gtrapprox50$ species) might be more redundant than smaller ones, as these have a greater probability of including trophically identical species as well as species that participate in just a single interaction.
-That is, species in larger networks might present lower position uniqueness than species in smaller networks.
-However, it is unclear to what extent uniqueness depends on species richness or is instead an intrinsic property of large ecological networks.
-For example, if a network of 10 species has a central node individually connected to the remaining 9 species, each of these 9 species will present very low position uniqueness.
-In contrast, if a network of 10 species is connected as two non-symmetric 5 species motifs, all 10 species will have high position uniqueness.
-This argument can be made for any number of species $n$.
-For the sake of this current study, nevertheless, all species show similar levels of position uniqueness, and if larger networks were significantly more redundant, we would expect to see lower levels of uniqueness in Fig.~2a from the main text.
-That is, controlling for species richness would not change the observed distribution in Fig.~2a in a way that would change our conclusions.
-
-\subsubsection*{Supplementary Note 2: Relationship between species' relative degree and abundances}
-To study the link between species' relative degree and abundance, we used the abundance estimates from \citet{caradonna2017interaction}.
-For plant species, these were measured by independent counting of flowers and flower heads within a series of transects at the study site.
-On the other side, weekly pollinator abundances were estimated using the number of observation periods in which these were observed during the collection of plant-pollinator interaction data (see \citealt{caradonna2017interaction} for details).
-Notice that since the abundance data for pollinator species was estimated from the interaction surveys from which their relative degree was calculated, one should expect the abundance measures to be at least somewhat dependent on the pollinators' observed number of interactions.
-
-We studied the relationship between species' relative degree and abundance using two different models.
-First, and for the sake of simplicity, we used a linear regression to understand the overall trend of the relationship between these variables.
-We found a positive relationship between relative degree of plants and pollinators and their estimated abundance (Supplementary Fig.~\ref{sfig:abundance}).
-The regressions presented $R^2$ values of $0.14$ and $0.34$ for plants and pollinators, respectively.
-Second, we used the Hill function to fit a more realistic scenario, where the relationship is nonlinear and absent species are expected to present 0 relative degree.
-In this case, we found the model to fit the data with a residual standard error of $0.2627$ on $388$ degrees of freedom for plants, and a residual standard error of $0.1928$ on $850$ degrees of freedom for pollinators (Supplementary Fig.~\ref{sfig:abundance}).
-Overall, although we observed a positive relationship between relative degree and abundance of species, there is considerable variation ($85.9\%$ and $66.5\%$ according to the linear regression for plants and pollinators, respectively) in species' relative degree that is not explained by abundance (Supplementary Fig.~\ref{sfig:abundance}).
-Linear regressions including the year of sampling as an additional fixed effect support this observation, with $R^2$ values of $0.17$ and $0.36$ for plants and pollinators, respectively.
-
-\clearpage
-
-\section*{Supplementary Figures and Tables}
-
 \begin{figure}[ht]
@@
     \vspace{0.5cm}
-    \includegraphics{figures/figure2-unweighted.png}
+    \includegraphics[width=1\textwidth]{figures/prior}
     	  \vspace{0.3cm}
-	   \caption[Analysis of species' positions across binary networks]{Analysis of species' positions across binary networks.
-(a) Uniqueness of the position of plant and pollinator species within networks.
-(b) Variability of species' position across networks, including only plant and pollinator species that appear in multiple networks during a season.}
-      \label{sfig:variability-positions}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[ht]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/brms-pollinators.pdf}
-        \vspace{0.3cm}
-	   \caption[Relationships between group and position properties]{Relationships between group and position properties.
-Every panel describes the comparison between the different properties studied (species' relative degree, neighbours' average relative degree and neighbours' maximum relative degree) for every pair of pollinator position groups ($A$, $B$, $C$).
-Each color and shape characterize the predictions of a different position group.
-The different columns of every panel describe the comparison of the predictions for every year of sampling.
-The points characterize the mean values, and the errorbars represent the credible intervals (2.5 and 97.5 percent quantiles).The scores along the x axis of every panel describe the mean effect size for each comparison (calculated as Cohen's \textit{d} using 1000 samples from the different posterior distributions; see Supplementary Methods for further details).
-Just as an example, the panel in the top-left corner of the figure compares the degree of species from groups $A$ and $B$, showing the predictions for every year of sampling.}
-      \label{sfig:multivariate}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-3-2013.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of pollinator positions found for the 2013 sampling season]{Distinguishing properties of the different groups of pollinator positions found for the 2013 sampling season.
-(a) Number of pollinator species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2013}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-3-2014.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of pollinator positions found for the 2014 sampling season]{Distinguishing properties of the different groups of pollinator positions found for the 2014 sampling season.
-(a) Number of pollinator species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2014}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-3-2013-plants.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of plant positions found for the 2013 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2013 sampling season.
-(a) Number of plant species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2013-plants}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-3-2014-plants.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of plant positions found for the 2014 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2014 sampling season.
-(a) Number of plant species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2014-plants}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-3-2015-plants.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of plant positions found for the 2014 sampling season]{Distinguishing properties of the different groups of plant positions found for the 2014 sampling season.
-(a) Number of plant species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2015-plants}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure3-modularity-7-2015.pdf}
-	  \vspace{0.3cm}
-	   \caption[Distinguishing properties of the different groups of species' position found for the 2015 sampling season using the multi-level community detection method]{Distinguishing properties of the different groups of species' position found for the 2015 sampling season using the multi-level community detection method (Table 1; \citealt{blondel2008fast}).
-(a) Number of pollinator species in each group of positions over time.
-Each color represents a different group of positions.
-(b) Different properties summarizing the species forming each group.
-The first panel depicts the relative degree distribution of the best connected neighbor of every species in each group.
-The second panel shows the neighbours' average relative degree for the species in each group.
-The third panel shows the relative degree distribution of species in each group.
-The colored segments depicted under the distributions characterize the mean of each distribution.}
-      \label{sfig:groups2015-modularity7}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[ht]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics[width=0.9\textwidth]{figures/abundance_degree.png}
-	   \caption{Relationship between relative degree and abundance of plants and pollinators.
-The points represent the information of a plant or pollinator for a given week and year.
-The lines characterize the fit of a linear regression and the Hill function to the data.
-The shaded areas describe the confidence intervals around the fitted models.
-The estimated slopes for the linear regression $\num{7.90e-05}\pm \num{9.90e-06}$ and $\num{2.70e-02}\pm \num{0.13}$ show a positive relationship between the relative degree and abundance of plants and pollinators, respectively.
-The estimated half-saturation density for the Hill function shows that the relationship vanishes at $453.01\pm 90.07$ flower counts and $5.47\pm 0.25$ pollinator observations.}
-
-      \label{sfig:abundance}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/WAIC-models.pdf}
-	  \vspace{0.3cm}
-	   \caption[WAIC models]{Comparison of the different models using WAIC values.
-Each row describes one of the models from the `Species dynamics across groups' section described above.
-Specifically, each row represents a model considering the following: i) constant transition probabilities; ii) constant transition probabilities accounting for sampling seasons as treatments to the intercept; iii) constant probabilities with sampling seasons as group-specific treatments; iv) time-dependent transition probabilities; v) time-dependent transition probabilities accounting for sampling seasons as treatments to the intercept; and vi) time-dependent probabilities with sampling seasons as group-specific treatments.
-(a) WAIC comparison of the different models.
-The filled points are the in-sample deviance of each model, and the open points are the estimated WAIC values.
-The dark line segments represent the standard error of each WAIC.
-The vertical line references the model with the lowest WAIC value.
-(b) Akaike weight for each model used to compare predictive accuracy of the different Bayesian models (see \citealt{mcelreath2018statisticalWAIC}).
-The open points are the mean Akaike weights and the line segments represent the corresponding standard error.
-The vertical line references the model with the highest Akaike weight.
-Note that the Akaike weights use the standard error of the difference between each WAIC and the best performing model to compare the relative predictive accuracy of the models.
-While the uncertainty in the WAIC values overlap between models, the measure shows that model iv) is the best model when accounting for the correlation between these uncertainties.
-The graphs were generated using some of the functions from the R package \textit{rethinking}\cite{mcelreath2018statisticalWAIC}.}
-      \label{sfig:WAIC}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/effectsizes-pollinators.pdf}
-	  \vspace{0.3cm}
-	   \caption[Effect sizes for the comparison of the transition probabilities reported in the Fig.~4 of the main text]{Effect sizes for the conclusions reported in the ``Results'' section of main text.
-Every panel represents the effect sizes for the comparison between the time-dependent transition probabilities of Fig.~4.
-These are calculated based on differences between means, using Cohen's \textit{d} (see Supplementary Methods for further details).
-Using the notation from Fig.~4, effect sizes are reported for the following comparisons: (a) `enter in A' relative to `enter in B' and `enter in C'; (b) `stay in A' relative to `A$\to$B', `A$\to$C' and `exit from A'; (c) `exit from A' relative to `A$\to$B' and `A$\to$C'; (d) `A$\to$B' relative to `A$\to$C'; (e) `stay in B' relative to `B$\to$A', `B$\to$C' and `exit from B'; (f) `B$\to$A' relative to `B$\to$C' and `exit from B'; (g) `C$\to$A' relative to `C$\to$B', `Stay in C' and `exit from C'; and (h) `exit from C' relative to `C$\to$B' and `Stay in C'.
-The different lines characterize the estimated effect size, and the shade of each line characterizes the first and third quantiles.
-The dotted gray line at 1.20 indicates the threshold for which the effect sizes are considered to be very large\cite{sawilowsky2009new}.}
-      \label{sfig:effectsizes}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/new-figure4-modularity-3-plants.pdf}
-	  \vspace{0.5cm}
-	   \caption[Movement of plant species across position groups]{Movement of plant species across position groups.
-Every panel in the matrix describes the transition probability between the different groups of positions (see model iv from Supplementary Methods).
-The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time.
-The different groups are those presented in Supplementary Figs.~\ref{sfig:groups2013-plants}--\ref{sfig:groups2015-plants}.
-The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups (the low values reflect the fact that plant species rarely leave the networks).
-The different lines in the graph represent the results found for every sampling year, where the shade of each line characterizes the first and third quantiles.
-The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles.
-Notice that the color is chosen based on the recipient group of the transition probability.}
-      \label{sfig:dynamics-plants}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/new-figure4-modularity-3-control_indepe.pdf}
-	  \vspace{0.5cm}
-	   \caption[Movement of pollinator species across position groups estimated using a model with constant transition probabilities]{Movement of pollinator species across position groups estimated using a model with constant transition probabilities (see model i from Supplementary Methods).
-Every panel in the matrix describes the transition probability between the different groups of positions.
-The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time.
-The different groups are those presented in Fig.~3 of the main text.
-The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups.
-The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles.
-Notice that the color is chosen based on the recipient group of the transition probability.}
-      \label{sfig:dynamics-constant}
-\end{figure}
-
-\clearpage
-
-%\begin{figure}[p]
-% \centering
-%    \vspace{0.5cm}
-%    \includegraphics{figures/new-figure4-modularity-3-control_abundance.pdf}
-%	  \vspace{0.5cm}
-%	   \caption[Movement of pollinator species across position groups estimated using a model with density-dependent transition probabilities]{Movement of pollinator species across position groups estimated using a model with density-dependent transition probabilities (Supplementary Methods).
-Every panel in the matrix describes the transition probability between the different groups of positions.
-The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time.
-The different groups are those presented in Fig.\ 3 of the main text.
-The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$ and $C$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups.
-The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles.
-Notice that the color is chosen based on the recipient group of the transition probability.}
-%      \label{sfig:dynamics-abundance}
-%\end{figure}
-
-\clearpage
-
-\begin{figure}[p]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics{figures/figure4-modularity-7.pdf}
-	  \vspace{0.5cm}
-	   \caption[Movement of pollinator species across position groups]{Movement of pollinator species across position groups.
-Every panel in the matrix describes the transition probability between the different groups of positions (see model iv from Supplementary Methods).
-The order of the matrix is such that it characterizes the transition probabilities from row groups to column groups over time.
-The different groups are those presented in Supplementary Fig.~\ref{sfig:groups2015-modularity7}.
-The shaded row describes the probabilities of species entering the network into the different groups ($A$, $B$, $C$ and $D$) whereas the shaded column describes the probabilities of species exiting the network from each of these groups.
-The different lines in the graph represent the mean transition probabilities and the shade of each line characterizes the first and third quantiles.
-Notice that the color is chosen based on the recipient group of the transition probability.}
-      \label{sfig:dynamics-modularity7}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[ht]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics[width=0.65\textwidth]{figures/sensitivity.png}
-    	  \vspace{0.3cm}
-	   \caption{Analysis of the sensitivity of species' relative degree and number of links to random removal of observations.
-(a) Sensitivity of species' relative degree to 5\%, 10\%, 25\% and 50\% random removal of observations for each different year.
-(b) Percent loss of links following the random removal of 5\%, 10\%, 25\% and 50\% of observations for each different year.
-The boxes in both panels characterize the median, and the the first and third quantiles; the line segments indicate the minima and maxima; and the whiskers represent the corresponding outliers.}
+	   \caption{Study of the priors for the baseline model presented in the main text.}
       \label{sfig:sensitivity}
@@
 
-\begin{table}[p]
-\begin{center}
-\def\arraystretch{1.3}
-\input{table1_SI.tex}
-\end{center}
-\caption[Community detection methods summary]{Community detection methods summary.
-The first column lists the community detection methods used here.
-The methods used are the following: greedy optimization (greedy\cite{clauset2004finding}); edge betweenness (betweenness\cite{newman2004finding}); short random walks (random walks\cite{pons2006computing});  map equation optimization (infomap\cite{rosvall2009map}); and, multi-level optimization (louvain\cite{blondel2008fast}).
-The second and third columns describe the number of pollinator (pol.) and plant (pl.) groups found with each method.
-The rest of the columns provide a pairwise comparison of the results found with the different methods by means of the normalized mutual information criteria\cite{danon2005comparing}.
-The methods highlighted are the ones used in the paper.}
-\label{stable:difference-community}
-\end{table}
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/figure1-secondaxis}
+    	  \vspace{0.3cm}
+	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
+      \label{sfig:sensitivity}
+\end{figure}
 
 \clearpage
+
 \bibliography{references_ab}

### 2021-01-22 code/paper-models/main.R

@@ -1,0 +1,255 @@
+# source("./prepare-data.R")
+source("./models.R")
+library(rethinking)
+library(rstan)
+
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+
+####
+## Run species distribution model for all species. I wrote the following models:
+# - baseline model
+####
+
+baseline <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+        # Fixing some of the options
+        variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
+        gp_type <- 2
+        ndim <- 2
+        pca <- T
+        
+        # File name extension
+        extension <- ""
+        
+        # If we are dealing with simulated data
+        if(simulated){
+                extension <- "-simulated"
+        }
+        extension <- paste(extension, as.character(gp_type), sep="")
+        
+        # Load the data
+        if(is.null(d)){
+                if(recompile){
+                        d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="gauss.gauss", min.occurrence=min.occurrence)
+                        # rename variables
+                        if(simulated){
+                                variables <- c("S1", "S2")
+                        }else{
+                                if(pca){
+                                        variables <- paste("PC", 1:ndim, sep="")
+                                }
+                        }
+                        filename <- paste("../../data/processed/jsdm/", paste(variables, collapse = ""), "data.rds", sep = "")
+
+                        if (file.exists(filename)){
+                                question <- askYesNo("Do you want to overwrite the file?", default = F, 
+                                                     prompts = getOption("askYesNo", gettext(c("Yes", "No", "Cancel"))))
+                                if(is.na(question)){question <- F}
+                                if(question){
+                                        saveRDS(d, file=filename)
+                                }else{
+                                        stop("you should add 'recompile=T'")
+                                }                                
+                        }else{
+                                saveRDS(d, file=filename)
+                        }
+                }else{
+                        # rename variables
+                        if(simulated){
[…]

### 2021-01-22 code/paper-models/models.R

@@ -1,0 +1,223 @@
+# GAM with binary data
+
+# This is the same as model 3.1 and 3.2 but much faster. 
+# The structure of the model is very different but it helps me vectorize the operations and reduce the size of the objects used
+base.model <- "
+functions{
+    matrix cov_GPL2(matrix x, real sq_alpha, real sq_rho, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+        for (i in 1:(N-1)) {
+          K[i, i] = sq_alpha + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = sq_alpha * exp(-sq_rho * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = sq_alpha + delta;
+        return K;
+    }
+}
+data{
+    int N;
+    int L;
+    int K;
+    int Y[L,N];
+    row_vector[N] X1;
+    row_vector[N] X2;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+
+}
+parameters{
+    vector[L] zalpha;
+    matrix[K,L] zbeta;
+    matrix[K,L] zgamma;
+    real alpha_bar;
+    vector[K] beta_bar;
+    vector[K] gamma_bar;
+    real<lower=0> sigma_a;
+    vector<lower=0>[K] sigma_b;
+    vector<lower=0>[K] sigma_g;
+    vector<lower=0>[K] etasq_b;
+    vector<lower=0>[K] rhosq_b;
+    vector<lower=0>[K] etasq_g;
+    vector<lower=0>[K] rhosq_g;
+}
+transformed parameters{
+    vector[L] alpha;
+    matrix[K,L] beta;
+    matrix[K,L] gamma;
+    matrix[L, L] L_SIGMA_b[K];
+    matrix[L, L] L_SIGMA_g[K];
+    for(i in 1:K){
+        L_SIGMA_b[i] = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b[i], rhosq_b[i], sigma_b[i]));
+        beta[i] = zbeta[i]*(L_SIGMA_b[i]') + beta_bar[i];
+    }
+    for(i in 1:K){
+        L_SIGMA_g[i] = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g[i], rhosq_g[i], sigma_g[i]));
[…]

### 2021-01-22 code/paper-models/run_inte.sh

@@ -1,0 +1,3 @@
+#!/usr/bin/env bash
+
+bsub -R "rusage[mem=4548]" -W 160:00 -n 15 "R --vanilla --slave < main.R"

### 2021-02-04 manuscript/manuscript.tex
Fixing convergence problem and figures

@@
 \section*{Introduction}
+Accounting for expert knowledge on species' environmental preferences to understand general distribution patterns.
+
+The use of species distribution models has grown a lot.
+These models try to estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
+Recent work on these modelling approaches has increasingly focused on estimating (ref) and accounting for (ref) biotic factors, such as competitive or facilitative relationships.
+The idea is that by understanding how all such factors shape species' distribution we will gain a mechanistic understanding on how these distributions are established and change over time.
+Unfortunately, while some of this approaches will certainly increase the predictive performance of distribution models (ref), the nature of some of the estimates have been shown to theoretically have some mishaps (ref).
+
 {\color{gray}
@@
 %\section*{Introduction}
+
+%%%%%%%%%%%%%%%%%%%%%%%
+% Paragraph on Rapoport's rule
+%%%%%%%%%%%%%%%%%%%%%%%
+% McCain
+%Understanding the factors that determine species range size dis- tributions is becoming increasingly urgent as more climate change assessments document the heightened risk for small- ranged species (e.g.
+Channell & Lomolino, 2000; Sekercioglu et al., 2008; La Sorte & Jetz, 2010; McCain & Colwell, 2011).
+%Rapoport’s rule is the positive relationship of species range sizes with increasing latitude (Stevens, 1989), elevation (Stevens, 1992) or water depth (Stevens, 1996).
+The latitudinal Rapoport’s rule (LRR) is the most examined in the literature, including testing empirical datasets (e.g.
+Rohde et al., 1993; Roy et al., 1994; Blackburn & Gaston, 1996; Lyons & Willig, 1997; Ruggiero & Lawton, 1998; Reed, 2003; Arita et al., 2005; Hausdorf, 2006), simulation modelling of the expected patterns (Colwell & Hurtt, 1994; Taylor & Gaines, 1999; Case & Taper, 2000; Arita, 2005; Stauffer & Rohde, 2006; Šizling et al., 2009), and theoretical and empirical evidence for the proposed mechanisms (e.g.
+Rohde, 1992; Kerr, 1999; Taylor & Gaines, 1999; Gaston & Chown, 1999a; Addo-Bediako et al., 2000; Parmesan et al., 2005).
+The various reviews of the LRR to date suggest that the overall support is weak (e.g.
+Rohde, 1996; Gaston et al., 1998; Rohde, 1999; Gaston & Chown, 1999b; Ribas & Schoereder, 2006), prin- cipally due to the high degree of variability in the fit to predic- tions (e.g.
+support: Blackburn & Gaston, 1996; Lyons & Willig, 1997; Price et al., 1997; Arita et al., 2005; no support: Rohde et al., 1993; Roy et al., 1994; Ruggiero & Lawton, 1998; Reed, 2003).
+%The elevational Rapoport’s rule (ERR) has received less examination in the literature than the LRR, but there have also been empirical tests of its predictions (e.g.
+Patterson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Sanders, 2002; Fu et al., 2004; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Bhattarai & Vetaas, 2006; Hausdorf, 2006) and underlying theory (e.g.
+Rohde, 1996; Fleishman et al., 1998; Gaston & Chown, 1999a; Hausdorf, 2006; Ribas & Schoereder, 2006).
+In the ERR, like the LRR, there is a high degree of variability in support from supportive (e.g.
+Pat- terson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Sanders, 2002; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Hausdorf, 2006; Ribas & Schoereder, 2006) to little or no support (e.g.
+Patterson et al., 1996; Price et al., 1997; Rahbek, 1997; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Fu et al., 2004; Bhattarai & Vetaas, 2006; Hausdorf, 2006; Ribas & Schoereder, 2006).
+Similarly, the processes proposed to underlie the ERR (Stevens, 1992) also show a variable amount of theoretical and empirical support (Patterson et al., 1996; Rohde, 1996; Rahbek, 1997; Fleishman et al., 1998; Gaston & Chown, 1999a; Almeida-Neto et al., 2006; Hausdorf, 2006; Colwell, 2011).
+Scarce data and little to no attempt to account for uncertainty in the predictions.
+Similar to rapoport's rule, we can also ask other questions regarding general geographical patterns of species distributions.
 

### 2021-02-04 code/paper-models/main.R

@@ -212,21 +212,15 @@ baseline.traits <- function(d = NULL, recompile = T, simulated=T, min.occurrence
                 alpha_bar = 0,
                 beta_bar = rep(0, dat_5.1$K),
                 gamma_bar = rep(0, dat_5.1$K),
-                zsigma_a = 0.1,
-                zetasq_a = 0,
-                zrhosq_a = 0,
-                zsigma_b = rep(0.1, dat_5.1$K),
-                zetasq_b = rep(0, dat_5.1$K),
-                zrhosq_b = matrix(0, dat_5.1$K, 2),
-                zsigma_g = rep(0.1, dat_5.1$K),
-                zetasq_g = rep(0, dat_5.1$K),
-                zrhosq_g = matrix(0, dat_5.1$K, 2),
-                sigma_sigma = 0.1,
-                sigma_mean = 0,
-                rho_sigma = 0.1,
-                rho_mean = 0,
-                eta_sigma = 0.1,
-                eta_mean = 0
+                sigma_a = 0.1,
+                etasq_a = 0.1,
+                rhosq_a = 0.1,
+                sigma_b = rep(0.1, dat_5.1$K),
+                etasq_b = rep(0.1, dat_5.1$K),
+                rhosq_b = matrix(0.1, dat_5.1$K, 2),
+                sigma_g = rep(0.1, dat_5.1$K),
+                etasq_g = rep(0.1, dat_5.1$K),
+                rhosq_g = matrix(0.1, dat_5.1$K, 2)
         )
         
         model_code=base.model.traits

### 2021-02-04 code/paper-models/models.R

@@ -134,29 +134,6 @@ parameters{
     real alpha_bar;
     vector[K] beta_bar;
     vector[K] gamma_bar;
-    real zsigma_a;
-    real zetasq_a;
-    real zrhosq_a;
-    vector[K] zsigma_b;
-    vector[K] zetasq_b;
-    matrix[K,2] zrhosq_b;
-    vector[K] zsigma_g;
-    vector[K] zetasq_g;
-    matrix[K,2] zrhosq_g;
-    real<lower=0> sigma_sigma;
-    real sigma_mean;
-    real<lower=0> rho_sigma;
-    real rho_mean;
-    real<lower=0> eta_sigma;
-    real eta_mean;
-}
-transformed parameters{
-    vector[L] alpha;
-    matrix[K,L] beta;
-    matrix[K,L] gamma;
-    matrix[L, L] L_SIGMA_a;
-    matrix[L, L] L_SIGMA_b[K];
-    matrix[L, L] L_SIGMA_g[K];
     real sigma_a;
     real etasq_a;
     real rhosq_a;
@@ -166,16 +143,14 @@ transformed parameters{
     vector[K] sigma_g;
     vector[K] etasq_g;
     matrix[K,2] rhosq_g;
-    
-    sigma_a = exp(zsigma_a * sigma_sigma + sigma_mean);
-    etasq_a = exp(zetasq_a * eta_sigma + eta_mean);    
-    rhosq_a = exp(zrhosq_a * rho_sigma + rho_mean);
-    sigma_b = exp(zsigma_b * sigma_sigma + sigma_mean);
-    etasq_b = exp(zetasq_b * eta_sigma + eta_mean);    
-    rhosq_b = exp(zrhosq_b * rho_sigma + rho_mean);
-    sigma_g = exp(zsigma_g * sigma_sigma + sigma_mean);
-    etasq_g = exp(zetasq_g * eta_sigma + eta_mean);    
-    rhosq_g = exp(zrhosq_g * rho_sigma + rho_mean);    
+}
+transformed parameters{
+    vector[L] alpha;
+    matrix[K,L] beta;
+    matrix[K,L] gamma;
+    matrix[L, L] L_SIGMA_a;
+    matrix[L, L] L_SIGMA_b[K];
+    matrix[L, L] L_SIGMA_g[K];
 
     L_SIGMA_a = cholesky_decompose(cov_GPL2_alpha( Dmat_t, etasq_a, rhosq_a, sigma_a));
[…]

### 2021-02-04 code/paper-models/visualizing-results.R

@@ -94,13 +94,17 @@ plot.scatter <- function(mu, variance, color="black", xlabel="-", ylabel="-", ti
   return(p)
 }
 
-plot.scatter2 <- function(mu, variance, label, color="black", xlabel="-", ylabel="-", tit=NULL, mar=margin(5.5,5.5,5.5,5.5)){
+plot.scatter2 <- function(mu, variance, label, color="black", alpha=NULL, xlabel="-", ylabel="-", tit=NULL, mar=margin(5.5,5.5,5.5,5.5)){
   
-  df <- data.frame(sp=1:length(mu), mean=mu, variance=variance, label=label)
+  if(is.null(alpha)){
+    alpha <- rep(0.7, length(mu))
+  }
+  df <- data.frame(sp=1:length(mu), mean=mu, variance=variance, label=label, alpha=alpha)
   
-  p <- ggplot(df, aes(x=variance, y=mean, color=label)) + 
-    geom_point(alpha=0.7)+
+  p <- ggplot(df, aes(x=variance, y=mean, color=label, alpha=alpha)) + 
+    geom_point()+
     scale_color_manual(values=color)+
+    guides(alpha = FALSE)+
     ylab(ylabel)+
     xlab(xlabel)+
     scale_x_continuous(expand = expansion(add = c(0, 0)))+
@@ -242,22 +246,27 @@ plot.actual.data <- function(model=NULL){
     ylim = c(min(ci_beta[[i]]), max(ci_beta[[i]]))
     xlim = c(-1, length(mu_beta[[i]])+1)
     posy = (ylim[2]-ylim[1])*0.1 + ylim[1]
-    posx = (xlim[2]-xlim[1])*0.796 + xlim[1]
+    posx = (xlim[2]-xlim[1])*0.8 + xlim[1]
     
     # Beta
-    plist[[1]] <- plot.ranking.x(mu_beta[[i]], ci_beta[[i]], color=colo[1], xlabel="species", ylabel="", ylim=ylim, xlim=xlim, mar=margin(5.5,5.5,5.5,5.5), additional_label = Tind_, posx=posx, posy=c(posy, ylim[2]-(posy-ylim[1])))
[…]

### 2021-02-09 manuscript/manuscript.tex
1d simulation getting ready to rock

@@
 \linenumbers
+
+\section*{Abstract}
+We know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
+However, much less is known about how species' distributions compare to each other.
+Here, we use a conceptually more conservative approach to instead understand and compare basic aspects regarding the shape of species' distribution along environmental gradients.
+
 \section*{Introduction}
-Accounting for expert knowledge on species' environmental preferences to understand general distribution patterns.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PARAGRAPH 1
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+One of the central goals of ecology is to understand the ways species are distributed across space (ref).
+Over the years, researchers have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions (ref).
+These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
+Moreover, recent work on these modelling approaches has focused on estimating (ref) and accounting for (ref) biotic factors, such as competitive or facilitative relationships.
+Overall, these methodological advances have provided with crucial ecological insights into the factors shaping species' distributions (ref); however, they have also faced some criticism as a result of the...
+For example, ...
 
@@
 
-{\color{gray}
-These are just a bunch of thoughts that I had on how to frame the introduction for the manuscript.
-I want to say something along the lines: \textit{we know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
-However, much less is known about how species' distributions compare to each other.
-Here, we use a conceptually more conservative approach to instead understand and compare basic aspects regarding the shape of species' distribution along environmental gradients.}}
+%Ecology has been described as the scientific understanding of factors determining the abundance and distribution of species (Smith 1966; Begon et al.
+1986).
+This understanding can hardly be achieved by studying species one by one since their abundances and distributions depend not only on their indi- vidual responses to the abiotic environment, but also on their interactions (Wisz et al.
+2013).
+Thus, a key aim in modern community ecology is to gain an integrative understanding of how biotic and abiotic factors mould local species pools at different spatiotemporal scales.
+Community ecology began as a descriptive science in which communities were classified based on the identities and sizes of local species pools (e.g.
+Clements 1936; Elton 1966).
+Mod- ern community ecology is progressing from the description of patterns towards a mechanistic perspective, which seeks to understand the processes determining the identities and abundances of the species from local to global spatiotemporal scales (Agrawal et al.
+2007; Logue et al.
+2011).
+During the last few decades, experimental ecologist have used observations and experiments to assess the relative influences of stochastic- ity, competition and niche differentiation (see Logue et al.
+2011), theoretical ecologists have developed models for pre- dicting community dynamics (e.g.
+Tilman 1990, 2004; Holt et al.
+1994; Bolker et al.
+2003; Leibold et al.
+2004; Holyoak et al.
+2005), and statistical ecologists have developed metrics for assessing compositional changes among local communities (e.g.
+Gauch 1982; ter Braak & Prentice 1988; Legendre & Legendre 2012).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PARAGRAPH 2
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges (ref).
+Accurately predicting how species are distributed across ranges, it is crucial for understanding the impacts and effect of global climate change.
+With that said, accounting for the mechanisms can come at the cost of modelling noise...
+However, much less attention is paid to how species' distributions compare to each other.
+
 %\citet{austin2002spatial} writes "\textit{there are three major components in any framework for statistical modelling in plant ecology.
@@
 However, much less attention is paid to how species' distributions compare to each other.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PARAGRAPH 3
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+A lot can be learned from the basic properties of species' realized niches.
+
+Accounting for expert knowledge on species' environmental preferences to understand general distribution patterns.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PARAGRAPH 4
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Here, we...Bayesian framework... This allow us to account for... as well as to tackle long-standing hypothesis regarding basic aspects of species distributions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% PARAGRAPH 5
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+In this work, we first....
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%   Stuff to realocate   %%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2021-02-09 code/paper-models/main.R

@@ -121,7 +121,7 @@ baseline <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, of
                            init=init_4.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
         
         
-        saveRDS(mfit_4.1, file = paste(ofolder, "baseline-model.rds", sep=""))
+        saveRDS(mfit_4.1, file = paste(ofolder, "baseline-model", extension,".rds", sep=""))
         return(mfit_4.1)
 }
 
@@ -239,11 +239,133 @@ baseline.traits <- function(d = NULL, recompile = T, simulated=T, min.occurrence
                            init=init_5.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
         
         
-        saveRDS(mfit_5.1, file = paste(ofolder, "baseline-model-traits.rds", sep=""))
+        saveRDS(mfit_5.1, file = paste(ofolder, "baseline-model-traits", extension,".rds", sep=""))
         return(mfit_5.1)
 }
 
+baseline.traits.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+        # Fixing some of the options
+        variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
+        gp_type <- 2
+        ndim <- 2
+        pca <- T
+        
+        # File name extension
+        extension <- ""
+        
+        # If we are dealing with simulated data
+        if(simulated){
+                extension <- "-1d-simulated"
+        }
+        extension <- paste(extension, as.character(gp_type), sep="")
+        
+        if(min.occurrence==10){
+                extension2 <- ""
+        }else{
+                extension2 <- "min30-"
+        }
+        
+        
+        # Load the data
+        if(is.null(d)){
+                if(recompile){
+                        d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="gauss.gauss", min.occurrence=min.occurrence)
+                        # rename variables
+                        if(simulated){
+                                variables <- c("S1", "S2")
+                        }else{
+                                if(pca){
+                                        variables <- paste("PC", 1:ndim, sep="")
+                                }
+                        }
+                        filename <- paste("../../data/processed/jsdm/", extension, paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+                        
+                        if (file.exists(filename)){
[…]

### 2021-02-09 code/paper-models/models.R

@@ -134,15 +134,15 @@ parameters{
     real alpha_bar;
     vector[K] beta_bar;
     vector[K] gamma_bar;
-    real sigma_a;
-    real etasq_a;
-    real rhosq_a;
-    vector[K] sigma_b;
-    vector[K] etasq_b;
-    matrix[K,2] rhosq_b;
-    vector[K] sigma_g;
-    vector[K] etasq_g;
-    matrix[K,2] rhosq_g;
+    real<lower=0> sigma_a;
+    real<lower=0> etasq_a;
+    real<lower=0> rhosq_a;
+    vector<lower=0>[K] sigma_b;
+    vector<lower=0>[K] etasq_b;
+    matrix<lower=0>[K,2] rhosq_b;
+    vector<lower=0>[K] sigma_g;
+    vector<lower=0>[K] etasq_g;
+    matrix<lower=0>[K,2] rhosq_g;
 }
 transformed parameters{
     vector[L] alpha;
@@ -191,3 +191,215 @@ model{
 }
 "
 
+# This is the same as model 3.1 and 3.2 but much faster. 
+# The structure of the model is very different but it helps me vectorize the operations and reduce the size of the objects used
+base.model.traits.1d <- "
+functions{
+    matrix cov_GPL2(matrix x, matrix y, real a, real b1, real b2, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b1 * square(x[i,j]) -b2 * square(y[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+    matrix cov_GPL2_alpha(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
[…]

### 2021-02-09 code/paper-models/prepare-data.R

@@ -6,6 +6,8 @@ library("rJava")
 library(SDMTools)
 source("../prepare-data/useful-tools.R")
 library(MASS)
+library(gridExtra)
+library(grid)
 
 # Define projection as global variable
 projection <- "+proj=somerc +init=world:CH1903"
@@ -104,15 +106,13 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
 simulated.data <- function(simulated.type="linear.corr"){
         
         # Define system dimensions
-        N <- 20
-        sites <- 100
+        N <- 50
+        sites <- 300
         
         # Environmental predictors for each site
         e1 <- rnorm(sites)
-        e2 <- rnorm(sites)
-        
+
         # uncorrelated coefficients for each species and parameters
-        alpha <- rnorm(N)
         sigma1 <- 0.3 # sd beta1
         mean1 <- -1 # mean beta1
         sigma2 <- 0.4 # sd beta2
@@ -124,76 +124,33 @@ simulated.data <- function(simulated.type="linear.corr"){
         s <- 0.5
         Dis <- (as.matrix(dist(1:N))/N)
         
-        if(simulated.type=="linear.corr"){
-                z1 <- rnorm(N)
-                z2 <- rnorm(N)
+        # coefficients for each species
+        Sigma <- nu*exp(-1/(s*s)*(Dis^2)) + diag(N)*sigma1*sigma1
+        z1 <- mvrnorm(mu = rep(mean1, times = N), Sigma = Sigma)
 
-                # Generate correlations
-                beta1 <- z1 * sigma1 + mean1
-                beta2 <- (rho*z1 + sqrt(1-rho^2)*z2)*sigma2 + mean2
-                
-        }else if (simulated.type=="linear.corr.gauss"){
-                
-                # coefficients for each species
-                Sigma <- nu*exp(-1/(s*s)*(Dis^2)) + diag(N)*sigma1*sigma1
-                z1_ <- mvrnorm(mu = rep(mean1, times = N), Sigma = Sigma)
-                Sigma <- nu*exp(-1/(s*s)*(Dis^2)) + diag(N)*sigma2*sigma2
-                z2_ <- mvrnorm(mu = rep(mean2, times = N), Sigma = Sigma)
-                
-                # remove correlation
-                rho_ <- cor(z1_, z2_)
-                z1 <- (z1_-mean(z1_))/sd(z1_)
[…]

### 2021-02-10 manuscript/manuscript.tex
Trying with more species because I don't know what else to do

@@
 These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
-Moreover, recent work on these modelling approaches has focused on estimating (ref) and accounting for (ref) biotic factors, such as competitive or facilitative relationships.
-Overall, these methodological advances have provided with crucial ecological insights into the factors shaping species' distributions (ref); however, they have also faced some criticism as a result of the...
+Recent work on these approaches has focused on estimating (ref) and accounting for (ref) biotic factors, such as competitive or facilitative relationships.
+The idea is that by understanding how such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
+Unfortunately, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
+For example...
+
+%Overall, these methodological advances have provided with crucial ecological insights into the factors shaping species' distributions (ref); however, they have also faced some criticism as a result of the...
 For example, ...
-
-The use of species distribution models has grown a lot.
+%
+%The use of species distribution models has grown a lot.
 These models try to estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
@@
 Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges (ref).
-Accurately predicting how species are distributed across ranges, it is crucial for understanding the impacts and effect of global climate change.
+Accurately predicting how species are distributed across ranges is crucial for understanding the impacts and effect of global climate change.
 With that said, accounting for the mechanisms can come at the cost of modelling noise...
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-A lot can be learned from the basic properties of species' realized niches.
+While the value of gaining a mechanistic understanding of species' distributions is unquestioned, a lot can be learned from shedding light on basic properties of species' realized niches.
 

### 2021-02-10 code/paper-models/models.R

@@ -221,8 +221,8 @@ data{
 }
 parameters{
     vector[L] zalpha;
-    row_vector[L] zbeta;
-    row_vector[L] zgamma;
+    vector[L] zbeta;
+    vector[L] zgamma;
     real alpha_bar;
     real beta_bar;
     real gamma_bar;
@@ -236,19 +236,20 @@ parameters{
 }
 transformed parameters{
     vector[L] alpha;
-    row_vector[L] beta;
-    row_vector[L] gamma;
+    vector[L] beta;
+    vector[L] gamma;
     matrix[L, L] L_SIGMA_b;
     matrix[L, L] L_SIGMA_g;
 
     alpha = exp(zalpha * sigma_a + alpha_bar);
 
     L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
-    beta = zbeta * (L_SIGMA_b') + beta_bar;
+    beta = L_SIGMA_b * zbeta + beta_bar;
 
     L_SIGMA_g = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g, rhosq_g, sigma_g));
-    gamma = zgamma * (L_SIGMA_g') + gamma_bar;
+    gamma = L_SIGMA_g * zgamma + gamma_bar;
     gamma = exp(gamma);
+    
 
 }
 model{
@@ -696,8 +697,7 @@ model{
 
     for ( i in 1:L ){
         p[i] = (X1 - beta[i]) * gamma[i];
-        p[i] = 0.5 * exp(-alpha[i] - columns_dot_self(p[i])) * (1 + erf(lambda * p[i] / sqrt2()));
-        Y[i] ~ binomial(1, p[i]);
+        Y[i] ~ binomial(1, 0.5 * exp(-alpha[i] - columns_dot_self(p[i])) * (1 + erf(lambda * p[i] / sqrt2())));
     }
 }
 "

### 2021-02-10 code/paper-models/prepare-data.R

@@ -115,7 +115,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
 simulated.data <- function(simulated.type="linear.corr"){
         
         # Define system dimensions
-        N <- 50
+        N <- 100
         sites <- 300
         
         # Environmental predictors for each site

### 2021-02-10 manuscript/manuscript.tex
Trying again with logistic... this sucks

@@
 
-One of the central goals of ecology is to understand the ways species are distributed across space (ref).
-Over the years, researchers have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions (ref).
+One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
+As a result, ecologists have developed multiple distribution models over the last two decades, to try to untangle the factors that play a role in defining such distributions (ref).
 These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
-Recent work on these approaches has focused on estimating (ref) and accounting for (ref) biotic factors, such as competitive or facilitative relationships.
+For example, ...
+More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species.
 The idea is that by understanding how such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
+However, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
+This was parsimoniously illustrated by \citealt{bla bla}, who ...
+
+%Comparisons of the predictive performance of such models are also extensive (ref), untangling the modelling approaches that better capture variation in species' distributions.
+
+% People comparing performance: Wilkinson, D.
+P., N.
+Golding, G.
+Guillera‐Arroita, R.
+Tingley, and M.
+A.
+McCarthy.
+2019.
+A comparison of joint species distribution models for presence–absence data.
+Methods in Ecology and Evolution 10:198–211; A comprehensive evaluation of predictive performance of 33 species distribution models at species and community levels
+
+%The last two decades have seen a proliferation of species distribution models (SDMs) addressing the challenge of predicting the occurrences of individual species (Guisan and Zimmermann 2000, Guisan and Thuiller 2005, Elith et al.
+2006, Leathwick et al.
+2006, Zimmermann et al.
+2010).
+Methodological advances in multiple‐species distribution modeling have lagged behind, but are recently experiencing a rapid expansion (Leathwick et al.
+2006, Dunstan et al.
+2011, Guisan and Rahbek 2011, Warton et al.
+2015, Wilkinson et al.
+2019).
+Many previous studies (Table 1) have compared the predictive performance of SDMs for single‐species analyses (Moisen and Frescino 2002, Thuiller et al.
+2003, Elith et al.
+2006, Leathwick et al.
+2006, Elith and Graham 2009, Guisan and Rahbek 2011).
+Some studies have compared single‐species and multi‐species distribution models (Araújo and Luoto 2007, Heikkinen et al.
+2007, Baselga and Araújo 2009, 2010, Elith and Leathwick 2009, Chapman and Purse 2011, Bonthoux et al.
+2013, Madon et al.
+2013, Maguire et al.
+2016, Harris et al.
+2018), while a few have examined the performance of alternative multiple species modeling approaches (Baselga and Araújo 2010, Madon et al.
+2013, Wilkinson et al.
+2019).
+
 Unfortunately, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured
+
 While the value of gaining a mechanistic understanding of species' distributions is unquestioned, a lot can be learned from shedding light on basic properties of species' realized niches.
@@
 Here, we...Bayesian framework... This allow us to account for... as well as to tackle long-standing hypothesis regarding basic aspects of species distributions.
+
+ For non-parametric models, these often do not account for species uncertainties.
+One can benefit a lot from a bayesian framework.
+First, it allows us to compare the parameter estimates ...
+It also help us incorporate prior information.
 
@@
 
-To decide among modelling approaches, we first needs to agree on what we know about the system.
+To decide among modelling approaches, we first need to agree on what we know about the system.
 We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.

### 2021-02-10 code/paper-models/models.R

@@ -241,7 +241,7 @@ transformed parameters{
     matrix[L, L] L_SIGMA_b;
     matrix[L, L] L_SIGMA_g;
 
-    alpha = exp(zalpha * sigma_a + alpha_bar);
+    alpha = zalpha * sigma_a + alpha_bar;
 
     L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
     beta = L_SIGMA_b * zbeta + beta_bar;
@@ -271,7 +271,7 @@ model{
 
     for ( i in 1:L ){
         // p[i] = exp(-alpha[i] - gamma[i] * columns_dot_self(X1 - beta[i]));
-        Y[i] ~ binomial(1, exp(-alpha[i] - gamma[i] * columns_dot_self(X1 - beta[i])));
+        Y[i] ~ binomial_logit(1, alpha[i] - gamma[i] * columns_dot_self(X1 - beta[i]));
     }
     // Y ~ binomial(1, to_vector(p));
 }

### 2021-02-10 code/paper-models/prepare-data.R

@@ -115,7 +115,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
 simulated.data <- function(simulated.type="linear.corr"){
         
         # Define system dimensions
-        N <- 100
+        N <- 50
         sites <- 300
         
         # Environmental predictors for each site
@@ -148,7 +148,7 @@ simulated.data <- function(simulated.type="linear.corr"){
         sigma_beta1 <- exp(mvrnorm(mu = rep(-1, times = N), Sigma = Sigma))
         # Sigma <- 1*exp(-1/(0.2*0.2)*(Dis^2)) + diag(N)*0.1
         # alpha <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
-        alpha <- exp(rnorm(N, -0.4,1))
+        alpha <- rnorm(N, 0,1)
 
         # Simulate data
         dataset <- expand.grid(site=1:sites, id=1:N)
@@ -157,7 +157,7 @@ simulated.data <- function(simulated.type="linear.corr"){
         dataset$alpha <- alpha[dataset$id]
         dataset$sigma_beta1 <- sigma_beta1[dataset$id]
         
-        dataset$p <- exp(-alpha[dataset$id] - sigma_beta1[dataset$id]*(beta1[dataset$id] - dataset$S1)**2)
+        dataset$p <- inv_logit(alpha[dataset$id] - sigma_beta1[dataset$id]*(beta1[dataset$id] - dataset$S1)**2)
         
         dataset$obs <- rbinom(n = length(dataset$S1), size = 1, prob = dataset$p)
         dataset <- data.frame(id=dataset$id, obs=dataset$obs, alpha=dataset$alpha, beta1=dataset$beta1, sigma_beta1=dataset$sigma_beta1, S1=dataset$S1)                

### 2021-02-10 code/paper-models/visualizing-results.R

@@ -15,12 +15,12 @@ library(cowplot)
 plot.simulated.data <- function(beta=T, gp_type = 2){
   # load data
   d <- readRDS(file = paste("../../data/processed/jsdm/1d-simulated2S1S2", "data.rds", sep = ""))
-  model <- readRDS(paste("../../results/models/baseline-model-1d1d-simulated2.rds", sep=""))
+  model_r <- readRDS(paste("../../results/models/baseline-model-1d1d-simulated2.rds", sep=""))
 
   # Extract variables from the model
-  alphas <- precis(model, pars = "alpha", depth=2)
-  betas <- precis(model, pars = "beta", depth=3)
-  sigmas <- precis(model, pars = "gamma", depth=3)
+  alphas <- precis(model_r, pars = "alpha", depth=2)
+  betas <- precis(model_r, pars = "beta", depth=3)
+  sigmas <- precis(model_r, pars = "gamma", depth=3)
   
   # Extract true values
   N <- length(unique(d$dataset$id))
@@ -238,7 +238,8 @@ plot.actual.data <- function(model=NULL){
     posx = (xlim[2]-xlim[1])*0.8 + xlim[1]
     
     # Beta
-    plist[[1]] <- plot.ranking.x(mu_beta[[i]], ci_beta[[i]], color=colo[1], xlabel="species", ylabel="", ylims=ylim, xlims=xlim, mar=margin(5.5,5.5,5.5,5.5), additional_label = Tind_, posx=posx, posy=c(posy, ylim[2]-(posy-ylim[1])))
+    plist[[1]] <- plot.ranking.x(mu_beta[[i]], ci_beta[[i]], color=colo[1], xlabel="species", ylabel="", ylims=ylim, xlims=xlim, mar=margin(5.5,5.5,5.5,5.5), additional_label = Tind_, posx=posx, posy=c(posy, ylim[2]-(posy-ylim[1])))+
+      coord_trans(x="identity")
 
     xlim = c(min(ci_gamma[[i]]), max(ci_gamma[[i]]))
     posx = exp((log(xlim[2])-log(xlim[1]))*0.796 + log(xlim[1]))
@@ -495,126 +496,7 @@ plot.actual.data.means <- function(model=NULL){
                 c(1,3,5),
                 c(2,4,6))
 
-  p <- grid.arrange(grobs=plist, ncol=3, nrow=3, heights=c(0.1,1,1), layout_matrix=hlay, widths=c(1,1,0.27)#, vp=viewport(width=1, height=1, clip = TRUE),
[…]

### 2021-02-12 manuscript/manuscript.tex
Trying to do anything to get this to converge

@@
 These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
-For example, ...
-More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species.
+And more recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species (ref).
 The idea is that by understanding how such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
 However, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
-This was parsimoniously illustrated by \citealt{bla bla}, who ...
+This was recently best parsimoniously illustrated by \citealt{bla bla}, who ...
+
+While the value of gaining a mechanistic understanding of species' distributions is unquestioned (ref), a lot can be learned from shedding light on basic properties of species' realized niches.
+Indeed, there is no general agreement the shape of species distributions.
+While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994) assume this to be unimodal and symmetric, some have warned that empirical distributions can take many different forms \citep{austin2002spatial}.
+
+There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems.
+The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies.
+Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
+Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.
+
+To decide among modelling approaches, we first need to agree on what we know about the system.
+We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.
+Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions.
+One could point out that we also know that many other factors might influence species' presence/absence---e.g.
+the influence of biotic interactions among species.
+However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions.
+As a result, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e.
+the one with the largest entropy---is that such distribution is a Gaussian.
 

### 2021-02-12 code/paper-models/main.R

@@ -573,15 +573,15 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
         
         # Set starting values for the parameters
         start_5.1 <- list(
-                lambda = rep(0, dat_5.1$L),
+                zlambda = rep(0, dat_5.1$L),
                 zalpha = rep(0, dat_5.1$L),
                 zbeta = rep(0, dat_5.1$L),
                 zgamma = rep(0, dat_5.1$L),
-                # lambda_bar = 0,
+                lambda_bar = 0,
                 alpha_bar = -1,
                 beta_bar = 0,
                 gamma_bar = 0,
-                # sigma_l = 0.1,
+                sigma_l = 0.1,
                 sigma_a = 0.1,
                 sigma_b = 0.1,
                 etasq_b = 0.1,
@@ -604,7 +604,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
                            chains=n_chains_5.1 ,
                            cores= n_chains_5.1 ,
                            warmup=1000, iter=2000,
-                           init=init_5.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
+                           init=init_5.1 , control = list(stepsize=0.01, adapt_delta=0.99, max_treedepth = 15))
         
         
         saveRDS(mfit_5.1, file = paste(ofolder, extension2, "skew-model-traits-1d", extension,".rds", sep=""))

### 2021-02-12 code/paper-models/models.R

@@ -633,6 +633,15 @@ functions{
     int sgn(real x) {
         return x < 0 ? -1 : x > 0;
     }
+    real findmax(real delta, real beta, real gamma, real lambda) {
+        real maxy;
+        real muz;
+        maxy = 0.5 * ( 4.0 - pi() ) * pow(3, delta * sqrt( 2.0 / pi() )) / pow(1.5, 1.0 - 2.0 * pow(2,delta) / pi() );
+        muz = sqrt( 2 / pi() ) * delta;
+        maxy = beta + (1.0 / sqrt( 2.0 * gamma)) * (muz - maxy * sqrt( 1.0 - muz * muz ) * 0.5 - 0.5 * sgn(lambda) * exp(- 2.0 * pi() / fabs(lambda) ));
+        maxy = exp(- gamma * pow(2, maxy - beta)) * (1.0 + erf((lambda * ( maxy - beta )) * sqrt( gamma ) ));
+        return maxy+0.0001
+    }
 }
 data{
     int N;
@@ -646,12 +655,12 @@ parameters{
     vector[L] zalpha;
     vector[L] zbeta;
     vector[L] zgamma;
-    vector[L] lambda;
+    vector[L] zlambda;
     real alpha_bar;
     real beta_bar;
     real gamma_bar;
-    //real lambda_bar;
-    //real<lower=0> sigma_l;
+    real lambda_bar;
+    real<lower=0> sigma_l;
     real<lower=0> sigma_a;
     real<lower=0> sigma_b;
     real<lower=0> etasq_b;
@@ -664,14 +673,12 @@ transformed parameters{
     vector[L] alpha;
     vector[L] beta;
     vector[L] gamma;
-    //vector[L] lambda;
+    vector[L] lambda;
     real delta;
-    real maxy;
-    real muz;
     matrix[L, L] L_SIGMA_b;
     matrix[L, L] L_SIGMA_g;
 
-    //lambda = zlambda * sigma_l + lambda_bar;
+    lambda = zlambda * sigma_l + lambda_bar;
     
     // delta = lambda ./ ( sqrt( 1 + (lambda .* lambda) ));
 
@@ -688,26 +695,18 @@ transformed parameters{
     //beta = beta - sqrt( 1 ./ (2 * gamma) ) .* (delta * sqrt(2/pi()));
     
[…]

### 2021-02-12 code/paper-models/models.R
fixing problem

@@ -640,7 +640,7 @@ functions{
         muz = sqrt( 2 / pi() ) * delta;
         maxy = beta + (1.0 / sqrt( 2.0 * gamma)) * (muz - maxy * sqrt( 1.0 - muz * muz ) * 0.5 - 0.5 * sgn(lambda) * exp(- 2.0 * pi() / fabs(lambda) ));
         maxy = exp(- gamma * pow(2, maxy - beta)) * (1.0 + erf((lambda * ( maxy - beta )) * sqrt( gamma ) ));
-        return maxy+0.0001
+        return maxy+0.0001;
     }
 }
 data{

### 2021-02-12 code/paper-models/main.R
fixing the problem... maybe

@@ -603,11 +603,11 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
                            data=dat_5.1 ,
                            chains=n_chains_5.1 ,
                            cores= n_chains_5.1 ,
-                           warmup=1000, iter=2000,
-                           init=init_5.1 , control = list(stepsize=0.01, adapt_delta=0.99, max_treedepth = 15))
+                           warmup=1, iter=2,
+                           init=init_5.1 , control = list(adapt_delta=0.95, max_treedepth = 15))
         
         
-        saveRDS(mfit_5.1, file = paste(ofolder, extension2, "skew-model-traits-1d", extension,".rds", sep=""))
+        # saveRDS(mfit_5.1, file = paste(ofolder, extension2, "skew-model-traits-1d", extension,".rds", sep=""))
         return(mfit_5.1)
 }
 
@@ -618,11 +618,12 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
 d <- readRDS(file = "../../data/processed/jsdm/skew-simulated2S1S2data.rds")
 skew.1d(d=d, simulated=T, recompile = F, min.occurrence = 10, ofolder="/cluster/scratch/bemora/plant-stan/")
 # baseline.1d(d=d, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")
-
-
-# skew.1d(d=NULL, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")
 # 
+# expose_stan_functions(mod)
 # 
+# # skew.1d(d=NULL, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")
+# # 
+# # 
 # skn <- function(x, alpha, sigma, beta, lambda){
 # 
 # 
@@ -634,17 +635,21 @@ skew.1d(d=d, simulated=T, recompile = F, min.occurrence = 10, ofolder="/cluster/
 #         # maxy_ <- dsn(mode_x, xi=beta, omega=sqrt(1/(2*sigma)), alpha=lambda)
 # 
[…]

### 2021-02-12 code/paper-models/models.R

@@ -282,7 +282,7 @@ generated quantities{
     k = 1;
     for ( i in 1:L ){
         for (j in 1:N){
-           log_lik[k] = binomial_lpmf(Y[i, j] | 1, exp(-alpha[i] - gamma[i] * pow(2, X1[j] - beta[i])));
+           log_lik[k] = binomial_lpmf(Y[i, j] | 1, exp(-alpha[i] - gamma[i] * pow(X1[j] - beta[i],2)));
            k = k + 1;
         }
     }
@@ -636,11 +636,11 @@ functions{
     real findmax(real delta, real beta, real gamma, real lambda) {
         real maxy;
         real muz;
-        maxy = 0.5 * ( 4.0 - pi() ) * pow(3, delta * sqrt( 2.0 / pi() )) / pow(1.5, 1.0 - 2.0 * pow(2,delta) / pi() );
-        muz = sqrt( 2 / pi() ) * delta;
+        maxy = 0.5 * ( 4.0 - pi() ) * pow((delta * sqrt( 2.0 / pi() )),3) / pow(1.0 - 2.0 * pow(delta,2) / pi() ,1.5);
+        muz = sqrt( 2.0 / pi() ) * delta;
         maxy = beta + (1.0 / sqrt( 2.0 * gamma)) * (muz - maxy * sqrt( 1.0 - muz * muz ) * 0.5 - 0.5 * sgn(lambda) * exp(- 2.0 * pi() / fabs(lambda) ));
-        maxy = exp(- gamma * pow(2, maxy - beta)) * (1.0 + erf((lambda * ( maxy - beta )) * sqrt( gamma ) ));
-        return maxy+0.0001;
+        maxy = exp(- gamma * pow(maxy - beta,2)) * (1.0 + erf((lambda * ( maxy - beta )) * sqrt( gamma ) ));
+        return maxy;
     }
 }
 data{
@@ -695,8 +695,8 @@ transformed parameters{
     //beta = beta - sqrt( 1 ./ (2 * gamma) ) .* (delta * sqrt(2/pi()));
     
     for (i in 1:L){
-        delta = lambda[i] / ( sqrt( 1.0 + (pow(2,lambda[i])) ));
-        gamma[i] = gamma[i] * (1.0 - (2 * pow(2, delta))/pi());
+        delta = lambda[i] / ( sqrt( 1.0 + (pow(lambda[i],2)) ));
[…]

### 2021-02-12 code/paper-models/main.R
adding more iterations

@@ -603,7 +603,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
                            data=dat_5.1 ,
                            chains=n_chains_5.1 ,
                            cores= n_chains_5.1 ,
-                           warmup=1, iter=2,
+                           warmup=1000, iter=2000,
                            init=init_5.1 , control = list(adapt_delta=0.95, max_treedepth = 15))
         
         

### 2021-02-12 code/paper-models/main.R
fixing initial param

@@ -573,15 +573,15 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
         
         # Set starting values for the parameters
         start_5.1 <- list(
-                zlambda = rep(0, dat_5.1$L),
+                lambda = 0,
                 zalpha = rep(0, dat_5.1$L),
                 zbeta = rep(0, dat_5.1$L),
                 zgamma = rep(0, dat_5.1$L),
-                lambda_bar = 0,
+                # lambda_bar = 0,
                 alpha_bar = -1,
                 beta_bar = 0,
                 gamma_bar = 0,
-                sigma_l = 0.1,
+                # sigma_l = 0.1,
                 sigma_a = 0.1,
                 sigma_b = 0.1,
                 etasq_b = 0.1,

### 2021-02-12 manuscript/manuscript.tex
fixing the problem... maybe
lambda removed

@@
 While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994) assume this to be unimodal and symmetric, some have warned that empirical distributions can take many different forms \citep{austin2002spatial}.
+XX showed that...
+Other work on...
+have also...
 
@@
 the one with the largest entropy---is that such distribution is a Gaussian.
+
+EXPLAIN WHY KNOWING THIS BASIC INFORMATION IS IMPORTANT 
+WHY DO YOU CARE
+ABUNDANCE CENTER HYPOTHESIS IDEA
+HOW STEEP THE EDGES ARE RAPIDLY EXPANDING
+https://eco.confex.com/eco/2020/meetingapp.cgi/Paper/86915
+https://advances.sciencemag.org/content/5/11/eaaz0414
+https://onlinelibrary.wiley.com/doi/full/10.1111/j.1466-8238.2009.00451.x
 

### 2021-02-12 code/paper-models/main.R

@@ -573,7 +573,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
         
         # Set starting values for the parameters
         start_5.1 <- list(
-                lambda = 0,
+                # lambda = rep(0, dat_5.1$L),
                 zalpha = rep(0, dat_5.1$L),
                 zbeta = rep(0, dat_5.1$L),
                 zgamma = rep(0, dat_5.1$L),

### 2021-02-12 code/paper-models/models.R

@@ -655,7 +655,7 @@ parameters{
     vector[L] zalpha;
     vector[L] zbeta;
     vector[L] zgamma;
-    real<lower=-10, upper=10> lambda;
+    //real<lower=-10, upper=10> lambda;
     real alpha_bar;
     real beta_bar;
     real gamma_bar;
@@ -673,15 +673,13 @@ transformed parameters{
     vector[L] alpha;
     vector[L] beta;
     vector[L] gamma;
-    //vector[L] lambda;
+    real lambda;
     real delta;
     matrix[L, L] L_SIGMA_b;
     matrix[L, L] L_SIGMA_g;
-
+    lambda = 5;
     //lambda = zlambda * sigma_l + lambda_bar;
     
-    // delta = lambda ./ ( sqrt( 1 + (lambda .* lambda) ));
-
     alpha = exp(zalpha * sigma_a + alpha_bar);
 
     L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
@@ -690,10 +688,7 @@ transformed parameters{
     L_SIGMA_g = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g, rhosq_g, sigma_g));
     gamma = L_SIGMA_g * zgamma + gamma_bar;
     gamma = exp(gamma);
-    //gamma = exp(gamma) .* (1 - (2 * (delta .* delta))/pi());
 
-    //beta = beta - sqrt( 1 ./ (2 * gamma) ) .* (delta * sqrt(2/pi()));
-    
     for (i in 1:L){
         delta = lambda / ( sqrt( 1.0 + (pow(lambda,2)) ));
         gamma[i] = gamma[i] * (1.0 - (2 * pow(delta,2))/pi());
@@ -718,7 +713,7 @@ model{
     zalpha ~ std_normal();
     zgamma ~ std_normal();
     zbeta ~ std_normal();
-    lambda ~ std_normal();
+    //lambda ~ std_normal();
 
     for ( i in 1:L ){
         Y[i] ~ binomial(1, exp(- alpha[i] - gamma[i] * columns_dot_self(X1 - beta[i])) .* (1 + erf((lambda * (X1 - beta[i])) * sqrt(gamma[i]) )));

### 2021-02-25 manuscript/manuscript.tex
Habemus introduction first draft

--- 2021-02-10 manuscript/manuscript.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PARAGRAPH 1
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Consider testing tails of distribution?
+I would use the Cauchy or generalized normal distribution for heavy tail distribution without skewness; and I would use an exponentially modified Gaussian distribution with heavy tails for a skew distribution with heavy tails.
 
 One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
-As a result, ecologists have developed multiple distribution models over the last two decades, to try to untangle the factors that play a role in defining such distributions (ref).
-These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
-For example, ...
-More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species.
-The idea is that by understanding how such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
-However, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
-This was parsimoniously illustrated by \citealt{bla bla}, who ...
+Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000, more}.
+These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
+More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}.
+The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
+However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been often questioned \citep{gotelliEmpiricalBayesApproach2010, harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019}.
+This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
+% Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018
+
+The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and species' dispersal ability in setting their range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
+That said, a lot can be learned from taking a phenomenological approach, focussing instead on the description of basic properties of species' realized niches.
+For example, the study of species' range sizes along environmental gradients can reveal general biodiversity patterns that are crucial from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992}.
+Differences in species' responses to the environment could shed light on how climatic processes and historical contingencies have shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}.
+Other properties, such as the skewness of species' distributions, can also reveal general underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
+More generally, understanding the shape of species' realized niches and the extend to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
+Indeed, there is no general agreement on the shape of species' distributions (ref).
+
+Many ecological textbooks (ref) assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}.
+In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY).
+This is useful because it simplifies the optimization process, but it comes with several statistical shortcomings.
+First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
+From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.
+their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
+This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
+Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
+Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.
+\citealt{landoltFloraIndicativaOkologische2010}).
+That said, this information is rarely accounted for in most modelling approaches, mainly because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}).
+Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{jamilGeneralizedLinearMixed2013, terbraakWeightedAveragingLogistic1986} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
+For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to capture following these model structures.
+
+% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured
+% While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994)
+%Niche theory predicts that species occurrence and abundance show non-linear, unimodal relationships with respect to environmental gradients (Austin, 1987; Palmer & Dixon, 1990; Whittaker, 1967)
+%\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species distributions.
+
+The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}.
+This has resulted in a plethora of models accounting for several biotic and abiotic factors into the predictions of species co-occurrence.
+Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
+In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology.
+We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions.
+Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns.
+First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}.
+While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
+Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient.
+Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref).
+\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citep{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
+Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
+Specifically, we are able to link plant physiological traits to the skewness of their distributions.
+Overall, we use models that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
+Using these model, we are able uncover the shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
+
+%%%%%%%%%%%%%%%%%%%%%%%FINAL
+
+% To decide among modelling approaches, we first need to agree on what we know about the system.
+We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.
+Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions.
+One could point out that we also know that many other factors might influence species' presence/absence---e.g.
+the influence of the aforementioned biotic interactions among species.
+However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions.
+Therefore, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e.
+the one with the largest entropy---is that such distribution is a Gaussian.
+
+%The strength of climate as a range limiting factor will becrucial for the extent to which species will retract from theirsouthern range margins and expand northwards under global warming.
+occurrence of a species along an environmentalgradient may primarily be limited by its physiological tolerancetowards the abiotically more stressful conditions, while otherfactors, such as biotic interactions, may assume a greater roletowards less stressful conditions (Connell, 1961; Austin, 1990;Brown  et al.
+, 1996); hereafter referred to as the asymmetricabiotic stress limitation (AASL) hypothesis.
+
+%abiotic stress is generally thought to increasetowards high latitudes and altitudes and to decrease in importancetowards the equatorial lowlands, where biotic interactions havebeen argued to increase in importance (Dobzhansky, 1950;Kaufman, 1995).
+Accordingly, a geographical prediction of theAASL hypothesis is that abiotic stress may be most importantin upper-latitudinal and upper-altitudinal range boundaries(MacArthur, 1972; Brown  et al.
+, 1996), but this has received littlestudy.
+Loehle (1998) suggested that the northern and southernrange limits of North American tree species are determined bycold tolerance and competitive ability, respectively.
+Similarly,abiotic stress appears to determine the upper-altitudinal rangelimit for some mountainous plant species, while biotic interactionsmay determine the lower limit (Guisan  et al ., 1998; Bruelheide &Scheidel, 1999; Vetaas, 2002).
+A second aspect of the AASL hypothesis proposes that theshape of species response curves vary along an environmentalgradient: species that occur under the physiologically moststressful conditions show skewed responses with a steep declinetowards the stressful conditions, while species that occur underless stressful conditions show symmetric responses (Austin,1990; also cf.
+Austin  et al ., 1994; Austin & Gaywood, 1994;Rydgren  et al.
+, 2003).
+In line with this, Kaufman (1995) suggeststhat the degree to which a species’ poleward range boundary islimited by abiotic stress increases the further north the species isdistributed.
+Analogously, we expect increasing stress limitationwith increasing elevation
+
+% Guisan found no evidence of that. 
+% No support for Kaufman’s (1995) geographical prediction ofthe AASL hypothesis,
+
+ %There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems.
+The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies.
+Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
+Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.
+
+%%%%%%%%%%%%%%% KEY PAPERS AND IDEAS REGARDING HYPOTHESES #################
+% EXPLAIN WHY KNOWING THIS BASIC INFORMATION IS IMPORTANT 
+% WHY DO YOU CARE
+% ABUNDANCE CENTER HYPOTHESIS IDEA
+% HOW STEEP THE EDGES ARE RAPIDLY EXPANDING
+% species’ thermal tolerance breadths 
+% https://eco.confex.com/eco/2020/meetingapp.cgi/Paper/86915
+% https://advances.sciencemag.org/content/5/11/eaaz0414
+% https://onlinelibrary.wiley.com/doi/full/10.1111/j.1466-8238.2009.00451.x
+% https://royalsocietypublishing.org/doi/10.1098/rspb.2010.1295
+
+%%%%%%%%%%%%%%   Species thermal tolerance breadths   %%%%%%%%%%%%%%%%%%%%%
+% Species' thermal tolerance increases with latitude (elevation). The asymmetricabiotic stress limitation (AASL) hypothesis
+% - mostly macroecological patterns. What about along elevational gradient?
+% - Taxonomic scope often restricted. What about plants?
+% - upper and lower thermal tolerance data are not always sampled from the same species. Bayes solve the problem?
 
@@
 
-Unfortunately, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref).
-For example...
-
 %Overall, these methodological advances have provided with crucial ecological insights into the factors shaping species' distributions (ref); however, they have also faced some criticism as a result of the...
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PARAGRAPH 2
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges (ref).
-Accurately predicting how species are distributed across ranges is crucial for understanding the impacts and effect of global climate change.
-With that said, accounting for the mechanisms can come at the cost of modelling noise...
-However, much less attention is paid to how species' distributions compare to each other.
-
 %\citet{austin2002spatial} writes "\textit{there are three major components in any framework for statistical modelling in plant ecology.
@@
 Indeed,
-
-Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges.
-However, much less attention is paid to how species' distributions compare to each other.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PARAGRAPH 3
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured
-
-While the value of gaining a mechanistic understanding of species' distributions is unquestioned, a lot can be learned from shedding light on basic properties of species' realized niches.
-
-Accounting for expert knowledge on species' environmental preferences to understand general distribution patterns.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PARAGRAPH 4
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Here, we...Bayesian framework... This allow us to account for... as well as to tackle long-standing hypothesis regarding basic aspects of species distributions.
-
- For non-parametric models, these often do not account for species uncertainties.
-One can benefit a lot from a bayesian framework.
-First, it allows us to compare the parameter estimates ...
-It also help us incorporate prior information.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PARAGRAPH 5
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In this work, we first....
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%   Stuff to realocate   %%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-There is no general agreement the shape of species distributions.
-While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994) assume this to be unimodal and symmetric, some have warned that empirical distributions can take many different forms \citep{austin2002spatial}.
-There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems.
-The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies.
-Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
-Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.
-
-To decide among modelling approaches, we first need to agree on what we know about the system.
-We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.
-Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions.
-One could point out that we also know that many other factors might influence species' presence/absence---e.g.
-the influence of biotic interactions among species.
-However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions.
-As a result, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e.
-the one with the largest entropy---is that such distribution is a Gaussian.
-
-%There is a growing literature on species' distribution models.
-For the most part, however, these models have often focussed on model performance and understanding the predictive power of certain covariates.
-This models often take a response variable, such as presence/absence or abundance data, and fit it with a set of environmental variables and species' ecological or phylogenetic relationships.
-There...
-The shape of distributions on systems of many species is rarely studied.
-That is, simple ...
-%model comparison... 
-%creating maps of relative suitabilities...
-%extrapolate to new data...
-
-%It will be argued in this paper that neglect of ecological knowledge is a limiting factor in the application of statistical modelling in ecology and conservation planning
-
-%The field has shifted towards understanding the effects of biotic factors on such distributions. And while the theory is advancing quickly, the...
-
-%This is a generalised additive model.
-% Results are presented for both simulated and empirical data.
-% But this approach understates the posterior uncertainty in inference. 
-%\section*{Introduction}
-
-%%%%%%%%%%%%%%%%%%%%%%%
-% Paragraph on Rapoport's rule
-%%%%%%%%%%%%%%%%%%%%%%%
-% McCain
-%Understanding the factors that determine species range size dis- tributions is becoming increasingly urgent as more climate change assessments document the heightened risk for small- ranged species (e.g.
-Channell & Lomolino, 2000; Sekercioglu et al., 2008; La Sorte & Jetz, 2010; McCain & Colwell, 2011).
-%Rapoport’s rule is the positive relationship of species range sizes with increasing latitude (Stevens, 1989), elevation (Stevens, 1992) or water depth (Stevens, 1996).
-The latitudinal Rapoport’s rule (LRR) is the most examined in the literature, including testing empirical datasets (e.g.
-Rohde et al., 1993; Roy et al., 1994; Blackburn & Gaston, 1996; Lyons & Willig, 1997; Ruggiero & Lawton, 1998; Reed, 2003; Arita et al., 2005; Hausdorf, 2006), simulation modelling of the expected patterns (Colwell & Hurtt, 1994; Taylor & Gaines, 1999; Case & Taper, 2000; Arita, 2005; Stauffer & Rohde, 2006; Šizling et al., 2009), and theoretical and empirical evidence for the proposed mechanisms (e.g.
-Rohde, 1992; Kerr, 1999; Taylor & Gaines, 1999; Gaston & Chown, 1999a; Addo-Bediako et al., 2000; Parmesan et al., 2005).
-The various reviews of the LRR to date suggest that the overall support is weak (e.g.
-Rohde, 1996; Gaston et al., 1998; Rohde, 1999; Gaston & Chown, 1999b; Ribas & Schoereder, 2006), prin- cipally due to the high degree of variability in the fit to predic- tions (e.g.
-support: Blackburn & Gaston, 1996; Lyons & Willig, 1997; Price et al., 1997; Arita et al., 2005; no support: Rohde et al., 1993; Roy et al., 1994; Ruggiero & Lawton, 1998; Reed, 2003).
-%The elevational Rapoport’s rule (ERR) has received less examination in the literature than the LRR, but there have also been empirical tests of its predictions (e.g.
-Patterson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Sanders, 2002; Fu et al., 2004; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Bhattarai & Vetaas, 2006; Hausdorf, 2006) and underlying theory (e.g.
-Rohde, 1996; Fleishman et al., 1998; Gaston & Chown, 1999a; Hausdorf, 2006; Ribas & Schoereder, 2006).
-In the ERR, like the LRR, there is a high degree of variability in support from supportive (e.g.
-Pat- terson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Sanders, 2002; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Hausdorf, 2006; Ribas & Schoereder, 2006) to little or no support (e.g.
-Patterson et al., 1996; Price et al., 1997; Rahbek, 1997; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Fu et al., 2004; Bhattarai & Vetaas, 2006; Hausdorf, 2006; Ribas & Schoereder, 2006).
-Similarly, the processes proposed to underlie the ERR (Stevens, 1992) also show a variable amount of theoretical and empirical support (Patterson et al., 1996; Rohde, 1996; Rahbek, 1997; Fleishman et al., 1998; Gaston & Chown, 1999a; Almeida-Neto et al., 2006; Hausdorf, 2006; Colwell, 2011).
-Scarce data and little to no attempt to account for uncertainty in the predictions.
-Similar to rapoport's rule, we can also ask other questions regarding general geographical patterns of species distributions.
 
@@
 \clearpage
-\bibliography{references}
+\bibliography{references2}
 

### 2021-02-25 code/paper-models/visualizing-results.R

@@ -704,6 +704,135 @@ plot.actual.data.means <- function(model=NULL){
   
 }
 
+plot.actual.data.means.pairwise <- function(model=NULL){
+  # Parameters for plots
+  extra <- 1
+  colo <- c("#e6ab02", "#1b9e77", "#d95f02", "#7570b3")
+  
+  # Load the data if not added  
+  if(is.null(model)){
+    model <- readRDS("../../results/models/min30-skew-model-traits-1d2.rds")
+  }
+  
+  #indicator values
+  Tind <- read.table("../../data/properties/codes/temperature_indicator_reindexed.csv", sep = ",")
+  Tend <- read.table("../../data/properties/codes/change-tendency_reindexed.csv", sep = ",")
+  NEO <- read.table("../../data/properties/codes/neophytes-list_reindexed.csv", sep = ",")
+  
+  # extract samples
+  post <- extract.samples(model, n = 1000, pars=c("alpha", "gamma", "beta", "lambda")) 
+  post <- transform.skew(post)
+  
+  # lambda
+  mu_lambda <- apply( post$lambda , 2 , mean )
+  ci_lambda <- apply( post$lambda , 2 , PI )
+  
+  # betas
+  mu_beta <- apply(post$beta,2,mean)
+  ci_beta <- apply(post$beta,2,PI)
+  
+  # gamma
+  mu_gamma <- apply( post$sigma , 2 , mean )
+  ci_gamma <- apply( post$sigma , 2 , PI )
+  
+  # alpha
+  mu_alpha <- apply(post$alpha,2,mean)
+  ci_alpha <- apply(post$alpha,2,PI)
+  
+  parameters_mu <- list(beta=mu_beta, gamma=mu_gamma, alpha=mu_alpha, lambda=mu_lambda)
+  parameters_ci <- list(beta=ci_beta, gamma=ci_gamma, alpha=ci_alpha, lambda=ci_lambda)
+
+  transformation <- c("identity", "log", "log", "identity")
+  label <- c(expression(beta), expression(gamma), expression(alpha), expression(lambda))
+    
+  plist = list()
+  k <- 1
+  for(i in 1:4){
+    for(j in 1:4){
+      ylim = c(min(parameters_mu[[i]]), max(parameters_mu[[i]]))
+      xlim = c(min(parameters_mu[[j]]), max(parameters_mu[[j]]))
+
[…]

### 2021-02-26 manuscript/manuscript.tex
Improving intro and restructuring methods

@@
 \title{\vspace{-1cm}
-Model simplicity breeds contempt: using simple models to answer basic questions on species' distributions}
+Model simplicity breeds contempt: simple models to compare properties of species' distributions}
 \author[1,*]{\normalsize Bernat Bramon Mora}
@@
 I would use the Cauchy or generalized normal distribution for heavy tail distribution without skewness; and I would use an exponentially modified Gaussian distribution with heavy tails for a skew distribution with heavy tails.
+% Option (1) for a long-tail distribution: Generalized normal distribution. We model it such that shape parameter beta can go from 1 to 8
+% Option (2) for a long-tail distribution: Student's t-distribution. In this case, we would need to define the shape parameter so that it is larger than 2 (otherwise, we would have undefined variance)
+% Option (1) for a long-tail skewed distribution.
+The exponentially modified Gaussian distribution has potential, but I am not sure I can fit a parameter that goes from like lambda if I want the skewness in either side.
+% Option (2) for a long-tail skewed distribution.
+Version 2 of the Generalized normal distribution.
+The problem here is that this function has a few discontinuities, which means that one needs to use a step function.
+I believe this to be quite problematic in Stan...
+% Option (3) for a long-tail skewed distribution. Skewed student-t distribution. This seems to be complex but quite what I need. It will be hard to fit but seems quite promising. See:
+%https://www.tandfonline.com/doi/abs/10.1080/01621459.1998.10474117
+%https://hess.copernicus.org/preprints/hess-2015-24/
+%https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2009WR008933
 
 One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
-Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000, more}.
+Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}.
 These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
@@
 
-The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and species' dispersal ability in setting their range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
+The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and dispersal ability in setting species' range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
 That said, a lot can be learned from taking a phenomenological approach, focussing instead on the description of basic properties of species' realized niches.
@@
 
-Many ecological textbooks (ref) assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}.
+Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}.
 In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY).
@@
 First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
-From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.
-their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
+From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
 This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
 Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution \citealt{entropy stuff}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
 Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
-Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.
-\citealt{landoltFloraIndicativaOkologische2010}).
+Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
 That said, this information is rarely accounted for in most modelling approaches, mainly because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}).
 Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{jamilGeneralizedLinearMixed2013, terbraakWeightedAveragingLogistic1986} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
-For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to capture following these model structures.
+For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.
 
-% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured
-% While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994)
-%Niche theory predicts that species occurrence and abundance show non-linear, unimodal relationships with respect to environmental gradients (Austin, 1987; Palmer & Dixon, 1990; Whittaker, 1967)
 %\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species distributions.
@@
 Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref).
-\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citep{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
+\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
 Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
 Specifically, we are able to link plant physiological traits to the skewness of their distributions.
-Overall, we use models that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
-Using these model, we are able uncover the shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
+Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constrains of the statistical framework.
+Using these models, we are able uncover the true shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
 
@@
 \subsubsection*{Distribution data}
-We studied the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}).
+We used data describing the distribution of 798 species across 912 sites covering most of the mountain region of the Western Alps in the Canton de Vaud (Switzerland; \citealt{scherrerEcologicalIndicatorValues2019}).
 Each of these sites is a $8\times 8\,\text{m}$ plot placed somewhere along an elevation range from $375\,\text{m}$ to $3210\,\text{m}$.
 In all sites, presence/absence data as well as Braun-Blanquet abundance-dominance classes were recorded for all species.
-Additionally, following 30 years (1961–1990) of meteorological data from national weather stations, \citet{scherrerEcologicalIndicatorValues2019} calculated multiple climatic variables for each site at high spatial resolution ($25\,\text{m}$).
-Here, we focussed on 9 climatic variables, including: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter.
+Additionally, we used meteorological data provided by \citet{scherrerEcologicalIndicatorValues2019}, containing multiple variables characterizing the climate in each site at high spatial resolution ($25\,\text{m}$).
+This dataset was compiled based on 30 years (1961–1990) of records from national weather stations.
+Since most of the data is highly correlated, we calculated the main axes of variation of the following variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig.
+1).
 %Missing TabsY
@@
 This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits.
-Species' environmental preferences in this database can be used to inform distribution models---e.g.
-as an informative prior in a Bayesian framework.
-These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found and a broad description of their range of variation.
+Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework.
+These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation.
 These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
-On the other hand, the information regarding species' physiological traits represent general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
+On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
 In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
@@
 
-{\color{gray}
-\subsubsection*{[Trait data]}
-This could be Tom's data if we end up using it.}
+\subsection*{Baseline model}
+There is a long list of model structures well suited to characterize species' distributions (see \citealt{norbergComprehensiveEvaluationPredictive2019}).
+As a baseline model, however, we were interested in a hierarchical model that does not make any assumptions regarding the shape of the distributions, and yet explicitly incorporates all information that we have regarding plant's environmental preferences.
+More specifically, we wanted to account for the climatic indicator values and range of variation registered in the floristic database for all plants in our dataset.
+These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions.
 
-\subsection*{Distribution model}
+\subsubsection*{Response curve}
+To choose an appropriate response curve, we first need to agree on what we truly know about the system.
+Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance.
+While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions.
+Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution.
+That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we define can species' responses to the environment as
+\begin{equation}
+\begin{split}
+y_{ij} & \sim F\left(p_{ij}\right) \\
+\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \gamma_{i} \left(x_{j}-\beta_{i}\right)^2 ,
+\end{split}
+\label{eq:baseline-response}
+\end{equation}
+where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the an environmental gradient, respectively.
+Notice that $F$ characterizes a Bernoulli distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods).
+For the sake of simplicity, we use only one environmental variable to characterize species' probability distribution.
+That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
+
+\subsubsection*{Model priors}
+The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database.
+To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
+\begin{equation} 
+\begin{split}
+\beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
+\text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
+\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
+\hat{\beta}, 
+\hat{\gamma}, 
+\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
+\sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
+\end{split}
+\label{eq:baseline-priors}
+\end{equation} 
+where parameters $\gamma_i$ and $\beta_i$ are expressed as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta}$ and $\Sigma^{\gamma}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
+We define these variance-covariance matrices as follows:
+\begin{equation} 
+\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
+\label{eq:covariance-baseline}
+\end{equation}
+where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$.
+Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
+In the floristic database, this information is represented by the set of ordinal specified for the different species.
+While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}).
+In each covariance matrix, the hyperparameter $\rho$ determines the rate of decline of the covariance between any two species, and $\eta$ defines its maximum value.
+The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species.
+For all these hyperparameters, we choose weekly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.
+
+\subsubsection*{Sampling the posterior}
+We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}.
+Sampling models like the ones described above can be computationally very expensive.
+This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
+Therefore, we focus on those species for which we have more than 30 occurrences when modelling ordinal data, which is the case for the majority of the results of this work.
+When using presence/absence data, we limit our study to those species for which have more than 10 occurrences.
+
+To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Supplementary Methods; Supplementary Fig.
+2).
+Notice that using the link function in Eq.
+\ref{eq:baseline-response} could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).
+
+\subsection*{Modifying the baseline model}
+We proposed a baseline model that is naive in terms of the structural assumptions regarding the data, and yet accounts for all information we truly have about the system.
+Modifying this model, we can now test hypotheses regarding the properties of empirical species' distributions.
+To propose new species' response curves, however, we want to ensure two key conditions: (i) the probability distribution must have defined variance, and (ii) the Gaussian shape must be a special case of the probability distribution.
+
+\subsubsection*{Heavy-tail response curve}
+
+\subsubsection*{Skewed response curve}
+
+\subsubsection*{Heavy-tail and skewed response curve}
+
+\subsection*{Alternative variance-covariance structures}
+The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information.
+Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}).
+For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.~two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
+One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.~parameter $\alpha_i$---such that
+\begin{equation} 
+\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
+\label{eq:covariance-complex}
+\end{equation}
+where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$.
+Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$.
+Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.
+
+Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
+\begin{equation} 
+\begin{split}
+y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
+\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
+\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
+\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
+\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
+\hat{\alpha}, 
+\hat{\lambda^{k}}, 
+\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
+\end{split}
+\label{eq:baseline}
+\end{equation}
+Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively\footnote[2]{I'll rewrite the likelihood function to an ordered categorical as soon as I get things to work properly with count data.}.
+While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
+More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
+Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.
+
+In all cases, all variance-covariance matrices are defined as follows:
+\begin{equation} 
+\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
+\label{eq:covariance}
+\end{equation}
+
+where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$.
+Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
+Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
+For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
+The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
+For any given hyperparameter, we choose adaptive priors across covariance structures.
+That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
+Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
+We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
+
+\subsection*{Covariance matrices from incomplete categorical and ordinal data}
+The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database.
+More specifically, both the ecological indicator values and range of variation are ordinal traits specified for all species, whereas plants' physiological data are characterized by categorical data containing multiple missing entries.
+These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the parameters of our Bayesian model.
+To do so, we need to compile the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
+
+We define these variance-covariance matrices are defined as follows:
+\begin{equation} 
+\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
+\label{eq:covariance}
+\end{equation}
+
+where $\Sigma_{ij}$ describes the covariance between any pair of species $i$ and $j$.
+Following this expression, such covariance declines exponentially with the square of the different $D_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
+Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
+For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
+The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
+For any given hyperparameter, we choose adaptive priors across covariance structures.
+That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
+Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
+We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
+
+---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
+In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
+Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
+
+The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
+In this model, such distance matrices characterize differences between plant species.
+In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits.
+More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
+In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
+Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
+
+More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
+One way to frame this problem is by using a network representation.
+Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
+Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$.
+That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
+
+Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
+In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
+For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
+Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
+Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
+Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
+\begin{equation}
+Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
+\end{equation}
+Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
+Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
+In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
+While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
+The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
+This algorithm initially assigns random group memberships to both species and traits.
+While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
+Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
+In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
+
+The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
+In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.~the extend to which a species is of one type or another.
+This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
+The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
+Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
+
+\subsection*{Distribution models}
 There is a long list of model structures well suited to characterize species' distributions (see XX for a review); however, we were interested in a model that explicitly incorporates all information regarding plant's environmental preferences found in the floristic database.
@@
 While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.
-Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
+More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
 Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.
@@
 The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
-In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.
-the extend to which a species is of one type or another.
+In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.~the extend to which a species is of one type or another.
 This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
@@
 Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}).
-For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.
-two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
-One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.
-parameter $\alpha_i$---such that
+For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.~two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
+One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.~parameter $\alpha_i$---such that
 \begin{equation} 
@@
 Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.
+
+% TODO list methods
+% 1 Pomodoro on long-tail distribution.
+Present both generalized Gaussian and Student-t and explain how the condition that we want is to have finite variance and a distribution for which the Gaussian is a special case
+% 1 Pomodoro on finishing baseline and MMSBM
+% 1 Pomodoro on skewed distribution.
+Here, we want the skewed normal distribution because we can measure skeweness without having discontinuities, the Gaussian is a special case, and the tails are normally distributed (check wiki)
+% 1 Pomodoro on finishing baseline and MMSBM
+% 1 Pomodoro on finishing long-tail
+% 1 Pomodoro on finishing skewed
+% 1 Pomodoro on section about variance covariance matrix. Not long, so I can spend time on revising other things
+% 1 Pomodoro on section about other possibilities? Calculate correlations? Add other predictors?
 

### 2021-02-26 manuscript/supplementary_information.tex
Improving intro and restructuring methods

@@
 
+% Full description of Binomial model with multiple environmental covariates
+% Full description of Categorical model with multiple environmental covariates
+% Using a MMSBM to calculate species similarity
+% Edge problem in stan and need to define minimum probability.
+
+\section*{Supplementary Methods}
+\subsubsection*{Distance matrices from incomplete categorical and ordinal data}
+The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database.
+More specifically, both the ecological indicator values and range of variation are ordinal traits specified for all species, whereas plants' physiological data are characterized by categorical data containing multiple missing entries.
+These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the parameters of our Bayesian model.
+To do so, we need to compile the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
+
+The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
+In this model, such distance matrices characterize differences between plant species.
+In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits.
+More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
+In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
+Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
+
+More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
+One way to frame this problem is by using a network representation.
+Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
+Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$.
+That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
+
+Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
+In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
+For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
+Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
+Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
+Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
+\begin{equation}
+Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
+\end{equation}
+Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
+Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
+In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
+While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
+The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
+This algorithm initially assigns random group memberships to both species and traits.
+While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
+Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
+In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
+
+The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
+In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.
+the extend to which a species is of one type or another.
+This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
+The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
+Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
+
+\section*{Supplementary Notes}
+\section*{Supplementary Figures}
+
 \begin{figure}[ht]

### 2021-02-26 code/paper-models/prepare-data.R

@@ -34,6 +34,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
         indicator <- read.table("../../data/properties/codes/temperature_indicator.csv", sep=",")
         neophytes <- read.table("../../data/properties/codes/neophytes-list.csv", sep=",")
         tendency <- read.table("../../data/properties/codes/change-tendency.csv", sep=",")
+        competitive <- read.table("../../data/properties/codes/competitive_indicator.csv", sep=",")
         
         # Check that there aren't unnexpected files
         if(!all(sort(files)==1:length(files))){
@@ -47,6 +48,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
         Tind <- c()
         NEO <- c()
         Tend <- c()
+        compet <- c()
         
         # Read observations
         for(idx in 1:length(files)){
@@ -60,6 +62,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
              Tind <- rbind(Tind, c(kdx, new.name, as.character(indicator$nflor.T[new.name==indicator$nflor.spnames])))
              NEO <- rbind(NEO, c(kdx, new.name, neophytes$neo[new.name==neophytes$names]))
              Tend <- rbind(Tend, c(kdx, new.name, tendency$decrease[new.name==tendency$names], tendency$decrease.low[new.name==tendency$names], tendency$increase[new.name==tendency$names], tendency$other[new.name==tendency$names]))
+             compet <- rbind(compet, c(kdx, new.name, as.character(competitive$nflor.KS[new.name==competitive$nflor.spnames])))
              
              name.idx <- c(name.idx,correlation_matrix_ids$V1[as.character(correlation_matrix_ids$V2)==new.name])
              ###
@@ -74,10 +77,12 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
                 write.table(Tind, "../../data/properties/codes/temperature_indicator_reindexed.csv", sep=",")
                 write.table(NEO, "../../data/properties/codes/neophytes-list_reindexed.csv", sep=",")
                 write.table(Tend, "../../data/properties/codes/change-tendency_reindexed.csv", sep=",")
+                write.table(compet, "../../data/properties/codes/competitive_reindexed.csv", sep=",")
         }else{
[…]

### 2021-02-26 code/paper-models/visualizing-results.R

@@ -163,6 +163,7 @@ compare.models <- function(){
   #indicator values
   Tind <- read.table("../../data/properties/codes/temperature_indicator_reindexed-30.csv", sep = " ", header=T)
   Tinvasive <- read.table("../../data/properties/codes/neophytes-list_reindexed-30.csv", sep = " ", header=T)
+  competitive <- read.table("../../data/properties/codes/competitive_reindexed-30.csv", sep = " ", header=T)
   
   # extract samples
   post <- extract.samples(m1, pars=c("alpha", "beta", "gamma", "lambda"))
@@ -739,6 +740,7 @@ plot.actual.data.means.pairwise <- function(model=NULL){
   mu_alpha <- apply(post$alpha,2,mean)
   ci_alpha <- apply(post$alpha,2,PI)
   
+  parameters_post <- list(beta=post$beta, gamma=post$sigma, alpha=post$alpha, lambda=post$lambda)
   parameters_mu <- list(beta=mu_beta, gamma=mu_gamma, alpha=mu_alpha, lambda=mu_lambda)
   parameters_ci <- list(beta=ci_beta, gamma=ci_gamma, alpha=ci_alpha, lambda=ci_lambda)
 
@@ -749,6 +751,11 @@ plot.actual.data.means.pairwise <- function(model=NULL){
   k <- 1
   for(i in 1:4){
     for(j in 1:4){
+      # correlation
+      corPI <- sapply(1:dim(parameters_post[[i]])[1], function(x) cor(parameters_post[[i]][x,], parameters_post[[j]][x,]))
+      meanPI <- mean(corPI)
+      sdPI <- sd(corPI)
+      corPIci <- round(sort(c(as.vector(PI(corPI)),meanPI)), 2)
       ylim = c(min(parameters_mu[[i]]), max(parameters_mu[[i]]))
       xlim = c(min(parameters_mu[[j]]), max(parameters_mu[[j]]))
 

### 2021-02-27 manuscript/manuscript.tex
Finished methods section

@@
 \title{\vspace{-1cm}
-Model simplicity breeds contempt: simple models to compare properties of species' distributions}
+Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions}
+% Model familiarity breeds contempt: simple models to compare properties of species' distributions
+% Model simplicity breeds contempt: simple models to compare properties of species' distributions
 \author[1,*]{\normalsize Bernat Bramon Mora}
@@
 In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY).
-This is useful because it simplifies the optimization process, but it comes with several statistical shortcomings.
+This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings.
 First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
@@
 This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
-Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
-The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution \citealt{entropy stuff}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
+Other factors might then condition species distributions to showcase fat-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
 Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
 Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
-That said, this information is rarely accounted for in most modelling approaches, mainly because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}).
-Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{jamilGeneralizedLinearMixed2013, terbraakWeightedAveragingLogistic1986} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
+That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}).
+Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
 For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.
@@
 Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constrains of the statistical framework.
-Using these models, we are able uncover the true shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
+Using these models, we are able uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
 
@@
 These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation.
-These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
+These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
 On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Methods; \citealt{landoltFloraIndicativaOkologische2010}).
+In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
+Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and growth tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent years).
+We describe this information in more detail in the Supplementary Table 1.
 
@@
 
+\subsubsection*{Alternative variance-covariance structures}
+The model structure defined above allows us to test how different sources of information characterize each of the different parameters.
+Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}).
+For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits.
+One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
+\begin{equation} 
+\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
+\label{eq:covariance-complex}
+\end{equation}
+where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.
+
 \subsubsection*{Sampling the posterior}
-We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}.
+We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}.
 Sampling models like the ones described above can be computationally very expensive.
@@
 2).
-Notice that using the link function in Eq.
-\ref{eq:baseline-response} could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).
+Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).
 
 \subsection*{Modifying the baseline model}
-We proposed a baseline model that is naive in terms of the structural assumptions regarding the data, and yet accounts for all information we truly have about the system.
-Modifying this model, we can now test hypotheses regarding the properties of empirical species' distributions.
-To propose new species' response curves, however, we want to ensure two key conditions: (i) the probability distribution must have defined variance, and (ii) the Gaussian shape must be a special case of the probability distribution.
+We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system.
+Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and the prior information.
+More specifically, to propose new species' response curves, we follow three key criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
 
-\subsubsection*{Heavy-tail response curve}
+\subsubsection*{Fat-tail response curve}
+Distributions with fat-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
+Indeed, fat-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have fat-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
+Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions \citep{ref}.
+To accommodate this feature into our baseline model, we could consider response curves that follow a generalized error or a non-standardized Student's t-distributions.
+In both cases, the normal distribution is a special case of them, and both contain parameters that regulate the kurtosis of the probability distribution (Supplementary Figure XX).
+
+Starting with the generalized error distribution, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
+\begin{equation}
+\begin{split}
+\text{log}\left(p_{ij}\right) = -\alpha_{i} - \gamma^{\prime}_{i}\, |x_{j}-\beta_{i}|^{\nu_{i}} ,
+\end{split}
+\label{eq:generalized-response}
+\end{equation}
+where $\gamma^{\prime}_{i} = g(\gamma_{i}, \nu_{i})$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(1, \infty\right)$.
+Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
+Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{stats book}).
+Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.
+
+In a similar fashion, we can use a non-standardized Student's t-distribution by modifying Eq.~(\ref{eq:baseline-response}). In this case, the response curve can be characterized as follows:
+\begin{equation}
+\begin{split}
+p_{ij}  =  \alpha^{\prime}_{i}\left(1+\frac{\gamma^{\prime}_{i}}{\nu_{i}}\, \left(x_{j}-\beta_{i}\right)^2\right)^{-\frac{\nu_{i}+1}{2}},
+\end{split}
+\label{eq:student-response}
+\end{equation}
+where $\alpha^{\prime}_{i} = f_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \nu_{i}\right)$, $\gamma^{\prime}_{i} = f_{2}\left(\gamma_{i}, \nu_{i}\right)$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(2, \infty\right)$.
+These boundary conditions are necessary in order for the distributions to have finite variance.
+Similar to the description above, we choose an adaptive prior such that $\log\left(\nu_{i}-2\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(1\right)$.
+Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
+In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:student-response}) when $\nu_i\to\infty$ \citep{stats book}.
 
 \subsubsection*{Skewed response curve}
-
-\subsubsection*{Heavy-tail and skewed response curve}
-
-\subsection*{Alternative variance-covariance structures}
-The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information.
-Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}).
-For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.~two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
-One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.~parameter $\alpha_i$---such that
-\begin{equation} 
-\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
-\label{eq:covariance-complex}
-\end{equation}
-where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$.
-Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$.
-Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.
-
-Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
+When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction.
+Likewise, this same skewed shape can also be the result of asymmetric environmental tolerance.
+One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX).
+As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness'.
+Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species' are distribution along the gradient.
+To test for the existence of this feature, we modified the species' response curve in Eq.~(\ref{eq:baseline-response}) as
 \begin{equation}
 \begin{split}
-y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
-\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
-\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
-\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
-\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
-\hat{\alpha}, 
-\hat{\lambda^{k}}, 
-\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
+p_{ij} & =\hat{p}_{ij} \left[1+\text{erf}\left(\lambda_{i}\left(x_{j}-\beta^{\prime}_{i}\right)\sqrt{\frac{\gamma^{\prime}_{i}}{2}}\right)\right] \\
+\text{log}\left(\hat{p}_{ij}\right) & = -\alpha^{\prime}_{i} - \gamma^{\prime}_{i} \left(x_{j}-\beta^{\prime}_{i}\right)^2 ,
 \end{split}
-\label{eq:baseline}
+\label{eq:skewed-response}
 \end{equation}
-Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively\footnote[2]{I'll rewrite the likelihood function to an ordered categorical as soon as I get things to work properly with count data.}.
-While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
-Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.
+where $\alpha^{\prime}_{i} = q_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma^{\prime}_{i}, \lambda_{i}\right)$, $\gamma^{\prime}_{i} = q_{3}\left(\gamma_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution, which we define as $\lambda_{i}\in\left(-10, 10\right)$.
+Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded.
+The function `erf' is the error function (see \citealt{stats book}).
+We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
+This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
+In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{stats book}.
 
-In all cases, all variance-covariance matrices are defined as follows:
-\begin{equation} 
-\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
-\label{eq:covariance}
-\end{equation}
+One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process.
+That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$.
+As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
 
-where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$.
-Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
-Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
-For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
-The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
-For any given hyperparameter, we choose adaptive priors across covariance structures.
-That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
-Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
-We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
-
-\subsection*{Covariance matrices from incomplete categorical and ordinal data}
-The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database.
-More specifically, both the ecological indicator values and range of variation are ordinal traits specified for all species, whereas plants' physiological data are characterized by categorical data containing multiple missing entries.
-These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the parameters of our Bayesian model.
-To do so, we need to compile the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
-
-We define these variance-covariance matrices are defined as follows:
-\begin{equation} 
-\Sigma_{ij} = \eta\,\text{exp}\left(-\rho {D_{ij}}^2\right) + \delta_{ij} \sigma ,
-\label{eq:covariance}
-\end{equation}
-
-where $\Sigma_{ij}$ describes the covariance between any pair of species $i$ and $j$.
-Following this expression, such covariance declines exponentially with the square of the different $D_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
-Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
-For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
-The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
-For any given hyperparameter, we choose adaptive priors across covariance structures.
-That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
-Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
-We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
-
----which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
-In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
-Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
-
-The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-In this model, such distance matrices characterize differences between plant species.
-In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits.
-More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
-In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
-Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
-
-More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
-One way to frame this problem is by using a network representation.
-Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
-Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$.
-That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
-
-Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
-In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
-For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
-Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
-Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
-Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
-\begin{equation}
-Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
-\end{equation}
-Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
-Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
-In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
-While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
-The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
-This algorithm initially assigns random group memberships to both species and traits.
-While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
-Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
-In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
-
-The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
-In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.~the extend to which a species is of one type or another.
-This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
-The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
-Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
-
-\subsection*{Distribution models}
-There is a long list of model structures well suited to characterize species' distributions (see XX for a review); however, we were interested in a model that explicitly incorporates all information regarding plant's environmental preferences found in the floristic database.
-More specifically, we wanted to account for the climatic indicator values and range of variation registered for all plants in our dataset.
-These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions.
-Therefore, we first formulated a baseline model that directly accounts for such prior information.
-
-\subsubsection*{Baseline model}
-Given $y_{ij}$ the presence/absence of any species $i$ in any given site $j$, and a set of $k$ environmental variables $x_{jk}$, we estimate species' distributions as:
-\begin{equation} 
-\begin{split}
-y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right)\\
-\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_{k} \lambda_{ik} \left(x_{jk}-\beta_{ik}\right)^2\\
-\text{log}(\alpha)  & \sim \text{MVNormal}\Big(\hat{\alpha}, \Sigma^{\alpha}\Big)\\
-\beta_{ik}  & \sim \text{MVNormal}\left(\hat{\beta}_{k}, \Sigma^{\beta_{k}}\right)\\
-\text{log}(\lambda_{ik})  & \sim \text{MVNormal}\left(\hat{\lambda_{k}}, \Sigma^{\lambda_{k}}\right)\\
-\hat{\alpha}, 
-\hat{\lambda^{k}}, 
-\hat{\beta^{k}}  & \sim \text{Normal}\left(0,1\right)
-\end{split}
-\label{eq:baseline}
-\end{equation}
-Notice that this model structure assumes all plants to have a uni-modal distributions along each environmental axis (see the model's behaviour in Supplementary Figure XX), where parameters $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the different environmental gradients, respectively\footnote[2]{I'll rewrite the likelihood function to an ordered categorical as soon as I get things to work properly with count data.}.
-While potentially sacrificing predictive accuracy, this model structure allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions via $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-More specifically, we express $\beta_i^k$ and $\log\left(\lambda_i^k\right)$ as multivariate normal distributions---i.e.~Gaussian processes---such that $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$ are variance-covariance matrices describing species' similarity in terms of their average climatic suitability and range of variation along the different environmental gradients, respectively.
-Likewise, $\log\left(\alpha\right)$ is characterized as a Gaussian Process, where the corresponding variance-covariance matrix $\Sigma^{\alpha}$ is designed to also incorporate some of the prior information that we have with regards to species' physiological traits.
-
-In all cases, all variance-covariance matrices are defined as follows:
-\begin{equation} 
-\Sigma^{\chi}_{ij} = \eta_{\chi}\,\text{exp}\left(-\rho_{\chi} {D^{\chi}_{ij}}^2\right) + \delta_{ij} \sigma_{\chi} ,
-\label{eq:covariance}
-\end{equation}
-
-where $\Sigma^{\chi}_{ij}$ describes the covariance between any pair of species $i$ and $j$ for any given parameter $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$.
-Following this expression, such covariance declines exponentially with the square of the different $D^{\chi}_{ij}$, which are distance measures computed using the prior information that we have regarding species' distributions.
-Specifically, given $\alpha_i$, $\beta_i^k$, and $\lambda_i^k$, the distance measures are calculated using plants' physiological traits, ecological indicator values and range of variation, respectively (see below for further details).
-For each covariance matrix, the hyperparameter $\rho_{\chi}$ determines the rate of decline of the covariance between any two species, and $\eta_{\chi}$ defines its maximum value.
-The hyperparameter $\sigma_{\chi}$ describes the additional covariance between the different observations for any given species.
-For any given hyperparameter, we choose adaptive priors across covariance structures.
-That is, and taking $\rho_{\chi}$ as an example, we choose a prior $\log\left(\rho_{\chi}\right)\sim \text{Normal}\left(\hat{\rho}, \sigma_{\rho}\right)$ such that $\hat{\rho}\sim \text{Normal}\left(0, 1\right)$ and $\sigma_{\rho}\sim \text{Exponential}\left(1\right)$.
-Similar priors were chosen for both $\eta_{\chi}$ and $\sigma_{\chi}$.
-We generated the posterior samples for the Bayesian models with the help of the R package `rstan' to \citep{rstan}.
-
-\subsubsection*{Distance matrices}
-The missing component in the description of model (\ref{eq:baseline}) is the distance matrices $D^{\chi}$ used to define the covariance matrices $\Sigma^{\alpha}$, $\Sigma^{\beta_{k}}$ and $\Sigma^{\lambda_{k}}$.
-In this model, such distance matrices characterize differences between plant species.
-In the floristic data, however, the prior information that we have for these differences is represented by a set of ordinal and categorical traits.
-More specifically, both the ecological indicator values and range of variation---which define the prior information that we have for $\beta_i^k$, and $\lambda_i^k$, respectively---are ordinal traits specified for all species.
-In contrast, the plants' physiological data---shaping the prior for the parameters $\alpha_i$---are characterized by categorical data containing multiple missing entries.
-Therefore, we need to carefully compile this data into distance matrices in order to be able to feed this prior information into the model.
-
-More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
-One way to frame this problem is by using a network representation.
-Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
-Notice that this provides as with multiple ways to account for the information---and lack thereof---contained in the different categorical and ordinal traits $M$.
-That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
-
-Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
-In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
-For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
-Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
-Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
-Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
-\begin{equation}
-Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
-\end{equation}
-Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
-Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
-In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
-While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
-The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
-This algorithm initially assigns random group memberships to both species and traits.
-While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
-Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
-In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
-
-The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
-In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.~the extend to which a species is of one type or another.
-This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
-The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
-Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
-
-\subsubsection*{Modifying the variance-covariance structures}
-The model structure defined in Eq.~(\ref{eq:baseline}) allows us to test the effect of adding new information.
-Specifically, we can do this by modifying Eq.~(\ref{eq:covariance}).
-For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---i.e.~two matrices characterizing ecological and environmental traits, or multiple matrices resulting from the different group memberships estimated using the MMSBM.
-One could modify Eq.~(\ref{eq:covariance}) for a particular parameter---e.g.~parameter $\alpha_i$---such that
-\begin{equation} 
-\Sigma^{\alpha}_{ij} = \eta_{\alpha}\,\text{exp}\left(-\sum_k\rho_{\alpha k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma_{\alpha} ,
-\label{eq:covariance-complex}
-\end{equation}
-where now $\rho_{\alpha k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\alpha_i$.
-Notice that the same is true for the covariance of parameters $\beta_i^k$ and $\lambda_i^k$.
-Finally, for all hyperparameters and as described for the baseline model, we use adaptive priors across covariance structures.
-
-% TODO list methods
-% 1 Pomodoro on long-tail distribution.
-Present both generalized Gaussian and Student-t and explain how the condition that we want is to have finite variance and a distribution for which the Gaussian is a special case
-% 1 Pomodoro on finishing baseline and MMSBM
-% 1 Pomodoro on skewed distribution.
-Here, we want the skewed normal distribution because we can measure skeweness without having discontinuities, the Gaussian is a special case, and the tails are normally distributed (check wiki)
-% 1 Pomodoro on finishing baseline and MMSBM
-% 1 Pomodoro on finishing long-tail
-% 1 Pomodoro on finishing skewed
-% 1 Pomodoro on section about variance covariance matrix. Not long, so I can spend time on revising other things
-% 1 Pomodoro on section about other possibilities? Calculate correlations? Add other predictors?
+%\subsubsection*{Fat-tail and skewed response curve}
+% Skewed generalized t distribution
+% https://en.wikipedia.org/wiki/Skewed_generalized_t_distribution
+% mcdonaldPartiallyAdaptiveRobust2016
 
 \section*{Results}
-%We first use the baseline model to estimate the distribution of all species with at least 10 occurrences using all the data that we have with regard to...
+% We first use the baseline model to understand how species are distributed along environmental gradient.
+As expected alpha declines with altitude, implying that the overall...
+study the relationship between mean and variance.
+We found positive relationship...
+Hypothesis!
+Similar results were found using binary data.
+Different species show different correlaitons, invasive present less prevasive...
+When adding additional covariate, such the second axis of our PCA, this relationship vanishes...
+(see Supplementary)
+
+%Other interesting comparisons...
+The value of adding plant physiology as prior information for both the mean and standard deviation of the distributions.
+Informative in both case, specially for...
+Capturing some of bla bla bla.
+
+%Then, we stud the tails of distributions by modifying the baseline model to account for different properties.
+Kurtois.
+Studying the parameter...
+Similar values were found for binary data and the model did not outperform the baseline model
+
+% Next we studied the skewness of distributions, outpreforming the other models.
+When studying the parameter, we found that the mean skewness of species distributions declined...
+Hypothesis!
+When accounting for plant physiology as priors for the skewness of distribution, we found this to be ...
+Figure and Hypothesis.
+The model also showcased the Rapopor's hypothesis
+
+\section*{Discussion}
+% Summary of results
+% Proper test of Rapopor's hypothesis. Different species follow different biogeographical patterns.
+% Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
+% What is the true shape of distributions? Missing bimodal curves
+%
+
+%%%%%%%%%%%%%%%%%%
+%%% Model runs
+%%%%%%%%%%%%%%%%%%
+% Finish categorical 1d for both skewed and normal (2 weeks)
+% Test for heavy tails 1 binomial and categorical (0.5 week)
+% Test for heavy tails 2 binomial and categorical (0.5 week)
+% Categorical 1d for heavy tails 1 (1 weeks)
+% Categorical 1d for heavy tails 2 (1 weeks)
+% Binomial 1d for heavy tails 1 (0.7 weeks)
+% Binomial 1d for heavy tails 2 (0.7 weeks)
+% Binomial 1d for skewed (0.7 weeks)
+% Binomial 1d normal two covariances (1 week)
+% Categorical 1d for skewed with physiology in lambda (1 week)
+% Additional tests for Binomial 1d skewed and normal (1 week)
+
+% With both heavy-tails ~ 10 weeks
+% With only one heavy-tail ~ 8 weeks
+
 \begin{figure}[ht]

### 2021-03-02 manuscript/manuscript.tex
Manuscript changes and data restructuring

@@
 \title{\vspace{-1cm}
-Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions}
+Model linearity breeds contempt: using non-linear models to uncover general biogeographical patterns}
 % Model familiarity breeds contempt: simple models to compare properties of species' distributions
 % Model simplicity breeds contempt: simple models to compare properties of species' distributions
+% Model complexity breeds contempt: using non-linear models to compare basic properties of species' distributions
 \author[1,*]{\normalsize Bernat Bramon Mora}
@@
 The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
-However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been often questioned \citep{gotelliEmpiricalBayesApproach2010, harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019}.
+However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been recently questioned \citep{harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019, poggiatoInterpretationsJointModeling2021}.
 This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
@@
 This dataset was compiled based on 30 years (1961–1990) of records from national weather stations.
-Since most of the data is highly correlated, we calculated the main axes of variation of the following variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig.
+Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig.
 1).
@@
 \end{equation}
-where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$.
+where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$, and $\delta_{ij}$ is the Kronecker delta.
 Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
@@
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have more than 30 occurrences when modelling ordinal data, which is the case for the majority of the results of this work.
-When using presence/absence data, we limit our study to those species for which have more than 10 occurrences.
+Therefore, we focus on those species for which we have more than 10 occurrences when modelling binary data, which is the case for the majority of the results of this work.
+When using ordinal data, we limit our study to those species for which have more than 50 occurrences.
 
@@
 As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
+Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.
 
@@
 \section*{Results}
-% We first use the baseline model to understand how species are distributed along environmental gradient.
-As expected alpha declines with altitude, implying that the overall...
-study the relationship between mean and variance.
-We found positive relationship...
-Hypothesis!
-Similar results were found using binary data.
-Different species show different correlaitons, invasive present less prevasive...
-When adding additional covariate, such the second axis of our PCA, this relationship vanishes...
-(see Supplementary)
+We studied the distribution data to characterize species' realized niches.
+To do so, we selected the main axis of variation of all environmental variables (Supplementary Fig.~XX).
+Then, using the presence or absence of species per site as response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
+Studying the relationship between the mean and variance of the distributions, we found clear evidence of the Rapopor's rule.
+Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}).
+The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Supplementary Fig.~XX).
+This pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
+The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.
+In particular, we found the amplitude of distributions to be positively and negatively correlated with the mean and the variance along the elevation gradient, respectively (Supplementary Fig.~XX).
 
+Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways.
+To do so, we changed the response curve of our Bayesian model to follow either a generalized error distribution (Eq.~\ref{eq:generalized-response}) or a non-standardized Student's t-distribution (Eq.~\ref{eq:student-response}).
+%The posterior distributions of the hyperparameters in Eq.~(\ref{eq:covariance-baseline}) revealed the value of this prior information on the different parameter estimates.
+to understand the value of our prior information.
+
+We then studied the level of information contained in physiological traits to explain the properties of distributions.
+Specifically, using Eq.~(\ref{eq:covariance-complex}), we accounted for both physiological and environmental traits in the variance-covariance structure for the mean and standard deviation of distributions.
+We information
+As expected, the posterior distributions also showcase a positive relationship between...
+This is expected as ...
+Likewise, studying ...
+Following this, we added additional ...
+using Eq.~(\ref{eq:covariance-complex}).
+When studying plant...
 %Other interesting comparisons...
@@
 Figure and Hypothesis.
-The model also showcased the Rapopor's hypothesis
+The model also showcased the Rapopor's hypothesis...
 

### 2021-03-02 code/paper-models/1d-baseline.R

@@ -1,4 +1,4 @@
-source("./prepare-data.R")
+# source("./prepare-data.R")
 source("./models.R")
 library(rethinking)
 library(rstan)
@@ -14,25 +14,21 @@ rstan_options(auto_write = TRUE)
 baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
   # Fixing some of the options
   variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
-  gp_type <- 2
   ndim <- 2
   pca <- T
-  
-  # File name extension
-  extension <- ""
   
   # If we are dealing with simulated data
   if(simulated){
     extension <- "1d-simulated"
+  }else{
+    extension <- "1d"
   }
-  extension <- paste(extension, as.character(gp_type), sep="")
   
   if(min.occurrence==10){
-    extension2 <- ""
+    extension2 <- "-"
   }else{
-    extension2 <- "min30-"
+    extension2 <- paste("min",min.occurrence,"-",sep="")
   }
-  
   
   # Load the data
   if(recompile){
@@ -45,7 +41,7 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
         variables <- paste("PC", 1:ndim, sep="")
       }
     }
-    filename <- paste("../../data/processed/jsdm/", extension, paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+    filename <- paste("../../data/processed/jsdm/", extension, "-",paste(variables, collapse = ""), extension2, "data.rds", sep = "")
     
     if (file.exists(filename)){
       question <- askYesNo("Do you want to overwrite the file?", default = F, 
@@ -69,7 +65,7 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
       }
     }
     if(is.null(d)){
-      d <- readRDS(file = paste("../../data/processed/jsdm/", extension, paste(variables, collapse = ""), extension2, "data.rds", sep = ""))
+      d <- readRDS(file = paste("../../data/processed/jsdm/", extension,"-", paste(variables, collapse = ""), extension2, "data.rds", sep = ""))
     }
   }
   
[…]

### 2021-03-02 code/paper-models/1d-categorical.R

@@ -6,29 +6,24 @@ library(rstan)
 options(mc.cores = parallel::detectCores())
 rstan_options(auto_write = TRUE)
 
-
 categorical.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
   # Fixing some of the options
   variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
-  gp_type <- 2
   ndim <- 2
   pca <- T
   
-  # File name extension
-  extension <- ""
-  
   # If we are dealing with simulated data
   if(simulated){
-    extension <- "simulated-categorical"
+    extension <- "1d-simulated-categorical"
+  }else{
+    extension <- "1d-categorical"
   }
-  extension <- paste(extension, as.character(gp_type), sep="")
   
   if(min.occurrence==10){
-    extension2 <- ""
+    extension2 <- "-"
   }else{
-    extension2 <- "min30-"
+    extension2 <- paste("min",min.occurrence,"-",sep="")
   }
-  
   
   # Load the data
   if(recompile){
@@ -41,7 +36,7 @@ categorical.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=
         variables <- paste("PC", 1:ndim, sep="")
       }
     }
-    filename <- paste("../../data/processed/jsdm/", extension, paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+    filename <- paste("../../data/processed/jsdm/", extension, "-",paste(variables, collapse = ""), extension2, "data.rds", sep = "")
     
     if (file.exists(filename)){
       question <- askYesNo("Do you want to overwrite the file?", default = F, 
@@ -65,7 +60,7 @@ categorical.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=
       }
     }
     if(is.null(d)){
-      d <- readRDS(file = paste("../../data/processed/jsdm/", extension, paste(variables, collapse = ""), extension2, "data.rds", sep = ""))
+      d <- readRDS(file = paste("../../data/processed/jsdm/", extension, "-", paste(variables, collapse = ""), extension2, "data.rds", sep = ""))
     }
   }
   
[…]

### 2021-03-02 code/paper-models/2d-baseline.R

@@ -242,3 +242,13 @@ baseline.traits <- function(d = NULL, recompile = T, simulated=T, min.occurrence
   saveRDS(mfit_5.1, file = paste(ofolder, "baseline-model-traits", extension,".rds", sep=""))
   return(mfit_5.1)
 }
+
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!
+# CHANGE FILE NAMES!!!

### 2021-03-02 code/paper-models/visualizing-results.R

@@ -152,7 +152,7 @@ plot.simulated.data <- function(beta=T, gp_type = 2){
 }
 
 compare.models <- function(){
-  # m1 <- readRDS(paste("../../results/models/min30-skew-model-traits-1d2.rds", sep=""))
+  m1 <- readRDS(paste("../../results/models/min30-skew-model-traits-1d2.rds", sep=""))
   # m2 <- readRDS(paste("../../results/models/min30-baseline-model-1d2.rds", sep=""))
   # comp <- rethinking::compare(m1, m2, refresh = 1)
   
@@ -425,6 +425,7 @@ plot.actual.data <- function(model=NULL){
     meanPI <- mean(corPI)
     sdPI <- sd(corPI)
     corPIci <- round(sort(c(as.vector(PI(corPI)),meanPI)), 2)
+    print(corPIci)
     
     # Generate empty list of plots
     plist = list()
@@ -513,8 +514,8 @@ plot.actual.data.alpha <- function(model=NULL){
   for(i in 1:2){
     if(i==1){
       target <- post$beta[,i,]      
-      mu_target <- mu_beta[[1]]
-      ci_target <- ci_beta[[1]]
+      mu_target <- mu_alpha[[1]]
+      ci_target <- ci_alpha[[1]]
     }else{
       target <- post$gamma[,i,]
       mu_target <- mu_gamma[[1]]
@@ -654,6 +655,20 @@ plot.actual.data.means <- function(model=NULL){
     xlab=""
     breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4))+1
     
+    if(i==1){
+      correlations <- c()
+      corPI <- sapply(1:dim(post$beta)[1], function(x) cor(post$beta[x,i,NEO$V3==1], post$gamma[x,1,NEO$V3==1]))
+      meanPI <- mean(corPI)
+      sdPI <- sd(corPI)
+      corPIci <- round(c(meanPI, sdPI), 2)
+      correlations <- rbind(correlations, c("neophytes", corPIci))
+      corPI <- sapply(1:dim(post$beta)[1], function(x) cor(post$beta[x,i,NEO$V3!=1], post$gamma[x,1,NEO$V3!=1]))
+      meanPI <- mean(corPI)
+      sdPI <- sd(corPI)
+      corPIci <- round(c(meanPI, sdPI), 2)
+      correlations <- rbind(correlations, c("native", corPIci))
+    }
+    
[…]

### 2021-03-02 manuscript/manuscript.tex
Preparing binomial models

@@
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have more than 10 occurrences when modelling binary data, which is the case for the majority of the results of this work.
+Therefore, we focus on those species for which we have at least 10 occurrences when modelling binary data, which is the case for the majority of the results of this work.
 When using ordinal data, we limit our study to those species for which have more than 50 occurrences.
+% A random subset of a 100 species of those with at least 10 occurrences.
 

### 2021-03-02 code/paper-models/1d-baseline.R

@@ -1,5 +1,5 @@
 # source("./prepare-data.R")
-source("./models.R")
+source("./models-binomial.R")
 library(rethinking)
 library(rstan)
 
@@ -82,6 +82,7 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   
   dat_5.1 <- list(N=N,
                   L=L,
+                  minp=1e-100,
                   Y=t(obs),
                   X1=X1,
                   Dmat_b=Dis_b,
@@ -199,6 +200,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
   
   dat_5.1 <- list(N=N,
                   L=L,
+                  minp=1e-100,
                   Y=t(obs),
                   X1=X1,
                   Dmat_b=Dis_b,
@@ -245,10 +247,13 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
 }
 
 min.occurrence <- 10
-d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2min",min.occurrence,"-data.rds", sep=""))
+if(min.occurrence==10){
+  d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2-data.rds", sep=""))
+}else{
+  d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2min",min.occurrence,"-data.rds", sep=""))
+}
 skew.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
-baseline.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
-
+# baseline.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 
 
 # # 

### 2021-03-02 code/paper-models/1d-categorical.R

@@ -112,7 +112,7 @@ categorical.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=
   model_code=categorical.model.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 15
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -121,7 +121,7 @@ categorical.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=
                      data=dat_5.1 ,
                      chains=n_chains_5.1 ,
                      cores= n_chains_5.1 ,
-                     warmup=1000, iter=1600,
+                     warmup=1000, iter=2000,
                      init=init_5.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
 
 
@@ -238,7 +238,7 @@ categorical.skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurr
   model_code=categorical.model.skew.1d
 
   # Initialize data structure
-  n_chains_5.1 <- 15
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -247,7 +247,7 @@ categorical.skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurr
                      data=dat_5.1 ,
                      chains=n_chains_5.1 ,
                      cores= n_chains_5.1 ,
-                     warmup=1000, iter=1600,
+                     warmup=1000, iter=2000,
                      init=init_5.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
 
 
@@ -256,6 +256,10 @@ categorical.skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurr
 }
 
 min.occurrence <- 10
-d <- readRDS(file = paste("../../data/processed/jsdm/1d-categorical-PC1PC2min",min.occurrence,"-data.rds", sep=""))
+if(min.occurrence==10){
+  d <- readRDS(file = paste("../../data/processed/jsdm/1d-categorical-PC1PC2-data.rds", sep=""))
+}else{
+  d <- readRDS(file = paste("../../data/processed/jsdm/1d-categorical-PC1PC2min",min.occurrence,"-data.rds", sep=""))
+}
[…]

### 2021-03-02 code/paper-models/models-binomial.R

@@ -1,0 +1,331 @@
+# Binomial 1d
+base.model.1d <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+}
+data{
+    int N;
+    int L;
+    real minp;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    row_vector[N] X1;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+}
+parameters{
+    vector[L] zalpha;
+    vector[L] zbeta;
+    vector[L] zgamma;
+    real alpha_bar;
+    real beta_bar;
+    real gamma_bar;
+    real<lower=0> sigma_a;
+    real<lower=0> sigma_b;
+    real<lower=0> etasq_b;
+    real<lower=0> rhosq_b;
+    real<lower=0> sigma_g;
+    real<lower=0> etasq_g;
+    real<lower=0> rhosq_g;
+}
+transformed parameters{
+    vector[L] alpha;
+    vector[L] beta;
+    vector[L] gamma;
+    matrix[L, L] L_SIGMA_b;
+    matrix[L, L] L_SIGMA_g;
+
+    alpha = exp(zalpha * sigma_a + alpha_bar);
+
+    L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
+    beta = L_SIGMA_b * zbeta + beta_bar;
+
+    L_SIGMA_g = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g, rhosq_g, sigma_g));
+    gamma = L_SIGMA_g * zgamma + gamma_bar;
+    gamma = exp(gamma);
+    
+
+}
+model{
+    // matrix[L,N] p;
+
+    sigma_a ~ exponential( 1 );
+    sigma_b ~ exponential( 1 );
+    sigma_g ~ exponential( 1 );
+    etasq_b ~ exponential( 1 );
+    etasq_g ~ exponential( 1 );
[…]

### 2021-03-02 code/paper-models/models.R

@@ -191,212 +191,6 @@ model{
 }
 "
 
-# Binomial 1d
-base.model.1d <- "
-functions{
-    matrix cov_GPL2(matrix x, real a, real b, real delta) {
-        int N = dims(x)[1];
-        matrix[N, N] K;
-
-        for (i in 1:(N-1)) {
-          K[i, i] = a + delta;
-          for (j in (i + 1):N) {
-            K[i, j] = a * exp(- b * square(x[i,j]) );
-            K[j, i] = K[i, j];
-          }
-        }
-        K[N, N] = a + delta;
-        return K;
-    }
-}
-data{
-    int N;
-    int L;
-    // int K;
-    // int Y[K];
-    int Y[L, N];
-    row_vector[N] X1;
-    matrix[L,L] Dmat_b;
-    matrix[L,L] Dmat_g;
-}
-parameters{
-    vector[L] zalpha;
-    vector[L] zbeta;
-    vector[L] zgamma;
-    real alpha_bar;
-    real beta_bar;
-    real gamma_bar;
-    real<lower=0> sigma_a;
-    real<lower=0> sigma_b;
-    real<lower=0> etasq_b;
-    real<lower=0> rhosq_b;
-    real<lower=0> sigma_g;
-    real<lower=0> etasq_g;
-    real<lower=0> rhosq_g;
-}
-transformed parameters{
-    vector[L] alpha;
-    vector[L] beta;
-    vector[L] gamma;
-    matrix[L, L] L_SIGMA_b;
-    matrix[L, L] L_SIGMA_g;
-
-    alpha = exp(zalpha * sigma_a + alpha_bar);
-
-    L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
-    beta = L_SIGMA_b * zbeta + beta_bar;
-
-    L_SIGMA_g = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g, rhosq_g, sigma_g));
-    gamma = L_SIGMA_g * zgamma + gamma_bar;
-    gamma = exp(gamma);
-    
-
-}
-model{
-    // matrix[L,N] p;
-
-    sigma_a ~ exponential( 1 );
-    sigma_b ~ exponential( 1 );
-    sigma_g ~ exponential( 1 );
-    etasq_b ~ exponential( 1 );
-    etasq_g ~ exponential( 1 );
[…]

### 2021-03-02 code/paper-models/visualizing-results.R

@@ -514,8 +514,8 @@ plot.actual.data.alpha <- function(model=NULL){
   for(i in 1:2){
     if(i==1){
       target <- post$beta[,i,]      
-      mu_target <- mu_alpha[[1]]
-      ci_target <- ci_alpha[[1]]
+      mu_target <- mu_beta[[1]]
+      ci_target <- ci_beta[[1]]
     }else{
       target <- post$gamma[,i,]
       mu_target <- mu_gamma[[1]]

### 2021-03-02 manuscript/manuscript.tex
Adding generalized error distributino

@@
 
-\section*{Abstract}
-We know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
+%\section*{Abstract}
+%We know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
 However, much less is known about how species' distributions compare to each other.
@@
 In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and growth tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent years).
+Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
 We describe this information in more detail in the Supplementary Table 1.
@@
 To do so, we selected the main axis of variation of all environmental variables (Supplementary Fig.~XX).
-Then, using the presence or absence of species per site as response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
+Then, using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
 Studying the relationship between the mean and variance of the distributions, we found clear evidence of the Rapopor's rule.
@@
 
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.9\textwidth]{figures/figure1}
+    	  \vspace{0.3cm}
+	   \caption{Relationship between mean and variance of species' distributions.
+These are the results for the main axis of variation for the climatic data (results for the second axis of variation presented in the Supplementary Fig.~2).}
+      \label{fig:correlation}
+\end{figure}
+
 Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways.
 To do so, we changed the response curve of our Bayesian model to follow either a generalized error distribution (Eq.~\ref{eq:generalized-response}) or a non-standardized Student's t-distribution (Eq.~\ref{eq:student-response}).
+In the case of a generalized error distribution, we found...
 %The posterior distributions of the hyperparameters in Eq.~(\ref{eq:covariance-baseline}) revealed the value of this prior information on the different parameter estimates.
@@
 
-\begin{figure}[ht]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics[width=0.9\textwidth]{figures/figure1}
-    	  \vspace{0.3cm}
-	   \caption{Relationship between mean and variance of species' distributions.
-These are the results for the main axis of variation for the climatic data (results for the second axis of variation presented in the Supplementary Fig.~2).}
-      \label{fig:correlation}
-\end{figure}
-
 \begin{figure}[h]

### 2021-03-02 code/paper-models/1d-baseline.R

@@ -128,6 +128,127 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   return(mfit_5.1)
 }
 
+generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+  # Fixing some of the options
+  variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
+  ndim <- 2
+  pca <- T
+  
+  # If we are dealing with simulated data
+  if(simulated){
+    extension <- "1d-generror-simulated"
+  }else{
+    extension <- "1d-generror"
+  }
+  
+  if(min.occurrence==10){
+    extension2 <- "-"
+  }else{
+    extension2 <- paste("min",min.occurrence,"-",sep="")
+  }
+  
+  # Load the data
+  if(recompile){
+    d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="gauss.gauss", min.occurrence=min.occurrence)
+    # rename variables
+    if(simulated){
+      variables <- c("S1", "S2")
+    }else{
+      if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+      }
+    }
+    filename <- paste("../../data/processed/jsdm/", extension, "-",paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+    
+    if (file.exists(filename)){
+      question <- askYesNo("Do you want to overwrite the file?", default = F, 
+                           prompts = getOption("askYesNo", gettext(c("Yes", "No", "Cancel"))))
+      if(is.na(question)){question <- F}
+      if(question){
+        saveRDS(d, file=filename)
+      }else{
+        stop("you should add 'recompile=T'")
+      }                                
+    }else{
+      saveRDS(d, file=filename)
+    }
+  }else{
+    # rename variables
+    if(simulated){
+      variables <- c("S1", "S2")
+    }else{
+      if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+      }
+    }
+    if(is.null(d)){
[…]

### 2021-03-02 code/paper-models/models-binomial.R

@@ -97,7 +97,115 @@ generated quantities{
 }
 "
 
+# Binomial 1d
+base.model.generror.1d <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
 
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+}
+data{
+    int N;
+    int L;
+    real minp;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    row_vector[N] X1;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+}
+parameters{
+    vector[L] zalpha;
+    vector[L] zbeta;
+    vector[L] zgamma;
+    vector[L] znu;
+    real alpha_bar;
+    real beta_bar;
+    real gamma_bar;
+    real nu_bar;
+    real<lower=0> sigma_a;
+    real<lower=0> sigma_n;
+    real<lower=0> sigma_b;
+    real<lower=0> etasq_b;
+    real<lower=0> rhosq_b;
+    real<lower=0> sigma_g;
+    real<lower=0> etasq_g;
+    real<lower=0> rhosq_g;
+}
+transformed parameters{
+    vector[L] alpha;
+    vector[L] beta;
+    vector[L] gamma;
+    vector[L] nu;
+    matrix[L, L] L_SIGMA_b;
+    matrix[L, L] L_SIGMA_g;
+
+    alpha = exp(zalpha * sigma_a + alpha_bar);
+    nu = exp(znu * sigma_n + nu_bar)+1;
+
+    L_SIGMA_b = cholesky_decompose(cov_GPL2(Dmat_b, etasq_b, rhosq_b, sigma_b));
+    beta = L_SIGMA_b * zbeta + beta_bar;
+
+    L_SIGMA_g = cholesky_decompose(cov_GPL2(Dmat_g, etasq_g, rhosq_g, sigma_g));
+    gamma = L_SIGMA_g * zgamma + gamma_bar;
+    gamma = exp(gamma);
+    
+    for (i in 1:L){
[…]

### 2021-03-02 code/paper-models/run_inte.sh
Final fixes for first run of fat-tail model

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=4548]" -W 200:00 -n 15 "R --vanilla --slave < 1d-categorical.R"
+bsub -R "rusage[mem=4548]" -W 120:00 -n 15 "R --vanilla --slave < 1d-baseline.R"

### 2021-03-05 manuscript/manuscript.tex
Comparing models function and changes to the manuscript

@@
 % Margin
-\usepackage[margin=2.9cm]{geometry}
+\usepackage[margin=2.8cm]{geometry}
 
@@
 \title{\vspace{-1cm}
-Model linearity breeds contempt: using non-linear models to uncover general biogeographical patterns}
+Model linearity breeds contempt: using Bayesian non-linear models to uncover general biogeographical patterns}
 % Model familiarity breeds contempt: simple models to compare properties of species' distributions
@@
 
-%\section*{Abstract}
-%We know a lot about the factors that could theoretically influence species' distributions, and a rapidly growing body of research have been primarily focused on trying to untangle some of such biotic and abiotic predictors---with an increasing effort placed in improving the predictive power of statistical models.
-However, much less is known about how species' distributions compare to each other.
-Here, we use a conceptually more conservative approach to instead understand and compare basic aspects regarding the shape of species' distribution along environmental gradients.
+\section*{Abstract}
+
+Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades.
+Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change.
+That said, uncovering the mechanisms shaping species' realized niches has been one of the main driving forces behind the development of these models.
+That is, recent efforts have been often focused on understanding which biotic and abiotic factors are good predictors of species' niches---with an increasing effort placed in improving the predictive power of the statistical models.
+However, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches.
+These models account for all prior knowledge we have regarding their shape, including expert knowledge on species' environmental preferences and physiology.
+With this approach, we are able to shed light on the true shape of empirical species' distributions.
+Moreover, they allow us to tackle long-standing hypothesis regarding general biogeographical patterns.
+In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients.
+Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species.
+Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
 
@@
 Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}.
-These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
+These models estimate species' realized niches using several covariates, including environmental variables \citep{guisanPredictingSpeciesDistribution2005}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
 More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}.
@@
 
-Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}.
-In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY).
+Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
+In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see \citealt{review} and \citealt{good example}).
 This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings.
@@
 From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
-This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
 Other factors might then condition species distributions to showcase fat-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
@@
 Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
-That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}).
+That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
 Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
@@
 
+%This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
 %\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species distributions.
@@
 where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the an environmental gradient, respectively.
-Notice that $F$ characterizes a Bernoulli distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods).
+Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods).
 For the sake of simplicity, we use only one environmental variable to characterize species' probability distribution.
@@
 Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
-In the floristic database, this information is represented by the set of ordinal specified for the different species.
+In the floristic database, this information is represented by the set of ordinal traits specified for the different species.
 While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}).
@@
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have at least 10 occurrences when modelling binary data, which is the case for the majority of the results of this work.
-When using ordinal data, we limit our study to those species for which have more than 50 occurrences.
+Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data, which corresponds to the majority of the results of this work.
+When using ordinal data, we limit our study to a random subset of a 100 species of those with at least 20 occurrences to facilitate the sampling of the models.
 % A random subset of a 100 species of those with at least 10 occurrences.
@@
 Indeed, fat-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have fat-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
-Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions \citep{ref}.
-To accommodate this feature into our baseline model, we could consider response curves that follow a generalized error or a non-standardized Student's t-distributions.
+Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions.
+To accommodate this feature into our baseline model, we could consider response curves that follow a generalized error or a non-standardized Student's t-distributions\footnote[3]{\label{note2} I might end up dropping the Student's t-distribution case, as I am having some trouble getting the sampling of the model to converge.
+I will give it another go nonetheless.}.
 In both cases, the normal distribution is a special case of them, and both contain parameters that regulate the kurtosis of the probability distribution (Supplementary Figure XX).
@@
 Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
-Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{stats book}).
+Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}).
 Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.
@@
 Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
-In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:student-response}) when $\nu_i\to\infty$ \citep{stats book}.
+In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:student-response}) when $\nu_i\to\infty$ \citep{jackmanBayesianAnalysisSocial2009}.
 
@@
 Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded.
-The function `erf' is the error function (see \citealt{stats book}).
+The function `erf' is the error function (see \citealt{ashourApproximateSkewNormal2010}).
 We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
 This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
-In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{stats book}.
+In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{ashourApproximateSkewNormal2010}.
 
@@
 Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}).
-The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Supplementary Fig.~XX).
+The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Fig.~\ref{fig:universality}).
 This pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
@@
     \vspace{0.5cm}
-    \includegraphics[width=0.9\textwidth]{figures/figure1}
+    \includegraphics[width=0.9\textwidth]{figures/figure1-prime}
     	  \vspace{0.3cm}
 	   \caption{Relationship between mean and variance of species' distributions.
-These are the results for the main axis of variation for the climatic data (results for the second axis of variation presented in the Supplementary Fig.~2).}
+Posterior distributions for parameters $\beta_i$ and $\gamma_i$ from Eq.~(\ref{eq:baseline-response}) across species, and the relationship between them.
+Panel (a) describes the relationship between range size and elevation.
+Every dot represents the relationship between the mean values for the $\beta_i$ and $\gamma_i$ estimates of the different species.
+The value in the top-right corner of the plot displays the Pearson's correlation between these parameters calculated across samples of the posterior distributions.
+Panel (b) describes the $\beta_i$ posterior distribution estimated for all species.
+Panel (c) describes the $\gamma_i$ posterior distribution estimated for all species.
+In (b) and (c), the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.}
+	   %7.5-5.5
       \label{fig:correlation}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
+    	  \vspace{0.3cm}
+	   \caption{Universality of the relationship between mean and variance of species' distributions.
+Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a.
+Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
+Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
+Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).
+{\color{red} I might move this to the Supplementary Information.}}
+	   %7.5-5.5
+      \label{fig:universality}
 \end{figure}
@@
 To do so, we changed the response curve of our Bayesian model to follow either a generalized error distribution (Eq.~\ref{eq:generalized-response}) or a non-standardized Student's t-distribution (Eq.~\ref{eq:student-response}).
-In the case of a generalized error distribution, we found...
+Studying the WAIC values, both non-linear regressions outperformed the baseline model (Supplementary Fig.~XX)\footnote[2]{\label{note1} I am still waiting on the results for this (currently running in the cluster), and this is just my prior expectation based on what I've seen in some of the other models I've been working with.}.
+In the case of a generalized error distribution, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with a Gaussian shape (Fig.~\ref{fig:kurtosis-skewness}).
+Similarly, we found the shape parameter in the non-standardized Student's t-distribution to present large values, also approximating the normal case\footref{note1}.
+However, in both cases, these parameter estimates displayed a lot of variation between species, which indicates that the shape of the tails of the distribution is species-specific (potentially other figures in the Supplementary).
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.8\textwidth]{figures/kurto-skew}
+    	  \vspace{0.3cm}
+	   \caption{Average kurtosis and skewness of species' distributions.
+Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively.
+Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution.
+Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution.
+In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
+      \label{fig:kurtosis-skewness}
+\end{figure}
+
+Finally, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}).
+Based on the estimates for the WAIC values, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions showed steeper declines towards stressful conditions (Fig.~\ref{fig:kurtosis-skewness}).
+That is, distributions are generally skewed towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
+Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species physiology (New Fig.~XX)\footref{note1}.
+
+{\color{gray}
+[Note: Reports on the hyperparameters describing the different covariance matrices missing at the moment.
+I think it's interesting, and I will certainly end up adding things here about this.
+I will also add a third figure showing how lambda varies with the corresponding variance-covariance matrix of species' environmental and physiological traits.]}
+
 %The posterior distributions of the hyperparameters in Eq.~(\ref{eq:covariance-baseline}) revealed the value of this prior information on the different parameter estimates.
@@
 
-We then studied the level of information contained in physiological traits to explain the properties of distributions.
+%We then studied the level of information contained in physiological traits to explain the properties of distributions.
 Specifically, using Eq.~(\ref{eq:covariance-complex}), we accounted for both physiological and environmental traits in the variance-covariance structure for the mean and standard deviation of distributions.
 We information
-As expected, the posterior distributions also showcase a positive relationship between...
+%As expected, the posterior distributions also showcase a positive relationship between...
 This is expected as ...
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%       DISCUSSION
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Discussion}
+%In this work, we used non-linear response curves to model the distribution of species across an environmental gradient.
+First, we used a baseline model that considered species to present bell-shaped distributions.
+Studying the parameter estimates across species, we were able to showcase how species follow the Rapopor's rule along an elevation gradient.
+Second, we used several...
+Structure of the discussion section:
+\begin{enumerate}
+\item Summary of results.
+\item Proper test of Rapopor's hypothesis. Different species follow different biogeographical patterns.
+\item Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
+\item What is the true shape of species' distributions? These display fat-tail and skewed distributions. 
+\item Future directions.
+Missing bimodal curves.
+Using this information to understand where jSDMs estimate interactions between species.
+Further test of the ability of traits to predict those parameter estimates.
+\end{enumerate}
 % Summary of results
@@
 
-\begin{figure}[h]
-  \centering
-    \vspace{0.5cm}
-    \includegraphics[width=0.9\textwidth]{figures/invasive}
-    	  \vspace{0.3cm}
-	   \caption{Are there clear geographical patterns for neophytes and for species with decreasing or increasing abundance?}
-      \label{fig:neophytes}
-\end{figure}
-
-%\section*{Discussion}
 \clearpage

### 2021-03-05 code/paper-models/1d-baseline.R

@@ -367,7 +367,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
   return(mfit_5.1)
 }
 
-min.occurrence <- 30
+min.occurrence <- 10
 if(min.occurrence==10){
   d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2-data.rds", sep=""))
 }else{

### 2021-03-05 code/paper-models/visualizing-results.R

@@ -153,8 +153,9 @@ plot.simulated.data <- function(beta=T, gp_type = 2){
 
 compare.models <- function(){
   m1 <- readRDS(paste("../../results/models/min30-skew-model-traits-1d2.rds", sep=""))
-  # m2 <- readRDS(paste("../../results/models/min30-baseline-model-1d2.rds", sep=""))
-  # comp <- rethinking::compare(m1, m2, refresh = 1)
+  m2 <- readRDS(paste("../../results/models/min30-baseline-model-1d2.rds", sep=""))
+  m3 <- readRDS(paste("../../results/models/min30-1d-generror.rds", sep=""))
+  comp <- rethinking::compare(m1, m2, m3, refresh = 1)
   
   # Parameters for plots
   extra <- 1
@@ -219,20 +220,22 @@ compare.models <- function(){
                       # top=textGrob("First axis", rot = 0, vjust = 0.9,gp=gpar(fontsize=10)),
   )
   print(p)
+
+  post <- extract.samples(m3, n = 1000, pars=c("nu_bar")) 
   
-  post <- extract.samples(m1, n = 1000, pars=c("lambda_bar")) 
-  
-  g <- ggplot(data.frame(x=post$lambda_bar), aes(x=x))+
+  plist <- list()
+  nu_hat <- exp(post$nu_bar)+1
+  nu_hat <- gamma(5/nu_hat)*gamma(1/nu_hat)/(gamma(3/nu_hat))**2-3
+  plist[[1]] <- ggplot(data.frame(x=nu_hat), aes(x=x))+
     stat_density(geom = "line", position = "identity") +    # geom_vline(aes(xintercept=1), color="black", linetype="dashed", size=0.5) +
     theme_bw() +
     geom_vline(xintercept = 0, colour = "red", linetype="dashed") +
-    ggtitle("average skewness of distributions")+
+    ggtitle("(a)  average kurtosis of distributions")+
     # scale_fill_grey() +
-    ylab("probability density")+
-    xlab(expression(lambda))+
-    scale_x_continuous(limits = c(-2.5, 0.1))+
+    ylab("density")+
+    xlab("kurtosis")+
+    scale_x_continuous(limits = c(-0.5, 0.5))+
     theme(text = element_text(size=10),
-          axis.title.y=element_blank(),
           panel.grid.major = element_blank(),
[…]

### 2021-03-05 code/paper-models/compare-models.R
Adding functions to compare models in the cluster

@@ -1,0 +1,9 @@
+library(rethinking)
+library(rstan)
+
+m1 <- readRDS(paste("/cluster/scratch/bemora/plant-stan/min30-skew-model-traits-1d2.rds", sep=""))
+m2 <- readRDS(paste("/cluster/scratch/bemora/plant-stan/min30-baseline-model-1d2.rds", sep=""))
+m3 <- readRDS(paste("/cluster/scratch/bemora/plant-stan/min30-1d-generror.rds", sep=""))
+comp <- rethinking::compare(m1, m2, m3)
+
+saveRDS(comp, file = "/cluster/scratch/bemora/plant-stan/comparision.rds")

### 2021-03-05 code/paper-models/run_compare.sh

@@ -1,0 +1,3 @@
+#!/usr/bin/env bash
+
+bsub -R "rusage[mem=68220]" "R --vanilla --slave < compare-models.R"

### 2021-03-05 manuscript/manuscript.tex
Adding test for generalized error

@@
 From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
-Other factors might then condition species distributions to showcase fat-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
 The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
@@
 
-\subsubsection*{Fat-tail response curve}
-Distributions with fat-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
-Indeed, fat-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have fat-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
+\subsubsection*{Heavy-tail response curve}
+Distributions with heavy-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
+Indeed, heavy-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
 Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions.
-To accommodate this feature into our baseline model, we could consider response curves that follow a generalized error or a non-standardized Student's t-distributions\footnote[3]{\label{note2} I might end up dropping the Student's t-distribution case, as I am having some trouble getting the sampling of the model to converge.
-I will give it another go nonetheless.}.
-In both cases, the normal distribution is a special case of them, and both contain parameters that regulate the kurtosis of the probability distribution (Supplementary Figure XX).
-
-Starting with the generalized error distribution, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
+To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Supplementary Figure XX).
+In particular, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
 \begin{equation}
@@
 
-In a similar fashion, we can use a non-standardized Student's t-distribution by modifying Eq.~(\ref{eq:baseline-response}). In this case, the response curve can be characterized as follows:
-\begin{equation}
-\begin{split}
-p_{ij}  =  \alpha^{\prime}_{i}\left(1+\frac{\gamma^{\prime}_{i}}{\nu_{i}}\, \left(x_{j}-\beta_{i}\right)^2\right)^{-\frac{\nu_{i}+1}{2}},
-\end{split}
-\label{eq:student-response}
-\end{equation}
-where $\alpha^{\prime}_{i} = f_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \nu_{i}\right)$, $\gamma^{\prime}_{i} = f_{2}\left(\gamma_{i}, \nu_{i}\right)$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(2, \infty\right)$.
-These boundary conditions are necessary in order for the distributions to have finite variance.
-Similar to the description above, we choose an adaptive prior such that $\log\left(\nu_{i}-2\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(1\right)$.
-Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
-In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:student-response}) when $\nu_i\to\infty$ \citep{jackmanBayesianAnalysisSocial2009}.
+Alternatively, we could have used other distributions that present heavy-tails and fulfil the selection criteria described above.
+For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Supplementary Fig.~XX).
+However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Supplementary Fig.~XX), and the sampling of the model is somewhat more challenging (ref).
 
@@
 As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness'.
-Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species' are distribution along the gradient.
+Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species are distributed along the gradient.
 To test for the existence of this feature, we modified the species' response curve in Eq.~(\ref{eq:baseline-response}) as
@@
 We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
-This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as above (see Supplementary Table 2; Code Availability section).
+This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as described for the baseline model (see Supplementary Table 2; Code Availability section).
 In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{ashourApproximateSkewNormal2010}.
@@
 
-%\subsubsection*{Fat-tail and skewed response curve}
-% Skewed generalized t distribution
-% https://en.wikipedia.org/wiki/Skewed_generalized_t_distribution
-% mcdonaldPartiallyAdaptiveRobust2016
+\subsubsection*{Heavy-tail and skewed response curve}
+Finally, one could also consider a response curve with both kurtosis and skewness.
+A convenient way to achieve this is by considering a response curve that follows a skewed generalized error distribution (also see XX).
+This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness.
+The skewed generalized error distribution can be considered by modifying the species' response curve in Eq.~(\ref{eq:baseline-response}) as
+\begin{equation}	
+\begin{split}
+\text{log}\left(p_{ij}\right) = -\alpha_{i} - \left(\frac{\gamma^{\prime}_{i}\,|x_{j}-\beta^{\prime}_{i}|}{1+\lambda_{i} \text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{\nu_{i}} ,
+\end{split}
+\label{eq:heavy-skewed-response}
+\end{equation}
+where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively.
+$\text{sgn}\left(x\right)$ is the sign function.
+We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}.
+However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(2\lambda_{i}-1\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
+Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as for the baseline model (see Supplementary Table 2; Code Availability section).
+Notice that the generalized error distribution (Eq.~\ref{eq:generalized-response}) and the skew normal distribution (Eq.~\ref{eq:skewed-response}) are special cases of Eq.~(\ref{eq:heavy-skewed-response}) when $\lambda_i=0$ and $\nu_i=2$, respectively.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
 Similarly, we found the shape parameter in the non-standardized Student's t-distribution to present large values, also approximating the normal case\footref{note1}.
-However, in both cases, these parameter estimates displayed a lot of variation between species, which indicates that the shape of the tails of the distribution is species-specific (potentially other figures in the Supplementary).
+However, in both cases, these parameter estimates displayed a lot of variation between species, which indicates that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
 
@@
 Finally, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}).
-Based on the estimates for the WAIC values, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
-Studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions showed steeper declines towards stressful conditions (Fig.~\ref{fig:kurtosis-skewness}).
-That is, distributions are generally skewed towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
-Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species physiology (New Fig.~XX)\footref{note1}.
+Based on the estimates for the WAIC values and Akaike weights, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions showed steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
+Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX)\footref{note1}.
 
@@
 \item Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
-\item What is the true shape of species' distributions? These display fat-tail and skewed distributions. 
+\item What is the true shape of species' distributions? These display heavy-tail and skewed distributions. 
 \item Future directions.

### 2021-03-05 code/paper-models/1d-baseline.R

@@ -149,7 +149,7 @@ generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   
   # Load the data
   if(recompile){
-    d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="gauss.gauss", min.occurrence=min.occurrence)
+    d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="generror", min.occurrence=min.occurrence)
     # rename variables
     if(simulated){
       variables <- c("S1", "S2")
@@ -368,12 +368,13 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
 }
 
 min.occurrence <- 10
-if(min.occurrence==10){
-  d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2-data.rds", sep=""))
-}else{
-  d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2min",min.occurrence,"-data.rds", sep=""))
-}
-generror.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
+# if(min.occurrence==10){
+#   d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2-data.rds", sep=""))
+# }else{
+#   d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2min",min.occurrence,"-data.rds", sep=""))
+# }
+d <- readRDS("../../data/processed/jsdm/1d-generror-simulated-S1S2-data.rds")
+generror.1d(d=d, simulated=T, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 # skew.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 # baseline.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 
[…]

### 2021-03-05 code/paper-models/prepare-data.R

@@ -172,6 +172,64 @@ simulated.data <- function(simulated.type="linear.corr"){
 }
 
 # Generate fake data to test the extent to which the model works
+simulated.generr.data <- function(){
+        
+        set.seed(30)
+        # Define system dimensions
+        N <- 20
+        sites <- 300
+        
+        # Environmental predictors for each site
+        e1 <- rnorm(sites)
+        e2 <- rnorm(sites)
+        
+        # uncorrelated coefficients for each species and parameters
+        sigma1 <- 0.3 # sd beta1
+        mean1 <- -1 # mean beta1
+        sigma2 <- 0.4 # sd beta2
+        mean2 <- 1.5 # mean beta2
+        rho <- 0.4 # correlation between betas
+        
+        # Coefficients for generating the variance-covariance matrix
+        nu <- 3
+        s <- 0.5
+        Dis <- (as.matrix(dist(1:N))/N)
+        
+        # coefficients for each species
+        Sigma <- nu*exp(-1/(s*s)*(Dis^2)) + diag(N)*sigma1
+        z1 <- mvrnorm(mu = rep(0, times = N), Sigma = Sigma)
+        
+        # Generate correlations
+        beta1 <- z1
+        
+        vec <- c(1:round(N*0.5), 1:round(N*0.5))
+        Dis_sigma <- as.matrix(dist(vec))+1-diag(N)
+        Dis_sigma <- (Dis_sigma/max(Dis_sigma))
+        
+        Sigma <- 1*exp(-1/(0.3*0.3)*(Dis_sigma^2)) + diag(N)*0.1
+        sigma_beta1 <- exp(mvrnorm(mu = rep(-1, times = N), Sigma = Sigma))
+        # Sigma <- 1*exp(-1/(0.2*0.2)*(Dis^2)) + diag(N)*0.1
+        # alpha <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
+        alpha <- exp(rnorm(N, 0,1))
+        nu <- exp(rnorm(N,0,0.5))+1
+        
+        # Simulate data
+        dataset <- expand.grid(site=1:sites, id=1:N)
+        dataset$S1 <- e1[dataset$site]
+        dataset$S2 <- e2[dataset$site]
+        dataset$beta1 <- beta1[dataset$id] 
+        dataset$alpha <- alpha[dataset$id]
+        dataset$nu <- nu[dataset$id]
[…]

### 2021-03-05 code/paper-models/visualizing-results.R

@@ -832,12 +832,12 @@ plot.actual.data.means2 <- function(model=NULL){
       transformation <- "log"
     }else{
       # alphas
-      mu_gamma <- apply( post$alpha , 2 , mean )
-      ci_gamma <- apply( post$alpha , 2 , PI )
+      mu_beta <- apply( post$alpha , 2 , mean )
+      ci_beta <- apply( post$alpha , 2 , PI )
       
-      mu_beta <- lapply(1:2, function(x) apply(post$beta[,x,],2,mean))[[1]]
-      ci_beta <- lapply(1:2, function(x) apply(post$beta[,x,],2,PI))[[1]]
-      transformation <- "identity"
+      mu_gamma <- lapply(1:2, function(x) apply(post$gamma[,x,],2,mean))[[1]]
+      ci_gamma <- lapply(1:2, function(x) apply(post$gamma[,x,],2,PI))[[1]]
+      transformation <- "log"
       
     }
     

### 2021-03-05 manuscript/manuscript.tex
change mean

@@
 \begin{split}
-\text{log}\left(p_{ij}\right) = -\alpha_{i} - \left(\frac{\gamma^{\prime}_{i}\,|x_{j}-\beta^{\prime}_{i}|}{1+\lambda_{i} \text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{\nu_{i}} ,
+\text{log}\left(p_{ij}\right) = -\alpha_{i} - \left(\frac{\gamma^{\prime}_{i}\,|x_{j}-\beta^{\prime}_{i}|}{1+\lambda_{i}\,\text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{\nu_{i}} ,
 \end{split}
@@
 We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}.
-However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(2\lambda_{i}-1\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
+However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
 Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as for the baseline model (see Supplementary Table 2; Code Availability section).

### 2021-03-05 code/paper-models/models-binomial.R

@@ -437,3 +437,126 @@ generated quantities{
      }
 }
 "
+
+# Binomial 1d
+base.model.skew.generror.1d <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+    int sgn(real x) {
+        return x < 0 ? -1 : x > 0;
+    }
+}
+data{
+    int N;
+    int L;
+    real minp;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    row_vector[N] X1;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+}
+parameters{
+    vector[L] zalpha;
+    vector[L] zbeta;
+    vector[L] zgamma;
+    vector[L] znu;
+    vector[L] zlambda;
+    real alpha_bar;
+    real beta_bar;
+    real gamma_bar;
+    real nu_bar;
+    real lambda_bar;
+    real<lower=0> sigma_a;
+    real<lower=0> sigma_n;
+    real<lower=0> sigma_l;
+    real<lower=0> sigma_b;
+    real<lower=0> etasq_b;
+    real<lower=0> rhosq_b;
+    real<lower=0> sigma_g;
+    real<lower=0> etasq_g;
+    real<lower=0> rhosq_g;
+}
+transformed parameters{
+    vector[L] lambda;
+    vector[L] alpha;
+    vector[L] beta;
+    vector[L] gamma;
+    vector[L] nu;
+    matrix[L, L] L_SIGMA_b;
+    matrix[L, L] L_SIGMA_g;
+
+    alpha = exp(zalpha * sigma_a + alpha_bar);
+    nu = exp(znu * sigma_n + nu_bar)+1;
+    lambda = inv_logit(zlambda*sigma_l + lambda_bar)*2-1;
+
[…]

### 2021-03-05 code/paper-models/prepare-data.R

@@ -211,7 +211,7 @@ simulated.generr.data <- function(){
         # Sigma <- 1*exp(-1/(0.2*0.2)*(Dis^2)) + diag(N)*0.1
         # alpha <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
         alpha <- exp(rnorm(N, 0,1))
-        nu <- exp(rnorm(N,0,0.5))+1
+        nu <- exp(rnorm(N,-1,0.5))+1
         
         # Simulate data
         dataset <- expand.grid(site=1:sites, id=1:N)
@@ -225,7 +225,7 @@ simulated.generr.data <- function(){
         dataset$p <- exp(-alpha[dataset$id] - sigma_beta1[dataset$id]*abs(beta1[dataset$id] - dataset$S1)**nu)
         
         dataset$obs <- rbinom(n = length(dataset$S1), size = 1, prob = dataset$p)
-        dataset <- data.frame(id=dataset$id, obs=dataset$obs, alpha=dataset$alpha, beta1=dataset$beta1, sigma_beta1=dataset$sigma_beta1, S1=dataset$S1, S2=dataset$S2)                
+        dataset <- data.frame(id=dataset$id, obs=dataset$obs, alpha=dataset$alpha, beta1=dataset$beta1, sigma_beta1=dataset$sigma_beta1,nu=dataset$nu, S1=dataset$S1, S2=dataset$S2)                
         return(list(dataset=dataset, corr=Dis, corr2=Dis_sigma, corr3=Dis))
 }
 

### 2021-03-05 code/paper-models/visualizing-results.R

@@ -151,6 +151,50 @@ plot.simulated.data <- function(beta=T, gp_type = 2){
   return(figure)
 }
 
+plot.simulated.data.tails <- function(beta=T, gp_type = 2){
+  # load data
+  d <- readRDS(file = paste("../../data/processed/jsdm/1d-generror-simulated-S1S2-", "data.rds", sep = ""))
+  model_r <- readRDS(paste("../../results/models/-1d-generror-simulated.rds", sep=""))
+  
+  # Extract variables from the model
+  alphas <- precis(model_r, pars = "alpha", depth=2)
+  betas <- precis(model_r, pars = "beta", depth=3)
+  sigmas <- precis(model_r, pars = "gamma", depth=3)
+  nu <- precis(model_r, pars = "nu", depth=3)
+
+  # Extract true values
+  N <- length(unique(d$dataset$id))
+  alpha_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$alpha[1])
+  nu_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$nu[1])
+  beta1_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$beta1[1])
+  sigma1_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$sigma_beta1[1])
+  
+  # Build data.frames for the plots
+  d_alpha <- data.frame(N=1:N, id= c(rep("real", length(alpha_r)), rep("estimated", length(alphas$mean))),value=c(alpha_r,alphas$mean) , sd=c(rep(0,length(alpha_r)),alphas$sd))
+  d_beta1 <- data.frame(N=1:N, id= c(rep("real", length(beta1_r)), rep("estimated", length(betas[1:N,]$mean))),value=c(beta1_r,betas[1:N,]$mean) , sd=c(rep(0,length(beta1_r)),betas[1:N,]$sd))
+  d_sigma1 <- data.frame(N=1:N, id= c(rep("real", length(sigma1_r)), rep("estimated", length(sigmas[1:N,]$mean))),value=c(sigma1_r,sigmas[1:N,]$mean) , sd=c(rep(0,length(sigma1_r)),sigmas[1:N,]$sd))
+  d_nu <- data.frame(N=1:N, id= c(rep("real", length(nu_r)), rep("estimated", length(nu$mean))),value=c(nu_r,nu$mean) , sd=c(rep(0,length(nu_r)),nu$sd))
+  
+  
[…]

### 2021-03-09 manuscript/manuscript.tex
Preparing final run

@@
 For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Supplementary Fig.~XX).
-However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Supplementary Fig.~XX), and the sampling of the model is somewhat more challenging (ref).
+However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Supplementary Fig.~XX), and the sampling of the model can be somewhat more challenging (ref).
 
@@
 Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways.
-To do so, we changed the response curve of our Bayesian model to follow either a generalized error distribution (Eq.~\ref{eq:generalized-response}) or a non-standardized Student's t-distribution (Eq.~\ref{eq:student-response}).
-Studying the WAIC values, both non-linear regressions outperformed the baseline model (Supplementary Fig.~XX)\footnote[2]{\label{note1} I am still waiting on the results for this (currently running in the cluster), and this is just my prior expectation based on what I've seen in some of the other models I've been working with.}.
-In the case of a generalized error distribution, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with a Gaussian shape (Fig.~\ref{fig:kurtosis-skewness}).
-Similarly, we found the shape parameter in the non-standardized Student's t-distribution to present large values, also approximating the normal case\footref{note1}.
-However, in both cases, these parameter estimates displayed a lot of variation between species, which indicates that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
+To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}).
+A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX)\footnote[2]{\label{note1} I am still waiting on the results for this (currently running in the cluster), and this is just my prior expectation based on what I've seen in some of the other models I've been working with.}.
+Studying the resulting posterior distributions, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with a Gaussian shape (Fig.~\ref{fig:kurtosis-skewness}).
+However, such parameter estimates displayed a lot of variation across species, which might indicate that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
 
@@
 
-Finally, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}).
+Next, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}).
 Based on the estimates for the WAIC values and Akaike weights, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.

### 2021-03-09 code/paper-models/1d-baseline.R

@@ -109,7 +109,7 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   model_code=base.model.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 3
+  n_chains_5.1 <- 5
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -227,7 +227,7 @@ generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   model_code=base.model.generror.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 3
+  n_chains_5.1 <- 5
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
   
@@ -347,7 +347,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
   model_code=skew.model.traits.1d
 
   # Initialize data structure
-  n_chains_5.1 <- 3
+  n_chains_5.1 <- 5
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -467,7 +467,7 @@ skew.generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrenc
   model_code=base.model.skew.generror.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 3
+  n_chains_5.1 <- 5
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
   
@@ -485,17 +485,25 @@ skew.generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrenc
 }
 
 
-min.occurrence <- 10
+min.occurrence <- 20
 # if(min.occurrence==10){
 #   d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2-data.rds", sep=""))
 # }else{
 #   d <- readRDS(file = paste("../../data/processed/jsdm/1d-PC1PC2min",min.occurrence,"-data.rds", sep=""))
 # }
-d <- readRDS("../../data/processed/jsdm/1d-skew-generror-simulated-S1S2-data.rds")
-skew.generror.1d(d=d, simulated=T, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
[…]

### 2021-03-09 code/paper-models/prepare-data.R

@@ -459,7 +459,7 @@ visua <- function(dataset){
         n <- length(id)
         qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
         col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
-        for(i in 1:3){
+        for(i in 1:n){
                 j <- id[i]
                 dat <- dataset[dataset$id==j,]
                 index <- sort(dat$S1, index.return=T)

### 2021-03-09 code/paper-models/run_inte.sh

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=4548]" -W 20:00 -n 15 "R --vanilla --slave < 1d-baseline.R"
+bsub -R "rusage[mem=4548]" -W 120:00 -n 15 "R --vanilla --slave < 1d-baseline.R"

### 2021-03-09 code/paper-models/plots/plot-real.R
Sorting plots a little bit

@@ -1,0 +1,816 @@
+# Import libraries
+library(rethinking)
+library(rstan)
+library(gridExtra)
+library(grid)
+library(ggpubr)
+library(hrbrthemes)
+library(plotly)
+library(cowplot)
+
+####
+# Visualizing the results of some of the models
+####
+
+plot.common.to.all <- function(p, fontsize=10, mar=margin(5.5,5.5,5.5,5.5)){
+  p <- p +
+    coord_cartesian(clip = 'off') +
+    theme_bw() + 
+    theme(legend.position = "none",
+          text = element_text(size=fontsize),
+          panel.grid.major = element_blank(),
+          panel.grid.minor = element_blank(),
+          panel.border = element_rect(colour = "black", fill=NA, size=0.4),
+          plot.margin = mar
+    )
+  
+  return(p)
+}
+
+plot.scatter <- function(mu, variance, color="black", xlabel="-", ylabel="-", tit=NULL, mar=margin(5.5,5.5,5.5,5.5), xlims=NA, ylims=NA){
+  
+  df <- data.frame(sp=1:length(mu), mean=mu, variance=variance)
+  
+  p <- ggplot(df, aes(x=variance, y=mean)) + 
+    geom_point(color=color, alpha=0.7)+
+    ylab(ylabel)+
+    xlab(xlabel)+
+    scale_x_continuous(expand = expansion(add = c(0, 0)), limits = xlims)+
+    scale_y_continuous(expand = expansion(add = c(0, 0)), limits = ylims)
+  
+  p <- plot.common.to.all(p, mar=mar)
+  if(!(is.null(tit))){
+    p <- p + ggtitle(tit)
+  }
+  
+  return(p)
+}
+
+plot.scatter2 <- function(mu, variance, label, color="black", alpha=NULL, xlabel="-", ylabel="-", tit=NULL, mar=margin(5.5,5.5,5.5,5.5)){
+  
+  if(is.null(alpha)){
+    alpha <- rep(0.7, length(mu))
+  }
+  df <- data.frame(sp=1:length(mu), mean=mu, variance=variance, label=label, alpha=alpha)
+  
+  p <- ggplot(df, aes(x=variance, y=mean, color=label, alpha=alpha)) + 
+    geom_point()+
+    scale_color_manual(values=color)+
+    guides(alpha = FALSE)+
+    ylab(ylabel)+
+    xlab(xlabel)+
[…]

### 2021-03-09 code/paper-models/plots/plot-simulated.R

@@ -1,0 +1,281 @@
+# Import libraries
+library(rethinking)
+library(rstan)
+library(gridExtra)
+library(grid)
+library(ggpubr)
+library(hrbrthemes)
+library(plotly)
+library(cowplot)
+
+####
+# Visualizing the results of some of the models
+####
+
+transform.skew <- function(post){
+  delta <- post$lambda/sqrt(1+post$lambda**2)
+  beta_hat <- post$beta + sqrt(1/(2*post$gamma)) * delta * sqrt(2/pi)
+  sigma_hat <- post$gamma * (1 - (2*(delta**2))/pi)^(-1)
+  
+  mu_z <- sqrt(2/pi)*delta
+  maxy = 0.5 * ( 4 - pi ) * (delta * sqrt(2/pi))**3 / (1 - 2 * delta**2 / pi )**(3 / 2.0);
+  maxy = post$beta + 1 / sqrt( 2 * sigma_hat) * (mu_z - maxy * sqrt(1 - mu_z**2 ) * 0.5 - 0.5 * sign(post$lambda) * exp(- 2 * pi / abs(post$lambda) ))
+  maxy = exp(- post$gamma * (maxy - post$beta)**2) * (1 + pracma::erf((post$lambda * (maxy - post$beta)) * sqrt(post$gamma) ))
+  
+  alpha_hat <- -log((maxy+0.0001)) + post$alpha
+  return(list(beta=beta_hat, sigma=sigma_hat, alpha=alpha_hat, lambda=post$lambda))
+}
+
+plot.simulated.compare <- function(){
+  m1 <- readRDS(paste("../../../results/models/skew-model-traits-1dskew-simulated2.rds", sep=""))
+  m2 <- readRDS(paste("../../../results/models/baseline-model-1d1d-simulated2.rds", sep=""))
+  rethinking::compare(m1, m2)
+  
+  d <- readRDS(file = paste("../../../data/processed/jsdm/skew-simulated2S1S2", "data.rds", sep = ""))
+  
+  for (i in 1:2){
+    if (i==1){
+      model_r <- m1
+      post <- extract.samples(model_r, pars=c("alpha", "beta", "gamma", "lambda"))
+      
+      post <- transform.skew(post)
+      
+      # Extract variables from the model
+      alphas <- precis(model_r, pars = "alpha", depth=2, )
+      betas <- precis(model_r, pars = "beta", depth=3)
[…]

### 2021-03-09 code/paper-models/plots/visualizing-results.R

@@ -27,11 +27,11 @@ transform.skew <- function(post){
 }
 
 plot.simulated.compare <- function(){
-  m1 <- readRDS(paste("../../results/models/skew-model-traits-1dskew-simulated2.rds", sep=""))
-  m2 <- readRDS(paste("../../results/models/baseline-model-1d1d-simulated2.rds", sep=""))
+  m1 <- readRDS(paste("../../../results/models/skew-model-traits-1dskew-simulated2.rds", sep=""))
+  m2 <- readRDS(paste("../../../results/models/baseline-model-1d1d-simulated2.rds", sep=""))
   rethinking::compare(m1, m2)
   
-  d <- readRDS(file = paste("../../data/processed/jsdm/skew-simulated2S1S2", "data.rds", sep = ""))
+  d <- readRDS(file = paste("../../../data/processed/jsdm/skew-simulated2S1S2", "data.rds", sep = ""))
 
   for (i in 1:2){
     if (i==1){
@@ -105,8 +105,8 @@ plot.simulated.compare <- function(){
 
 plot.simulated.data <- function(beta=T, gp_type = 2){
   # load data
-  d <- readRDS(file = paste("../../data/processed/jsdm/skew-simulated-workingexample-S1S2", "data.rds", sep = ""))
-  model_r <- readRDS(paste("../../results/models/skew-workingexample-simulated.rds", sep=""))
+  d <- readRDS(file = paste("../../../data/processed/jsdm/skew-simulated-workingexample-S1S2", "data.rds", sep = ""))
+  model_r <- readRDS(paste("../../../results/models/skew-workingexample-simulated.rds", sep=""))
 
   # Extract variables from the model
   alphas <- precis(model_r, pars = "alpha", depth=2)
@@ -152,8 +152,8 @@ plot.simulated.data <- function(beta=T, gp_type = 2){
 
 plot.simulated.data.tails <- function(beta=T, gp_type = 2){
   # load data
-  d <- readRDS(file = paste("../../data/processed/jsdm/1d-generror-simulated-S1S2-", "data.rds", sep = ""))
-  model_r <- readRDS(paste("../../results/models/-1d-generror-simulated.rds", sep=""))
+  d <- readRDS(file = paste("../../../data/processed/jsdm/1d-generror-simulated-S1S2-", "data.rds", sep = ""))
+  model_r <- readRDS(paste("../../../results/models/-1d-generror-simulated.rds", sep=""))
   
[…]

### 2021-03-18 manuscript/manuscript.tex
Lowering iterations again

@@
 These models account for all prior knowledge we have regarding their shape, including expert knowledge on species' environmental preferences and physiology.
-With this approach, we are able to shed light on the true shape of empirical species' distributions.
+With this approach, we are able to uncover the true shape of empirical species' distributions.
 Moreover, they allow us to tackle long-standing hypothesis regarding general biogeographical patterns.
@@
 In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes)---and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
+Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
 We describe this information in more detail in the Supplementary Table 1.
@@
 While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions.
-Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution.
-That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we define can species' responses to the environment as
+Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution (Fig~\ref{fig:response}a).
+That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
 \begin{equation}
@@
 That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=1\textwidth]{figures/all-distributions}
+	   \caption{Different response curves.
+Panel (a) shows the probability density function characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\gamma$, when $\mu=0$.
+Panel (b) shows the probability density function characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\nu$, when $\gamma=1$ and $\mu=0$.
+Panel (c) shows the probability density function characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\lambda$, when $\gamma=1$ and $\mu=0$.
+Panel (d) shows the probability density function characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\lambda$ and $\nu$, when $\gamma=1$ and $\mu=0$.
+Notice that for each case, we chose $\alpha$ values that normalize the probability distributions.}
+      \label{fig:response}
+\end{figure}
 
@@
 We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}.
-Sampling models like the ones described above can be computationally very expensive.
+Sampling models like the ones described above can be computationally very intensive.
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data, which corresponds to the majority of the results of this work.
+Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data.
 When using ordinal data, we limit our study to a random subset of a 100 species of those with at least 20 occurrences to facilitate the sampling of the models.
-% A random subset of a 100 species of those with at least 10 occurrences.
 
@@
 We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system.
-Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and the prior information.
-More specifically, to propose new species' response curves, we follow three key criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
+Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information.
+To propose new species' response curves, we follow three criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
 
@@
 Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions.
-To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Supplementary Figure XX).
+To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Fig~\ref{fig:response}b).
 In particular, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
@@
 Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
-Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the corresponding prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}).
+Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}).
 Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.
@@
 Alternatively, we could have used other distributions that present heavy-tails and fulfil the selection criteria described above.
-For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Supplementary Fig.~XX).
-However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Supplementary Fig.~XX), and the sampling of the model can be somewhat more challenging (ref).
+For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Fig~\ref{fig:response}b).
+However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Fig~\ref{fig:response}b), and the sampling of the model can be somewhat more challenging (ref).
 
@@
 When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction.
-Likewise, this same skewed shape can also be the result of asymmetric environmental tolerance.
+Likewise, this same skewed shape can also emerge as a result of species' asymmetric environmental tolerance.
 One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX).
-As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness'.
+As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness' (Fig~\ref{fig:response}c).
 Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species are distributed along the gradient.
-To test for the existence of this feature, we modified the species' response curve in Eq.~(\ref{eq:baseline-response}) as
+To test for the existence of this feature, we modified Eq.~(\ref{eq:baseline-response}) as
 \begin{equation}
@@
 That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$.
-As described before, this variance-covariance matrix can be estimated by the model following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
+As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
 Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.
@@
 \subsubsection*{Heavy-tail and skewed response curve}
-Finally, one could also consider a response curve with both kurtosis and skewness.
-A convenient way to achieve this is by considering a response curve that follows a skewed generalized error distribution (also see XX).
-This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness.
+Finally, one could consider a response curve with both kurtosis and skewness.
+A convenient way to achieve this is by using a response curve that follows a skewed generalized error distribution (also see XX).
+This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness (Fig~\ref{fig:response}d).
 The skewed generalized error distribution can be considered by modifying the species' response curve in Eq.~(\ref{eq:baseline-response}) as
@@
 where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively.
-$\text{sgn}\left(x\right)$ is the sign function.
+The function $\text{sgn}\left(x\right)$ characterizes the sign function.
 We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}.
 However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
-Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as for the baseline model (see Supplementary Table 2; Code Availability section).
+Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as in the baseline model (see Supplementary Table 2; Code Availability section).
 Notice that the generalized error distribution (Eq.~\ref{eq:generalized-response}) and the skew normal distribution (Eq.~\ref{eq:skewed-response}) are special cases of Eq.~(\ref{eq:heavy-skewed-response}) when $\lambda_i=0$ and $\nu_i=2$, respectively.
@@
   \centering
-    \vspace{0.5cm}
     \includegraphics[width=0.9\textwidth]{figures/figure1-prime}
-    	  \vspace{0.3cm}
+    	  \vspace{0.1cm}
 	   \caption{Relationship between mean and variance of species' distributions.
@@
   \centering
-    \vspace{0.5cm}
     \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
-    	  \vspace{0.3cm}
+    	  \vspace{0.1cm}
 	   \caption{Universality of the relationship between mean and variance of species' distributions.
@@
   \centering
-    \vspace{0.5cm}
     \includegraphics[width=0.8\textwidth]{figures/kurto-skew}
-    	  \vspace{0.3cm}
+    	  \vspace{0.1cm}
 	   \caption{Average kurtosis and skewness of species' distributions.
@@
 
-Next, we studied the skewness of species distributions using the skewed response curve in Eq.~(\ref{eq:skewed-response}).
-Based on the estimates for the WAIC values and Akaike weights, this model clearly outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
-Studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions showed steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
+Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species distributions.
+Based on the estimates for the WAIC values, this model outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Perhaps most importantly, studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions present steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
 Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX)\footref{note1}.
 
-{\color{gray}
-[Note: Reports on the hyperparameters describing the different covariance matrices missing at the moment.
-I think it's interesting, and I will certainly end up adding things here about this.
-I will also add a third figure showing how lambda varies with the corresponding variance-covariance matrix of species' environmental and physiological traits.]}
+Finally, we used a model that allowed for heavy-tails and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}) to understand the relationship between the kurtosis and skewness of distributions.
+This model outperformed the rest, presenting Akaike weights close to 1 when comparing models (Supplementary Fig.~XX).
+Comparing the posterior distribution for parameters $\nu_i$ and $\lambda_i$, we found this to be uncorrelated\footref{note1}.
+However, different types of species seem to present characteristically different distributions based on their kurtosis and skewness.
+Importantly, this model identified similar relationships between parameters as the baseline model (Supplementary Fig.~XX), and measured similar average levels of kurtosis and skewness than the the other two models (Fig.~\ref{fig:kurtosis-skewness})\footref{note1}.
 

### 2021-03-18 code/paper-models/1d-baseline.R

@@ -109,7 +109,7 @@ baseline.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   model_code=base.model.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 5
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -227,7 +227,7 @@ generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10,
   model_code=base.model.generror.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 5
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
   
@@ -326,7 +326,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
   
   # Set starting values for the parameters
   start_5.1 <- list(
-    zlambda = rep(0, dat_5.1$L),
+    lambda = rep(0, dat_5.1$L),
     zalpha = rep(0, dat_5.1$L),
     zbeta = rep(0, dat_5.1$L),
     zgamma = rep(0, dat_5.1$L),
@@ -347,7 +347,7 @@ skew.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofo
   model_code=skew.model.traits.1d
 
   # Initialize data structure
-  n_chains_5.1 <- 5
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
@@ -467,7 +467,7 @@ skew.generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrenc
   model_code=base.model.skew.generror.1d
   
   # Initialize data structure
-  n_chains_5.1 <- 5
+  n_chains_5.1 <- 3
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
   

### 2021-03-18 code/paper-models/models-binomial.R

@@ -360,7 +360,7 @@ parameters{
     vector[L] zalpha;
     vector[L] zbeta;
     vector[L] zgamma;
-    vector<lower=-10, upper=10>[L] zlambda;
+    vector<lower=-10, upper=10>[L] lambda;
     real alpha_bar;
     real beta_bar;
     real gamma_bar;
@@ -378,12 +378,12 @@ transformed parameters{
     vector[L] alpha;
     vector[L] beta;
     vector[L] gamma;
-    vector[L] lambda;
+    //vector[L] lambda;
     real delta;
     matrix[L, L] L_SIGMA_b;
     matrix[L, L] L_SIGMA_g;
     
-    lambda = zlambda * sigma_l + lambda_bar;
+    //lambda = zlambda * sigma_l + lambda_bar;
     
     alpha = exp(zalpha * sigma_a + alpha_bar);
 
@@ -418,7 +418,7 @@ model{
     zalpha ~ std_normal();
     zgamma ~ std_normal();
     zbeta ~ std_normal();
-    zlambda ~ std_normal();
+    lambda ~ normal( lambda_bar , sigma_l );
 
     for ( i in 1:L ){
         Y[i] ~ binomial(1, exp(- alpha[i] - gamma[i] * columns_dot_self(X1 - beta[i])) .* (1.0 + erf((lambda[i] * (X1 - beta[i])) * sqrt(gamma[i]) )) + minp);

### 2021-03-18 code/paper-models/plots/other-plots.R

@@ -114,4 +114,84 @@ prior_p <- function(N=2, sd_=1){
   print(p)
 }
 
+distribution.type <- function(){
+  x <- seq(-5, 5, length.out = 2000)
+  generrskew <- function(x, a, mu, sigma, lambda, p){
+    v <- sqrt((pi*gamma(1/p))/(pi*(1+3*lambda**2)*gamma(3/p)-(16**(1/p))*lambda*lambda*(gamma(1/2+1/p)**2)*gamma(1/p)))
+    m <- (2**(2/p))*v*sigma*lambda*gamma(1/2+1/p)/sqrt(pi)
+    return((p*sigma/(2*v*gamma(1/p)))*exp(-(sigma*abs(x-mu+m)/(v*(1+lambda*sign(x-mu+m))))**p))
+  }
+  
+  plist <- list()
+  
+  for(i in 1:4){
+    if(i==1){
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,0.6,1.5)
+      lambda <- rep(0,3)
+      nu <- rep(2,3)
+      title <- "(a)  normal"
+      labs <- c(expression(gamma==1), expression(gamma<1), expression(gamma>1))
+    }else if (i==2){
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,1,1)
+      lambda <- rep(0,3)
+      nu <- c(2,1.2,9)
+      labs <- c(expression(nu==2), expression(nu<2), expression(nu>2))
+      title <- "(b)  heavy-tails"
+    }else if (i==3){
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,1,1)
+      lambda <- c(0,-0.5, 0.5)
+      nu <- rep(2,3)
+      labs <- c(expression(lambda==0), expression(lambda<0), expression(lambda>0))
+      title <- "(c)  skewed"
+    }else{
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,1, 1)
+      lambda <- c(0, -0.75, 0.75)
+      nu <- c(2,1.2,9)
+      title <- "(d)  heavy-tail and skewed"
+      labs <- c(expression(list(lambda==0, nu==2)), expression(list(lambda<0, nu<2)), expression(list(lambda>0, nu>2)))
+      
+    }
[…]

### 2021-03-18 code/paper-models/plots/plot-real.R

@@ -270,7 +270,7 @@ plot.actual.data <- function(model=NULL){
   
   # Load the data if not added  
   if(is.null(model)){
-    model <- readRDS("../../../results/models/baseline-model.rds")
+    model <- readRDS("../../../results/models-old/baseline-model.rds")
   }
   
   #indicator values
@@ -278,7 +278,8 @@ plot.actual.data <- function(model=NULL){
   
   # extract samples
   post <- extract.samples(model, n = 1000, pars=c("alpha", "gamma", "beta")) 
-  post$beta <- post$beta*(-1)
+  # post$beta <- post$beta*(-1)
+  # post$gamma <- 1/(post$gamma*2)
   
   # alphas
   mu_alpha <- apply( post$alpha , 2 , mean )
@@ -307,7 +308,7 @@ plot.actual.data <- function(model=NULL){
     Tind_ <- as.numeric(as.character(Tind[idx,3]))
     Tind_ <- Tind_[!(is.na(Tind_))]
     Tind_ <- mean(Tind_[1:round(length(Tind_)*0.5)])>mean(Tind_[round(length(Tind_)*0.5):length(Tind_)])
-    Tind_ <- c("low elevation", "high elevation")[c(Tind_*1+1, (!(Tind_))*1+1)]
+    Tind_ <- c("high elevation", "low elevation")[c(Tind_*1+1, (!(Tind_))*1+1)]
     
     ylim = c(min(ci_beta[[i]]), max(ci_beta[[i]]))
     xlim = c(-1, length(mu_beta[[i]])+1)
@@ -469,7 +470,7 @@ plot.actual.data.means <- function(model=NULL){
   
   # Load the data if not added  
   if(is.null(model)){
-    model <- readRDS("../../../results/models/baseline-model.rds")
+    model <- readRDS("../../../results/models-old/baseline-model.rds")
   }
   
   #indicator values

### 2021-03-18 code/paper-models/run_inte.sh

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=4548]" -W 120:00 -n 15 "R --vanilla --slave < 1d-baseline.R"
+bsub -R "rusage[mem=4300]" -W 200:00 -n 30 "R --vanilla --slave < 1d-baseline.R"

### 2021-03-29 code/paper-models/1d-baseline.R
Multithread attempt and manuscript

@@ -479,6 +479,24 @@ skew.generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrenc
                      warmup=1000, iter=2000,
                      init=init_5.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
   
+  # dat_5.1 <- list(N=N,
+  #                 L=L,
+  #                 minp=1e-100,
+  #                 Y=t(obs),
+  #                 indices=1:L,
+  #                 X1=X1,
+  #                 Dmat_b=Dis_b,
+  #                 Dmat_g=Dis_g
+  # )
+  # model_code = base.model.skew.generror.1d.multithread
+  # generror1d <- cmdstan_model(write_stan_file(model_code), cpp_options = list(stan_threads = TRUE))
+  # mfit_5.1 <- generror1d$sample(data = dat_5.1,
+  #                               init = init_5.1,
+  #                               chains = 4,
+  #                               threads_per_chain = 10,
+  #                               parallel_chains = 4,
+  #                               refresh = 1000)
+  # mfit_5.1$save_object(file = paste(ofolder, extension2, "", extension,".rds", sep=""))
   
   saveRDS(mfit_5.1, file = paste(ofolder, extension2, "", extension,".rds", sep=""))
   return(mfit_5.1)

### 2021-03-29 code/paper-models/models-binomial.R

@@ -559,3 +559,141 @@ generated quantities{
 }
 "
 
+# Binomial 1d
+base.model.skew.generror.1d.multithread <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+    int sgn(real x) {
+        return x < 0 ? -1 : x > 0;
+    }
+
+    real partial_sum(int[] y_slice,
+                   int start, int end,
+                   int N, vector lambda, vector alpha, vector beta, vector gamma, vector nu, int[ , ] Y, row_vector X1, real minp ) {
+        real lp = 0.0;
+        real m;
+        real v;
+        for( i in start:end){
+            v = gamma[i] * sqrt((pi()*(1+3*pow(lambda[i],2))*tgamma(3.0/nu[i])-pow(16,(1.0/nu[i]))*pow(lambda[i],2)*tgamma(0.5+1.0/nu[i])*tgamma(0.5+1.0/nu[i])*tgamma(1.0/nu[i]))/(pi()*tgamma(1.0/nu[i])));
+            m = beta[i] - pow(2, 2.0/nu[i])*lambda[i]*tgamma(0.5+1.0/nu[i])/(sqrt(pi())*gamma[i]);
+            for( j in 1:N ){
+                lp += binomial_lpmf( Y[i, j] | 1 , exp(-alpha[i] - pow(v * fabs(X1[j] - m)/(1+lambda[i]*sgn(X1[j] - m)), nu[i])) + minp);
+            }
+        }
+        return lp;
+    }
+}
+data{
+    int N;
+    int L;
+    real minp;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    int indices[L];
+    row_vector[N] X1;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+}
+parameters{
+    vector[L] zalpha;
+    vector[L] zbeta;
[…]

### 2021-03-29 code/paper-models/plots/plot-real.R

@@ -267,10 +267,11 @@ plot.actual.data <- function(model=NULL){
   # Parameters for plots
   extra <- 1
   colo <- c("#1b9e77", "#d95f02", "#7570b3")
+  n <- 1
   
   # Load the data if not added  
   if(is.null(model)){
-    model <- readRDS("../../../results/models-old/baseline-model.rds")
+    model <- readRDS("../../../results/models/baseline-min20-1d.rds")
   }
   
   #indicator values
@@ -279,24 +280,30 @@ plot.actual.data <- function(model=NULL){
   # extract samples
   post <- extract.samples(model, n = 1000, pars=c("alpha", "gamma", "beta")) 
   # post$beta <- post$beta*(-1)
-  # post$gamma <- 1/(post$gamma*2)
+  # post$gamma <- 1/(2*post$gamma)
   
   # alphas
   mu_alpha <- apply( post$alpha , 2 , mean )
   ci_alpha <- apply( post$alpha , 2 , PI )
   
-  # betas
-  mu_beta <- lapply(1:2, function(x) apply(post$beta[,x,],2,mean))
-  ci_beta <- lapply(1:2, function(x) apply(post$beta[,x,],2,PI))
-  
-  # gammas
-  mu_gamma <- lapply(1:2, function(x) apply(post$gamma[,x,],2,mean))
-  ci_gamma <- lapply(1:2, function(x) apply(post$gamma[,x,],2,PI))
-  
   # Beta plots
-  for(i in 1:1){
+  for(i in 1:n){
+    
+    if(n==1){
+      mu_beta <- apply( post$beta , 2 , mean )
+      ci_beta <- apply( post$beta , 2 , PI )
+      mu_gamma <- apply( post$gamma , 2 , mean )
+      ci_gamma <- apply( post$gamma , 2 , PI )
+      corPI <- sapply(1:dim(post$beta)[1], function(x) cor(post$beta[x,], post$gamma[x,]))
+    }else{
+      mu_beta <- apply( post$beta[,i,] , 2 , mean )
+      ci_beta <- apply( post$beta[,i,] , 2 , PI )
+      mu_gamma <- apply( post$gamma[,i,] , 2 , mean )
+      ci_gamma <- apply( post$gamma[,i,] , 2 , PI )
[…]

### 2021-04-17 code/paper-models/plots/plot-real.R
Some changes t plots and some changes to the processing of the data

@@ -135,13 +135,46 @@ plot.ranking.y <- function(mu, ci, color="black", xlabel="-", ylabel="-", mu_ord
   return(p)
 }
 
+kurtosis.skewness <- function(lambda){
+  v <- ((pi-8*lambda*lambda+3*pi*lambda*lambda)/(2*pi))**(-0.5)
+  kurtosis <- (v**4)*(-192*(lambda**4)+16*pi*(lambda**2)*(-5+lambda**2)+3*pi*pi*(1+10*lambda*lambda+5*(lambda**4)))/(4*(pi**2))-3
+  return(kurtosis)
+}
+kurtosis.generror <- function(nu){
+  kurtosis <- gamma(5/nu)*gamma(1/nu)/((gamma(3/nu))**2)-3
+  return(kurtosis)
+}
+kurtosis.skew.generror <- function(nu, lambda){
+  v <- ((pi*(1+3*lambda*lambda)*gamma(3/nu)-((16)**(1/nu))*lambda*lambda*(gamma(0.5+1/nu)**2)*gamma(1/nu))/(pi*gamma(1/nu)))**(-1/2)
+  kurtosis <- ((v**4)/(pi*pi*gamma(1/nu)))*(-3*(256**(1/nu))*(lambda**4)*(gamma(0.5+1/nu)**4)*gamma(1/nu) +
+                                                  3*(2**((4+nu)/nu))*pi*(lambda**2)*(1+3*(lambda**2))*(gamma(0.5+1/nu)**2)*gamma(3/nu)-
+                                                  (2**(4+2/nu))*(pi**(3/2))*(lambda**2)*(1+lambda**2)*gamma(0.5+1/nu)*gamma(4/nu)+
+                                                  (pi**2)*(1+10*(lambda**2)+5*(lambda**4))*gamma(5/nu))-3
+  return(kurtosis)
+}
+
+skewness.skew <- function(lambda){
+  v <- ((pi-8*lambda*lambda+3*pi*lambda*lambda)/(2*pi))**(-0.5)
+  skewness <- (v**3)*lambda*(16*lambda*lambda-pi*(-1+5*lambda*lambda))/(pi**(3/2))
+  return(skewness)
+}
+skewness.skew.generror <- function(nu, lambda){
[…]

### 2021-04-17 code/prepare-data/process-data.R

@@ -37,7 +37,7 @@ prepare.data <- function(){
   ## Extract all places and species names to loop over, and rename stuff
   sp.weighted <- colnames(dat.weighted)
   sp.codes <- cbind(sp.weighted, 1:length(sp.weighted),colSums(1*(dat.weighted>0)))
-  places.codes <- cbind(rownames(dat.weighted), 1:length(rownames(dat.weighted)), rowSums(1*(dat.weighted>0)), dat_SP$Easting, dat_SP$Northing)
+  places.codes <- cbind(rownames(dat.weighted), 1:length(rownames(dat.weighted)), rowSums(1*(dat.weighted>0)), dat_SP$Easting, dat_SP$Northing, elevation)
   
   ## Rename rows and columns
   colnames(dat.weighted) <- 1:length(sp.weighted)
@@ -45,7 +45,7 @@ prepare.data <- function(){
 
   ## Save species and places codes
   write.table(sp.codes, file = "../../data/properties/codes/sp_codes.csv", quote = F, row.names = F, col.names = c("sp", "id", "range"), sep=",")
-  write.table(places.codes, file = "../../data/properties/codes/places_codes.csv", quote = F, row.names = F, col.names = c("place", "id", "richness", "easting", "northing"), sep=",")
+  write.table(places.codes, file = "../../data/properties/codes/places_codes.csv", quote = F, row.names = F, col.names = c("place", "id", "richness", "easting", "northing", "elevation"), sep=",")
   write.table(file="../../data/processed/distribution/distribution.csv", x=1*(dat.weighted>0), sep = ",",row.names = F, col.names = F, quote = F )
 
   return(list(dat = dat.weighted, coord = dat_SP, places = places.codes, sp = sp.codes))
@@ -65,7 +65,10 @@ prepare.data.sdm <- function(absences=T){
   }
 }
 
-prepare.data.jsdm <- function(){
+prepare.sliding <- function(){
+  dat <- prepare.data()
+  
+  pi <- rethinking::PI(elevation, prob = c((1:9)*0.1))
   
 }
 

### 2021-05-01 manuscript/manuscript.tex
Multithread attempt and manuscript
Some changes t plots and some changes to the processing of the data
Sliding data and code

@@
 \title{\vspace{-1cm}
-Model linearity breeds contempt: using Bayesian non-linear models to uncover general biogeographical patterns}
+Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns}
 % Model familiarity breeds contempt: simple models to compare properties of species' distributions
@@
 
-%Consider testing tails of distribution?
-I would use the Cauchy or generalized normal distribution for heavy tail distribution without skewness; and I would use an exponentially modified Gaussian distribution with heavy tails for a skew distribution with heavy tails.
-% Option (1) for a long-tail distribution: Generalized normal distribution. We model it such that shape parameter beta can go from 1 to 8
-% Option (2) for a long-tail distribution: Student's t-distribution. In this case, we would need to define the shape parameter so that it is larger than 2 (otherwise, we would have undefined variance)
-% Option (1) for a long-tail skewed distribution.
-The exponentially modified Gaussian distribution has potential, but I am not sure I can fit a parameter that goes from like lambda if I want the skewness in either side.
-% Option (2) for a long-tail skewed distribution.
-Version 2 of the Generalized normal distribution.
-The problem here is that this function has a few discontinuities, which means that one needs to use a step function.
-I believe this to be quite problematic in Stan...
-% Option (3) for a long-tail skewed distribution. Skewed student-t distribution. This seems to be complex but quite what I need. It will be hard to fit but seems quite promising. See:
+% REFERENCES TO LOOK AT FROM JAKE
+% DATE EMAIL: 16.03.2021
+
 %https://www.tandfonline.com/doi/abs/10.1080/01621459.1998.10474117
@@
 %https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2009WR008933
+
+%############# NEW INTRO ##############
+
+% Describe different aspects of shape
+
+One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
+Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}.
+The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
+While these models are fundamental to the scientific community for predicting changes in species geographic distributions and the effects of environmental disturbances, distribution models ...
+can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019},
+
+While the value of these models is unquestionable, these approaches often rely on broad simplifications regarding the shape of species' realized niches. Describing
+
+ These models estimate species' realized niches using several covariates, including environmental variables \citep{guisanPredictingSpeciesDistribution2005}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
+More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}.
+The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
+However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been recently questioned \citep{harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019, poggiatoInterpretationsJointModeling2021}.
+This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
+
+From a phenomenological perspective, one can focus on several aspects of species' distributions. For example,
+
+Two aspects that are far less explored: the shape and the relationship between the properties of species distributions.
+
+Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric.
+However, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
+From a purely statistical perspective, there is a strong argument to be made in support of bell shaped distributions.
+That is, if all that we are willing to assume about a species' distributions is that these occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model their distribution as Gaussian \citep{frankCommonPatternsNature2009}.
+That said, many natural processes can lead to non-normal distributions, and heavy-tails and skewed shapes are very common across fields.
+For example, heavy-tailed distributions can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
+Indeed, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
+Likewise, several processes can lead to skewed distributions.
+For example, when species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction; and this same skewed shape can also emerge as a result of species' asymmetric environmental tolerance.
+
+ For example, species' distributions are shaped by multiple stochastic events, such as di distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+
+% Relate them to macroecological hypothesis
+
+That said, a lot can be learned about species' geographic distributions from taking a phenomenological approach, focussing instead on the description of basic properties of their realized niches.
+For example, the study of species' range sizes along environmental gradients can reveal general biodiversity patterns that are crucial from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992}.
+Differences in species' responses to the environment could shed light on how climatic processes and historical contingencies have shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}.
+Other properties, such as the skewness of species' distributions, can also reveal general underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
+More generally, understanding the shape of species' realized niches and the extend to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
+Indeed, there is no general agreement on the shape of species' distributions (ref).
+
+% Describe limitations in previous methods to characterize shape of Realized Niche: prior knowledge; simultaneous estimation of many (also rare); uncertainty
+
+Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
+In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see \citealt{review} and \citealt{good example}).
+This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings.
+First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
+From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
+Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
+Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
+Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
+That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
+Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
+For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.
+
+% Intro new method
+Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
+In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology.
+We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions.
+Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns.
+
+% Questions
+
+%############# OLD INTRO ##############
 
@@
 The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and dispersal ability in setting species' range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
-That said, a lot can be learned from taking a phenomenological approach, focussing instead on the description of basic properties of species' realized niches.
+That said, a lot can be learned about species' geographic distributions from taking a phenomenological approach, focussing instead on the description of basic properties of their realized niches.
 For example, the study of species' range sizes along environmental gradients can reveal general biodiversity patterns that are crucial from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992}.
@@
 \begin{split}
-p_{ij} & =\hat{p}_{ij} \left[1+\text{erf}\left(\lambda_{i}\left(x_{j}-\beta^{\prime}_{i}\right)\sqrt{\frac{\gamma^{\prime}_{i}}{2}}\right)\right] \\
-\text{log}\left(\hat{p}_{ij}\right) & = -\alpha^{\prime}_{i} - \gamma^{\prime}_{i} \left(x_{j}-\beta^{\prime}_{i}\right)^2 ,
+\text{log}\left(p_{ij}\right) = -\alpha_{i} - \gamma^{\prime}_{i}\left(\frac{x_{j}-\beta^{\prime}_{i}}{1+\lambda_{i}\,\text{sgn}\left(x_{j}-\beta^{\prime}_{i}\right)}\right)^{2} ,
 \end{split}
@@
 \end{equation}
-where $\alpha^{\prime}_{i} = q_{1}\left(\alpha_{i}, \gamma^{\prime}_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma^{\prime}_{i}, \lambda_{i}\right)$, $\gamma^{\prime}_{i} = q_{3}\left(\gamma_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution, which we define as $\lambda_{i}\in\left(-10, 10\right)$.
-Notice that these boundary conditions for $\lambda_{i}$ are arbitrarily set for computational purposes, as this parameter is theoretically not bounded.
-The function `erf' is the error function (see \citealt{ashourApproximateSkewNormal2010}).
-We choose an adaptive prior such that $\lambda_{i}\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
-This model can be re-parametrized following $q_1$, $q_2$ and $q_3$, and set the rest of the prior distributions as described for the baseline model (see Supplementary Table 2; Code Availability section).
+where $\gamma^{\prime}_{i} = q_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution such that $\lambda_{i}\in\left(-1, 1\right)$.
+The function $\text{sgn}\left(x\right)$ characterizes the sign function.
+We chose $\lambda_i$ to have an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
+Notice that this model can be re-parametrized following $q_1$ and $q_2$, allowing us to set the rest of the prior distributions as described for the baseline model (see Supplementary Table 2; Code Availability section).
 In this case, the Gaussian distribution is a special case of Eq.~(\ref{eq:skewed-response}) when $\lambda_{i}=0$ \citep{ashourApproximateSkewNormal2010}.
-
-One way to test the extend to which species' physiological traits inform the skewness of the distribution of any species $i$ is by characterizing $\lambda_{i}$ as a Gaussian process.
-That is, we can consider the prior distribution for $\lambda_{i}$ as a multivariate normal with a variance covariance matrix $\Sigma^{\lambda}$.
-As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
-Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.
 
@@
 where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively.
-The function $\text{sgn}\left(x\right)$ characterizes the sign function.
-We define $\nu_i$ and its prior as in Eq.~\ref{eq:generalized-response}.
-However, $\lambda_i$ is now defined as $\lambda_{i}\in\left(-1, 1\right)$, with an adaptive prior such that $\text{logit}\left(\frac{\lambda_{i}+1}{2}\right)\sim \text{Normal}\left(\hat{\lambda}, \sigma_{\lambda}\right)$, where $\hat{\lambda}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\lambda}\sim\text{Exponential}\left(1\right)$.
+We define $\nu_i$, $\lambda_i$ and their prior distributions as in Eq.~\ref{eq:generalized-response} and \ref{eq:skewed-response}, respectively.
 Again, we can re-parametrize the model following $f_1$ and $f_2$, and set the rest of the prior distributions as in the baseline model (see Supplementary Table 2; Code Availability section).
 Notice that the generalized error distribution (Eq.~\ref{eq:generalized-response}) and the skew normal distribution (Eq.~\ref{eq:skewed-response}) are special cases of Eq.~(\ref{eq:heavy-skewed-response}) when $\lambda_i=0$ and $\nu_i=2$, respectively.
+
+One way to test the extend to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes.
+That is, we can consider the prior distribution for these parameters as multivariate normal with a variance covariance matrix $\Sigma$.
+As described before, this variance-covariance matrix can be estimated following Eq.~(\ref{eq:covariance-baseline}), and its structure can shed light on how informative the physiological traits are.
+Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.
 
@@
 Studying the relationship between the mean and variance of the distributions, we found clear evidence of the Rapopor's rule.
-Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}).
-The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Fig.~\ref{fig:universality}).
-This pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
+Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}; Supplementary Fig.~XX).
+The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), but this pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
 The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.
@@
 
+Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways.
+To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}).
+A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX).
+Studying the resulting posterior distributions, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with longer tails than a Gaussian (Fig.~\ref{fig:kurtosis-skewness}).
+However, such parameter estimates displayed a lot of variation across species, which might indicate that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
+Similarly, using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species distributions.
+Based on the estimates for the WAIC values, this model outperformed the previous models (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Perhaps most importantly, studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions generally present steeper declines towards higher elevations (Fig.~\ref{fig:kurtosis-skewness}).
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.9\textwidth]{figures/kurto-skew}
+    	  \vspace{0.1cm}
+	   \caption{Average kurtosis and skewness of species' distributions.
+Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively.
+Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution.
+Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution.
+In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
+      \label{fig:kurtosis-skewness}
+      %7.5-3
+\end{figure}
+
+Finally, we used a model that allowed for heavy-tails and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}) to understand the relationship between the kurtosis and skewness of distributions.
+This model outperformed the rest, presenting Akaike weights close to 1 when comparing models (Supplementary Fig.~XX).
+Comparing the posterior distribution for parameters $\nu_i$ and $\lambda_i$, we revealed the general constrains on the distribution of alpine plants studied.
+That is, we uncovered the relationship between the mean, variance, amplitude, kurtosis and skewness of empirical distribution of plants along an elevational gradient (Figs.~XX and YY).
+Moreover, different types of species seem to present characteristically different distributions (Fig.~XX).
+For example, invasive species...
+Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area (bootstrapping...
+Supplementary Fig.~XX).
+Likewise, the posterior distributions obtained using binary data were strikingly similar to those obtained using categorical data.
+This is important because...
+is model identified similar relationships between parameters as the baseline model (Supplementary Fig.~XX), and measured similar average levels of kurtosis and skewness than the the other two models (Fig.~\ref{fig:kurtosis-skewness}).
+
+%, and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Fig.~\ref{fig:universality}). T
+%Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX).
+
 \begin{figure}[ht]
@@
 {\color{red} I might move this to the Supplementary Information.}}
-	   %7.5-5.5
+	   %7.5-4
       \label{fig:universality}
 \end{figure}
-
-Maintaining the symmetry of species' distributions, we then allowed the kurtosis---or shape of the tails---of these to vary in different ways.
-To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}).
-A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX)\footnote[2]{\label{note1} I am still waiting on the results for this (currently running in the cluster), and this is just my prior expectation based on what I've seen in some of the other models I've been working with.}.
-Studying the resulting posterior distributions, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with a Gaussian shape (Fig.~\ref{fig:kurtosis-skewness}).
-However, such parameter estimates displayed a lot of variation across species, which might indicate that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
-
-\begin{figure}[ht]
-  \centering
-    \includegraphics[width=0.8\textwidth]{figures/kurto-skew}
-    	  \vspace{0.1cm}
-	   \caption{Average kurtosis and skewness of species' distributions.
-Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively.
-Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution.
-Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution.
-In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
-      \label{fig:kurtosis-skewness}
-\end{figure}
-
-Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species distributions.
-Based on the estimates for the WAIC values, this model outperformed the rest (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
-Perhaps most importantly, studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions present steeper declines towards lower elevations (Fig.~\ref{fig:kurtosis-skewness}).
-Moreover, species' physiology seems to strongly shape this parameter, which suggest that distribution skewness is an intrinsic property of species (New Fig.~XX)\footref{note1}.
-
-Finally, we used a model that allowed for heavy-tails and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}) to understand the relationship between the kurtosis and skewness of distributions.
-This model outperformed the rest, presenting Akaike weights close to 1 when comparing models (Supplementary Fig.~XX).
-Comparing the posterior distribution for parameters $\nu_i$ and $\lambda_i$, we found this to be uncorrelated\footref{note1}.
-However, different types of species seem to present characteristically different distributions based on their kurtosis and skewness.
-Importantly, this model identified similar relationships between parameters as the baseline model (Supplementary Fig.~XX), and measured similar average levels of kurtosis and skewness than the the other two models (Fig.~\ref{fig:kurtosis-skewness})\footref{note1}.
 
@@
 
+First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}.
+While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
+Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient.
+Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref).
+\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
+Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
+Specifically, we are able to link plant physiological traits to the skewness of their distributions.
+
 %%%%%%%%%%%%%%%%%%

### 2021-05-01 code/paper-models/plots/plot-real.R

@@ -966,3 +966,97 @@ plot.actual.data.distributions <- function(model=NULL){
   
 }
 
+ll.analysis <- function(model){
+  # post <- extract.samples(model, n = 1000, pars=c("log_lik", "gamma", "beta"))
+  # post$gamma <- 1/(2*post$gamma)
+  d <- readRDS("../../../data/processed/jsdm/1d-categorical-PC1PC2min20-data.rds")
+  N <- sum(d$dataset$id==1)
+  L <- length(unique(d$dataset$id))
+  dat <- d$dataset
+  X <- matrix(dat$PC1, N, L)[,1]
+  
+  pb <- txtProgressBar(min = 0, max = L, initial = 0, style = 2)
+  for(i in 1:L){
+    idxs <- sample(1:dim(post$beta)[1],10)
+    beta <- post$beta[idxs,i]
+    gamma <- post$gamma[idxs,i]
+    ll <- post$log_lik[idxs,((i-1)*N+1):(i*N)]
+    diffs <- outer(X, beta, "-")
+    diffs <- t(diffs %*% diag(1/gamma))
+    
+    diffs <- as.vector(diffs)
+    ll <- as.vector(ll)
+    if (i==1){
+      data <- data.frame( x=diffs, y=ll )
+    }else{
+      dat <- data.frame( x=diffs, y=ll )
+      data <- rbind(data, dat)
+    }
+    setTxtProgressBar(pb, i)
+  }
+  close(pb)
+  
+  p <- ggplot(data, aes(x=x, y=y) ) +
+    geom_hex() +
+    theme_bw()
+  print(p)
+}
+
+abundance.center.HP <- function(model){
+  post <- extract.samples(model, n = 1000, pars=c("log_lik","alpha", "gammav", "betam", "lambda", "nu"))
+  # post$gamma <- 1/(2*post$gamma)
+  d <- readRDS("../../../data/processed/jsdm/1d-categorical-PC1PC2min20-data.rds")
+  N <- sum(d$dataset$id==1)
+  L <- length(unique(d$dataset$id))
+  dat <- d$dataset
+  X <- matrix(dat$PC1, N, L)[,1]
+  obs <- matrix(dat$obs, N, L)
+
+  data_max <- list()
+  data_min <- list()
+  data <- list
+  for(i in 1:L){
+    idxs <- sample(1:dim(post$beta)[1],10)
+    loglik <- post$log_lik[idxs,i]
[…]

### 2021-05-01 code/sliding-paper-models/1d-baseline.R

@@ -1,0 +1,146 @@
+source("./prepare-data-sliding.R")
+source("./models-binomial.R")
+library(rethinking)
+library(rstan)
+library(cmdstanr)
+
+options(mc.cores = parallel::detectCores())
+rstan_options(auto_write = TRUE)
+
+####
+## Run species distribution model for all species. I wrote the following models:
+# - baseline model
+####
+
+skew.generror.1d <- function(partition=1, d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+    # Fixing some of the options
+    variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
+    ndim <- 2
+    pca <- T
+  
+  
+  # If we are dealing with simulated data
+  if(simulated){
+    extension <- paste("sliding-",as.character(partition),"-skew-generror-simulated", sep="")
+  }else{
+    extension <- paste("sliding-",as.character(partition),"-skew-generror", sep="")
+  }
+  
+  extension2 <- ""
+  
+  # Load the data
+  if(recompile){
+    d <- species_distribution.data(partition=partition, variables=variables, pca=pca, ndim = ndim, simulated=F, elevation = F, simulated.type="skew.generror", min.occurrence=min.occurrence)
+    # rename variables
+
+    if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+    }
+    
+    filename <- paste("../../data/processed/jsdm/jsdm-sliding/", extension, "-",paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+    
+    if (file.exists(filename)){
+      question <- askYesNo("Do you want to overwrite the file?", default = F, 
+                           prompts = getOption("askYesNo", gettext(c("Yes", "No", "Cancel"))))
+      if(is.na(question)){question <- F}
+      if(question){
+        saveRDS(d, file=filename)
+      }else{
+        stop("you should add 'recompile=T'")
+      }                                
+    }else{
+      saveRDS(d, file=filename)
+    }
+  }else{
+    # rename variables
+      if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+      }
+    
[…]

### 2021-05-01 code/sliding-paper-models/models-binomial.R

@@ -1,0 +1,121 @@
+# Binomial 1d
+base.model.skew.generror.1d.multithread <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
+          K[i, i] = a + delta;
+          for (j in (i + 1):N) {
+            K[i, j] = a * exp(- b * square(x[i,j]) );
+            K[j, i] = K[i, j];
+          }
+        }
+        K[N, N] = a + delta;
+        return K;
+    }
+    int sgn(real x) {
+        return x < 0 ? -1 : x > 0;
+    }
+    real partial_sum(int[] y_slice,
+                   int start, int end,
+                   int N, vector lambda, vector alpha, vector beta, vector gamma, vector nu, int[ , ] Y, row_vector X1, real minp ) {
+        real lp = 0.0;
+        real m;
+        real v;
+        for( i in start:end){
+            v = gamma[i] * sqrt((pi()*(1+3*pow(lambda[i],2))*tgamma(3.0/nu[i])-pow(16,(1.0/nu[i]))*pow(lambda[i],2)*tgamma(0.5+1.0/nu[i])*tgamma(0.5+1.0/nu[i])*tgamma(1.0/nu[i]))/(pi()*tgamma(1.0/nu[i])));
+            m = beta[i] - pow(2, 2.0/nu[i])*lambda[i]*tgamma(0.5+1.0/nu[i])/(sqrt(pi())*v);
+            for( j in 1:N ){
+                lp += binomial_lpmf( Y[i, j] | 1 , exp(-alpha[i] - pow(v * fabs(X1[j] - m)/(1+lambda[i]*sgn(X1[j] - m)), nu[i])) + minp);
+            }
+        }
+        return lp;
+    }
+}
+data{
+    int N;
+    int L;
+    real minp;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    int indices[L];
+    row_vector[N] X1;
+    matrix[L,L] Dmat_b;
+    matrix[L,L] Dmat_g;
+}
+parameters{
+    vector[L] zalpha;
+    vector[L] zbeta;
+    vector[L] zgamma;
+    vector[L] znu;
+    vector[L] zlambda;
+    real alpha_bar;
+    real beta_bar;
[…]

### 2021-05-01 code/sliding-paper-models/prepare-data-sliding.R

@@ -1,0 +1,167 @@
+library("sp")
+library("raster")
+library("maptools")
+library("dismo")
+library("rJava")
+library(SDMTools)
+source("../prepare-data/useful-tools.R")
+library(MASS)
+library(gridExtra)
+library(grid)
+
+# Define projection as global variable
+projection <- "+proj=somerc +init=world:CH1903"
+
+prepare.data <- function(partition=1, variables = c("bio5_", "bio6_","bio12_"), min.occurrence=0, elevation=F){
+        
+        # Determine geographic extent of our data
+        places <- read.csv(file = paste("../../data/properties/codes/sdm-sliding/places_codes_",as.character(partition),".csv", sep=""), header = T)
+        # places$elevation <- as.vector(scale(places$elevation))
+        max.lat <- max(places$northing)
+        min.lat <- min(places$northing)
+        max.lon <- max(places$easting)
+        min.lon <- min(places$easting)
+        geographic.extent <- extent(x = c(min.lon, max.lon, min.lat, max.lat))
+        
+        # Load data for all species
+        files <- as.numeric(gsub(".csv", "", list.files(paste("../../data/processed/sdm-sliding/partition-",as.character(partition),"/", sep=""))))
+        sp_codes <- read.table(paste("../../data/properties/codes/sdm-sliding/sp_codes_",as.character(partition),".csv", sep=""), sep=",", header = T)
+        dictionary <- read.table("../../data/properties/codes/dictionary.csv", sep=",", header = T)
+        correlation_matrix_ids <- read.table("../../data/properties/codes/correlation_matrix_ids.csv", sep="\t", header = F)
+        denvironment <- as.matrix(read.table("../../data/properties/distance-matrices/environment.csv", sep=","))
+        dvariation <- as.matrix(read.table("../../data/properties/distance-matrices/variation.csv", sep=","))
+        dtraits <- as.matrix(read.table("../../data/properties/distance-matrices/trait.csv", sep=","))
+        
+        indicator <- read.table("../../data/properties/codes/temperature_indicator.csv", sep=",")
+        neophytes <- read.table("../../data/properties/codes/neophytes-list.csv", sep=",")
+        tendency <- read.table("../../data/properties/codes/change-tendency.csv", sep=",")
+        competitive <- read.table("../../data/properties/codes/competitive_indicator.csv", sep=",")
+        
+        # Check that there aren't unnexpected files
+        if(!all(sort(files)==1:length(files))){
+            stop("Odd files in the folder")    
+        }
+        
[…]

### 2021-05-01 code/sliding-paper-models/run_inte.sh

@@ -1,0 +1,3 @@
+#!/usr/bin/env bash
+
+bsub -R "rusage[mem=2840]" -W 200:00 -n 45 "R --vanilla --slave < 1d-baseline.R"

### 2021-05-01 code/sliding-paper-models/1d-baseline.R
Fixing little things

@@ -112,30 +112,30 @@ skew.generror.1d <- function(partition=1, d = NULL, recompile = T, simulated=T, 
   init_5.1 <- list()
   for ( i in 1:n_chains_5.1 ) init_5.1[[i]] <- start_5.1
 
-  # 
-  # dat_5.1$indices <- 1:L
-  # 
-  # model_code = base.model.skew.generror.1d.multithread
-  # generror1d <- cmdstan_model(write_stan_file(model_code), cpp_options = list(stan_threads = TRUE))
-  # mfit_5.1 <- generror1d$sample(data = dat_5.1,
-  #                               init = init_5.1,
-  #                               chains = 3,
-  #                               threads_per_chain = 15,
-  #                               parallel_chains = 3,
-  #                               max_treedepth = 15,
-  #                               max_depth = 15,
-  #                               iter_sampling = 1000,
-  #                               #adapt_delta = 0.95,
-  #                               refresh = 500)
-  # mfit_5.1$save_object(file = paste(ofolder, extension2, "", extension,"-cdmstan.rds", sep=""))
-  # saveRDS(rstan::read_stan_csv(mfit_5.1$output_files()), file = paste(ofolder, extension2, "", extension,".rds", sep=""))
-  # print(mfit_5.1$cmdstan_diagnose())
-  # 
-  # return(mfit_5.1)
+
+  dat_5.1$indices <- 1:L
+
+  model_code = base.model.skew.generror.1d.multithread
+  generror1d <- cmdstan_model(write_stan_file(model_code), cpp_options = list(stan_threads = TRUE))
+  mfit_5.1 <- generror1d$sample(data = dat_5.1,
+                                init = init_5.1,
+                                chains = 3,
+                                threads_per_chain = 10,
+                                parallel_chains = 3,
+                                max_treedepth = 15,
+                                max_depth = 15,
+                                iter_sampling = 1000,
+                                #adapt_delta = 0.95,
+                                refresh = 500)
+  mfit_5.1$save_object(file = paste(ofolder, extension2, "", extension,"-cdmstan.rds", sep=""))
+  saveRDS(rstan::read_stan_csv(mfit_5.1$output_files()), file = paste(ofolder, extension2, "", extension,".rds", sep=""))
+  print(mfit_5.1$cmdstan_diagnose())
+
+  return(mfit_5.1)
 }
 
 args <- commandArgs(trailingOnly=T) #Getting your input arguments
-i <- args[1]
+i <- as.numeric(args[1])
 
 min.occurrence <- 20
 

### 2021-05-01 code/sliding-paper-models/run_inte.sh

@@ -1,3 +1,4 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=2840]" -W 200:00 -n 45 "R --vanilla --slave < 1d-baseline.R"
+--args \$LSB_JOBINDEX $1 <
+bsub -R "rusage[mem=2500]" -W 24:00 -n 35 "R --vanilla --slave --args $1 < 1d-baseline.R"

### 2021-05-01 code/sliding-paper-models/1d-baseline.R
Preparing code for the cluster

@@ -1,4 +1,4 @@
-source("./prepare-data-sliding.R")
+# source("./prepare-data-sliding.R")
 source("./models-binomial.R")
 library(rethinking)
 library(rstan)
@@ -142,5 +142,5 @@ min.occurrence <- 20
 # Basic analyses
 d <- readRDS(paste("../../data/processed/jsdm/jsdm-sliding/sliding-",as.character(i),"-skew-generror-PC1PC2data.rds", sep=""))
 
-skew.generror.1d(partition=i, d=NULL, simulated=F, recompile = T, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
+skew.generror.1d(partition=i, d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 

### 2021-05-13 manuscript/manuscript.tex
New intro and new categorical models

@@
 \section*{Abstract}
+Species' realized niches are classically pictured as bell-shaped probability distributions.
+These distributions, however, can actually take many different forms.
+For example, fat-tailed or skewed responses are very common across fields, as these can naturally emerge as a result of several ecological processes.
+While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies have shaped ecological communities.
+Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to uncover such properties.
+These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies.
+With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space.
+In particular, we studied the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients.
+Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts.
+Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
 
-Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades.
+%Species' realized niches are often visualized as bell-shaped probability distributions.
+These, however, can take many different forms.
+For example, fat-tailed or skewed distributions are very common across fields, as these can naturally emerge as a result of several ecological processes.
+While one does not need to know the shape of species' distributions to effectively model them, there is a lot that can be learned about how climatic processes and historical contingencies differently affected species' distributions by studying the basic form.
+Indeed, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to uncover such properties.
+These models account for all prior knowledge we have regarding species' niches, including expert knowledge of their environmental preferences and ecological strategies.
+With this approach, we are able to study the shape empirical plant distributions, revealing general patterns regarding the way communities are assembled across space.
+In particular, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns that are present along environmental gradients.
+We found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation, or their general level of skewness.
+Perhaps most importantly, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical context.
+Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
+
+%The shape of species' distributions along environmental gradients is part of old debates in ecology and biogeography.
+These have led the development of many statistical models centred around predicting the presence and absence of species over space and time.
+Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change.
+However, there is a lot that can be learned about how climatic processes and historical contingencies have differently shaped species' distributions by studying instead their basic shape.
+Indeed, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches.
+These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and ecological strategies.
+With this approach, we are able to uncover the general properties characterizing empirical plant distributions, revealing general patterns regarding the way communities are assembled.
+That is, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns present along environmental gradients.
+In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients.
+Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species.
+Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
+
+%Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades.
 Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change.
@@
 Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches.
-These models account for all prior knowledge we have regarding their shape, including expert knowledge on species' environmental preferences and physiology.
+These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and physiology.
 With this approach, we are able to uncover the true shape of empirical species' distributions.
-Moreover, they allow us to tackle long-standing hypothesis regarding general biogeographical patterns.
+Moreover, they allow us to tackle long-standing hypothesis regarding broad biogeographical patterns.
 In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients.
@@
 %############# NEW INTRO ##############
+%Bell shape is a good option, but it is unclear what the real shape is
+One of the central goals of ecology is to understand the ways species are distributed across space and time.
+While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, abundant-centre-hypothesis}.
+In practice, there is a strong argument to be made in favour of assuming these to be bell shaped.
+Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}).
+That said, there is currently no general agreement on the basic shape of species' realized niches.
+Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.
 
-% Describe different aspects of shape
+%That said, there is currently no general agreement on the general shape of species' realized niches, as several natural processes can lead to non-normal distributions.
+%That said, there is currently no general agreement on the general shape of species' realized niches.
+Several natural processes can lead to non-normal distributions, and every new shape encompasses different underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.
+%That said, there is currently no general agreement on the general shape of species' realized niches, as several natural processes can lead to non-normal distributions.
+%That said, several natural processes can lead to non-normal distributions, and there is currently no general agreement on the general shape of species' realized niches, and.
 
-One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
-Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}.
-The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
-While these models are fundamental to the scientific community for predicting changes in species geographic distributions and the effects of environmental disturbances, distribution models ...
-can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019},
+%Skewed and fat-tails are common and they could well describe niches
+Fat-tailed and skewed distributions are very common across fields.
+The former naturally emerges as a result of processes involving seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
+Indeed, species' dispersal patterns have been shown to have fat tails due to the natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
+This is important because one might expect environmental and individual variation to also be crucial factors determining the presence and absence of species along gradients, and fat tails are therefore a plausible property of species' realized niches.
+Similarly, several processes can lead to skewed distributions.
+For example, species might present asymmetric environmental tolerances along altitudinal gradients, allowing them to withstand different temperature extremes \citep{sundayGlobalAnalysisThermal2011}.
+Species might also experience abiotic and biotic pressures that increase or decrease along a temperature gradient, which could result in species' distributions presenting steeper declines towards warmer or colder environments \citep{normandImportanceAbioticStress2009}.
+Overall, many different properties could characterize species' realized niches, and every new shape entails different underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.
 
-While the value of these models is unquestionable, these approaches often rely on broad simplifications regarding the shape of species' realized niches. Describing
+%they can also reveal differences among species regarding the way they assemble along a gradient and broad macroecological patterns defining natural constraints shaping communities. 
+%The general properties of species' distributions are valuable because they point to important ecological processes shaping species' realized niches. 
+%For example, a relationship or lack thereof between range size and elevation would speak to classical ecological hypotheses (i.e.~Rapopor's rule;  \citealt{stevensElevationalGradientAltitudinal1992}) and could hint at the existence of general biogeographical constraints that shape species' distributions along gradients.
+%Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
+Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, bernat}.
+For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992, bernat}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients.
+This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments.
+That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}.
+Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve.
+Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
+%Understanding the shape of species' realized niches and the extent to which these vary across species is therefore a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
+%Understanding the shape of species' realized niches and the extent to which these vary across species is however challenging statistical problem.
+That is, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
+%With all of that being said, measuring these properties is a challenging statistical problem, and we do not have an effective way to parsimoniously compare the shape of the realized niches of many species.
 
- These models estimate species' realized niches using several covariates, including environmental variables \citep{guisanPredictingSpeciesDistribution2005}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}.
-More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}.
-The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time.
-However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been recently questioned \citep{harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019, poggiatoInterpretationsJointModeling2021}.
-This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
+Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000}.
+These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances.
+Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}).
+This is useful because it simplifies the optimization process but might not be ideal when studying and comparing the shape of species' distributions along environmental gradients.
+First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified.
+While this might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}.
+Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates.
+Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
+That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
+Finally, \citet{huismanHierarchicalSetModels1993} proposed several non-linear structures to characterize several features of individual species' response curves.
+Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following these model structures, these were not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g.
+sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).
 
-From a phenomenological perspective, one can focus on several aspects of species' distributions. For example,
-
-Two aspects that are far less explored: the shape and the relationship between the properties of species distributions.
-
-Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric.
-However, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
-From a purely statistical perspective, there is a strong argument to be made in support of bell shaped distributions.
-That is, if all that we are willing to assume about a species' distributions is that these occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model their distribution as Gaussian \citep{frankCommonPatternsNature2009}.
-That said, many natural processes can lead to non-normal distributions, and heavy-tails and skewed shapes are very common across fields.
-For example, heavy-tailed distributions can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
-Indeed, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
-Likewise, several processes can lead to skewed distributions.
-For example, when species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction; and this same skewed shape can also emerge as a result of species' asymmetric environmental tolerance.
-
- For example, species' distributions are shaped by multiple stochastic events, such as di distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
-
-% Relate them to macroecological hypothesis
-
-That said, a lot can be learned about species' geographic distributions from taking a phenomenological approach, focussing instead on the description of basic properties of their realized niches.
-For example, the study of species' range sizes along environmental gradients can reveal general biodiversity patterns that are crucial from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992}.
-Differences in species' responses to the environment could shed light on how climatic processes and historical contingencies have shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}.
-Other properties, such as the skewness of species' distributions, can also reveal general underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
-More generally, understanding the shape of species' realized niches and the extend to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
-Indeed, there is no general agreement on the shape of species' distributions (ref).
-
-% Describe limitations in previous methods to characterize shape of Realized Niche: prior knowledge; simultaneous estimation of many (also rare); uncertainty
-
-Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
-In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see \citealt{review} and \citealt{good example}).
-This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings.
-First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
-From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
-Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
-The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
-Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
-Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
-That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
-Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging.
-For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to understand altogether following these model structures.
-
-% Intro new method
-Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
-In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology.
-We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions.
-Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns.
-
-% Questions
+In this work, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
+In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies.
+We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions.
+Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns.
+Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or neophytes, might respond to the environment.
+More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
 
@@
 
-One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
+%One of the central goals of ecology is to understand the ways species are distributed across space and time (ref).
 Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000}.
@@
 This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
-% Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018
+%%%Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018
 
-The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and dispersal ability in setting species' range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
+%The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and dispersal ability in setting species' range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}.
 That said, a lot can be learned about species' geographic distributions from taking a phenomenological approach, focussing instead on the description of basic properties of their realized niches.
@@
 Other properties, such as the skewness of species' distributions, can also reveal general underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
-More generally, understanding the shape of species' realized niches and the extend to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
+More generally, understanding the shape of species' realized niches and the extent to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species.
 Indeed, there is no general agreement on the shape of species' distributions (ref).
 
-Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
+%Many ecological textbooks \citep{krebsEcologyExperimentalAnalysis1972} assume the shape of species' distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}.
 In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see \citealt{review} and \citealt{good example}).
 This is useful because it simplifies the optimization process, but it comes with some statistical shortcomings.
-First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified.
+First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified.
 From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e.~their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}.
-Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
+Other factors might then condition species' distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}.
 The starting point, nevertheless, should be the one that makes the fewest assumptions (i.e.~the maximum entropy distribution; \citealt{frankCommonPatternsNature2009}), and every new shape will imply a hypotheses on how communities are distributed \citep{damenSpatialPredictionsCommunity2017}.
-Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates.
-Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
+Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates.
+Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
 That said, this information is rarely accounted for in most modelling approaches, potentially because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
@@
 
-%This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
-%\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species distributions.
+%%%%This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs).
+%%%%\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species' distributions.
 
-The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}.
+%The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}.
 This has resulted in a plethora of models accounting for several biotic and abiotic factors into the predictions of species co-occurrence.
 Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
-In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology.
+In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant physiology.
 We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions.
@@
 Specifically, we are able to link plant physiological traits to the skewness of their distributions.
-Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constrains of the statistical framework.
+Overall, we use models that are solely constrained by the empirical information that we truly have regarding our system, relaxing as much as possible the structural constraints of the statistical framework.
 Using these models, we are able uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
@@
 We know that species occupy a geographic range; therefore, we know that their distributions have finite variance.
-Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions.
+Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' distributions.
 One could point out that we also know that many other factors might influence species' presence/absence---e.g.
@@
 The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies.
-Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework.
+Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constraints of the statistical framework.
 Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.
@@
 2019.
-A comparison of joint species distribution models for presence–absence data.
-Methods in Ecology and Evolution 10:198–211; A comprehensive evaluation of predictive performance of 33 species distribution models at species and community levels
+A comparison of joint species' distribution models for presence–absence data.
+Methods in Ecology and Evolution 10:198–211; A comprehensive evaluation of predictive performance of 33 species' distribution models at species and community levels
 
-%The last two decades have seen a proliferation of species distribution models (SDMs) addressing the challenge of predicting the occurrences of individual species (Guisan and Zimmermann 2000, Guisan and Thuiller 2005, Elith et al.
+%The last two decades have seen a proliferation of species' distribution models (SDMs) addressing the challenge of predicting the occurrences of individual species (Guisan and Zimmermann 2000, Guisan and Thuiller 2005, Elith et al.
 2006, Leathwick et al.
@@
 2010).
-Methodological advances in multiple‐species distribution modeling have lagged behind, but are recently experiencing a rapid expansion (Leathwick et al.
+Methodological advances in multiple‐species' distribution modeling have lagged behind, but are recently experiencing a rapid expansion (Leathwick et al.
 2006, Dunstan et al.
@@
 %
-%The use of species distribution models has grown a lot.
+%The use of species' distribution models has grown a lot.
 These models try to estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref).
@@
 \subsection*{Empirical data}
-We studied the distribution of alpine plant communities along an elevation gradient.
-To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.
+We studied the distribution of plant communities along an elevation gradient.
+To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps \citep{randinClimateChangePlant2009}, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.
 
@@
 \subsubsection*{Floristic data}
-To complement the aforementioned distribution data, we used a floristic database of most vegetation across Switzerland.
-This database was build based on expert knowledge and field experience of botanists and ecologists, and contains information regarding species' environmental preferences and physiological traits.
+To complement the aforementioned distribution data, we used a floristic database of around 5500 vascular plants across Switzerland.
+Some of the information in this database has been previously shown to account for unexplained variation when used as explanatory variables in species' distribution models \citep{scherrerEcologicalIndicatorValues2019}.
+It was built based on expert knowledge and phytosociological field experience of botanists and ecologists, and contains information regarding plants' environmental preferences and ecological strategies.
+
 Species' environmental preferences in this database can be used to inform distribution models---e.g.~as an informative prior in a Bayesian framework.
 These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation.
-These values are provided for a range of 10 climatic variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-On the other hand, the information regarding species' physiological traits represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify more than $120$ binary traits that characterize the physiology of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species types (e.g.~identifying those species that are neophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
+These values are provided for a range of 10 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
+On the other hand, the information regarding species' ecological strategies represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
+In total, we identify more than $120$ binary traits that characterize the physiology and life-history of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
+Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species introduction status (e.g.~identifying those species that are neophytes or archaeophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
 We describe this information in more detail in the Supplementary Table 1.
@@
 \subsection*{Baseline model}
-There is a long list of model structures well suited to characterize species' distributions (see \citealt{norbergComprehensiveEvaluationPredictive2019}).
+There is a long list of model structures well suited to characterizing species' distributions (see \citealt{norbergComprehensiveEvaluationPredictive2019}).
 As a baseline model, however, we were interested in a hierarchical model that does not make any assumptions regarding the shape of the distributions, and yet explicitly incorporates all information that we have regarding plant's environmental preferences.
@@
 \end{equation}
-where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the an environmental gradient, respectively.
+where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively.
 Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods).
-For the sake of simplicity, we use only one environmental variable to characterize species' probability distribution.
+For the sake of simplicity, we use only one environmental variable to characterize the                                                                             species' probability distribution.
 That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
@@
 The hyperparameter $\sigma$ describes the additional covariance between the different observations for any given species.
-For all these hyperparameters, we choose weekly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.
+For all these hyperparameters, we choose weakly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.
 
@@
 \subsubsection*{Sampling the posterior}
-We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R package `rstan' to \citep{standevelopentteamRStanInterfaceStan2021}.
+We generated the posterior samples for the Bayesian models with the Hamiltonian Monte Carlo algorithm implementation provided by the R packages `rstan' and `cmdstanr' \citep{standevelopentteamRStanInterfaceStan2021}.
 Sampling models like the ones described above can be computationally very intensive.
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have at least 20 occurrences when modelling binary data.
-When using ordinal data, we limit our study to a random subset of a 100 species of those with at least 20 occurrences to facilitate the sampling of the models.
+Therefore, we focus on those species for which we have at least 20 occurrences when modelling both binary data and ordinal data.
 
@@
 Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).
+To perform the data analysis and generate the figures, we used some of the functions available with the R package `rethinking' \citep{mcelreathStatisticalRethinkingBayesian2020}.
 
@@
 Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information.
-To propose new species' response curves, we follow three criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
+We focused on two properties: fat-tailed and skewed responses.
 
-\subsubsection*{Heavy-tail response curve}
-Distributions with heavy-tails are very common across fields, as they can capture processes such as seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
-Indeed, heavy-tail distributions are pervasive in ecology; for example, species' dispersal patterns have been shown to have heavy-tails due to natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
-Therefore, one might expect these properties to also emerge in alpine communities, where seasonality and dispersal patterns are crucial factors determining species' distributions.
-To accommodate this feature into our baseline model, we could consider a response curve that follows a generalized error distribution, since the Gaussian shape is a special case of it and contains parameter that regulates the level of kurtosis (Fig~\ref{fig:response}b).
+%To modify our baseline model, we propose new species' response curves we follow three criteria: (i) the probability distribution must have a defined variance and mean, (ii) the Gaussian shape must be a special case of the probability distribution, and (iii) there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
+
+We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system.
+Now, we want to modify this model to test the extent to which empirical species' distributions showcase different shapes.
+We focused on two properties: fat-tailed and skewed responses.
+While there are several model structures that could account for these properties, we propose new species' response curves following three criteria.
+First, the probability distribution of a species along an environmental gradient must have a defined mean and variance.
+This is important because we know that species naturally have different environmental preferences as well as finite geographic ranges.
+Second, the Gaussian shape must be a special case of the probability distribution, allowing species to showcase variation regarding the presence (or lack thereof) of any given pattern.
+Finally, there must be a re-parametrization of the model that allows us to keep the same prior information and interpretable parameters.
+
+\subsubsection*{Fat-tailed response curve}
+Fat-tailed distributions represent distributions with relatively high representation of extreme events.
+While many different distributions exhibit this property, we decided to accommodate this feature into our baseline model by considering a response curve that follows a generalized error distribution.
+Such a distribution is useful because the Gaussian shape is a special case of it, and it contains a parameter that regulates the level of kurtosis---ranging from longer to shorter tails than the Gaussian case (Fig~\ref{fig:response}b).
 In particular, we can adapt Eq.~(\ref{eq:baseline-response}) to present this non-linear form as follows:
@@
 
-Alternatively, we could have used other distributions that present heavy-tails and fulfil the selection criteria described above.
-For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for heavy tails without generating a cusp at the center (see Fig~\ref{fig:response}b).
+Alternatively, we could have used other distributions that present fat tails and fulfil the selection criteria described above.
+For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for fat tails without generating a cusp at the center (see Fig~\ref{fig:response}b).
 However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Fig~\ref{fig:response}b), and the sampling of the model can be somewhat more challenging (ref).
@@
 \subsubsection*{Skewed response curve}
-When species experience abiotic or biotic pressures that increase or decrease along an environmental gradient, one might expect their distributions to be skewed in one direction.
-Likewise, this same skewed shape can also emerge as a result of species' asymmetric environmental tolerance.
-One way to accommodate this feature to our models is by considering a skewed normal distribution (Supplementary Figure XX).
-As for the two cases described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness' (Fig~\ref{fig:response}c).
+Skewed responses present steeper declines towards either side of the distribution.
+One way to accommodate this feature in our models is by considering a skewed normal distribution (Supplementary Figure XX).
+As for the case described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness' (Fig~\ref{fig:response}c).
 Importantly, this distribution presents normal-like tails; therefore, the added skewness does not make additional assumptions regarding how species are distributed along the gradient.
@@
 
-\subsubsection*{Heavy-tail and skewed response curve}
+\subsubsection*{Fat-tailed and skewed response curve}
 Finally, one could consider a response curve with both kurtosis and skewness.
-A convenient way to achieve this is by using a response curve that follows a skewed generalized error distribution (also see XX).
+A convenient way to achieve this is by using a response curve that follows a skewed generalized error distribution.
 This is a combination of the two distributions described above, containing two parameters that control for both the level and direction of kurtosis and skewness (Fig~\ref{fig:response}d).
@@
 \end{equation}
-where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, $\nu_{i}$ and $\lambda_i$ are parameters that describe the kurtosis and skewness of the distribution, respectively.
+where $\gamma^{\prime}_{i} = f_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = f_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, and $\nu_{i}$ and $\lambda_i$ are parameters that control the kurtosis and skewness of the distribution, respectively.
 We define $\nu_i$, $\lambda_i$ and their prior distributions as in Eq.~\ref{eq:generalized-response} and \ref{eq:skewed-response}, respectively.
@@
 
-One way to test the extend to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes.
+One way to test the extent to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes.
 That is, we can consider the prior distribution for these parameters as multivariate normal with a variance covariance matrix $\Sigma$.
@@
 \section*{Results}
-We studied the distribution data to characterize species' realized niches.
-To do so, we selected the main axis of variation of all environmental variables (Supplementary Fig.~XX).
-Then, using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
-Studying the relationship between the mean and variance of the distributions, we found clear evidence of the Rapopor's rule.
-Specifically, we found that species' range sizes decline with elevation and temperature (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model are negatively correlated; Fig.~\ref{fig:correlation}; Supplementary Fig.~XX).
-The relationship was also found when using instead ordinal data (Supplementary Fig.~XX), but this pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
+We studied the distribution data to characterize species' realized niches along the main axis of variation of all environmental variables.
+Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
+This allowed us to map the center and variance of species' distributions along the environmental gradient (Fig.~\ref{fig:correlation}).
+Studying the relationship between these properties, we found these to be negatively correlated (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model were positively correlated; Fig.~\ref{fig:correlation}).
+This means that species found at lower elevations have generally wider distributions than those at higher elevations.
+The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~XX) as well as when using ordinal data (Supplementary Fig.~XX); however, the pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
 The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.
-In particular, we found the amplitude of distributions to be positively and negatively correlated with the mean and the variance along the elevation gradient, respectively (Supplementary Fig.~XX).
+In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and the variance, respectively (i.e.
+$\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~XX).
+This implies that, at higher elevations, species' distributions generally have lower amplitudes.
 
@@
 A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX).
-Studying the resulting posterior distributions, we found the parameter controlling for the kurtosis to be centred around $\nu_i \sim 2$, which corresponds to a distribution with longer tails than a Gaussian (Fig.~\ref{fig:kurtosis-skewness}).
-However, such parameter estimates displayed a lot of variation across species, which might indicate that the shape of the tails of the distribution is species-specific (Supplementary Fig.~XX).
-Similarly, using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species distributions.
-Based on the estimates for the WAIC values, this model outperformed the previous models (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
-Perhaps most importantly, studying the mean value of the skewness across species (i.e.~parameter $\hat{\lambda}$), we found that species' distributions generally present steeper declines towards higher elevations (Fig.~\ref{fig:kurtosis-skewness}).
+Studying the resulting posterior distributions, we found the average kurtosis of the distributions to be slightly greater than zero, which corresponds to distributions with longer tails than the Gaussian case (Fig.~\ref{fig:kurtosis-skewness}).
+However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~XX), which might indicate that the shape of the tails is species-specific.
 
@@
 
-Finally, we used a model that allowed for heavy-tails and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}) to understand the relationship between the kurtosis and skewness of distributions.
-This model outperformed the rest, presenting Akaike weights close to 1 when comparing models (Supplementary Fig.~XX).
-Comparing the posterior distribution for parameters $\nu_i$ and $\lambda_i$, we revealed the general constrains on the distribution of alpine plants studied.
-That is, we uncovered the relationship between the mean, variance, amplitude, kurtosis and skewness of empirical distribution of plants along an elevational gradient (Figs.~XX and YY).
-Moreover, different types of species seem to present characteristically different distributions (Fig.~XX).
-For example, invasive species...
-Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area (bootstrapping...
-Supplementary Fig.~XX).
-Likewise, the posterior distributions obtained using binary data were strikingly similar to those obtained using categorical data.
-This is important because...
-is model identified similar relationships between parameters as the baseline model (Supplementary Fig.~XX), and measured similar average levels of kurtosis and skewness than the the other two models (Fig.~\ref{fig:kurtosis-skewness}).
+Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species' distributions.
+Based on the estimates for the WAIC values, this model outperformed the first two (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Perhaps most importantly, studying the mean value of the skewness across species, we found this to be consistently below zero (Fig.~\ref{fig:kurtosis-skewness}).
+This indicates that species' distributions generally present steeper declines towards higher elevations (i.e.
+$\hat{\lambda}<0$; Fig.~\ref{fig:response}).
+The same was true when using a model that allowed for both fat-tailed and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}).
+This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~XX), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}).
 
+The model characterizing fat-tailed and skewed distributions allowed us to study the posterior distributions for the parameters describing the mean, variance, amplitude, kurtosis and skewness of species realized niches altogether.
+We observed that different type of species seem to present characteristically different distributions (Fig.~XX).
+For example, focussing on the aforementioned negative correlation between the mean and variance of species' distributions, we found neophyte and archaeophyte species to escape this apparent macroecological constraint (Fig.~XX).
+Similarly, neophytes' distributions also appear to showcase steeper declines towards lower elevations, which is in contrast to the general pattern found across species.
+Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area, as the same model produced comparable results when using simulated and bootstrapped data (Supplementary Information and Supplementary Fig.~XX).
+The comparison across other types of species also provided us with additional interesting mappings.
+For example,...
+Likewise, other pairwise comparisons across parameters also revealed additional macroecological constrains (Supplementary Fig.~XX).
+
+%The same was not true when clustering species based on their changes in abundance over the years (Fig.~XX). 
 %, and it does not seem to depend on species' type or reflect species' abundance change tendency over the years (Fig.~\ref{fig:universality}). T
@@
 Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
-Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).
-{\color{red} I might move this to the Supplementary Information.}}
+Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
 	   %7.5-4
@@
 % Next we studied the skewness of distributions, outpreforming the other models.
-When studying the parameter, we found that the mean skewness of species distributions declined...
+When studying the parameter, we found that the mean skewness of species' distributions declined...
 Hypothesis!
@@
 \section*{Discussion}
-%In this work, we used non-linear response curves to model the distribution of species across an environmental gradient.
-First, we used a baseline model that considered species to present bell-shaped distributions.
-Studying the parameter estimates across species, we were able to showcase how species follow the Rapopor's rule along an elevation gradient.
-Second, we used several...
+In this work, we used non-linear response curves to model the distribution of species across an environmental gradient.
+First, we used a baseline model that considered these as bell-shaped, and we studied the relationship between the basic parameters characterizing them.
+We found both the amplitude and variance of distributions to be negatively correlated with elevation.
+Considering more complex response curves, we then found species' distributions to present non-normal tails and skewed shapes.
+That is, species' distributions were generally characterized by fat tails and steeper declines towards higher elevations.
+That said, the nature of these distributions was not homogeneous across species, as some species presented singularly different properties.
+This is the case of neophyte, which present distributions ...
+%Finally, ...
+
+The statistical framework used here allowed us to parsimoniously compare the shape of the species' realized niches along an altitude gradient, testing for the existence of several interesting macroecological patterns.
+We first studied the relationship between the mean and variance of species' distributions.
+The Rapoport's rule predicts wider ranges of species at higher latitudes and altitudes \citep{stevensElevationalGradientAltitudinal1992}; and therefore, one might expect a positive correlation between these properties.
+A common explanation for the Rapoport's rule is that climatic variability selects for species with greater climatic tolerances.
+However, while this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}, contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
+Our results seem to contradict the predictions of the Rapoport's rule, as we observed a negative correlation between species' range and elevation.
+
+Similarly, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis \citep{Loutlain 2015 TREE}.
+\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
+Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
+Specifically, we are able to link plant physiological traits to the skewness of their distributions.
+
 Structure of the discussion section:
@@
 Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient.
-Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref).
+% Why previous tests fail: geographic vs environmental; wrong models; latitude vs altitude
+
+Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis \citep{Loutlain 2015 TREE}.
 \citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
@@
 
+There is a lot of variation that remains unexplained regarding the shape of species' distribution, and Fig... gives us glimpse of it. 
 %%%%%%%%%%%%%%%%%%
@@
 \clearpage
-\bibliography{references2}
+\bibliography{references}
 

### 2021-05-13 code/paper-models/models-categorical.R

@@ -1,3 +1,68 @@
+categorical.line.multithread <- "
+functions{
+    real partial_sum(int[] y_slice,
+                   int start, int end,
+                   int N, int M, vector beta, vector phi, int[ , ] Y, row_vector X1) {
+        real lp = 0.0;
+
+        for( i in start:end){
+            for( j in 1:N ){
+                lp += ordered_logistic_lpmf( Y[i, j] | beta[i] * X1[j], phi);
+            }
+        }
+        return lp;
+    }
+}
+data{
+    int N;
+    int L;
+    int M;
+    // int K;
+    // int Y[K];
+    int Y[L, N];
+    int indices[L];
+    row_vector[N] X1;
+}
+parameters{
+    ordered[M] zphi;
+    vector[L] zbeta;
+    real beta_bar;
+    real<lower=0> sigma_b;
+}
+transformed parameters{
+    vector[L] beta;
+    vector<lower=0>[M] phi;
+    beta = zbeta * sigma_b + beta_bar;
+}
+model{
+    // matrix[L,N] p;
+
+    sigma_b ~ exponential( 1 );
+    beta_bar ~ std_normal();
+    zbeta ~ std_normal();
+    zphi ~ std_normal();
+    
+    int grainsize = 1;
+
+    target += reduce_sum(partial_sum, indices,
+                     grainsize,
+                     N, M, beta, phi, Y, X1);
+}
+generated quantities{
+    vector[L*N] log_lik;
+    int k;
+
+    k = 1;
+    for ( i in 1:L ){
+        for (j in 1:N){
+           log_lik[k] = ordered_logistic_lpmf( Y[i, j] | beta[i] * X1[j], phi);
+           k = k + 1;
+        }
+    }
+}
+"
+
+
 # Binomial 1d
 categorical.model.1d <- "
 functions{
@@ -91,6 +156,148 @@ generated quantities{
     for ( i in 1:L ){
         for (j in 1:N){
            log_lik[k] = binomial_lpmf(Y[i, j] | 1, exp(-alpha[i] - gamma[i] * pow(X1[j] - beta[i],2)) + minp);
+           k = k + 1;
+        }
+    }
+}
+"
+
+categorical.baseline.multithread <- "
+functions{
+    matrix cov_GPL2(matrix x, real a, real b, real delta) {
+        int N = dims(x)[1];
+        matrix[N, N] K;
+
+        for (i in 1:(N-1)) {
[…]

### 2021-05-13 code/paper-models/plots/plot-real.R

@@ -7,6 +7,7 @@ library(ggpubr)
 library(hrbrthemes)
 library(plotly)
 library(cowplot)
+library(dplyr)
 
 ####
 # Visualizing the results of some of the models
@@ -248,6 +249,12 @@ compare.models <- function(){
 
   kurtosis.median.m4 <- kurtosis.skew.generror(post.m4$nu, post.m4$lambda)
   nu_bar.m4 <- apply(kurtosis.median.m4, 1, mean)
+  
+  # kurtosis.median.m3 <- post.m3$nu
+  # nu_bar.m3 <- apply(kurtosis.median.m3, 1, median)
+  # 
+  # kurtosis.median.m4 <- post.m4$nu
+  # nu_bar.m4 <- apply(kurtosis.median.m4, 1, median)
 
   # nu_bar.m3 <- apply(post.m3$nu, 1, median)
   # nu_bar.m4 <- apply(post.m4$nu, 1, median)
@@ -293,9 +300,15 @@ compare.models <- function(){
 
   skewness.mean.m1 <- skewness.skew(post.m1$lambda)
   lambda_bar.m1 <- apply(skewness.mean.m1, 1, mean)
-  
+
   skewness.mean.m4 <- skewness.skew.generror(post.m4$nu, post.m4$lambda)
   lambda_bar.m4 <- apply(skewness.mean.m4, 1, mean)
+  
+  # skewness.mean.m1 <- post.m1$lambda
+  # lambda_bar.m1 <- apply(skewness.mean.m1, 1, median)
+  # 
+  # skewness.mean.m4 <- post.m4$lambda
+  # lambda_bar.m4 <- apply(skewness.mean.m4, 1, median)
   
   # nu_bar.m3 <- apply(post.m3$nu, 1, median)
   # nu_bar.m4 <- apply(post.m4$nu, 1, median)
@@ -343,11 +356,16 @@ compare.models <- function(){
 
 distributions.hyperparameters <- function(model=NULL){
   if(is.null(model)){
-    model <- readRDS("../../../results/models/skew-traits-min20-1d.rds")
+    model <- readRDS("../../../results/models/skew-variation-min20-1d.rds")
   }
+  # model <- readRDS("../../../results/models/skew-generror-min20-1d.rds")
   
-  post <- extract.samples(model, n = 1000, pars=c("etasq_l", "sigma_l"))
-
+  
+  # post <- extract.samples(model, n = 1000, pars=c("etasq_l", "sigma_l"))
[…]

### 2021-05-13 code/paper-models/prepare-data.R

@@ -36,6 +36,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
         neophytes <- read.table("../../data/properties/codes/neophytes-list.csv", sep=",")
         tendency <- read.table("../../data/properties/codes/change-tendency.csv", sep=",")
         competitive <- read.table("../../data/properties/codes/competitive_indicator.csv", sep=",")
+        detailed <- read.table("../../data/properties/codes/neophytes-detailed.csv", sep=",")
         
         # Check that there aren't unnexpected files
         if(!all(sort(files)==1:length(files))){
@@ -50,6 +51,7 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
         NEO <- c()
         Tend <- c()
         compet <- c()
+        deta <- c()
         
         # Read observations
         for(idx in 1:length(files)){
@@ -64,6 +66,17 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
              NEO <- rbind(NEO, c(kdx, new.name, neophytes$neo[new.name==neophytes$names]))
              Tend <- rbind(Tend, c(kdx, new.name, tendency$decrease[new.name==tendency$names], tendency$decrease.low[new.name==tendency$names], tendency$increase[new.name==tendency$names], tendency$other[new.name==tendency$names]))
              compet <- rbind(compet, c(kdx, new.name, as.character(competitive$nflor.KS[new.name==competitive$nflor.spnames])))
+             deta <- rbind(deta, c(kdx, new.name, detailed$i[new.name==detailed$names],
+                                   detailed$n[new.name==detailed$names],
+                                   detailed$nw[new.name==detailed$names],
+                                   detailed$a[new.name==detailed$names],
+                                   detailed$ja[new.name==detailed$names], 
+                                   detailed$jn[new.name==detailed$names],
+                                   detailed$isn[new.name==detailed$names],
+                                   detailed$isa[new.name==detailed$names],
+                                   detailed$isj[new.name==detailed$names],
+                                   detailed$isasn[new.name==detailed$names],
+                                   detailed$asn[new.name==detailed$names]))
              
              name.idx <- c(name.idx,correlation_matrix_ids$V1[as.character(correlation_matrix_ids$V2)==new.name])
              ###
[…]

### 2021-05-13 code/paper-models/1d-categorical.R
Added two more categorical

@@ -7,6 +7,243 @@ library(cmdstanr)
 options(mc.cores = parallel::detectCores())
 rstan_options(auto_write = TRUE)
 
+categorical.line.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+  # Fixing some of the options
+  variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
+  ndim <- 2
+  pca <- T
+  
+  # If we are dealing with simulated data
+  if(simulated){
+    extension <- "1d-categorical-line"
+  }else{
+    extension <- "1d-categorical-line"
+  }
+  
+  if(min.occurrence==10){
+    extension2 <- "-"
+  }else{
+    extension2 <- paste("min",min.occurrence,"-",sep="")
+  }
+  
+  # Load the data
+  if(recompile){
+    d <- species_distribution.data(variables=variables, pca=pca, ndim = ndim, simulated=simulated, simulated.type="categorical", min.occurrence=min.occurrence)
+    # rename variables
+    if(simulated){
+      variables <- c("S1", "S2")
+    }else{
+      if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+      }
+    }
+    filename <- paste("../../data/processed/jsdm/", extension, "-",paste(variables, collapse = ""), extension2, "data.rds", sep = "")
+    
+    if (file.exists(filename)){
+      question <- askYesNo("Do you want to overwrite the file?", default = F, 
+                           prompts = getOption("askYesNo", gettext(c("Yes", "No", "Cancel"))))
+      if(is.na(question)){question <- F}
+      if(question){
+        saveRDS(d, file=filename)
+      }else{
+        stop("you should add 'recompile=T'")
+      }                                
+    }else{
+      saveRDS(d, file=filename)
+    }
+  }else{
+    # rename variables
+    if(simulated){
+      variables <- c("S1", "S2")
+    }else{
+      if(pca){
+        variables <- paste("PC", 1:ndim, sep="")
+      }
+    }
+    if(is.null(d)){
+      d <- readRDS(file = paste("../../data/processed/jsdm/", extension,"-", paste(variables, collapse = ""), extension2, "data.rds", sep = ""))
+    }
+  }
+  
[…]

### 2021-05-13 code/paper-models/run_inte.sh

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=2840]" -W 200:00 -n 45 "R --vanilla --slave < 1d-baseline.R"
+bsub -R "rusage[mem=3500]" -W 200:00 -n 35 "R --vanilla --slave < 1d-categorical.R"

### 2021-05-31 manuscript/manuscript.tex
New version of the draft

@@
 \author[1,*]{\normalsize Bernat Bramon Mora}
+\author[2,3]{\normalsize Antoine Guisan}
 \author[1]{\normalsize Jake M.\ Alexander}
 \affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
+\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
+\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
+\affil[3]{\footnotesize Institute of Earth Surface Dynamics, University of Lausanne, Lausanne, Switzerland}
 \affil[*]{\footnotesize  bernat.bramon@gmail.com}
@@
 With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space.
-In particular, we studied the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients.
+In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients.
 Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts.
@@
 One of the central goals of ecology is to understand the ways species are distributed across space and time.
-While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, abundant-centre-hypothesis}.
+While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, sagarinMovingAssumptionsUnderstand2006}.
 In practice, there is a strong argument to be made in favour of assuming these to be bell shaped.
@@
 %Moreover, properties such as the skewness of species' distributions might capture other underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}.
-Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, bernat}.
-For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992, bernat}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients.
-This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments.
-That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}.
+Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, channellDynamicBiogeographyConservation2000}.
+For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients.
+This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}.
+That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, helmuthBIOPHYSICSPHYSIOLOGICALECOLOGY2004, siefertHowClimateDispersal2015}.
 Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve.
@@
 Such frameworks, however, commonly assume an underlying linear relationship between covariates (but see `semiparametric models'; \citealt{norbergComprehensiveEvaluationPredictive2019}).
-This is useful because it simplifies the optimization process but might not be ideal when studying and comparing the shape of species' distributions along environmental gradients.
+This is useful because it simplifies the optimization process, but it might not be ideal when studying and comparing the shape of species' distributions along environmental gradients.
 First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified.
@@
 That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
-Finally, \citet{huismanHierarchicalSetModels1993} proposed several non-linear structures to characterize several features of individual species' response curves.
-Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following these model structures, these were not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g.
+Finally, some have proposed several non-linear structures to characterize several features of individual species' response curves \citep{huismanHierarchicalSetModels1993}.
+Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g.
 sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).
 
-In this work, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
+In this work, we rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
 In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies.
@@
 These values are provided for a range of 10 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-On the other hand, the information regarding species' ecological strategies represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify more than $120$ binary traits that characterize the physiology and life-history of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-Finally, and in addition to species' environmental preferences and physiological traits, the floristic data also contains information on species introduction status (e.g.~identifying those species that are neophytes or archaeophytes) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
+In addition to species' environmental preferences, the floristic data also contains information on species introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
 We describe this information in more detail in the Supplementary Table 1.
 
-%we independently consider those traits that describe growth and life strategies of the plants---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
-In total, we identify $\sim120$ binary traits that characterize the physiology of species.
-
-%comprising a total of 10 environmental indicator values with 10 types of associations for each of them (i.e.
-the $r$ associations defined above).
-Finally, we additionally independently consider species' range of variations for the such indicator values, also defining 10 `traits' with 3 types of associations (I need to see if I can added these to the analyses above).
-
-%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19.
-Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland.
-There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
-%Based on 30 years (1961–1990) of meteorological data from national weather stations, fourteen climatic variables (Table S5) were calculated according to the method described in Zimmermann & Kienast44 and Zimmerman et al.55.
-A digital elevation model (DEM) was used to spatially interpolate the climatic data and to create topographic variables (Table S5).
-Additionally, we used a soil pH map based on Buri et al.43.
-All environmental data had a spatial resolution of 25 m, a resolution similar to the areas used to calculate the ecological indicator values (EIV; see below).
-This allowed a straightforward comparison of different sets of predictors composed from environmental variables and EIVs.
-
-%In this study we used the ecological indicator values (EIVs) first developed by Landolt17 and later extended by Landolt et al.19.
-Landolt’s EIVs are largely similar to the more renown ones proposed by Ellenberg et al.18, but specifically adapted to the flora and environmental conditions of Switzerland.
-There are a large number of EIVs (for details see Landolt et al.19), but in this study we focused on six of the most commonly used, namely: T for temperature;, M for soil moisture/water availability;, L for light;, K for continentality;, R for soil pH; and N for soil nutrients.
+%On the other hand, the information regarding species' ecological strategies represents general descriptions of species' growth and life strategies---examples include their growth forms, nature of the storage organs, dispersal ability and pollinator agents.
+In total, we identify more than $120$ binary traits that characterize the physiology and life-history of species (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
+Finally, and i
 
@@
 That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
 	   \caption{Different response curves.
-Panel (a) shows the probability density function characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\gamma$, when $\mu=0$.
-Panel (b) shows the probability density function characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\nu$, when $\gamma=1$ and $\mu=0$.
-Panel (c) shows the probability density function characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\lambda$, when $\gamma=1$ and $\mu=0$.
-Panel (d) shows the probability density function characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\lambda$ and $\nu$, when $\gamma=1$ and $\mu=0$.
-Notice that for each case, we chose $\alpha$ values that normalize the probability distributions.}
+Panel (a) shows the distribution shapes characterized by Eq.~(\ref{eq:baseline-response}) for different values of $\beta$, $\alpha$ and $\gamma$.
+Panel (b) shows the distribution shapes characterized by Eq.~(\ref{eq:generalized-response}) for different values of $\beta$, $\alpha$ and $\nu$, when $\gamma=1$.
+Panel (c) shows the distribution shapes characterized by Eq.~(\ref{eq:skewed-response}) for different values of $\beta$, $\alpha$ and $\lambda$, when $\gamma=1$.
+Panel (d) shows the distribution shapes characterized by Eq.~(\ref{eq:heavy-skewed-response}) for different values of $\beta$, $\alpha$, $\lambda$ and $\nu$, when $\gamma=1$.
+Notice that for all panels, we chose $\alpha$ values such that $\text{p}_i = \exp(-\alpha_i)$.
+	   }
       \label{fig:response}
@@
 where $\Sigma_{ij}$ characterizes the covariance between any pair of species $i$ and $j$, and $\delta_{ij}$ is the Kronecker delta.
-Notice that such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
+Such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
 In the floristic database, this information is represented by the set of ordinal traits specified for the different species.
@@
 For all these hyperparameters, we choose weakly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.
-
-\subsubsection*{Alternative variance-covariance structures}
-The model structure defined above allows us to test how different sources of information characterize each of the different parameters.
-Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}).
-For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits.
-One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
-\begin{equation} 
-\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
-\label{eq:covariance-complex}
-\end{equation}
-where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.
+Notice that other structures can be used to define the covariance matrices of the different Gaussian processes \citep{mcelreathStatisticalRethinkingBayesian2020}, including structures that account for multiple distance matrix $D_{ij}$ for any given parameter.
 
@@
 \subsection*{Modifying the baseline model}
-We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system.
+%We proposed a baseline model that is naive regarding how the data is distributed, and yet accounts for all prior information that we have about the system.
 Now, we want to modify this model to test the extent to which empirical species' distributions showcase different properties, while preserving both the interpretation of the parameter estimates and prior information.
@@
 
-One way to test the extent to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes.
+%One way to test the extent to which species' physiological traits inform the kurtosis or skewness of the distribution of any species $i$ is by characterizing $\nu_{i}$ or $\lambda_{i}$ as Gaussian processes.
 That is, we can consider the prior distribution for these parameters as multivariate normal with a variance covariance matrix $\Sigma$.
@@
 Likewise, we could instead use Eq.~(\ref{eq:covariance-complex}) and test the effectiveness of other prior information in explaining the skewness of species' distributions.
+
+\subsection*{Evaluating the log-likelihood}
+One way to understand the shape of distributions is by evaluating the computed log-likelihood values across the environmental gradient.
+Comparing these values, one can understand what aspects of the shape of distributions are missing.
+To do so, for every sample of the model, we computed the log-likelihood values and the normalized probability distribution.
+This normalized probability is defined such that its maximum is set to 1 for all species in our dataset.
+In particular, for a heavy-tailed and skewed response, the normalized probability distribution was calculated for every sample of the Bayesian model using Eq.~(\ref{eq:heavy-skewed-response}), where $\alpha_i$ was set to 0 for any value of $x_{j}$.
+Notice that the normalized probability distribution is interesting when comparing the log-likelihood values across species because it can be used to understand whether the model errors are at the tails of the distributions or their center.
 
@@
 This means that species found at lower elevations have generally wider distributions than those at higher elevations.
-The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~XX) as well as when using ordinal data (Supplementary Fig.~XX); however, the pattern was not present along the second axis of variation of the environmental variables (Supplementary Fig.~XX).
+The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~XX) as well as when using ordinal data (Supplementary Fig.~XX); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~XX).
 The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.
-In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and the variance, respectively (i.e.
+In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e.
 $\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~XX).
@@
     	  \vspace{0.1cm}
-	   \caption{Relationship between mean and variance of species' distributions.
-Posterior distributions for parameters $\beta_i$ and $\gamma_i$ from Eq.~(\ref{eq:baseline-response}) across species, and the relationship between them.
-Panel (a) describes the relationship between range size and elevation.
-Every dot represents the relationship between the mean values for the $\beta_i$ and $\gamma_i$ estimates of the different species.
-The value in the top-right corner of the plot displays the Pearson's correlation between these parameters calculated across samples of the posterior distributions.
-Panel (b) describes the $\beta_i$ posterior distribution estimated for all species.
-Panel (c) describes the $\gamma_i$ posterior distribution estimated for all species.
+	   \caption{Relationship between the posterior distributions for parameters $\beta_i$ and $\gamma_i$ from Eq.~(\ref{eq:baseline-response}) across species.
+Panel (a) describes the relationship between the mean ($\beta_i$) and variance ($\propto\gamma_i^{-1}$) of distributions.
+Every dot characterizes the average value of the corresponding posterior distributions for any given species.
+The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.
+Panel (b) displays the estimates for the center of species' distributions along the environmental gradient.
+Panel (c) displays the estimates for the variance of species' distributions along the environmental gradient.
 In (b) and (c), the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.}
@@
 	   \caption{Average kurtosis and skewness of species' distributions.
-Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions, respectively.
-Panel (a) displays the results obtained by using a response curve that follows a generalized error distribution.
-Panel (b) displays the results obtained by using a response curve that follows a skewed normal distribution.
-In both case, the red dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
+Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions.
+Panel (a) displays the results obtained using response curves that follow a generalized error distribution (black line) and a skewed generalized error distribution (yellow line).
+Panel (b) displays the results obtained using response curves that follow a skewed normal distribution (black line) and a skewed generalized error distribution (yellow line).
+In both cases, the gray dotted line indicates the conditions by which species are normally distributed along the environmental axis.}
       \label{fig:kurtosis-skewness}
@@
 This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~XX), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}).
+A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note XX.
+%Comparing the prior and posterior distributions for the hyperparameters in the.
 
 The model characterizing fat-tailed and skewed distributions allowed us to study the posterior distributions for the parameters describing the mean, variance, amplitude, kurtosis and skewness of species realized niches altogether.
-We observed that different type of species seem to present characteristically different distributions (Fig.~XX).
-For example, focussing on the aforementioned negative correlation between the mean and variance of species' distributions, we found neophyte and archaeophyte species to escape this apparent macroecological constraint (Fig.~XX).
-Similarly, neophytes' distributions also appear to showcase steeper declines towards lower elevations, which is in contrast to the general pattern found across species.
-Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area, as the same model produced comparable results when using simulated and bootstrapped data (Supplementary Information and Supplementary Fig.~XX).
-The comparison across other types of species also provided us with additional interesting mappings.
+We observed that different types of species seem to present characteristically different distributions (Fig.~\ref{fig:pairwise}).
+Focussing on the negative correlation between the mean and variance of species' distributions, we found some species to escape some of the aforementioned macroecological constraints (Supplementary Fig.~XX).
+Moreover, recent and historical range expanders are often found at lower altitudes, presenting higher amplitudes and distributions that appear to showcase steeper declines towards lower elevations.
+Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area, as the same model produced comparable results when using simulated and bootstrapped data (Supplementary Note XX and Supplementary Fig.~XX).
+Moreover, these results did not substantially change when using ordinal data (Supplementary Fig.~XX).
+
+%The comparison across other types of species also provided us with additional interesting mappings.
 For example,...
@@
   \centering
-    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
+    \includegraphics[width=0.9\textwidth]{figures/figure3}
     	  \vspace{0.1cm}
-	   \caption{Universality of the relationship between mean and variance of species' distributions.
-Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a.
-Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
-Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
-Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
-	   %7.5-4
-      \label{fig:universality}
+	   \caption{Comparing the distributions of different types of species.
+Focussing on the differences between indigenous, historical range expanding and recent range expanding species, the panels describe the relationship between the basic properties of their distributions.
+Panels (a-d) characterize the relationship between the mean, and the variance, amplitude, skewness and kurtosis of the species' distributions (Supplementary Table XX).
+The points in every panel are calculated as the average value across all samples of the model.
+%The analogous results found for species classified accroding to their population change tendency over the years is presented in Supplementary Fig~XX.
+The ellipses characterize the 1 standard deviation ellipses for each species type.
+	   }
+	   %7-4.5
+      \label{fig:pairwise}
+\end{figure}
+
+Finally, we wanted to identify what aspects of the shape of distributions we were still missing.
+We used the computed log-likelihood values and normalized probabilities to understand where our best performing model fails to capture the variation in empirical plant distributions.
+We found most data points to be located at the tails of distributions ($\text{normalized probability}\,\approx\,0$) and to present high log-likelihood values (Supplementary Fig.~XX).
+This is not surprising as the study area spans an extensive altitude gradient, and species' distributions are generally narrow relative to it; therefore, the model accurately predicts that species are usually absent in those sampling sites that fall relatively far from the center of their distributions.
+However, studying instead only those points for which the model did not perform well (with a $\text{likelihood}\leq 0.5$), we found these to generally be associated with high normalized probabilities (Fig.~\ref{fig:loglik}).
+This indicates that the unexplained variation is often found at the center of species' distributions.
+Similar results were found when using ordinal data (Supplementary Fig.~XX).
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figures/loglik-notshow-bin}
+    	  \vspace{0.1cm}
+	   \caption{Studying the distribution of log-likelihood values.
+The graph maps the log-likelihood and normalized probability values for all species across all samples.
+Notice that in this figure there are only displayed those points that present a likelihood smaller than $0.5$.
+The mapping of all log-likelihood values is presented in Supplementary Fig.~XX.}
+	   %5-4
+      \label{fig:loglik}
 \end{figure}
@@
 We found both the amplitude and variance of distributions to be negatively correlated with elevation.
-Considering more complex response curves, we then found species' distributions to present non-normal tails and skewed shapes.
-That is, species' distributions were generally characterized by fat tails and steeper declines towards higher elevations.
-That said, the nature of these distributions was not homogeneous across species, as some species presented singularly different properties.
-This is the case of neophyte, which present distributions ...
-%Finally, ...
+Considering more complex response curves, we then found species' distributions to also present non-normal tails and skewed shapes.
+Specifically, we found species' distributions to generally be characterized by fat tails and steeper declines towards higher elevations.
+That said, the nature of these distributions was not homogeneous across species, as some of them presented singularly different properties.
+This is the case of rapid and historical range expanders, often found in warmer environments, with distributions presenting higher amplitudes and skewed responses towards high altitudes.
+Finally, we studied the variation that remained unexplained by the best performing model.
+We found this unexplained variation to be generally located at the center of distributions, which identified potential general properties of empirical distributions that were missed by our model.
+Putting this all together, our results uncovered several aspects of the shape of empirical plant distributions and revealed crucial differences between the way species are assembled along environmental gradients.
 
-The statistical framework used here allowed us to parsimoniously compare the shape of the species' realized niches along an altitude gradient, testing for the existence of several interesting macroecological patterns.
-We first studied the relationship between the mean and variance of species' distributions.
-The Rapoport's rule predicts wider ranges of species at higher latitudes and altitudes \citep{stevensElevationalGradientAltitudinal1992}; and therefore, one might expect a positive correlation between these properties.
+Our approach allowed us to parsimoniously compare the shape of the species' realized niches along an altitude gradient, testing for the existence of several macroecological patterns.
+For example, the Rapoport's rule predicts wider ranges of species at higher latitudes and altitudes \citep{stevensElevationalGradientAltitudinal1992}; and therefore, one might expect a positive correlation between the mean and variance of species distributions.
 A common explanation for the Rapoport's rule is that climatic variability selects for species with greater climatic tolerances.
-However, while this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}, contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
+But while this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}, contrasting evidence suggests that this rule is not pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
 Our results seem to contradict the predictions of the Rapoport's rule, as we observed a negative correlation between species' range and elevation.
+Moreover, other properties of species' distributions---such as their amplitude---were also significantly correlated with species' ranges.
+This is interesting because it hints at the existence of some general macroecological constraints that dictate the way different species assemble across environments.
+That said, our results also suggest that species such as neophytes and archeophytes might not obey this same constraints, as these were singularly positioned along the gradient (Supplementary Fig.~XX).
 
-Similarly, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis \citep{Loutlain 2015 TREE}.
-\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
-Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
-Specifically, we are able to link plant physiological traits to the skewness of their distributions.
+The level of skewness of species' distribution as well as the variability in the shape of theirs tails diverged from traditionally assumed bell-shaped curves.
+This allowed us to focus on other interesting macroecological hypotheses.
+For instance, the so-called abiotic stress limitation hypothesis predicts species' distributions to present steeper declines towards stressful conditions \citep{austinCommunityTheoryCompetition1990}.
+\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis (but see \citealt{ziffer-bergerSpatialPatternsProvide2014}).
+Our results, however, showcased species' distributions to generally present steeper declines towards higher elevations, providing clear evidence of this geographical pattern.
+Moreover, we were able to highlight the degree to which different species might present different levels of decline towards stressful conditions, as plants found at low elevations---such as recent and historical range expanding species---displayed contrasting levels of skewness.
+This is important because it could provide glimpses of the different stages of species' assembly processes, with range expanders' distributions trending towards higher elevations.
 
-Structure of the discussion section:
-\begin{enumerate}
-\item Summary of results.
-\item Proper test of Rapopor's hypothesis. Different species follow different biogeographical patterns.
-\item Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
-\item What is the true shape of species' distributions? These display heavy-tail and skewed distributions. 
-\item Future directions.
-Missing bimodal curves.
+There are many other properties characterizing empirical distributions that might not have been captured by the different models.
+One possible way to untangle these properties is by studying the unexplained variation in the empirical data.
+We observed that this variation is often located at the center of distributions, which suggests that the aspects of their shape not picked up by the models involve those points at the peak of the distributions.
+This observation is directly linked to another macroecological pattern: the so called abundant-center hypothesis \citep{sagarinAbundantCentreDistribution2002}.
+This hypothesis predicts species to be most abundant at the center of their distributions, and it is an implicit assumption at the core of most modelling approaches.
+Namely, if one is only willing to assume that species have finite geographic ranges, the abundant-center hypothesis is a consequence our state of ignorance (i.e.~the maximum entropy distribution).
+That said, several studies have pointed out that the abundant-center hypothesis is not pervasive in empirical distributions \citep{wagnerSimilarPerformanceCentral2011, pirononGeographicVariationGenetic2017, dallasSpeciesAreNot2017}, suggesting that population abundance could often be more strongly driven by interactions and community structure than the environment \citep{dallasSpeciesAreNot2017}.
+Our results, for both binary and ordinal data, support these observations, suggesting that the species' probability of appearance---as well as likelihood of presenting high abundance at a given site---might not ubiquitously be highest at the center of their distributions.
+Allowing species to showcase other distribution shapes, such as those including multimodal or plateau peaks, could potentially resolve some of the unexplained variation.
+Indeed, studying the tails of species' distributions, we observed several species presenting low kurtosis levels.
+While this implied that these distributions had shorter tails than normal, it also reflected plateau-shaped response curves (e.g.~$\nu>2$ in Fig.~\ref{fig:response}b).
+
+The different hypotheses regarding the shape of species' distributions address central topics in ecology and evolution \citep{sagarinAbundantCentreDistribution2002}.
+These distributions are the result of environmental variability \citep{helmuthClimateChangeLatitudinal2002, butterfieldEnvironmentalFilteringIncreases2015}, biotic interactions \citep{hastingsUnexpectedSpatialPatterns1997} and historical contingencies \citep{frickEmergingDiseaseCauses2010}, and their shape determines gene flow \citep{haldaneRelationDensityRegulation1956, lesicaWhenArePeripheral1995, pirononGeographicVariationGenetic2017} and energy balances along gradients \citep{hallDistributionAbundanceOrganisms1992}.
+Perhaps most importantly, the shape of species' distributions will influence their responses to environmental changes \citep{channellDynamicBiogeographyConservation2000}, and it could therefore be used as an ecological compass to inform conservation and management decisions \citep{channellTrajectoriesExtinctionSpatial2000}.
+In this context, we identify two areas we feel represent key steps from which to move forward.
+First, trait data could crucially inform the different parameters controlling the shape of distributions.
+For example, if the skewness of species' distributions is the result of uneven environmental tolerances along the gradient \citep{sundayGlobalAnalysisThermal2011}, this information should be accounted for analogously to the way we used the expert knowledge on plants' environmental preferences.
+The same is true for species' ecological strategies, with aspects regarding their competitive ability potentially informing the shape of distributions.
+Second, from a performance standpoint, the models presented here will likely do a worse job at predicting species' occurrences than some of the distribution models developed over the recent years \citep{norbergComprehensiveEvaluationPredictive2019}, including those accounting for spatial autocorrelation \citep{ovaskainenUncoveringHiddenSpatial2016}, associations between species \citep{tikhonovJointSpeciesDistribution2020}, and some non-parametric approximations \citep{harrisGeneratingRealisticAssemblages2015}.
+However, our models have clear interpretable parameters, and can be used to directly compare the shape of species' realized niches.
+These comparisons could be used to generate hypotheses regarding where and when different species might strongly interact with one another along an environmental gradient \citep{louthanWhereWhenSpecies2015}, making ecologically-informed predictions regarding the presence and absence of these relationships \citep{SIASH}.
+
+%A consequence of Ecological interactions (Hochberg Ives 1998), energy balances along gradients (Hall Standford Hauer 1992), 
+% From article: These hypotheses directly address many fundamental issues in ecology and evolution, such as how genes flow between populations, as well as applied ecological issues, such as how populations will respond to climate change and what populations should be the focus oflimited conservation resources.
+If species are mostabundant at the edge of their range, genes and individuals may flow from the edges to the centre, leading to the opposite conclusions of abundant centre-based hypotheses.
+Alternatively, if species were most abundant at only one edge of the range (e.g.
+species became gradually more abundant toward the south edge of the range), the direction and magnitude of gene swamping or population outbreaks could not be predicted without identifying specific study locations.
+\citep{sagarinAbundantCentreDistribution2002}.
+% From article: The abundant centre distribution, if present, also simplifes many hypo-theses about population dynamics, conservation biology,and factors that set species' ranges \citep{sagarinAbundantCentreDistribution2002}.
+% From article: This theoretical work relates to some of the most fundamental and practical issues in ecology, including questions of where to situate nature reserves, where pest outbreaks are expected, how species will respond to climate change, and the genetic structure of population \citep{sagarinAbundantCentreDistribution2002}.
+%From article: Understanding how interspeci-ﬁc interactions, natural enemies, environmental forces anddispersal barriers inﬂuence the existence of distance–abundancerelationship remains an open question; one, when answered,may provide an underlying basis for the emergence of themacroecological pattern \citep{dallasSpeciesAreNot2017}
+
+%Understanding the extend to which this new properties are across species could potentially
+
+%Refs for center-abundance hypo: Hengeveld & Haeck1982; Brown 1984; Holt et al. 1997; McGill & Collins 2003)
+
+%From article: species interactions andcommunity structure may be more important in regulatingpopulation abundance than the environment Second, the spa-tial distribution of abundance, and subsequent distance–abun-dance relationships, may be limited by dispersal boundaries orunmeasured ecological interactions.
+For instance, coasts andmountain ranges represent obvious barriers to species spread.Species abundance may be highest at the barrier (Brown et al.1996), with the putative explanation being directional disper-sal against a barrier, and an environment capable of sustain-ing relatively high species abundance.
+\citep{dallasSpeciesAreNot2017}
+% From article: tandem action of niche and dispersal processes appears to play an important role in constraining this species’ northern latitudinal range \citep{baerDecliningDemographicPerformance2019}.
+
+%Finally, we identify three areas we feel represent key steps from which to move forward.
+First,...
+Is species' physiology informative to explain the pattern?
+Further test of the ability of traits to predict those parameter estimates.
+Second,...
 Using this information to understand where jSDMs estimate interactions between species.
-Further test of the ability of traits to predict those parameter estimates.
-\end{enumerate}
-% Summary of results
-% Proper test of Rapopor's hypothesis. Different species follow different biogeographical patterns.
-% Proper test of skewed towards high altitude. Is species' physiology informative to explain the pattern? 
-% What is the true shape of distributions? Missing bimodal curves
-%
-
-First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}.
-While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
-Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient.
-% Why previous tests fail: geographic vs environmental; wrong models; latitude vs altitude
-
-Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis \citep{Loutlain 2015 TREE}.
-\citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citeauthor{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis.
-Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions.
-Specifically, we are able to link plant physiological traits to the skewness of their distributions.
-
-There is a lot of variation that remains unexplained regarding the shape of species' distribution, and Fig... gives us glimpse of it. 
-%%%%%%%%%%%%%%%%%%
-%%% Model runs
-%%%%%%%%%%%%%%%%%%
-% Finish categorical 1d for both skewed and normal (2 weeks)
-% Test for heavy tails 1 binomial and categorical (0.5 week)
-% Test for heavy tails 2 binomial and categorical (0.5 week)
-% Categorical 1d for heavy tails 1 (1 weeks)
-% Categorical 1d for heavy tails 2 (1 weeks)
-% Binomial 1d for heavy tails 1 (0.7 weeks)
-% Binomial 1d for heavy tails 2 (0.7 weeks)
-% Binomial 1d for skewed (0.7 weeks)
-% Binomial 1d normal two covariances (1 week)
-% Categorical 1d for skewed with physiology in lambda (1 week)
-% Additional tests for Binomial 1d skewed and normal (1 week)
-
-% With both heavy-tails ~ 10 weeks
-% With only one heavy-tail ~ 8 weeks
+Third,...
+understanding the dynamical implications of certain properties...
 
 \clearpage
-\bibliography{references}
+\bibliography{references2}
 

### 2021-05-31 manuscript/supplementary_information.tex
Finished methods section
New version of the draft

@@
 % Full description of Categorical model with multiple environmental covariates
+% Full description of Categorical model with heavy-tails---generalized error distribution
+% Full description of Categorical model with heavy-tails---Student's t-distribution
+% Full description of Categorical model with skewed---skewed normal distribution
 % Using a MMSBM to calculate species similarity
@@
 
+\subsubsection*{Alternative variance-covariance structures}
+The model structure defined above allows us to test how different sources of information characterize each of the different parameters.
+Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}).
+For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits.
+One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
+\begin{equation} 
+\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
+\label{eq:covariance-complex}
+\end{equation}
+where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.
+
 \section*{Supplementary Notes}
@@
 
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
+    	  \vspace{0.1cm}
+	   \caption{Universality of the relationship between mean and variance of species' distributions.
+Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a.
+Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
+Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
+Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
+	   %7.5-4
+      \label{sfig:universality}
+\end{figure}
+
 \clearpage

### 2021-05-31 code/paper-models/plots/other-plots.R

@@ -194,4 +194,276 @@ distribution.type <- function(){
   
 }
 
+distribution.type_2 <- function(){
+  x <- seq(-5, 5, length.out = 2000)
+  generrskew <- function(x, a, mu, sigma, lambda, p, normalized=T){
+    v <- sqrt((pi*gamma(1/p))/(pi*(1+3*lambda**2)*gamma(3/p)-(16**(1/p))*lambda*lambda*(gamma(1/2+1/p)**2)*gamma(1/p)))
+    m <- (2**(2/p))*v*sigma*lambda*gamma(1/2+1/p)/sqrt(pi)
+    if(normalized){
+      return((p*sigma/(2*v*gamma(1/p)))*exp(-(sigma*abs(x-mu+m)/(v*(1+lambda*sign(x-mu+m))))**p))
+    }else{
+      return(exp(-(sigma*abs(x-mu+m)/(v*(1+lambda*sign(x-mu+m))))**p))
+    }
+  }
+  
+  plist <- list()
+  
+  for(i in 1:6){
+    normalization <- rep(1,3)
+    if(i==2){
+      a <- rep(0,3)
+      mu <- c(-1.3,0,1.3)
+      sigma <- c(1,1,1)
+      lambda <- rep(0,3)
+      nu <- rep(2,3)
+      title <- "(b)  mean"
+      labs <- c(expression(beta<0), expression(beta==0), expression(beta>0))
+      xlab_ <- ""
+      ylab_ <- ""
+      normalized <- T
+    }else if(i==1){
+      normalization <- c(exp(-1),exp(-0.5),exp(0))
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,1,1)
+      lambda <- rep(0,3)
+      nu <- rep(2,3)
+      xlab_ <- ""
+      ylab_ <- ""
+      title <- "(a)  amplitude"
+      labs <- c(expression(alpha==1), expression(alpha==0.5), expression(alpha==0))
+      normalized <- F
+    }else if(i==3){
+      a <- rep(0,3)
+      mu <- rep(0,3)
+      sigma <- c(1,0.6,1.5)
+      lambda <- rep(0,3)
+      nu <- rep(2,3)
+      xlab_ <- ""
+      ylab_ <- ""
+      title <- "(c)  variance"
+      labs <- c(expression(gamma==1), expression(gamma<1), expression(gamma>1))
+      normalized <- T
[…]

### 2021-05-31 code/paper-models/plots/plot-real.R

@@ -923,6 +923,10 @@ plot.actual.data.pairs <- function(model=NULL, typ=1){
   post <- extract.samples(model, n = 1000, pars=params) 
   post$gamma <- 1/(2*post$gamma)
   post$alpha <- exp(-post$alpha)
+  kurtosis <- kurtosis.skew.generror(post$nu, post$lambda)
+  skewness <- skewness.skew.generror(post$nu, post$lambda)
+  post$nu <- kurtosis
+  post$lambda <- skewness
   
   plist <- list()
   hlay <- matrix(NA, nrow=length(params), ncol=length(params))
@@ -1043,6 +1047,177 @@ plot.actual.data.pairs <- function(model=NULL, typ=1){
   return(p)
 }
 
+plot.pairs.manuscript <- function(model=NULL, typ=1){
+  # Parameters for plots
+  extra <- 1
+  colo <- c("#e6ab02", "#1b9e77", "#d95f02", "#7570b3")
+  
+  # Load the data if not added  
+  if(is.null(model)){
+    model <- readRDS(paste("../../../results/models/skew-generror-min20-elevation.rds", sep=""))
+  }
+  # log10_rev_trans <- trans_new(
+  #   "log10_rev",
+  #   function(x) exp(-exp(x)),
+  #   log_breaks(10),
+  #   domain = c(-Inf, Inf)
+  # )
+  
+  #indicator values
+  Tind <- read.table("../../../data/properties/codes/temperature_indicator_reindexed-20.csv", sep = " ")
+  Tend <- read.table("../../../data/properties/codes/change-tendency_reindexed-20.csv", sep = " ")
+  NEO <- read.table("../../../data/properties/codes/neophytes-list_reindexed-20.csv", sep = " ")
+  deta <- read.table("../../../data/properties/codes/neophytes-detailed_reindexed-20.csv", sep = " ")
+  competitive <- read.table("../../../data/properties/codes/competitive_reindexed-20.csv", sep = " ")
+  competitive$c <- (competitive$V3=="ccc" | grepl("cc",  competitive$V3))
+  competitive$r <- (competitive$V3=="rrr" | grepl("rr",  competitive$V3))
+  competitive$s <- (competitive$V3=="sss" | grepl("ss",  competitive$V3))
[…]

### 2021-06-22 manuscript/manuscript.tex
simulations for baseline

@@
 This dataset was compiled based on 30 years (1961–1990) of records from national weather stations.
-Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (see Supplementary Methods; Supplementary Fig.
-1).
+Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (Supplementary Fig.~1).
 %Missing TabsY
@@
 These are characterized following the ecological indicator values developed by \citet{landoltFloraIndicativaOkologische2010}, providing both an estimate of the average conditions in which a species can be found as well as a broad description of their range of variation.
-These values are provided for a range of 10 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
+These values are provided for a range of 8 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
 In addition to species' environmental preferences, the floristic data also contains information on species introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
@@
 
-To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Supplementary Methods; Supplementary Fig.
-2).
-Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Supplementary Methods and the Code Availability section).
+To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Code Availability section).
+Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Code Availability section).
 To perform the data analysis and generate the figures, we used some of the functions available with the R package `rethinking' \citep{mcelreathStatisticalRethinkingBayesian2020}.
@@
 Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
-Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the prior distributions (see Supplementary Table 2; Code Availability section; \citealt{nadarajahGeneralizedNormalDistribution2005}).
+Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the prior distributions (see Supplementary Table 2; \citealt{nadarajahGeneralizedNormalDistribution2005}).
 Notice that the Gaussian distribution will naturally emerge when $\nu_i=2$.
@@
 Skewed responses present steeper declines towards either side of the distribution.
-One way to accommodate this feature in our models is by considering a skewed normal distribution (Supplementary Figure XX).
+One way to accommodate this feature in our models is by considering a skewed normal distribution.
 As for the case described above, the Gaussian is a special case of this distribution, and it contains a parameter that controls for the level and direction of `skewness' (Fig~\ref{fig:response}c).
@@
 \end{equation}
-where $\gamma^{\prime}_{i} = q_{1}\left(\gamma_{i}, \nu_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma_{i}, \beta_{i}, \nu_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution such that $\lambda_{i}\in\left(-1, 1\right)$.
+where $\gamma^{\prime}_{i} = q_{1}\left(\gamma_{i}, \lambda_{i}\right)$, $\beta^{\prime}_{i} = q_{2}\left(\gamma_{i}, \beta_{i}, \lambda_{i}\right)$, and $\lambda_{i}$ is a parameter that describes the skewness of the distribution such that $\lambda_{i}\in\left(-1, 1\right)$.
 The function $\text{sgn}\left(x\right)$ characterizes the sign function.

### 2021-06-22 manuscript/supplementary_information.tex
simulations for baseline

--- 2021-01-22 manuscript/supplementary_information.tex
@@
 % Maths tools
-\usepackage{amsmath}
+\usepackage[tbtags]{amsmath}
 \usepackage{amssymb}
@@
 \usepackage{siunitx}
+
+% Landscape page
+\usepackage{pdflscape}
+\usepackage{afterpage}
 
@@
 
+\usepackage{makecell}
+
 \usepackage{titlesec}% http://ctan.org/pkg/titlesec
@@
 
+\renewcommand{\arraystretch}{1.3}
+
 %Title paper
@@
 
-\author{\textit{Bramon Mora and Alexander}}
+\author{\textit{Bramon Mora et.\,al.}}
 
@@
 \thispagestyle{empty}
+
+\clearpage
+
+% Full description of Binomial model with multiple environmental covariates
+% Full description of Categorical model with multiple environmental covariates
+% Full description of Categorical model with heavy-tails---generalized error distribution
+% Full description of Categorical model with heavy-tails---Student's t-distribution
+% Full description of Categorical model with skewed---skewed normal distribution
+% Using a MMSBM to calculate species similarity
+% Edge problem in stan and need to define minimum probability.
+
+\section*{Supplementary Methods}
+\subsubsection*{Full description of the baseline model}
+While the baseline model described in the main text included only one predictor, the same model can be extended to include multiple environmental variables.
+For example, imagine that we had $k$ predictors describing the environmental conditions of any given site $j$.
+The baseline model in Eq.~(1) of the main text could then be rewritten as:
+\begin{equation}
+\begin{split}
+y_{ij} & \sim \text{Binomial}\left(1, p_{ij}\right) \\
+\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_k \gamma_{ik} \left(x_{jk}-\beta_{ik}\right)^2,
+\end{split}
+\label{eq:baseline-binary}
+\end{equation}
+where $y_ij$ describes the presence and absence of species, and the prior distributions for the different parameters are analogous to the ones described in the main text.
+
+Similarly, if the response variable $y_ij$ represented ordinal data (e.g.~Braun-Blanquet abundance-dominance classes), characterized by $K$ different classes.
+Then, one can adapt Supplementary Eq.~(\ref{eq:baseline-binary}) as:
+\begin{equation}
+\begin{split}
+y_{ij} & \sim \text{Categorical}\left(\mathbf{q}\right) \\
+q_{1} & = 1- p_{ij}\phi_1 \\
+q_{k} & =  p_{ij}\phi_{k-1} -  p_{ij}\phi_{k} \\
+q_{K} & =  p_{ij}\phi_{K-1} \\
+\text{log}\left(p_{ij}\right) & = -\alpha_{i} - \sum_k \gamma_{ik} \left(x_{jk}-\beta_{ik}\right)^2 \\
+\phi_1 & = \hat{\phi_1} \\
+\phi_k & = \hat{\phi_k} \,\, \hat{\phi_{k-1}} \\
+\hat{\phi} & = \text{Beta}(1,1),
+\end{split}
+\label{eq:baseline-categorical}
+\end{equation}
+where the prior distributions for the rest of parameters are analogous to the ones described in the main text.
+
+\subsubsection*{Distance matrices from incomplete categorical and ordinal data}
+The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database.
+More specifically, both the ecological indicator values and range of variation are ordinal traits, containing potential missing entries (Supplementary Table \ref{stab:EI}).
+These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the relevant parameters of our Bayesian model.
+One way to do so is by compiling the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
+To accomplish this, we need to turn the set of ecological indicator values and range of variation into the distance matrices $D$ presented in the text.
+%, whereas plants' physiological data are characterized by categorical data containing multiple missing entries.
+
+More generally, we want to understand the way $N$ species are characterized by $M$ categorical traits.
+One way to frame this problem is by using a network representation.
+Following the ideas presented by \citet{godoy-loriteAccurateScalableSocial2016}, we assume that species can be connected to each of these traits by an interaction $\left(i, j\right)$ that can be of any type $r\in R$.
+Notice that this provides us with multiple ways to account for the information---and lack thereof---contained in the different ordinal traits $M$.
+That is, the $R$ types of interactions can represent the lack of information for a particular link $\left(i, j\right)$, the absence or presence of such interaction, and any type of association between $i$ and $j$.
+Then, given a set of interactions $R^{*}$ between $N$ and $M$, we can use a Mixed Membership Stochastic Block Model (MMSBM) to understand the differences between species.
+In particular, these models consider that the $N$ plants and $M$ traits can be classified into $K$ and $L$ groups types, respectively.
+Then, they estimate `group-membership vectors', describing the extend to which the different plants and traits belong to the different group types.
+
+Here, we use the MMSBM presented by \citet{godoy-loriteAccurateScalableSocial2016}, and classify plants into 5 types of species, and traits into 5 trait groups.
+Then, we estimate the group-membership vectors $\vec{\theta}_{\text{plants}}$, and calculate the pairwise distances $D_{ij}$ between plants as the euclidean distance between their membership vectors.
+We run these models independently for species' ecological indicator values and their range of variation (Supplementary Table \ref{stab:EI}) to generate the two distance matrices for parameters $\beta_i$ and $\gamma_i$.
+Notice that the MMSBM uses an expectation-maximization algorithm to estimate the different membership vectors.
+Therefore, in order to have good estimates of the different pairwise distances $D_{ij}$ for $\beta_i$ and $\gamma_i$, one needs to run the MMSBM several times (1000 times) and average the resulting pairwise distances \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
+
+This type of classification of ordinal variables that describe different species can also be done with other algorithms, including categorical PCA and matrix factorization methods.
+However, the MMSBM provides a convenient tool to classify very different types of traits, including categorical variables or diverse data containing multiple missing entries.
+Moreover, MMSBM have been shown to better capture the patterns of this sort of data \citet{godoy-loriteAccurateScalableSocial2016}.
+
+%Given a set of interactions $R^{*}$ between $N$ and $M$, we use a Mixed Membership Stochastic Block Model (MMSBM) to characterize these.
+In particular, we consider that plants and traits can be classified into $K$ and $L$ groups, respectively.
+For every species $i$, we assume that there is a probability $\theta_{i\alpha}$ for it to belong to any of the $K$ species groups.
+Likewise, we also assume that any trait $j$ has a probability $\phi_{j\beta}$ of belonging to any of the $L$ trait groups.
+Finally, we define $p_{\alpha\beta}\left(r\right)$ as the probability of a species from group $\alpha$ interacting with a trait from group $\beta$ by an association type $r$.
+Putting these together, the probability of an interaction $\left(i, j\right)$ of type $r$ can be calculated as:
+%\begin{equation}
+%Pr[r_{ij}=r] = \sum_{\alpha \beta} \theta_{i\alpha} \phi_{j\beta} p_{\alpha\beta}\left(r\right)
+%\end{equation}
+%Following this definition, we want to find the group memberships that maximize the likelihood $P\left(R^{*}|\theta, \phi, p\right)$.
+Doing so is difficult optimization problem; however, it has been shown that one can estimate the different $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$ parameters by maximizing the likelihood using an expectation-maximization algorithm \citep{godoy-loriteAccurateScalableSocial2016, tarres-deulofeuTensorialBipartiteBlock2019}.
+In simple terms, one can iteratively find multiple local minima for the likelihood, and average over the estimated the parameter values \citep{godoy-loriteAccurateScalableSocial2016}\footnote[2]{
+%While this averaging is trivial for the estimated probabilities $Pr[r_{ij}=r]$, it is non-trivial if one wants to find averages for the group memberships.
+The reason for this is related to the stochastic nature of the expectation-maximization algorithm.
+This algorithm initially assigns random group memberships to both species and traits.
+While this random labelling is irrelevant when studying the probabilities $Pr[r_{ij}=r]$, it is instead crucial for averaging $\theta_{i\alpha}$, $\phi_{j\beta}$, and $p_{\alpha\beta}\left(r\right)$.
+Therefore, before averaging the group membership estimates, one needs to find the bijective relationship for the labellings of different iterations of the optimization algorithm.
+In a nutshell, for every iteration, I do this by using a simulated annealing algorithm on the estimated $p_{\alpha\beta}\left(r\right)$, matching the corresponding labelling to a reference iteration.}.
+
+%The average estimates for the group memberships provide us with a different scale to classify species based on the traits these have.
+In short, for any species $i$, we can estimate a $K$-dimensional vector $\vec{\theta}_{i}$ that describes the extend to which $i$ belong to each group membership---i.e.
+the extend to which a species is of one type or another.
+This classification is useful because it can be used to compare species, defining a way to measure the distance between species based on an arbitrary---and potentially incomplete---set of categorical or ordinal traits $M$.
+The simplest case is to define the distance as $D_{ij} = |\vec{\theta}_{i}-\vec{\theta}_{j}|$.
+Alternatively, one could also define $K$ distance matrices based on the different group memberships $D^{\alpha}_{ij} = |\theta_{i\alpha}-\theta_{j\alpha}|$.
+
+%\subsubsection*{Sampling the posterior: simulated data and additional information}
+%A good way to understand the model behaviour and the choice of prior distributions is by using simulated data.
+Here, we only present the simulations for the baseline model---the simplest model---but similar simulations were done for the others (see Code Availability section of the main text).
+%
+%To simulate the data, we first generated two distance matrices relating the different parameters
+
+\subsubsection*{Alternative variance-covariance structures}
+The model structure defined above allows us to test how different sources of information characterize each of the different parameters.
+Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}).
+For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits.
+One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
+\begin{equation} 
+\Sigma_{ij} = \eta\,\text{exp}\left(-\sum_k\rho_{k} {D^{k}_{ij}}^2\right) + \delta_{ij} \sigma ,
+\label{eq:covariance-complex}
+\end{equation}
+where now $\rho_{k}$ are separate relevance hyperparameters for each distance matrix in the total variance of $\beta_i$.
+
+\section*{Supplementary Notes}
+\section*{Supplementary Figures}
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/variances}
+    	  \vspace{0.3cm}
+	   \caption{PCA and variance explained.}
+      \label{sfig:pca}
+\end{figure}
 
@@
 
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/pca-elevation-temperature}
+    	  \vspace{0.3cm}
+	   \caption{The same as in Fig.~1 but with elevation and temperature.}
+      \label{sfig:temperature-elevation}
+\end{figure}
+
+\clearpage
+
 \begin{figure}[h]
@@
 
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
+    	  \vspace{0.1cm}
+	   \caption{Universality of the relationship between mean and variance of species' distributions.
+Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a.
+Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
+Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
+Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
+	   %7.5-4
+      \label{sfig:universality}
+\end{figure}
+
 \clearpage
 
-\bibliography{references_ab}
+\section*{Supplementary Tables}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{ p{8em} p{6em} l p{16em}} 
+\textbf{Information} & \textbf{Variable type} & \textbf{Variable} & \textbf{Description} \\
+\hline
+\multirow{3}{7em}{ecological indicator values} & \multirow{3}{6em}{climate indicators} & temperature & \textit{10 classes ranging from alpine climates to very warm collines} \\
+& & continentality & \textit{5 classes ranging from oceanic to continental climates}\\ 
+& & light & \textit{5 classes ranging from deep shade to full light}\\ 
+\cline{2-4}
+& \multirow{5}{6em}{soil indicators} & moisture & \textit{10 classes ranging from very dry to flooded soil} \\
+& & reaction & \textit{5 classes ranging from extremely acid to alkaline soil} \\
+& & nutrients & \textit{5 classes ranging from very infertile to over-rich soils} \\ 
+& & humus & \textit{5 classes ranging from little to high humus content} \\
+& & aeration & \textit{5 classes ranging from bad to good aeration} \\ 
+\hline
+\multirow{3}{7em}{range of variation} & \multirow{3}{6em}{climate indicators} & temperature & \multirow{3}{14em}{\textit{3 classes indicating small, large and unlimited variation of the different climatic indicators}} \\
+& & continentality & \\ 
+& & light & \\ 
+\cline{2-4}
+& \multirow{3}{6em}{soil indicators} & moisture & \multirow{5}{14em}{\textit{3 classes indicating small, large and unlimited variation of the different soil indicators}}  \\ 
+& & reaction & \\ 
+& & nutrients & \\ 
+& & humus & \\ 
+& & aeration & \\ 
+\hline
+\multirow{3}{7em}{species types} & \multirow{2}{6em}{time of immigration} & invasion status & \textit{classified as indigenous, historical range expanders and recent range expanders} \\  
+& \makecell[l]{ \\change\\tendency} & abundance & \textit{classified as stable, decreasing, decreasing at low elevations, and increasing in abundance}\\ 
+\hline  
+\end{tabular}
+\end{center}
+\caption{Indicator values \citep{landoltFloraIndicativaOkologische2010}}
+\label{stab:EI}   
+\end{table}
+
+\begin{table}[h]
+\begin{center}
+\begin{tabular}{  p{6em} | p{1em} l l } 
+\textbf{response curve} & &  \(\beta^{\prime}_i\) & \(\gamma^{\prime}_i\) \\
+\\[-1em]
+\hline
+\\[-1em]
+baseline & & \(\beta_i\) & \(\gamma_i\) \\
+\\[-1em]
+
+\\[-1em]
+fat-tailed & & \(\beta_i\) & \( g = \sqrt{\frac{\gamma_i \Gamma\left(\frac{3}{\nu_i}\right)}{\Gamma\left(\frac{1}{\nu_i}\right)}}\) \\
+\\[-1em]
+
+\\[-1em]
+skewed &  & \( q_2 = \beta_i - \frac{2 \lambda_i}{\sqrt{\pi} \gamma^{\prime}_i}  \)  & \( q_1 = \gamma_i  \sqrt{\frac{\pi\left(1+3 \lambda^{2}_{i}\right)-8\lambda^2_i}{2\pi}}  \)  \\
+\\[-1em]
+
+\\[-1em]
+\multirow{-1.25}{6em}{skewed and fat-tailed} & & \( f_2 = \beta_i - \frac{2^{\frac{2}{\nu_i}}\lambda_i \Gamma\left(\frac{1}{2}+\frac{1}{\nu_i}\right)}{\sqrt{\pi}\gamma^{\prime}_i}   \)  & \( f_1 = \gamma_i \sqrt{\frac{\pi \left(1+3\lambda^2_i\right) \Gamma\left(\frac{3}{\nu_i}\right)-16^{\frac{1}{\nu_i}} \lambda^{2}_{i} \Gamma\left(\frac{1}{2}+\frac{1}{\nu_i}\right)^2 \Gamma\left(\frac{1}{\nu_i}\right)}{\pi \Gamma\left(\frac{1}{\nu_i}\right)}} \) \\
+\\[-1em]
+
+\end{tabular}
+\end{center}
+\caption{cite vignette}
+\label{stab:formulas}   
+\end{table}
+
+\bibliography{references2}
 

### 2021-06-22 code/paper-models/1d-baseline.R

@@ -844,7 +844,7 @@ skew.generror.1d <- function(d = NULL, recompile = T, simulated=T, min.occurrenc
 min.occurrence <- 20
 
 # Basic analyses
-d <- readRDS("../../data/processed/jsdm/1d-PC1PC2min20-data.rds")
+# d <- readRDS("../../data/processed/jsdm/1d-PC1PC2min20-data.rds")
 # d <- readRDS("../../data/processed/jsdm/skew-generror-simulated-data.rds")
 # d <- readRDS("../../data/processed/jsdm/skew-simulated-data.rds")
 # d <- readRDS("../../data/processed/jsdm/generror-simulated-data.rds")
@@ -868,4 +868,8 @@ d <- readRDS("../../data/processed/jsdm/1d-PC1PC2min20-data.rds")
 # baseline.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, elevation=F, temperature=T, ofolder="/cluster/scratch/bemora/plant-stan/")
 
 # Line regression
-line.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
+# line.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
+
+# Simulated
+d <- readRDS("../../data/processed/jsdm/1d-simulated-S1S2-data.rds")
+baseline.1d(d=d, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")

### 2021-06-22 code/paper-models/plots/functions_useful.R

@@ -1,0 +1,264 @@
+library('igraph')
+library("TeachingDemos")
+library("grid")
+library("extrafont")
+library(cluster, lib.loc = .Library)
+library("factoextra")
+library("magrittr")
+library("NbClust")
+library(pracma)
+library(vegan)
+library(ggplot2)
+library(amap)
+
+
+`plot.betadisper` <- function(x, axes = c(1,2), cex = 0.7, pch = seq_len(ng),
+                              col = NULL, lty = "solid", lwd = 1, hull = TRUE,
+                              ellipse = FALSE, conf,
+                              segments = TRUE, seg.col = "grey",
+                              seg.lty = lty, seg.lwd = lwd,
+                              label = TRUE, label.cex = 1,
+                              ylab, xlab, main, sub, ...)
+{
+  localAxis <- function(..., col, bg, pch, cex, lty, lwd) axis(...)
+  localBox <- function(..., col, bg, pch, cex, lty, lwd) box(...)
+  localTitle <- function(..., col, bg, pch, cex, lty, lwd) title(...)
+  Ellipse <- function(scrs, centres, conf, col, lty, lwd, ...) {
+    mat <- cov.wt(scrs, center = centres)
+    if (mat$n.obs == 1)
+      mat$cov[] <- 0
+    xy <- if (mat$n.obs > 1) {
+      veganCovEllipse(mat$cov, mat$center, conf)
+    } else {
+      scrs
+    }
+    ordiArgAbsorber(xy, FUN = lines, col = col, lty = lty, lwd = lwd, ...)
+  }
+  if(missing(main))
+    main <- deparse(substitute(x))
+  if(missing(sub))
+    sub <- paste("method = \"", attr(x, "method"), "\"", sep = "")
+  if(missing(xlab))
+    xlab <- paste("PCoA", axes[1])
+  if(missing(ylab))
+    ylab <- paste("PCoA", axes[2])
+  t <- if (missing(conf)) {
+    1
+  } else {
+    sqrt(qchisq(conf, df = 2))
+  }
+  g <- scores(x, choices = axes)
+  ng <- length(levels(x$group))
+  lev <- levels(x$group)
+  ## sort out colour vector if none supplied
+  if (is.null(col)) {
+    col <- palette()
+  }
[…]

### 2021-06-22 code/paper-models/plots/plot-real.R

@@ -1047,7 +1047,108 @@ plot.actual.data.pairs <- function(model=NULL, typ=1){
   return(p)
 }
 
-plot.pairs.manuscript <- function(model=NULL, typ=1){
+plot.correlations <- function(model=NULL, typ=1){
+  # Parameters for plots
+  extra <- 1
+  colo <- c("#e6ab02", "#1b9e77", "#d95f02", "#7570b3")
+  
+  # Load the data if not added  
+  if(is.null(model)){
+    model <- readRDS(paste("../../../results/models/skew-generror-min20-elevation.rds", sep=""))
+  }
+  
+  #indicator values
+  Tind <- read.table("../../../data/properties/codes/temperature_indicator_reindexed-20.csv", sep = " ")
+  Tend <- read.table("../../../data/properties/codes/change-tendency_reindexed-20.csv", sep = " ")
+  NEO <- read.table("../../../data/properties/codes/neophytes-list_reindexed-20.csv", sep = " ")
+  deta <- read.table("../../../data/properties/codes/neophytes-detailed_reindexed-20.csv", sep = " ")
+  competitive <- read.table("../../../data/properties/codes/competitive_reindexed-20.csv", sep = " ")
+  competitive$c <- (competitive$V3=="ccc" | grepl("cc",  competitive$V3))
+  competitive$r <- (competitive$V3=="rrr" | grepl("rr",  competitive$V3))
+  competitive$s <- (competitive$V3=="sss" | grepl("ss",  competitive$V3))
+  competitive$mix <- (competitive$c==0 & competitive$r==0 & competitive$s==0)  
+  
+  params <- c("alpha", "gamma", "beta", "lambda", "nu")
+  label <- c("amplitude", "variance", "mean", "skewness", "kurtosis")
+  trans <- c("log", "log", "identity", "identity", "identity")
+  
+  com <- combn(c(1:length(params)), 2)
+  # extract samples
+  post <- extract.samples(model, n = 1000, pars=params) 
+  post$gamma <- 1/(2*post$gamma)
+  post$alpha <- exp(-post$alpha)
+  kurtosis <- kurtosis.skew.generror(post$nu, post$lambda)
+  skewness <- skewness.skew.generror(post$nu, post$lambda)
+  post$nu <- kurtosis
+  post$lambda <- skewness
+  
+  plist <- list()
[…]

### 2021-06-22 code/paper-models/plots/plot-simulated.R

@@ -27,7 +27,7 @@ transform.skew <- function(post){
 }
 
 plot.simulated.compare <- function(){
-  m1 <- readRDS(paste("../../../results/models-old/skew-model-traits-1dskew-simulated2.rds", sep=""))
+  m1 <- readRDS(paste("../../../results/models/skew-generror-simulated.rdsskew-model-traits-1dskew-simulated2.rds", sep=""))
   m2 <- readRDS(paste("../../../results/models-old/baseline-model-1d1d-simulated2.rds", sep=""))
   rethinking::compare(m1, m2)
   
@@ -101,6 +101,64 @@ plot.simulated.compare <- function(){
     print(figure)
   }
   
+}
+
+plot.simulated.baseline <- function(beta=T, gp_type = 2){
+  # load data
+  d <- readRDS("../../../data/processed/jsdm/backup-data/1d-simulated2S1S2data.rds")
+  model_r <- readRDS(paste("../../../results/models/", sep=""))
+  
+  post <- extract.samples(model_r, pars=c("gammav", "lambda"))
+  post$sigma <- 1/(post$gammav**2)
+  
+  # Extract variables from the model
+  alphas <- precis(model_r, pars = "alpha", depth=2, prob = 0.89)
+  betas <- precis(model_r, pars = "beta", depth=3, prob = 0.89)
+  sigmas <- precis(model_r, pars = "gamma", depth=3, prob = 0.89)
+  lambdas <- precis(model_r, pars = "lambda", depth=3, prob = 0.89)
+  precis(model_r, pars = "lambda_bar", depth=3)
+  precis(model_r, pars = "sigma_l", depth=3)
+  
+  sigmas$mean <- sapply(1:dim(post$sigma)[2], function(x) mean(post$sigma[,x]), USE.NAMES = F)
+  sigmas$`5.5%` <- sapply(1:dim(post$sigma)[2], function(x) PI(post$sigma[,x], prob = c(0.890))[1], USE.NAMES = F)
+  sigmas$`94.5%` <- sapply(1:dim(post$sigma)[2], function(x) PI(post$sigma[,x], prob = c(0.890))[2], USE.NAMES = F)
+  
+  # Extract true values
+  N <- length(unique(d$dataset$id))
[…]

### 2021-06-22 code/paper-models/prepare-data.R

@@ -146,8 +146,10 @@ prepare.data <- function(variables = c("bio5_", "bio6_","bio12_"), min.occurrenc
 # Generate fake data to test the extent to which the model works
 simulated.data <- function(simulated.type="linear.corr"){
         
+        set.seed(3)
+        
         # Define system dimensions
-        N <- 50
+        N <- 25
         sites <- 300
         
         # Environmental predictors for each site
@@ -173,15 +175,16 @@ simulated.data <- function(simulated.type="linear.corr"){
         # Generate correlations
         beta1 <- z1
 
-        vec <- c(1:round(N*0.5), 1:round(N*0.5))
-        Dis_sigma <- as.matrix(dist(vec))+1-diag(N)
-        Dis_sigma <- (Dis_sigma/max(Dis_sigma))
+        # vec <- c(1:round(N*0.5), 1:round(N*0.5))
+        # Dis_sigma <- as.matrix(dist(vec))+1-diag(N)
+        # Dis_sigma <- (Dis_sigma/max(Dis_sigma))
+        Dis_sigma <- (as.matrix(dist(1:N))/N)
         
         Sigma <- 1*exp(-1/(0.3*0.3)*(Dis_sigma^2)) + diag(N)*0.1
         sigma_beta1 <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
         # Sigma <- 1*exp(-1/(0.2*0.2)*(Dis^2)) + diag(N)*0.1
         # alpha <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
-        alpha <- exp(rnorm(N, 0,1))
+        alpha <- exp(rnorm(N, -0.5,1))
 
         # Simulate data
         dataset <- expand.grid(site=1:sites, id=1:N)

### 2021-06-22 code/paper-models/run_inte.sh

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=3500]" -W 200:00 -n 35 "R --vanilla --slave < 1d-categorical.R"
+bsub -R "rusage[mem=3500]" -W 10:00 -n 35 "R --vanilla --slave < 1d-baseline.R"

### 2021-06-22 manuscript/manuscript.tex
baseline model 2d

@@
 This means that species found at lower elevations have generally wider distributions than those at higher elevations.
-The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~XX) as well as when using ordinal data (Supplementary Fig.~XX); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~XX).
+The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~3) as well as when using ordinal data (Supplementary Fig.~4); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~5).
 The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.

### 2021-06-22 code/paper-models/2d-baseline.R

@@ -1,4 +1,4 @@
-source("./prepare-data.R")
+#source("./prepare-data.R")
 source("./models.R")
 library(rethinking)
 library(rstan)
@@ -11,7 +11,7 @@ rstan_options(auto_write = TRUE)
 # - baseline model
 ####
 
-baseline <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, ofolder="../../results/models/"){
+baseline <- function(d = NULL, recompile = F, simulated=F, min.occurrence=20, ofolder="../../results/models/"){
   # Fixing some of the options
   variables=c("bio5_", "bio6_","bio12_", "gdd5_", "bio1_","bio15_","bio17_", "bio8_", "TabsY_")
   gp_type <- 2
@@ -81,6 +81,7 @@ baseline <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, of
   dat_4.1 <- list(N=N,
                   L=L,
                   K=2,
+                  minp=1e-100,
                   Y=t(obs),
                   X1=X1,
                   X2=X2,
@@ -121,7 +122,7 @@ baseline <- function(d = NULL, recompile = T, simulated=T, min.occurrence=10, of
                      init=init_4.1 , control = list(adapt_delta = 0.95, max_treedepth = 15))
   
   
-  saveRDS(mfit_4.1, file = paste(ofolder, "baseline-model", extension,".rds", sep=""))
+  saveRDS(mfit_4.1, file = paste(ofolder, "2d-baseline-model", extension,".rds", sep=""))
   return(mfit_4.1)
 }
 
@@ -242,6 +243,10 @@ baseline.traits <- function(d = NULL, recompile = T, simulated=T, min.occurrence
   saveRDS(mfit_5.1, file = paste(ofolder, "baseline-model-traits", extension,".rds", sep=""))
   return(mfit_5.1)
 }
+
+d <- readRDS("../../data/processed/jsdm/1d-PC1PC2min20-data.rds")
+
+baseline(d=d, recompile = F, simulated=F, min.occurrence=20, ofolder="/cluster/scratch/bemora/plant-stan/")
 
 # CHANGE FILE NAMES!!!
 # CHANGE FILE NAMES!!!

### 2021-06-22 code/paper-models/models.R

@@ -22,6 +22,7 @@ data{
     int N;
     int L;
     int K;
+    real minp;
     int Y[L,N];
     row_vector[N] X1;
     row_vector[N] X2;
@@ -78,7 +79,7 @@ model{
     to_vector( zbeta ) ~ std_normal();
 
     for ( i in 1:L ){
-        Y[i] ~ bernoulli(exp(-alpha[i] - gamma[1,i] * columns_dot_self(X1 - beta[1, i]) - gamma[2,i] * columns_dot_self(X2 - beta[2, i])));
+        Y[i] ~ bernoulli(exp(-alpha[i] - gamma[1,i] * columns_dot_self(X1 - beta[1, i]) - gamma[2,i] * columns_dot_self(X2 - beta[2, i])) + minp);
     }
 }
 "

### 2021-06-22 code/paper-models/run_inte.sh
run_inte

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=3500]" -W 10:00 -n 35 "R --vanilla --slave < 1d-baseline.R"
+bsub -R "rusage[mem=3500]" -W 10:00 -n 35 "R --vanilla --slave < 2d-baseline.R"

### 2021-06-23 manuscript/manuscript.tex
New simulations

@@
 
-To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (see Code Availability section).
+To test the performance of the model as well as our choice of prior distributions, we modelled simulated data and compared the sampled posterior distributions to the data-generating parameters (e.g.~Supplementary Fig.~2; see Code Availability section).
 Notice that using the link function in Eq.~(\ref{eq:baseline-response}) could cause problems when sampling the model, and some adjustments need to be made when specifying the model (see Code Availability section).
@@
 In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e.
-$\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~XX).
+$\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~6).
 This implies that, at higher elevations, species' distributions generally have lower amplitudes.
@@
 To do so, we changed the response curve of our Bayesian model to follow a generalized error distribution (Eq.~\ref{eq:generalized-response}).
-A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~XX).
+A comparison of the WAIC values showed this non-linear regression to outperform the baseline model (Supplementary Fig.~7).
 Studying the resulting posterior distributions, we found the average kurtosis of the distributions to be slightly greater than zero, which corresponds to distributions with longer tails than the Gaussian case (Fig.~\ref{fig:kurtosis-skewness}).
-However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~XX), which might indicate that the shape of the tails is species-specific.
+However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~8), which might indicate that the shape of the tails is species-specific.
 
@@
 	   \caption{Average kurtosis and skewness of species' distributions.
-Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table XX), the two panels describe the average (a) kurtosis and (b) skewness of distributions.
+Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table 2 and \citealt{vignette}), the two panels describe the average (a) kurtosis and (b) skewness of distributions.
 Panel (a) displays the results obtained using response curves that follow a generalized error distribution (black line) and a skewed generalized error distribution (yellow line).
@@
 Using Eq.~(\ref{eq:skewed-response}), we next studied the skewness of species' distributions.
-Based on the estimates for the WAIC values, this model outperformed the first two (Supplementary Fig.~XX), which sheds light on the naturally skewed nature of species' distributions.
+Based on the estimates for the WAIC values, this model outperformed the first two (Supplementary Fig.~7), which sheds light on the naturally skewed nature of species' distributions.
 Perhaps most importantly, studying the mean value of the skewness across species, we found this to be consistently below zero (Fig.~\ref{fig:kurtosis-skewness}).
@@
 The same was true when using a model that allowed for both fat-tailed and skewed response curves (Eq.~\ref{eq:heavy-skewed-response}).
-This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~XX), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}).
-A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note XX.
+This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~7), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}).
+A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note 1 and Supplementary Fig.~9.
 %Comparing the prior and posterior distributions for the hyperparameters in the.

### 2021-06-23 manuscript/supplementary_information.tex
baseline model 2d
New simulations

@@
 
+%\begin{figure}[ht]
+%  \centering
+%    \vspace{0.5cm}
+%    \includegraphics[width=1\textwidth]{figures/prior}
+%    	  \vspace{0.3cm}
+%	   \caption{Study of the priors for the baseline model presented in the main text.}
+%     \label{sfig:sensitivity}
+%\end{figure}
+%\clearpage
+
 \begin{figure}[ht]
@@
     \vspace{0.5cm}
-    \includegraphics[width=1\textwidth]{figures/prior}
+    \includegraphics[width=1\textwidth]{figures/simulations-comparison}
     	  \vspace{0.3cm}
-	   \caption{Study of the priors for the baseline model presented in the main text.}
-      \label{sfig:sensitivity}
+	   \caption{Studying the simulations.
+Transparency is a way to indicate the different sample sizes.
+Say that for both $\beta$ and $gamma$, the posterior distributions for the different species are correlated so that the variance between species increases as they are further apart.}
+      \label{sfig:simulations}
 \end{figure}
@@
   \centering
-    \vspace{0.5cm}
-    \includegraphics[width=1\textwidth]{figures/figure1-secondaxis}
-    	  \vspace{0.3cm}
-	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
-      \label{sfig:sensitivity}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-    \includegraphics[width=0.9\textwidth]{figures/means-Rapopor}
+    \includegraphics[width=0.9\textwidth]{figures/categorical-figure1}
     	  \vspace{0.1cm}
@@
 	   %7.5-4
-      \label{sfig:universality}
+      \label{sfig:categorical-baseline}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/figure1-secondaxis}
+    	  \vspace{0.3cm}
+	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
+      \label{sfig:secondaxis}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.9\textwidth]{figures/alpha-vs-gamma-beta}
+    	  \vspace{0.3cm}
+	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
+      \label{sfig:alpha-vs-gammabeta}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.8\textwidth]{figures/WAIC-plot}
+    	  \vspace{0.3cm}
+	   \caption{Comparison of the different models using WAIC values.
+Each row describes one of the models described in the main text.
+(a) WAIC comparison of the different models.
+The filled points are the in-sample deviance of each model, and the open points are the estimated WAIC values.
+The dark line segments represent the standard error of each WAIC.
+The standard error of the difference between each WAIC and the best performing model (top row) is characterized by the gray triangles and line segments.
+The vertical line references the model with the lowest WAIC value.
+(b) Akaike weight for each model used to compare predictive accuracy of the different Bayesian models (see \citealt{mcelreathStatisticalRethinkingBayesian2020}).
+According to this measure, the model with fat-tails and skewed response is the best model.
+The open points are the mean Akaike weights and the line segments represent the corresponding standard error.
+The vertical line references the model with the highest Akaike weight.
+These graphs were generated using some of the functions from the R package \textit{rethinking} \citep{mcelreathStatisticalRethinkingBayesian2020}.}
+      \label{sfig:WAIC}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.8\textwidth]{figures/nu-kurtosis}
+    	  \vspace{0.3cm}
+	   \caption{Variability in the tails.}
+      \label{sfig:nu-kurtosis}
 \end{figure}

### 2021-06-23 code/paper-models/1d-baseline.R

@@ -871,5 +871,5 @@ min.occurrence <- 20
 # line.1d(d=d, simulated=F, recompile = F, min.occurrence = min.occurrence, ofolder="/cluster/scratch/bemora/plant-stan/")
 
 # Simulated
-d <- readRDS("../../data/processed/jsdm/1d-simulated-S1S2-data.rds")
-baseline.1d(d=d, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")
+d <- readRDS("../../data/processed/jsdm/1d-skew-generror-simulated-S1S2-data.rds")
+skew.generror.1d(d=d, simulated=T, recompile = F, ofolder="/cluster/scratch/bemora/plant-stan/")

### 2021-06-23 code/paper-models/plots/other-plots.R

@@ -11,6 +11,11 @@ library(cowplot)
 ####
 # Visualizing the results of some of the models
 ####
+shade_curve <- function(MyDF, zstart, zend, fill = "red", alpha = .5, mean.1=0, sd.1=1){
+  geom_area(data = subset(MyDF, x >= mean.1 + zstart*sd.1
+                          & x < mean.1 + zend*sd.1),
+            aes(y=y), fill = fill, color = NA, alpha = alpha)
+}
 
 prior_p <- function(N=2, sd_=1){
   set.seed(2)

### 2021-06-23 code/paper-models/plots/plot-real.R

@@ -411,11 +411,11 @@ plot.actual.data <- function(model=NULL){
   # Parameters for plots
   extra <- 1
   colo <- c("#1b9e77", "#d95f02", "#7570b3")
-  n <- 1
+  n <- 2
   
   # Load the data if not added  
   if(is.null(model)){
-    model <- readRDS("../../../results/models/baseline-min20-1d.rds")
+    model <- readRDS("../../../results/models/baseline-min20-2d.rds")
   }
   
   #indicator values
@@ -475,8 +475,8 @@ plot.actual.data <- function(model=NULL){
     
     # Scatter plot
     plist[[2]] <- plot.scatter(mu=mu_beta, variance=mu_gamma, color=colo[3], xlabel="", ylabel=expression(paste("mean ", beta)), mar=margin(5.5,5.5,5.5,5.5), ylims=ylim, xlims=xlim)
-    # breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4))+1
-    breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4), digits = 1)
+    breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4))+1
+    # breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4), digits = 1)
     plist[[2]] <- plist[[2]] + coord_trans(x="log")+
       scale_x_continuous(breaks = breaks, labels = breaks, limits = xlim)
     plist[[2]] <- plist[[2]] + annotate("text", x=posx, y = posy, label=paste0(round(meanPI,2), " ± ", round(sdPI,2)), size=3)
@@ -1155,7 +1155,7 @@ plot.pairs.manuscript <- function(model=NULL, typ=0){
   
   # Load the data if not added  
   if(is.null(model)){
-    model <- readRDS(paste("../../../results/models/skew-generror-min20-elevation.rds", sep=""))
+    model <- readRDS(paste("../../../results/models/skew-generror-min20-1d.rds", sep=""))
   }
   # log10_rev_trans <- trans_new(
   #   "log10_rev",
@@ -1892,3 +1892,248 @@ plot.elevation.temperature <- function(model=NULL){
   
 }
 
+plot.supplementary.abundanceplot <- function(model=NULL, typ=0){
+  # Parameters for plots
[…]

### 2021-06-23 code/paper-models/plots/plot-simulated.R

@@ -105,19 +105,16 @@ plot.simulated.compare <- function(){
 
 plot.simulated.baseline <- function(beta=T, gp_type = 2){
   # load data
-  d <- readRDS("../../../data/processed/jsdm/backup-data/1d-simulated2S1S2data.rds")
-  model_r <- readRDS(paste("../../../results/models/", sep=""))
+  d <- readRDS("../../../data/processed/jsdm/1d-simulated-S1S2-data.rds")
+  model_r <- readRDS(paste("../../../results/models/simulated-baseline-1d.rds", sep=""))
   
-  post <- extract.samples(model_r, pars=c("gammav", "lambda"))
-  post$sigma <- 1/(post$gammav**2)
+  post <- extract.samples(model_r, pars=c("gamma", "beta", "alpha"))
+  post$sigma <- post$gamma
   
   # Extract variables from the model
   alphas <- precis(model_r, pars = "alpha", depth=2, prob = 0.89)
   betas <- precis(model_r, pars = "beta", depth=3, prob = 0.89)
   sigmas <- precis(model_r, pars = "gamma", depth=3, prob = 0.89)
-  lambdas <- precis(model_r, pars = "lambda", depth=3, prob = 0.89)
-  precis(model_r, pars = "lambda_bar", depth=3)
-  precis(model_r, pars = "sigma_l", depth=3)
   
   sigmas$mean <- sapply(1:dim(post$sigma)[2], function(x) mean(post$sigma[,x]), USE.NAMES = F)
   sigmas$`5.5%` <- sapply(1:dim(post$sigma)[2], function(x) PI(post$sigma[,x], prob = c(0.890))[1], USE.NAMES = F)
@@ -126,37 +123,63 @@ plot.simulated.baseline <- function(beta=T, gp_type = 2){
   # Extract true values
   N <- length(unique(d$dataset$id))
   alpha_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$alpha[1])
-  lambda_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$lambda[1])
   beta1_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$beta1[1])
   sigma1_r <- sapply(1:N, function(x) d$dataset[d$dataset$id==x,]$sigma_beta1[1])
-  
[…]

### 2021-06-23 code/paper-models/plots/visualizing-results.R

@@ -1468,12 +1468,12 @@ shade_curve <- function(MyDF, zstart, zend, fill = "red", alpha = .5, mean.1=0, 
             aes(y=y), fill = fill, color = NA, alpha = alpha)
 }
 
-prior_p <- function(N=2, sd_=1){
+prior_p <- function(N=2, sd_=2){
   set.seed(2)
   xmin=-2
   xmax=2
-  a <- rnorm(1e6, 0,sd_)
-  c <- rnorm(1e6, 0,sd_)
+  a <- rnorm(1e6, rnorm(1e6, 0,1.3),rexp(1e6, 1))
+  c <- rnorm(1e6, rnorm(1e6, 0,1.3),rexp(1e6, 1))
   b <- rnorm(N, 0,sd_)
   x <- seq(from=xmin, to=xmax, length.out = 1000)
   
@@ -1493,7 +1493,7 @@ prior_p <- function(N=2, sd_=1){
           MyDF$y <- MyDF$y/max(MyDF$y)
           
           plist[[l]] <- ggplot(MyDF, aes(x = x, y = y)) + geom_line() +
-            shade_curve(MyDF = MyDF, zstart = lims_c[(j-1)], zend = lims_c[(j)], fill = my_col, alpha = .3, sd.1 = sd_) +
+            shade_curve(MyDF = MyDF, zstart = lims_c[(j-1)], zend = lims_c[(j)], fill = "red", alpha = .3, sd.1 = sd_) +
             theme_bw()+
             annotate("text", x=0,y=0.4, label = expression(log(lambda)), size = 3)+
             theme(text = element_text(size=10),
@@ -1512,7 +1512,7 @@ prior_p <- function(N=2, sd_=1){
         MyDF$y <- MyDF$y/max(MyDF$y)
         
         plist[[l]] <- ggplot(MyDF, aes(x = x, y = y)) + geom_line() +
-          shade_curve(MyDF = MyDF, zstart = lims_a[(i-1)], zend = lims_a[(i)], fill = my_col, alpha = .3, sd.1 = sd_) +
+          shade_curve(MyDF = MyDF, zstart = lims_a[(i-1)], zend = lims_a[(i)], fill = "red", alpha = .3, sd.1 = sd_) +
           theme_bw()+
           annotate("text", x=0,y=0.5, label = expression(log(alpha)), size = 3)+
           coord_flip()+

### 2021-06-23 code/paper-models/prepare-data.R

@@ -263,10 +263,10 @@ simulated.generr.data <- function(){
 # Generate fake data to test the extent to which the model works
 simulated.skew.generr.data <- function(){
         
-        set.seed(1)
+        set.seed(2)
         # Define system dimensions
-        N <- 20
-        sites <- 300
+        N <- 50
+        sites <- 900
         
         # Environmental predictors for each site
         e1 <- rnorm(sites)
@@ -291,15 +291,17 @@ simulated.skew.generr.data <- function(){
         # Generate correlations
         beta1 <- z1
         
-        vec <- c(1:round(N*0.5), 1:round(N*0.5))
-        Dis_sigma <- as.matrix(dist(vec))+1-diag(N)
-        Dis_sigma <- (Dis_sigma/max(Dis_sigma))
+        # vec <- c(1:round(N*0.5), 1:round(N*0.5))
+        # Dis_sigma <- as.matrix(dist(vec))+1-diag(N)
+        # Dis_sigma <- (Dis_sigma/max(Dis_sigma))
+        Dis_sigma <- (as.matrix(dist(1:N))/N)
         
         Sigma <- 1*exp(-1/(0.3*0.3)*(Dis_sigma^2)) + diag(N)*0.2
-        sigma_beta1 <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
+        sigma_beta1 <- exp(mvrnorm(mu = rep(0.5, times = N), Sigma = Sigma))
+        sigma_test <- sigma_beta1
         # Sigma <- 1*exp(-1/(0.2*0.2)*(Dis^2)) + diag(N)*0.1
         # alpha <- exp(mvrnorm(mu = rep(0, times = N), Sigma = Sigma))
-        alpha <- exp(rnorm(N, -0.5,0.7))
+        alpha <- exp(rnorm(N, -0.5,0.5))
         nu <- abs(rnorm(N,0.5, 0.1))+1
         lambda <- inv_logit(rnorm(N, 1.1,0.4))*2-1
         

### 2021-06-23 code/paper-models/run_inte.sh
changing bash file for simulations

@@ -1,3 +1,3 @@
 #!/usr/bin/env bash
 
-bsub -R "rusage[mem=3500]" -W 10:00 -n 35 "R --vanilla --slave < 2d-baseline.R"
+bsub -R "rusage[mem=3500]" -W 10:00 -n 35 "R --vanilla --slave < 1d-baseline.R"

### 2021-07-15 manuscript/manuscript.tex
manuscript changes

@@
 \title{\vspace{-1cm}
-Model linearity breeds contempt: using Bayesian non-linear models to uncover broad macroecological patterns}
+Using Bayesian non-linear models to uncover broad macroecological patterns}
 % Model familiarity breeds contempt: simple models to compare properties of species' distributions
@@
 \affil[1]{\footnotesize Institute of Integrative Biology, ETH Zürich, Zürich, Switzerland}
-\affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
 \affil[2]{\footnotesize Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
@@
 \section*{Abstract}
+%Species' realized niches are classically pictured as bell-shaped probability distributions.
+These distributions, however, can actually take many different forms.
+For example, fat-tailed or skewed responses are very common in nature, as these can naturally emerge as a result of several ecological processes.
+While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies shape ecological communities.
+Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to quantify such properties in empirical plant distributions.
+These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies.
+With this approach, we are able to distil the shape of plant distributions, tackling long-standing hypotheses regarding the way ecological communities are assembled across space.
+In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients.
+We also shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species or the result of their historical contexts.
+Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide a novel perspective on the way systems of many species are distributed along environmental gradients.
+
+% Am Nat abstract (200)
 Species' realized niches are classically pictured as bell-shaped probability distributions.
-These distributions, however, can actually take many different forms.
-For example, fat-tailed or skewed responses are very common across fields, as these can naturally emerge as a result of several ecological processes.
-While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies have shaped ecological communities.
-Unfortunately, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients.
-Here, we use a set of Bayesian non-linear models to uncover such properties.
-These models account for all prior knowledge we have regarding species' realized niches, including expert knowledge of their environmental preferences and ecological strategies.
-With this approach, we are able to distil the shape of empirical plant distributions, which helps us tackle long-standing hypotheses regarding the way ecological communities are assembled across space.
-In particular, we study the relationship between several properties of distributions, such as the link between species' range size and elevation, revealing the existence of broad macroecological patterns along environmental gradients.
-Moreover, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical contexts.
-Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
-
-%Species' realized niches are often visualized as bell-shaped probability distributions.
-These, however, can take many different forms.
-For example, fat-tailed or skewed distributions are very common across fields, as these can naturally emerge as a result of several ecological processes.
-While one does not need to know the shape of species' distributions to effectively model them, there is a lot that can be learned about how climatic processes and historical contingencies differently affected species' distributions by studying the basic form.
-Indeed, we still lack a general understanding of the basic properties describing the shape of species' distributions, and much less is known about how these compare to each other across gradients.
-Here, we use a set of Bayesian non-linear models to uncover such properties.
-These models account for all prior knowledge we have regarding species' niches, including expert knowledge of their environmental preferences and ecological strategies.
-With this approach, we are able to study the shape empirical plant distributions, revealing general patterns regarding the way communities are assembled across space.
-In particular, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns that are present along environmental gradients.
-We found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation, or their general level of skewness.
-Perhaps most importantly, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species' historical context.
-Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
-
-%The shape of species' distributions along environmental gradients is part of old debates in ecology and biogeography.
-These have led the development of many statistical models centred around predicting the presence and absence of species over space and time.
-Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change.
-However, there is a lot that can be learned about how climatic processes and historical contingencies have differently shaped species' distributions by studying instead their basic shape.
-Indeed, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients.
-Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches.
-These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and ecological strategies.
-With this approach, we are able to uncover the general properties characterizing empirical plant distributions, revealing general patterns regarding the way communities are assembled.
-That is, by learning about the shape of distributions, we tackle long-standing hypothesis regarding broad macroecological patterns present along environmental gradients.
-In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients.
-Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species.
-Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
-
-%Species' distribution models have emerged as one of the most influential methodological advances in ecology and biogeography of the last decades.
-Useful to understand how populations of species will change along environmental gradients, they have become ecologists' compass to predict the effects of global climate change.
-That said, uncovering the mechanisms shaping species' realized niches has been one of the main driving forces behind the development of these models.
-That is, recent efforts have been often focused on understanding which biotic and abiotic factors are good predictors of species' niches---with an increasing effort placed in improving the predictive power of the statistical models.
-However, we still lack a general understanding of the shape of species' distributions, and much less is known about how these distributions compare to each other across gradients.
-Here, we use a set of Bayesian non-linear models to uncover the shape of species' realized niches.
-These models account for all prior knowledge we have regarding their shape, including expert knowledge of species' environmental preferences and physiology.
-With this approach, we are able to uncover the true shape of empirical species' distributions.
-Moreover, they allow us to tackle long-standing hypothesis regarding broad biogeographical patterns.
-In particular, we found conclusive evidence of the relationship between several properties of distributions, including the link between species' range size and elevation and their skewness along gradients.
-Finally, we are able to shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties of species.
-Overall, our approach offers a useful statistical framework to understand the shape of species' distributions, and our results provide an unprecedented perspective of the way systems of many species are distributed along environmental gradients.
+These distributions, however, can actually take many different forms, including fat-tailed or skewed responses.
+While one does not need to know the shape of species' distributions to effectively model them, studying their basic form can teach us a lot about the ways climatic processes and historical contingencies shape ecological communities.
+Unfortunately, we still lack a general understanding of the properties describing the shape of species' distributions, and how these compare to each other across gradients.
+Here, we use a set of Bayesian non-linear models to quantify such properties in empirical plant distributions.
+With this approach, we are able to distil the shape of plant distributions, tackling long-standing hypotheses regarding the way ecological communities are assembled across space.
+Studying the relationship between distribution properties, we revealed the existence of broad macroecological patterns along environmental gradients.
+We also shed light on the extent to which some aspects of the shape of observed realized niches---such as kurtosis and skewness of the distributions---could be intrinsic properties or the result of species’ historical contexts.
+Overall, our results provide a novel perspective on the way systems of many species are distributed along environmental gradients.
 
@@
 
-% REFERENCES TO LOOK AT FROM JAKE
-% DATE EMAIL: 16.03.2021
-
-%https://www.tandfonline.com/doi/abs/10.1080/01621459.1998.10474117
-%https://hess.copernicus.org/preprints/hess-2015-24/
-%https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2009WR008933
-
-%############# NEW INTRO ##############
-%Bell shape is a good option, but it is unclear what the real shape is
 One of the central goals of ecology is to understand the ways species are distributed across space and time.
-While many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002, sagarinMovingAssumptionsUnderstand2006}.
+Many ecological textbooks assume the shape of species' realized niches to be unimodal and symmetric along environmental gradients \citep{krebsEcologyExperimentalAnalysis1972}.
 In practice, there is a strong argument to be made in favour of assuming these to be bell shaped.
 Namely, if all that we are willing to assume about species' distributions is that these occupy finite geographic ranges, the most conservative statistical approach is to model their distribution as Gaussian (i.e.~the corresponding maximum entropy distribution; \citealt{frankCommonPatternsNature2009}).
-That said, there is currently no general agreement on the basic shape of species' realized niches.
+However, some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austinSpatialPredictionSpecies2002}, and there is currently no general agreement on the basic shape of species' realized niches \citep{sagarinAbundantCentreDistribution2002, , sagarinMovingAssumptionsUnderstand2006}.
 Indeed, many factors can play a role in defining their shape, and several natural processes can lead to non-normal distributions.
@@
 %Skewed and fat-tails are common and they could well describe niches
-Fat-tailed and skewed distributions are very common across fields.
-The former naturally emerges as a result of processes involving seasonality (e.g.~in communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or some stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
+Fat-tailed and skewed distributions are very common across scientific fields.
+The former naturally emerges as a result of processes involving intermittent (e.g.~in email communications patterns; \citealt{malmgrenPoissonianExplanationHeavy2008}) or stochastic events (e.g.~in the spread of infectious diseases; \citealt{wongEvidenceThatCoronavirus2020}).
 Indeed, species' dispersal patterns have been shown to have fat tails due to the natural variability among individuals \citep{petrovskiiDispersalStatisticallyStructured2009}.
@@
 Similarly, several processes can lead to skewed distributions.
-For example, species might present asymmetric environmental tolerances along altitudinal gradients, allowing them to withstand different temperature extremes \citep{sundayGlobalAnalysisThermal2011}.
+For example, species might have asymmetric environmental tolerances along altitudinal gradients, allowing them to withstand different temperature extremes \citep{sundayGlobalAnalysisThermal2011}.
 Species might also experience abiotic and biotic pressures that increase or decrease along a temperature gradient, which could result in species' distributions presenting steeper declines towards warmer or colder environments \citep{normandImportanceAbioticStress2009}.
-Overall, many different properties could characterize species' realized niches, and every new shape entails different underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.
+Overall, many different properties could characterize species' realized niches, and every new shape entails different and potentially competing underlying hypotheses regarding the way communities are assembled over time \citep{damenSpatialPredictionsCommunity2017}.
 
@@
 Comparing these properties across species allows us to study broad macroecological patterns that could be critical from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992, channellDynamicBiogeographyConservation2000}.
-For example, the Rapoport's rule, a classsic biogeographical hypothesis, predicts species' ranges to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients.
-This sort of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}.
+For example, Rapoport's rule, a classic biogeographical hypothesis, predicts species' range size to increase with latitude or elevation \citep{stevensElevationalGradientAltitudinal1992}, hinting at the existence of general biogeographical constraints that shape species' distributions along gradients.
+These sorts of macroecological patterns are interesting because they provide insights into the way different species assemble and establish in different environments \citep{linderGeographicRangeStructure2000}.
 That is, the differences in species' responses to the environment can shed light on how climatic processes and historical contingencies have differently shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, helmuthBIOPHYSICSPHYSIOLOGICALECOLOGY2004, siefertHowClimateDispersal2015}.
-Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem to solve.
+Uncovering the shape of species' realized niches and the extent to which these vary across species is nevertheless a challenging statistical problem.
 Indeed, to this date, we do not have an effective way to parsimoniously compare the shape of the realized niches of many species along environmental gradients.
@@
 
-Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000}.
+Over the last two decades, ecologists have developed a plethora of distribution models to try to untangle the factors that play a role in defining species' realized niches \citep{guisanPredictiveHabitatDistribution2000, zurellEcologicalNicheModeling2019}.
 These models are fundamental to the scientific community for predicting changes in species' geographic distributions and the effects of environmental disturbances.
@@
 This is useful because it simplifies the optimization process, but it might not be ideal when studying and comparing the shape of species' distributions along environmental gradients.
-First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints that might not be biologically justified.
-While this might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}.
-Second, the aforementioned structural constraints also limit our ability to include any prior information to our parameter estimates.
-Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
+First and foremost, a linear relationship between covariates often comes with a set of implicit mathematical constraints.
+These might not hinder the predictive performance of the models \citep{norbergComprehensiveEvaluationPredictive2019}, but a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{terbraakWeightedAveragingLogistic1986, jamilGeneralizedLinearMixed2013} to fat-tailed or skewed responses \citep{huismanHierarchicalSetModels1993}.
+Second, these mathematical constraints also limit our ability to include any prior information to our parameter estimates.
+Observations of species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge of species' realized niches (e.g.~\citealt{landoltFloraIndicativaOkologische2010}).
 That said, this information is rarely accounted for in most modelling approaches, likely because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986, ovaskainenHowMakeMore2017}).
 Finally, some have proposed several non-linear structures to characterize several features of individual species' response curves \citep{huismanHierarchicalSetModels1993}.
-Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, taking full advantage of modern statistical approaches (e.g.
+Setting aside the fact that the interpretation and comparison of parameter estimates becomes challenging following most of these model structures, these are generally not designed to jointly study different species, and therefore taking full advantage of modern statistical approaches (e.g.
 sharing information among species or accounting for parameter uncertainty; \citealt{evansProcessbasedRangeModeling2016}).
 
-In this work, we rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy.
-In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies.
+In this work, we develop a set of Bayesian hierarchical distribution models to study the shape of empirical plant distributions.
+We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions.
+This allows us to measure different properties of species' realized niches while accounting for all prior information regarding these distributions, including expert knowledge of species' environmental indicator values, range sizes, and plant ecological strategies.
+Using our statistical framework, we study the distribution of plant species along an elevation gradient in the Swiss Alps, revisiting some classic hypotheses in ecology and biogeoraphy.
+Specifically, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns.
+Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or non-native, might respond to the environment.
+More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
+
+%In this work, we rethink traditional modelling approaches and develop a conceptually simple statistical framework to revisit some classic hypotheses in ecology and biogeoraphy.
+Specifically, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of plant species along an elevation gradient in the Swiss Alps, including expert knowledge of species environmental indicator values, range sizes, and plant ecological strategies.
 We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow non-linear responses characterizing skewed and long-tailed distributions.
 Using this statistical framework, we are able to compare the basic properties of the realized niche of multiple species, testing for the existence of broad macroecological patterns.
-Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or neophytes, might respond to the environment.
+Comparing the posterior distribution of those parameters that control for the shape of distributions, we are also able to showcase variation in the way different types of species, such as native or non-native, might respond to the environment.
 More generally, we are able to uncover the approximate shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.
@@
 We studied the distribution of plant communities along an elevation gradient.
-To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps \citep{randinClimateChangePlant2009}, and ii) an extensive floristic database containing environmental and physiological traits for all vegetation across Switzerland \citep{landoltFloraIndicativaOkologische2010}.
+To do so, we combined two different datasets: i) one describing the co-occurrence of species across multiple open grasslands in the Swiss Alps \citep{randinClimateChangePlant2009}, and ii) an extensive plant-attribute database containing environmental and life-history traits for all plant species across Switzerland \citep{landoltFloraIndicativaOkologische2010}.
 
@@
 This dataset was compiled based on 30 years (1961–1990) of records from national weather stations.
-Since most of the data is highly correlated, we calculated the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (Supplementary Fig.~1).
-%Missing TabsY
+Since most of the data are highly correlated, we used a Principal Component Analysis to calculate the main axes of variation of the following scaled variables: daily minimum, maximum and average temperature; sum of growing degree-days above $5^{\circ}\text{C}$; mean temperature of wettest quarter; annual precipitation, precipitation seasonality, and precipitation of driest quarter (Supplementary Fig.~1).
+Notice that the first axis of variation is positively correlated with elevation and negatively correlated to temperature, while the second axis is positively correlation to precipitation seasonality (Supplementary Fig.~1).
+% TabsY
 
-\subsubsection*{Floristic data}
-To complement the aforementioned distribution data, we used a floristic database of around 5500 vascular plants across Switzerland.
-Some of the information in this database has been previously shown to account for unexplained variation when used as explanatory variables in species' distribution models \citep{scherrerEcologicalIndicatorValues2019}.
+\subsubsection*{Plant attributes}
+To complement the aforementioned distribution data, we used a attribute database of around 5500 vascular plants across Switzerland.
+Notice that some of the information in this database has been previously shown to account for unexplained variation when used as explanatory variables in species' distribution models \citep{scherrerEcologicalIndicatorValues2019}.
 It was built based on expert knowledge and phytosociological field experience of botanists and ecologists, and contains information regarding plants' environmental preferences and ecological strategies.
@@
 These values are provided for a range of 8 environmental variables, including temperature, continentality, light conditions, as well as moisture, acidity and nutrient content of the soil (see a full list and description of the ecological indicators in the Supplementary Table 1; \citealt{landoltFloraIndicativaOkologische2010}).
-In addition to species' environmental preferences, the floristic data also contains information on species introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
-We describe this information in more detail in the Supplementary Table 1.
+In addition to species' environmental preferences, the attribute data also contain information on species' introduction status (e.g.~identifying those species that are recent and historical range expanders) and change tendency (e.g.~indicating species that have shown decline or increase in their populations over the recent decades).
+We describe this information in more detail in Supplementary Table 1.
 
@@
 As a baseline model, however, we were interested in a hierarchical model that does not make any assumptions regarding the shape of the distributions, and yet explicitly incorporates all information that we have regarding plant's environmental preferences.
-More specifically, we wanted to account for the climatic indicator values and range of variation registered in the floristic database for all plants in our dataset.
-These two values provide basic information regarding plant's optimal environmental conditions and width of their distributions.
+More specifically, we wanted to account for the climatic indicator values and range of variation registered in the attribute database for all plants in our dataset.
+These two values provide basic information regarding plants' optimal environmental conditions and the width of their distributions.
 
@@
 Given the prior information that we have about the system, we know that species occupy specific geographic ranges; therefore, we know that their distributions have finite variance.
-While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among specie in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions.
-Therefore, for this baseline model, if all that we are willing to assume about species' realized niches is that these have finite variance, the most conservative assumption and the safest bet---i.e.~the one with the largest entropy---is that they follow a Gaussian distribution (Fig~\ref{fig:response}a).
+While we could also assume that many other factors might influence species' presence in a given site---e.g.~the biotic interactions among species in the site---we do not necessarily have an \textit{a priori} expectation of how exactly these factors will influence the shape of species' distributions.
+Therefore, for this baseline model, we choose the maximum entropy distribution with finite variance: a Gaussian distribution (Fig~\ref{fig:response}a).
 That is, given the presence/absence or abundance $y_{ij}$ of any species $i$ in any given site $j$, and an environmental variable $x_{j}$, we can define species' responses to the environment as
@@
 \end{equation}
-where $F$ is the likelihood function, and $\alpha_i$, $\beta_i^k$, and $\gamma_i$ describe amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively.
+where $F$ is the likelihood function, and $\alpha_i$, $\beta_i$, and $\gamma^{-1}_i$ describe the amplitude of the probability $p_{ij}$, species' average climatic suitability and range of variation along the environmental gradient, respectively.
 Notice that $F$ characterizes a Binomial distribution when considering binary data, and it characterizes an ordered categorical likelihood function when we consider Braun-Blanquet abundance-dominance classes as response variables (see the full description of both models in the Supplementary Methods).
@@
 That said, this model can easily be generalized to account for multiple predictors (see Supplementary Methods).
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
@@
 \subsubsection*{Model priors}
-The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the floristic database.
+The model structure described above allows us to explicitly incorporate all prior knowledge that we have regarding species' distributions contained in the attribute database.
 To do so, we define the prior distributions for the parameters in model (\ref{eq:baseline-response}) as:
@@
 \begin{split}
+\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
 \beta_{i}  & \sim \text{MVNormal}\left(\hat{\beta}, \Sigma^{\beta}\right)\\
 \text{log}(\gamma_{i})  & \sim \text{MVNormal}\left(\hat{\gamma}, \Sigma^{\gamma}\right)\\
-\text{log}(\alpha_{i}) & \sim \text{Normal}\left(\hat{\alpha}, \sigma_{\alpha}\right)\\
+\hat{\alpha}, 
 \hat{\beta}, 
-\hat{\gamma}, 
-\hat{\alpha}  & \sim \text{Normal}\left(0,1\right)\\
+\hat{\gamma}  & \sim \text{Normal}\left(0,1\right)\\
 \sigma_{\alpha}  & \sim \text{Exponential}\left(1\right)
@@
 Such a covariance structure declines exponentially with the square of a distance matrix $D_{ij}$, which characterize differences between species computed using our prior information.
-In the floristic database, this information is represented by the set of ordinal traits specified for the different species.
+In the attribute database, this information is represented by the set of ordinal traits specified for the different species.
 While there are many different ways to turn ordinal data into distance matrices, we choose to use a mixed-membership stochastic block model because it allows us to deal with cases of missing data (see Supplementary Methods for extended details; \citealt{godoy-loriteAccurateScalableSocial2016}).
@@
 For all these hyperparameters, we choose weakly informative priors such that $\sigma , \eta \sim \text{Exponential}\left(1\right)$ and $\rho\sim \text{Exponential}\left(0.5\right)$.
-Notice that other structures can be used to define the covariance matrices of the different Gaussian processes \citep{mcelreathStatisticalRethinkingBayesian2020}, including structures that account for multiple distance matrix $D_{ij}$ for any given parameter.
+Notice that other structures can be used to define the covariance matrices of the different Gaussian processes \citep{mcelreathStatisticalRethinkingBayesian2020}, including structures that account for multiple distance matrix $D_{ij}$ for any given parameter (Supplementary Methods).
 
@@
 This is especially true when using ordered categorical likelihood functions (see \citealt{standevelopmentteamStanModelingLanguage2021}).
-Therefore, we focus on those species for which we have at least 20 occurrences when modelling both binary data and ordinal data.
+Therefore, we focus on those species for which we have at least 20 occurrences when modelling both binary data and ordinal data ($251$ species in total).
 
@@
 where $\gamma^{\prime}_{i} = g(\gamma_{i}, \nu_{i})$, and $\nu_{i}$ is a parameter that describes the kurtosis of the distribution, which we define as $\nu_{i}\in\left(1, \infty\right)$.
-Following this, we choose an adaptive prior for this set of new parameter such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
+Following this, we choose an adaptive prior for this set of new parameters such that $\log\left(\nu_{i}-1\right)\sim \text{Normal}\left(\hat{\nu}, \sigma_{\nu}\right)$, where $\hat{\nu}\sim\text{Normal}\left(0, 1\right)$ and $\sigma_{\nu}\sim\text{Exponential}\left(2\right)$.
 Given the relationship between $\gamma^{\prime}_{i}$ and $\gamma_{i}$,  we can re-parametrize the model and follow Eq.~(\ref{eq:baseline-priors}) to define the prior distributions (see Supplementary Table 2; \citealt{nadarajahGeneralizedNormalDistribution2005}).
@@
 For example, the non-standardized Student's t-distributions is an interesting distribution because, as opposed to the generalized error distribution, it allows for fat tails without generating a cusp at the center (see Fig~\ref{fig:response}b).
-However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Fig~\ref{fig:response}b), and the sampling of the model can be somewhat more challenging (ref).
+However, we avoided using the non-standardized Student's t-distributions because it does not allow for tails that are lighter than normal (e.g.~$\nu_i>2$ in Eq.~\ref{eq:generalized-response}; Fig~\ref{fig:response}b), and the sampling of the model can be somewhat more challenging.
 
@@
 \subsection*{Evaluating the log-likelihood}
-One way to understand the shape of distributions is by evaluating the computed log-likelihood values across the environmental gradient.
-Comparing these values, one can understand what aspects of the shape of distributions are missing.
+The log-likelihood values measure the goodness of fit of a statistical model to any data point, for a given sample of the posterior distributions.
+These values can be used to understand where the models fail to capture the variation of our empirical data.
+For example, high log-likelihood values indicate those data points that are well captured by a given distribution model, while low values signal those points that are instead unexpected by the model.
+Therefore, one could use these values to understand what aspects of the shape of distributions that are missing.
 To do so, for every sample of the model, we computed the log-likelihood values and the normalized probability distribution.
@@
 We studied the distribution data to characterize species' realized niches along the main axis of variation of all environmental variables.
-Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the floristic database regarding species' indicator values and range of variation.
+Using the presence and absence of species across sites as the response variable, we sampled the posterior distributions of the baseline model, accounting for the information in the attribute database regarding species' indicator values and range of variation.
 This allowed us to map the center and variance of species' distributions along the environmental gradient (Fig.~\ref{fig:correlation}).
 Studying the relationship between these properties, we found these to be negatively correlated (i.e.~$\beta_{i}$ and $\gamma_{i}$ in the baseline model were positively correlated; Fig.~\ref{fig:correlation}).
-This means that species found at lower elevations have generally wider distributions than those at higher elevations.
-The same relationship was found when using instead elevation or mean temperature as explanatory variables (Supplementary Fig.~3) as well as when using ordinal data (Supplementary Fig.~4); however, the pattern was not present along the second main axis of variation of our environmental variables (Supplementary Fig.~5).
+This means that species found at the lower end of the gradient (i.e.~higher temperature and lower elevation) have generally wider distributions than those found at the higher end (i.e.~lower temperature and higher elevation).
+The same relationship was found when independently using elevation or mean temperature as explanatory variables (Supplementary Fig.~3) as well as when using ordinal data (Supplementary Fig.~4); however, the pattern was not present along the second main axis of variation of our environmental variables (i.e.~correlated to precipitation seasonality; Supplementary Fig.~1 and 5).
 The comparison between the other parameter estimates revealed additional, somewhat more expected, relationships.
-In particular, we found the amplitude of distributions to be positively and negatively correlated with their mean and variance, respectively (i.e.
+In particular, we found the amplitude of distributions (i.e.~height of the distributions) to be positively and negatively correlated with their mean and variance, respectively (i.e.
 $\alpha_{i}$ is positively correlated with $\beta_{i}$ and $\gamma_{i}$; Supplementary Fig.~6).
-This implies that, at higher elevations, species' distributions generally have lower amplitudes.
+This implies that, at higher elevations, species' distributions generally have lower amplitudes (i.e.~lower overall probability of occurrence).
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
@@
 Panel (a) describes the relationship between the mean ($\beta_i$) and variance ($\propto\gamma_i^{-1}$) of distributions.
-Every dot characterizes the average value of the corresponding posterior distributions for any given species.
+Each point represents the average value of the corresponding posterior distributions for any given species.
 The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.
@@
 Panel (c) displays the estimates for the variance of species' distributions along the environmental gradient.
-In (b) and (c), the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.}
+In (b) and (c), the points represent the mean of the posterior distributions, the corresponding lines characterize the 89\% confidence intervals, and species are sorted according to the mean of the posterior distributions along the x and y axes, respectively.}
 	   %7.5-5.5
@@
 Studying the resulting posterior distributions, we found the average kurtosis of the distributions to be slightly greater than zero, which corresponds to distributions with longer tails than the Gaussian case (Fig.~\ref{fig:kurtosis-skewness}).
-However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~8), which might indicate that the shape of the tails is species-specific.
+However, the parameter controlling for the kurtosis $\nu_i$ displayed a lot of variation across species (Supplementary Fig.~8), which might indicate that the shape of the tails is species-specific and potentially explained by species' ecological traits.
 
@@
 	   \caption{Average kurtosis and skewness of species' distributions.
-Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table 2 and \citealt{vignette}), the two panels describe the average (a) kurtosis and (b) skewness of distributions.
+Calculated using the posterior distributions of parameters $\hat{\nu}$ and $\hat{\lambda}$ from the models (see Supplementary Table 2 and \citealt{kermanSkewnessKurtosisBounds2013}), the two panels describe the average (a) kurtosis and (b) skewness of distributions.
 Panel (a) displays the results obtained using response curves that follow a generalized error distribution (black line) and a skewed generalized error distribution (yellow line).
@@
 This model outperformed the rest, presenting Akaike weights close to 1 (Supplementary Fig.~7), suggesting that both the kurtosis and skewness are useful properties to describe empirical distributions (Fig.~\ref{fig:kurtosis-skewness}).
-A study of how the prior knowledge we had regarding species' environmental preferences, range of variation and ecological strategies informed the different parameters of the model is presented in the Supplementary Note 1 and Supplementary Fig.~9.
+Notice that these properties are also not fully independent, as the consideration of both fat-tails and skewed responses substantially change the results obtained using the previous two models (Eq.~\ref{eq:generalized-response} and \ref{eq:skewed-response}; Fig.~\ref{fig:kurtosis-skewness}).
+A study of how the prior knowledge we had regarding species' environmental preferences and range of variation informed the different parameters of the model is presented in the Supplementary Note 1 and Supplementary Fig.~9.
 %Comparing the prior and posterior distributions for the hyperparameters in the.
 
-The model characterizing fat-tailed and skewed distributions allowed us to study the posterior distributions for the parameters describing the mean, variance, amplitude, kurtosis and skewness of species realized niches altogether.
+The model characterizing fat-tailed and skewed distributions allowed us to study the posterior distributions for the parameters describing the mean, variance, amplitude, kurtosis and skewness of species' realized niches simultaneously.
 We observed that different types of species seem to present characteristically different distributions (Fig.~\ref{fig:pairwise}).
-Focussing on the negative correlation between the mean and variance of species' distributions, we found some species to escape some of the aforementioned macroecological constraints (Supplementary Fig.~XX).
-Moreover, recent and historical range expanders are often found at lower altitudes, presenting higher amplitudes and distributions that appear to showcase steeper declines towards lower elevations.
-Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area, as the same model produced comparable results when using simulated and bootstrapped data (Supplementary Note XX and Supplementary Fig.~XX).
-Moreover, these results did not substantially change when using ordinal data (Supplementary Fig.~XX).
+Focussing on the negative correlation between the mean and variance of species' distributions, we found some species to escape such macroecological constraints (Supplementary Fig.~10).
+Moreover, recent and historical range expanders are often found at the lower end of the environmental axis (i.e.~higher temperature and lower elevation), presenting higher amplitudes, and distributions that appear to showcase steeper declines towards lower elevations (Fig.~\ref{fig:pairwise})---potentially showing these species to spread towards the higher elevations.
+Notice the nature of these results does not depend on the presence or absence of a species at the edge of the sampling area, as the same model produced comparable results when using simulated and bootstrapped data (Supplementary Note 2 and Supplementary Fig.~11).
+Moreover, these results did not substantially change when using ordinal data (Supplementary Fig.~12).
 
@@
 Focussing on the differences between indigenous, historical range expanding and recent range expanding species, the panels describe the relationship between the basic properties of their distributions.
-Panels (a-d) characterize the relationship between the mean, and the variance, amplitude, skewness and kurtosis of the species' distributions (Supplementary Table XX).
+Panels (a-d) characterize the relationship between the mean, and the variance, amplitude, skewness and kurtosis of the species' distributions (Supplementary Table 2 and \citealt{kermanSkewnessKurtosisBounds2013}).
 The points in every panel are calculated as the average value across all samples of the model.
@@
 
-Finally, we wanted to identify what aspects of the shape of distributions we were still missing.
+Finally, we wanted to identify what aspects of the shape of distributions were still missing.
 We used the computed log-likelihood values and normalized probabilities to understand where our best performing model fails to capture the variation in empirical plant distributions.
-We found most data points to be located at the tails of distributions ($\text{normalized probability}\,\approx\,0$) and to present high log-likelihood values (Supplementary Fig.~XX).
-This is not surprising as the study area spans an extensive altitude gradient, and species' distributions are generally narrow relative to it; therefore, the model accurately predicts that species are usually absent in those sampling sites that fall relatively far from the center of their distributions.
+We found most data points to be located at the tails of distributions ($\text{normalized probability}\,\approx\,0$) and to present high log-likelihood values (Supplementary Fig.~13).
+This is not surprising as the study area spans an extensive elevation gradient, and species' distributions are generally narrow relative to it; therefore, the model accurately predicts that species are usually absent in those sampling sites that fall relatively far from the center of their distributions.
 However, studying instead only those points for which the model did not perform well (with a $\text{likelihood}\leq 0.5$), we found these to generally be associated with high normalized probabilities (Fig.~\ref{fig:loglik}).
 This indicates that the unexplained variation is often found at the center of species' distributions.
-Similar results were found when using ordinal data (Supplementary Fig.~XX).
+Similar results were found when using ordinal data (Supplementary Fig.~14).
 
@@
 The graph maps the log-likelihood and normalized probability values for all species across all samples.
+The colour characterizes the overall percentage of points falling within a given hexagon.
 Notice that in this figure there are only displayed those points that present a likelihood smaller than $0.5$.
-The mapping of all log-likelihood values is presented in Supplementary Fig.~XX.}
+The mapping of all log-likelihood values is presented in Supplementary Fig.~13.}
 	   %5-4
@@
 But while this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}, contrasting evidence suggests that this rule is not pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}.
-Our results seem to contradict the predictions of the Rapoport's rule, as we observed a negative correlation between species' range and elevation.
+Our results seem to contradict the predictions of the Rapoport's rule, as we observed a negative correlation between species' range  width and elevation.
 Moreover, other properties of species' distributions---such as their amplitude---were also significantly correlated with species' ranges.
 This is interesting because it hints at the existence of some general macroecological constraints that dictate the way different species assemble across environments.
-That said, our results also suggest that species such as neophytes and archeophytes might not obey this same constraints, as these were singularly positioned along the gradient (Supplementary Fig.~XX).
+That said, our results also suggest that species such as neophytes and archeophytes might not obey this same constraints, as these were singularly positioned along the gradient (Supplementary Fig.~10).
 
@@
 This hypothesis predicts species to be most abundant at the center of their distributions, and it is an implicit assumption at the core of most modelling approaches.
-Namely, if one is only willing to assume that species have finite geographic ranges, the abundant-center hypothesis is a consequence our state of ignorance (i.e.~the maximum entropy distribution).
+Namely, if one is only willing to assume that species have finite geographic ranges, the abundant-center hypothesis is a consequence of our state of ignorance (i.e.~the maximum entropy distribution).
 That said, several studies have pointed out that the abundant-center hypothesis is not pervasive in empirical distributions \citep{wagnerSimilarPerformanceCentral2011, pirononGeographicVariationGenetic2017, dallasSpeciesAreNot2017}, suggesting that population abundance could often be more strongly driven by interactions and community structure than the environment \citep{dallasSpeciesAreNot2017}.
@@
 However, our models have clear interpretable parameters, and can be used to directly compare the shape of species' realized niches.
-These comparisons could be used to generate hypotheses regarding where and when different species might strongly interact with one another along an environmental gradient \citep{louthanWhereWhenSpecies2015}, making ecologically-informed predictions regarding the presence and absence of these relationships \citep{SIASH}.
+These comparisons could be used to generate hypotheses regarding where and when different species might strongly interact with one another along an environmental gradient \citep{louthanWhereWhenSpecies2015}, making ecologically-informed predictions regarding the presence and absence of these relationships \citep{callawayPositiveInteractionsAlpine2002, heGlobalShiftsPositive2013}.
 

### 2021-07-15 manuscript/supplementary_information.tex
manuscript changes

@@
 \title{\vspace{-1cm} \normalsize Supplementary Information\\\vspace{0.2cm}
-\LARGE Basic aspects of species' distributions}
+\LARGE Using Bayesian non-linear models to uncover broad macroecological patterns}
 
@@
 \subsubsection*{Distance matrices from incomplete categorical and ordinal data}
-The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the floristic database.
+The prior information that we have regarding species' distributions is represented by the set of ordinal and categorical traits found in the attributes database.
 More specifically, both the ecological indicator values and range of variation are ordinal traits, containing potential missing entries (Supplementary Table \ref{stab:EI}).
 These data could be directly used as covariates in any given distribution model; however, we want this information to be accounted for as a prior for the relevant parameters of our Bayesian model.
-One way to do so is by compiling the traits in the floristic database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
+One way to do so is by compiling the traits in the attributes database into variance-covariance matrices characterizing the \textit{a priori} similarity between species.
 To accomplish this, we need to turn the set of ecological indicator values and range of variation into the distance matrices $D$ presented in the text.
@@
 The model structure defined above allows us to test how different sources of information characterize each of the different parameters.
-Specifically, we can do this by modifying Eq.~(\ref{eq:covariance-baseline}).
 For example, imagine that we have multiple matrices $D^k$ characterizing species' differences along different axis of variation---e.g.~two matrices characterizing physiological and environmental traits.
-One can modify Eq.~(\ref{eq:covariance-baseline}) for a particular parameter---e.g.~$\beta_{i}$---such that
+One can modify Eq.~(3) of the main text for a particular parameter---e.g.~$\beta_{i}$---such that
 \begin{equation} 
@@
 \section*{Supplementary Notes}
+\subsubsection*{Supplementary Note 1: studying the hyperparameters of the Gaussian processes}
+As described in the main text, we use Gaussian processes to account for the prior information regarding a given parameter (e.g.~$\beta_i$), characterized by a distance matrix $D$ (see `Distance matrices from incomplete categorical and ordinal data' section of the Supplementary Methods).
+The advantage of using a Gaussian processes is that one can assess the value of the prior information by studying the posterior distributions of the corresponding hyperparameters.
+As described in Eq.~(3) of the main text, the hyperparameter $\rho$ of a Gaussian process determines the rate of decline of the covariance between any two species, $\eta$ defines its maximum value, and $\sigma$ describes the additional covariance between the different observations for any given species.
+Therefore, the higher the $\eta$ values relative to $\sigma$ (for $\rho>0$), the more the prior knowledge informs the parameter values of the different species.
+
+In all models, we use species' environmental indicator values and range of variation to inform the parameters $\beta_i$ and $\gamma_i$, respectively.
+Supplementary Fig.~\ref{sfig:hyperparameters1} displays the posterior distributions for $\eta$, $\rho$ and $\sigma$ for each parameter, calculated using the model with fat-tailed and skewed shape.
+Notice that the environmental indicator values are much more informative to the $\beta_{i}$ parameters than the species' range of variation is to the $\gamma_{i}$ parameters.
+This is potentially due to the fact the environmental indicator values have higher resolution that the information regarding species' range of variation in the attributes database.
+
+The information obtained in Supplementary Fig.~\ref{sfig:hyperparameters1} is not particularly surprising, as we already expected our prior knowledge to be somewhat informative to both $\gamma_{i}$ and $\beta_{i}$ parameters.
+However, this figure highlights the potential for understanding the extend to which different sources of data inform the different aspects of the shape of distributions.
+For example, imagine that we had trait information on the different species.
+We could use a Gaussian process to characterize the $\lambda_i$ parameters, using this information as our prior knowledge.
+The hyperparameters of this Gaussian processes could then reveal the extend to which some ecological traits explain the skewness of species distributions.
+
+\subsubsection*{Supplementary Note 3: validation of $\lambda$ estimates}
+
+One could hypothesise that the level skewness of distributions depends on the fact that some species can be found near the edge of the sampling area.
+However, the advantage of using a Bayensian approach to study such distributions is that we obtain full descriptions of the parameter posterior distributions, and we don't make any assumptions regarding the nature of the distributions outside of the sampling area.
+Therefore, a species that is near the edge of the sampling area should only result in larger confidence intervals for some of the parameter estimates, as it only implies less certainty regarding the true shape of such species' distribution.
+To showcase this, we used two strategies: studying subsets of the original data and using simulated data.
+
+To study subsets of the data, we considered different windows of elevation containing 25\% of the data.
+To do so, given the distribution of site elevation, we divided our dataset into 5\% subsets $s_{i}$.
+This means that, for example, $s_1$ and $s_{20}$ contain the 5\% of sites at the lowest and highest elevations, respectively.
+Using these subsets of data, we grouped them to consider 16 elevation windows containing 25\% of the data, so that the first window of elevation contains all subsets of data between $s_1$ and $s_5$, the second window contains the subsets between $s_2$ and $s_6$; and so on.
+Then, we used the model characterized by fat-tails and a skewed response curve, and analysed the distribution of species in each elevation window, independently (considering those species with at least 20 occurrences).
+Finally, focussing on the parameter characterizing the skewness of distributions $\lambda_i$, we compared the posterior distributions obtained for each window of elevation with those obtained using the full dataset.
+In particular, we focused on whether or not the mean parameter values obtained using the full dataset fell within the confidence intervals of the posterior distributions obtained using the different windows of elevation.
+As expected, we found this to be true in $>99\%$ of the cases, which highlights the fact that sampling edges might add uncertainty to our estimates but does not make additional assumptions regarding the nature of the data.
+
+Another way to validate the predictions of the model is by using simulated data.
+To do so, one can generate species' distribution data using a skewed generalized error distribution (see Code availability section), and compare the true values to the estimated posterior distributions.
+Supplementary Fig.~\ref{sfig:simulated-lambdas} displays the results for 50 simulated species distributions for $\beta_i$ and $\lambda_i$.
+Notice that the model does a good job characterizing the true $\lambda_i$ values in most cases (with the exception of some simulated species with low number of occurrences $<20$).
+This is important because it shows that with limited data, the model does a good job predicting the true values of $\lambda_i$ despite the variation in $\alpha_i$, $\gamma_i$, $\beta_i$, $\nu_i$.
+Perhaps most importantly, we generated the simulated data assuming that $\text{logit}\left(\frac{\lambda_i+1}{2}\right)\sim\text{Normal}(1, 0.5)$ (see Code Availability section).
+That is, the simulated lambdas are generally positive (i.e.~skewed in the same direction).
+Notice that the posterior distributions are not affected by the $\beta_i$ parameters (i.e.~the center of the distributions), and are consistently predicted to be above zero.
+
 \section*{Supplementary Figures}
@@
     	  \vspace{0.3cm}
-	   \caption{PCA and variance explained.}
+	   \caption{Principal components analysis of the different environmental variables used.
+Panel (a) displays the percentage of variation explained by each principal component.
+Panel (b) provides a visualization of the contribution of each variable to the first two axes of variation.
+These graphs were generated using R functions from the package `factoextra' \citep{kassambaraFactoextraExtractVisualize2017}}.
       \label{sfig:pca}
@@
     	  \vspace{0.3cm}
-	   \caption{Studying the simulations.
-Transparency is a way to indicate the different sample sizes.
-Say that for both $\beta$ and $gamma$, the posterior distributions for the different species are correlated so that the variance between species increases as they are further apart.}
+	   \caption{Testing the baseline model's performance with simulated data.
+We generated 25 bell-shaped distributions along an environmental axis, randomly sampling the different parameter values $\alpha_i$, $\beta_i$ and $\gamma_i$.
+More specifically, the $\text{log}\left(\alpha_i\right)$ parameters were sampled from a normal distribution with mean $-0.5$ and variance $1$; the $\beta_i$ and $log(\gamma_i)$ parameters were instead sampled from two multivariate normal distributions so that the covariance between species 1 and 2 is larger than the covariance between species 1 and 3 (see Code Availability section for further information).
+Then we used these distributions to simulate presence/absence data along the environmental axis using a binomial distribution, and used the baseline model to estimate the posterior distributions for the different parameters.
+The different panels showcase the comparison between real parameter values (black) and the posterior distributions (red) for (a) $\alpha_i$, (b) $\beta_i$, and (c) $\gamma_i$.
+The transparency for the points describing the real parameter values provides an indication of the sample size of each simulated species (number of occurrences).
+The red dots and red bars characterize the mean and the $89\%$ confidence intervals for the different posterior distributions, respectively.}
       \label{sfig:simulations}
@@
     	  \vspace{0.3cm}
-	   \caption{The same as in Fig.~1 but with elevation and temperature.}
+	   \caption{Relationship between the posterior distributions for parameters $\beta_i$ and $\gamma_i$ across species considering different environmental variables.
+Panel (a) describes the relationship between the mean ($\beta_i$) and variance ($\propto\gamma_i^{-1}$) of distributions along the main axis of variation in Supplementary Fig.~\ref{sfig:pca}.
+Panel (b) and (c) describes the same relationship along elevation and average temperature.
+Each point represents the average value of the corresponding posterior distributions for any given species.
+The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.}
       \label{sfig:temperature-elevation}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
-    \includegraphics[width=0.9\textwidth]{figures/categorical-figure1}
+    \includegraphics[width=0.95\textwidth]{figures/categorical-figure1}
     	  \vspace{0.1cm}
-	   \caption{Universality of the relationship between mean and variance of species' distributions.
-Comparison between how different types of species are mapped in Fig.~\ref{fig:correlation}a.
-Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
-Panel (b) shows the differences between neophytes and native species in the way these are distributed along the environmental gradient.
-Panel (c) shows the same differences for species that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
+	   \caption{Relationship between the posterior distributions for parameters $\beta_i$ and $\gamma_i$ across species.
+These results are analogous results to those displayed in Fig.~2 of the main text, but they were obtained using ordinal measures of abundance instead of presence/absence data.
+Panel (a) describes the relationship between the mean ($\beta_i$) and variance ($\propto\gamma_i^{-1}$) of distributions.
+Each point represents the average value of the corresponding posterior distributions for any given species.
+The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.
+Panel (b) displays the estimates for the center of species' distributions along the environmental gradient.
+Panel (c) displays the estimates for the variance of species' distributions along the environmental gradient.
+In (b) and (c), the points represent the mean of the posterior distributions, the corresponding lines characterize the 89\% confidence intervals, and species are sorted according to the mean of the posterior distributions along the x and y axes, respectively.}
 	   %7.5-4
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
@@
     	  \vspace{0.3cm}
-	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
+	   \caption{Relationship between the posterior distributions for parameters $\beta_i$ and $\gamma_i$ across species along the second axis of variation in Supplementary Fig.~\ref{sfig:pca}.
+Panel (a) describes the relationship between the mean ($\beta_i$) and variance ($\propto\gamma_i^{-1}$) of distributions.
+Each point represents the average value of the corresponding posterior distributions for any given species.
+The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.
+Panel (b) displays the estimates for the center of species' distributions along the environmental gradient.
+Panel (c) displays the estimates for the variance of species' distributions along the environmental gradient.
+In (b) and (c), the points represent the mean of the posterior distributions, the corresponding lines characterize the 89\% confidence intervals, and species are sorted according to the mean of the posterior distributions along the x and y axes, respectively.}
       \label{sfig:secondaxis}
@@
     	  \vspace{0.3cm}
-	   \caption{Relationship between mean and variance of species' distributions. These are the results for the second axis of variation for the climatic data.}
+	   \caption{Relationship between the posterior distributions for parameters $\alpha_i$, $\gamma_i$ and $\beta_i$ across species.
+Panel (a) describes the relationship between the amplitude ($e^{\alpha_i}$) and variance ($\propto\gamma_i^{-1}$) of distributions.
+Panel (b) describes the relationship between the amplitude ($e^{\alpha_i}$) and mean ($\beta_i$) of distributions.
+Each point represents the average value of the corresponding posterior distributions for any given species.
+The value in the bottom-left corner of the plot displays the Pearson's correlation coefficient between the parameters calculated across all samples.}
       \label{sfig:alpha-vs-gammabeta}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
     \vspace{0.5cm}
-    \includegraphics[width=0.8\textwidth]{figures/WAIC-plot}
+    \includegraphics[width=0.85\textwidth]{figures/WAIC-plot}
     	  \vspace{0.3cm}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
     \vspace{0.5cm}
-    \includegraphics[width=0.8\textwidth]{figures/nu-kurtosis}
+    \includegraphics[width=0.85\textwidth]{figures/nu-kurtosis}
     	  \vspace{0.3cm}
-	   \caption{Variability in the tails.}
+	   \caption{Study of the estimates for $\nu_i$ and kurtosis of the distributions.
+Panel (a) displays the posterior distributions for $\nu_i$ of the 251 species studied.
+Panel (b) represents the posterior distributions for the kurtosis of the distributions.
+In both panels, the points represent the mean of the posterior distributions, and the corresponding lines characterize the 89\% confidence intervals.
+The black dotted lines indicates the $\nu_i$ and kurtosis values corresponding to a normal distribution.}
       \label{sfig:nu-kurtosis}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/hyperparameters-supplementary}
+    	  \vspace{0.3cm}
+	   \caption{Posterior distributions for the hyperparameters $\eta$, $\rho$ and $\sigma$ characterizing the Gaussian processes in Equations (2) and (3) of the main text.
+Panels (a) and (b) display the results for $\beta_i$ and $\gamma_i$, respectively.}
+      \label{sfig:hyperparameters1}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=1\textwidth]{figures/Rapaport-species-differences}
+    	  \vspace{0.3cm}
+	   \caption{Universality of the relationship between mean and variance of species' distributions.
+Comparison between how different types of species are mapped in Fig.~2a of the main text.
+Panel (a) describes the correlation coefficient between $\beta_i$ and $\gamma_i$ for each type of species.
+Panel (b) shows the differences in the way indigenous species, historical expanders and recent expanders are distributed along the environmental gradient.
+Panel (c) shows the same differences for species with populations that have decreased, decreased in low elevations, increase and remain stable over the last decades (see Supplementary Table 1 for further details).}
+      \label{sfig:rapaport-types}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \vspace{0.5cm}
+    \includegraphics[width=0.85\textwidth]{figures/lambda-test}
+    	  \vspace{0.3cm}
+	   \caption{Testing the model's ability to estimate $\lambda_i$ with simulated data.
+We generated 50 fat-tailed and skewed distributions along an environmental axis, randomly sampling the different parameter values $\alpha_i$, $\beta_i$, $\gamma_i$, $\nu_i$, and $\lambda_i$ (see Code Availability section for further information).
+Then we used these distributions to simulate presence/absence data along the environmental axis using a binomial distribution, and used the model described in Eq.~(6) of the main text to estimate the posterior distributions for the different parameters.
+The different panels showcase the comparison between real parameter values (black) and the posterior distributions (red) for (a) $\beta_i$ and (b) $\lambda_i$.
+The transparency for the points describing the real parameter values provides an indication of the sample size of each simulated species (number of occurrences).
+The red dots and red bars characterize the mean and the $89\%$ confidence intervals for the different posterior distributions, respectively.
+The black-dotted line indicates the $\lambda_i$ value characterizing a normal distribution.}
+      \label{sfig:simulated-lambdas}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.9\textwidth]{figures/categorical-figure3}
+    	  \vspace{0.1cm}
+	   \caption{Comparing the distributions of different types of species.
+These results are analogous results to those displayed in Fig.~4 of the main text, but they were obtained using ordinal measures of abundance instead of presence/absence data.
+Focussing on the differences between indigenous, historical range expanding and recent range expanding species, the panels describe the relationship between the basic properties of their distributions.
+Panels (a-d) characterize the relationship between the mean, and the variance, amplitude, skewness and kurtosis of the species' distributions (Supplementary Table 2 and \citealt{kermanSkewnessKurtosisBounds2013}).
+The points in every panel are calculated as the average value across all samples of the model.
+%The analogous results found for species classified accroding to their population change tendency over the years is presented in Supplementary Fig~XX.
+The ellipses characterize the 1 standard deviation ellipses for each species type.
+	   }
+	   %7-4.5
+      \label{sfig:figure3-categorical}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figures/loglik-all-bin}
+    	  \vspace{0.1cm}
+	   \caption{Studying the distribution of log-likelihood values.
+The graph maps the log-likelihood and normalized probability values for all species across all samples.
+The colour characterizes the overall percentage of points falling within a given hexagon.}
+	   %5-4
+      \label{sfig:loglik-all}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
+  \centering
+    \includegraphics[width=0.5\textwidth]{figures/loglik-notshow}
+    	  \vspace{0.1cm}
+	   \caption{Studying the distribution of log-likelihood values.
+These results are analogous results to those displayed in Fig.~5 of the main text, but they were obtained using ordinal measures of abundance instead of presence/absence data.
+The graph maps the log-likelihood and normalized probability values for all species across all samples.
+The colour characterizes the overall percentage of points falling within a given hexagon.
+Notice that in this figure there are only displayed those points that present a likelihood smaller than $0.5$.}
+	   %5-4
+      \label{sfig:loglik-all-categorical}
 \end{figure}
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \begin{center}
@@
 \end{center}
-\caption{Indicator values \citep{landoltFloraIndicativaOkologische2010}}
+\caption{Description of the different plant attributes obtained form \citealt{landoltFloraIndicativaOkologische2010}.
+Each row describe one of the variables used, including the variable type, name in the database, and number of classes for such variable.}
 \label{stab:EI}   
@@
 
-\begin{table}[h]
+\clearpage
+
+\begin{table}[ht]
 \begin{center}
@@
 \end{center}
-\caption{cite vignette}
+Additional definitions for the different models used.
+Except for the baseline model, all models are described in terms of $\beta^{\prime}_i$ and $\gamma^{\prime}_i$.
+However these models are parametrized using instead $\alpha_i$, $\beta_i$, $\gamma_i$, $\nu_i$ and $\lambda_i$, which are parameters that are comparable across models.
+The relationship between all parameters is described in the table (see \citealt{kermanSkewnessKurtosisBounds2013} for further details, including the analytical expressions for the kurtosis and skewness of all distributions).
 \label{stab:formulas}   
 \end{table}
+
+\clearpage
 

### 2021-07-15 code/paper-models/plots/plot-real.R

@@ -8,7 +8,7 @@ library(hrbrthemes)
 library(plotly)
 library(cowplot)
 library(dplyr)
-
+library(tidyverse)
 ####
 # Visualizing the results of some of the models
 ####
@@ -753,7 +753,7 @@ plot.actual.data.means <- function(model=NULL){
 plot.actual.data.means.detailed <- function(model=NULL){
   # Parameters for plots
   extra <- 1
-  colo <- c("#e6ab02", "#1b9e77", "#d95f02", "#7570b3")
+  colo <- c("#1b9e77", "#7570b3", "#e7298a", "#e6ab02", "#d95f02", "#666666", "#a6761d")
   
   # Load the data if not added  
   if(is.null(model)){
@@ -802,9 +802,9 @@ plot.actual.data.means.detailed <- function(model=NULL){
   xlab=""
   breaks <- round(seq(from=xlim[1], to=xlim[2], length.out = 4), 1)
   
-  labs <- c("indigenous", "neophytes", "archaeophytes")
+  labs <- c("indigenous", "recent\nexpanders", "historical\nexpanders")
   labels <- labs[deta$i+(deta$n+deta$jn+deta$isn)*2+(deta$a+deta$ja+deta$isa)*3]
-  color <- c(colo[2],colo[4], "#e7298a")
+  color <- colo[1:3]
   alphas <- c(0.6,0.7,0.7)
   alphas <- alphas[deta$i+(deta$n+deta$jn+deta$isn)*2+(deta$a+deta$ja+deta$isa)*3]
 
@@ -814,12 +814,12 @@ plot.actual.data.means.detailed <- function(model=NULL){
   meanPI <- mean(corPI)
   sdPI <- sd(corPI)
   corPIci <- round(c(meanPI, sdPI), 2)
-  correlations <- rbind(correlations, c("archaeophytes", corPIci))
+  correlations <- rbind(correlations, c("historical\nexpanders", corPIci))
   corPI <- sapply(1:dim(post$beta)[1], function(x) cor(post$beta[x,(deta$n+deta$jn+deta$isn)==1], post$gamma[x,(deta$n+deta$jn+deta$isn)==1]))
   meanPI <- mean(corPI)
   sdPI <- sd(corPI)
   corPIci <- round(c(meanPI, sdPI), 2)
-  correlations <- rbind(correlations, c("neophytes", corPIci))
+  correlations <- rbind(correlations, c("recent\nexpanders", corPIci))
[…]

### 2021-07-15 code/paper-models/plots/plot-simulated.R

@@ -136,14 +136,14 @@ plot.simulated.baseline <- function(beta=T, gp_type = 2){
   L <- length(unique(d$dataset$id))
   obs <- matrix(d$dataset$obs, N, L)
   size <- colSums(obs)
-  size <- ((size-min(size))/(max(size)-min(size)))+1
 
+  plist <- list()
   d_alpha$size <- c(size, rep(max(size), length(size)))
-  p1 <- ggplot(d_alpha, aes(x=N, y=value, group=id, color=id, alpha=size)) + 
+  plist[[1]] <- ggplot(d_alpha, aes(x=N, y=value, group=id, color=id, alpha=size)) + 
     geom_pointrange(aes(ymin=lower, ymax=upper)) + theme_linedraw() +
     scale_color_manual(values = c("real" = "gray20", "estimated"="#fb8072")) +
     ylab(expression(alpha)) + xlab("species") + theme_bw() + ggtitle("(a)") +
-    scale_alpha(range = c(0.25, 1)) +
+    scale_alpha_continuous(limits = c(0,max(d_alpha$size)), breaks=round(seq(from=min(d_alpha$size), to=max(d_alpha$size), length.out = 4)))+
     theme(text = element_text(size=10),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
@@ -152,10 +152,10 @@ plot.simulated.baseline <- function(beta=T, gp_type = 2){
         panel.border = element_rect(colour = "black", fill=NA, size=0.5),
         plot.title = element_text(size=10))
   d_beta1$size <- c(size, rep(max(size), length(size)))
-  p2 <- ggplot(d_beta1, aes(x=N, y=value, group=id, color=id, alpha=size)) + 
+  plist[[2]] <- ggplot(d_beta1, aes(x=N, y=value, group=id, color=id, alpha=size)) + 
     geom_pointrange(aes(ymin=lower, ymax=upper)) + theme_linedraw() +
     scale_color_manual(values = c("real" = "gray20", "estimated"="#fb8072")) +
-    scale_alpha(range = c(0.25, 1)) +
[…]

### 2021-07-15 code/paper-models/prepare-data.R

@@ -552,17 +552,44 @@ species_distribution.data <- function(variables=c("bio5_", "bio6_","bio12_", "gd
                         # Prepare full dataset
                         dataset = cbind(dat$obs.data, pca.clim$x[,c(1:ndim)])
                         
-                        # p1 <- fviz_eig(pca.clim)+
+                        # plist <- list()
+                        # plist[[1]] <- fviz_eig(pca.clim, barcolor = "gray", barfill = "gray")+
                         #         scale_y_continuous(expand = expansion(add = c(0, 0)))+
                         #         theme_bw()+
-                        #         theme(plot.title = element_blank(), text = element_text(size=10))
-                        # p2 <- fviz_pca_var(pca.clim,
+                        #         xlab("dimensions")+ylab("percentage of explained variances")+
+                        #         theme(plot.title = element_blank(), text = element_text(size=11),
+                        #               panel.grid.major = element_blank(),
+                        #               panel.grid.minor = element_blank())
+                        # p <- fviz_pca_var(pca.clim,
                         #              col.var = "contrib", # Color by contributions to the PC
                         #              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
                         #              repel = T,     # Avoid text overlapping
-                        # ) + theme(plot.title = element_blank(), text = element_text(size=10))
+                        # ) +
+                        #         xlab("dim 1 (82.6%)")+ylab("dim 2 (9.5%)")+
+                        #         theme(plot.title = element_blank(), text = element_text(size=11),
+                        #           panel.grid.major = element_blank(),
+                        #           panel.grid.minor = element_blank(), legend.title = element_blank())
+                        # plist[[2]] <- p+theme(legend.position = "none")
                         # 
-                        # p <- grid.arrange(grobs=list(p1, p2), ncol=2, nrow=1, widths=c(0.65,1))
+                        # grobs <- list()
+                        # widths <- list()
+                        # heights <- list()
+                        # 
+                        # for (k in 1:length(plist)){
+                        #         grobs[[k]] <- ggplotGrob(plist[[k]])
+                        #         widths[[k]] <- grobs[[k]]$widths[2:5]
+                        #         heights[[k]] <- grobs[[k]]$heights[2:5]
+                        # }
+                        # maxwidth <- do.call(grid::unit.pmax, widths)
+                        # maxheight <- do.call(grid::unit.pmax, heights)
+                        # for (k in 1:length(grobs)){
[…]
